---
title: "Interaction Effects"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 8 | Interaction Effects"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
library(modeldata)
options(scipen = 999)
```


# Why Interactions?

## Quick Review: What Dummies Do {.smaller}

- Last unit we saw that **dummy variables** shift the intercept:

. . .

$$\widehat{score} = \hat{\beta}_0 + \hat{\beta}_1(hsgpa) + \hat{\beta}_2(calculus)$$

. . .

- Calculus-takers: $(\hat{\beta}_0 + \hat{\beta}_2) + \hat{\beta}_1(hsgpa)$
- Non-calculus-takers: $\hat{\beta}_0 + \hat{\beta}_1(hsgpa)$

. . .

::: {.callout-important}
## The Limitation
Both groups have the **same slope** — the effect of GPA is assumed identical for both groups. Is that always realistic?
:::


## What If the Slopes Differ? {.smaller}

- Maybe calculus preparation changes *how much* GPA matters
- Maybe education has different returns for men vs. women

. . .

- We need a way to let the **effect** of one variable **depend on** the value of another

. . .

> **Interaction effects** let us model this by including the *product* of two variables in our regression


# Interactions Between Two Dummies

## Setting Up the Model {.smaller}

- Suppose we want to understand how **gender** and **marital status** jointly affect wages

- We can estimate this relationship using an **interaction term**:

. . .

$$\ln(wage) = \beta_0 + \beta_1(female) + \beta_2(married) + \beta_3(female \times married) + \mu$$

. . .

- $female \times married$ is the **interaction term**
- It equals **1** only when *both* are true (the person is female AND married)
- It equals **0** for everyone else

## Identifying the Baseline {.smaller}

- With dummy variables, we always need a **baseline group**: the group where all dummies equal zero

. . .

- Here, the baseline is **unmarried men** ($female = 0$, $married = 0$)

. . .

- $\beta_0$ is the average $\ln(wage)$ for this baseline group
- Every other coefficient is an effect **relative to unmarried men**

. . .

::: {.callout-tip}
## Key Habit
Before interpreting any coefficient, always ask: *who is the baseline?*
:::


## What Does Each Coefficient Mean? {.smaller}

Once we add the interaction, each coefficient has a **specific** meaning:

. . .

- $\beta_1$ = effect of being female **when $married = 0$** (relative to unmarried men)

. . .

- $\beta_2$ = effect of being married **when $female = 0$** (relative to unmarried men)

. . .

- $\beta_3$ = the **additional** effect of being *both* female and married, beyond the sum of the individual effects (relative to unmarried men)

## Plugging in the Values

| Group | female | married | Predicted $\ln(wage)$ |
|:------|:------:|:-------:|:----------------------|
| Unmarried men (baseline) | 0 | 0 | $\beta_0$ |

## Plugging in the Values

| Group | female | married | Predicted $\ln(wage)$ |
|:------|:------:|:-------:|:----------------------|
| Unmarried men (baseline) | 0 | 0 | $\beta_0$ |
| Unmarried women | 1 | 0 | $\beta_0 + \beta_1$ |

## Plugging in the Values

| Group | female | married | Predicted $\ln(wage)$ |
|:------|:------:|:-------:|:----------------------|
| Unmarried men (baseline) | 0 | 0 | $\beta_0$ |
| Unmarried women | 1 | 0 | $\beta_0 + \beta_1$ |
| Married men | 0 | 1 | $\beta_0 + \beta_2$ |


## Plugging in the Values

| Group | female | married | Predicted $\ln(wage)$ |
|:------|:------:|:-------:|:----------------------|
| Unmarried men (baseline) | 0 | 0 | $\beta_0$ |
| Unmarried women | 1 | 0 | $\beta_0 + \beta_1$ |
| Married men | 0 | 1 | $\beta_0 + \beta_2$ |
| Married women | 1 | 1 | $\beta_0 + \beta_1 + \beta_2 + \beta_3$ |




## The Data {.smaller}

```{r}
#| echo: true

wage_data <- wooldridge::wage1 |>
    mutate(female_married = female * married)
```

```{r}
#| echo: false

wage_data |>
    mutate(ln_wage = round(log(wage), 3)) |>
    select(ln_wage, female, married, female_married) |>
    head(8) |>
    knitr::kable(col.names = c("ln(wage)", "female", "married", "female × married"),
                 align = "c")
```

Notice: `female × married` = 1 **only** when *both* dummies equal 1.


## Let's Estimate It {.smaller}

```{r}
#| echo: true

lm_interact <- lm(log(wage) ~ female + married + female_married, data = wage_data)
coef(lm_interact) |> round(3)
```

. . .

- **Unmarried women** earn `r paste0(round(100*coef(lm_interact)[2], 1))`% less than unmarried men
- **Married men** earn `r paste0(round(100*coef(lm_interact)[3], 1))`% more than unmarried men
- The interaction term ($\hat{\beta}_3 \approx `r round(coef(lm_interact)[4], 3)`$): marriage premium is **much smaller** for women


## Visualizing the Components {.smaller auto-animate=true}

```{r}
#| echo: false

coefs <- coef(lm_interact)

# Helper function for the progressive waterfall plot
# show_groups: how many groups to reveal (1-4)
plot_waterfall <- function(coefs, show_groups = 4) {

    # Full waterfall data for all 4 groups
    all_data <- tribble(
        ~group, ~component, ~start, ~end, ~coef_label, ~group_num,
        "Unmarried\nmen", "β0 (intercept)", 0, coefs[1], round(coefs[1], 2), 1,
        "Unmarried\nwomen", "β0 (intercept)", 0, coefs[1], NA_real_, 2,
        "Unmarried\nwomen", "β1 (female)", coefs[1], coefs[1] + coefs[2], round(coefs[2], 2), 2,
        "Married\nmen", "β0 (intercept)", 0, coefs[1], NA_real_, 3,
        "Married\nmen", "β2 (married)", coefs[1], coefs[1] + coefs[3], round(coefs[3], 2), 3,
        "Married\nwomen", "β0 (intercept)", 0, coefs[1], NA_real_, 4,
        "Married\nwomen", "β1 (female)", coefs[1], coefs[1] + coefs[2], NA_real_, 4,
        "Married\nwomen", "β2 (married)", coefs[1] + coefs[2], coefs[1] + coefs[2] + coefs[3], NA_real_, 4,
        "Married\nwomen", "β3 (interaction)", coefs[1] + coefs[2] + coefs[3], coefs[1] + coefs[2] + coefs[3] + coefs[4], round(coefs[4], 2), 4
    ) |>
        mutate(
            group = factor(group, levels = c("Unmarried\nmen", "Unmarried\nwomen", "Married\nmen", "Married\nwomen")),
            component = factor(component, levels = c("β0 (intercept)", "β1 (female)", "β2 (married)", "β3 (interaction)")),
            # Handle negative segments: swap so ymin < ymax
            ymin = pmin(start, end),
            ymax = pmax(start, end)
        )

    # Filter to only groups we want to show
    plot_data <- all_data |> filter(group_num <= show_groups)

    # Totals for each visible group (the final predicted value)
    totals <- plot_data |>
        group_by(group) |>
        summarise(total = last(end), .groups = "drop")

    # Split into two layers so adjustment segments are always visible on top
    intercept_bars <- plot_data |> filter(component == "β0 (intercept)")
    adjustment_bars <- plot_data |> filter(component != "β0 (intercept)")

    ggplot(plot_data, aes(x = group)) +
        # Layer 1: intercept bars (background)
        geom_rect(data = intercept_bars,
                  aes(xmin = as.numeric(group) - 0.35,
                      xmax = as.numeric(group) + 0.35,
                      ymin = ymin, ymax = ymax,
                      fill = component),
                  color = "white", linewidth = 0.5) +
        # Layer 2: adjustment bars (foreground — drawn on top of intercepts)
        geom_rect(data = adjustment_bars,
                  aes(xmin = as.numeric(group) - 0.35,
                      xmax = as.numeric(group) + 0.35,
                      ymin = ymin, ymax = ymax,
                      fill = component),
                  color = "white", linewidth = 0.5) +
        geom_text(data = plot_data |> filter(!is.na(coef_label)),
                  aes(x = group,
                      y = (ymin + ymax) / 2,
                      label = coef_label),
                  fontface = "bold", size = 4, color = "white") +
        geom_segment(data = totals,
                     aes(x = as.numeric(group) - 0.38,
                         xend = as.numeric(group) + 0.38,
                         y = total, yend = total),
                     linewidth = 1, color = "black") +
        geom_text(data = totals, aes(x = group, y = total,
                  label = paste0("= ", round(total, 2))),
                  hjust = -0.8,
                  vjust = -1, fontface = "bold", size = 4.5) +
        scale_fill_manual(
            values = c("β0 (intercept)" = "#002169",
                       "β1 (female)" = "#D55E00",
                       "β2 (married)" = "#0072B2",
                       "β3 (interaction)" = "#EEB808"),
            name = "Component",
            drop = FALSE
        ) +
        scale_x_discrete(drop = FALSE) +
        labs(x = NULL, y = "Predicted ln(wage)",
             title = "Building Up Predicted Wages from Coefficients") +
        coord_cartesian(ylim = c(0, 2.1)) +
        theme_minimal(base_size = 14) +
        theme(legend.position = "bottom",
              panel.grid.major.x = element_blank())
}
```

**Step 1:** Start with the baseline — **unmarried men** get only $\beta_0$

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

plot_waterfall(coefs, show_groups = 1)
```


## Visualizing the Components {.smaller auto-animate=true}

**Step 2:** **Unmarried women** = $\beta_0 + \beta_1$ — the [orange]{style="color: #D55E00; font-weight: bold;"} segment pulls their bar *down*

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

plot_waterfall(coefs, show_groups = 2)
```


## Visualizing the Components {.smaller auto-animate=true}

**Step 3:** **Married men** = $\beta_0 + \beta_2$ — the [blue]{style="color: #0072B2; font-weight: bold;"} marriage premium pushes their bar *up*

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

plot_waterfall(coefs, show_groups = 3)
```


## Visualizing the Components {.smaller auto-animate=true}

**Step 4:** **Married women** = $\beta_0 + \beta_1 + \beta_2 + \beta_3$ — all four components

- The [gold]{style="color: #EEB808; font-weight: bold;"} interaction ($\beta_3$) nearly wipes out the [blue]{style="color: #0072B2; font-weight: bold;"} marriage premium

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

plot_waterfall(coefs, show_groups = 4)
```

## Practice: Turn to Your Neighbor {.smaller .nonincremental}

Consider this model of home prices in Sacramento:

$$\ln(price) = \beta_0 + \beta_1(sqft) + \beta_2(Family) + \beta_3(Suburb) + \beta_4(Family \times Suburb) + \mu$$

where $Family = 1$ if the unit is detached, single-family residential house and $0$ if not, $Suburb = 1$ if the unit is outside the city limits of Sacramento

. . .

::: columns
::: {.column width="50%"}
**Discuss with your neighbor:**

1. What is the baseline group?
2. What does $\beta_2$ represent?
3. What does $\beta_3$ represent?
4. What is the suburb premium for single family homes?
5. What is the total effect of being a suburban single-family home (vs. baseline)?
:::
::: {.column width="50%"}
**Answers:**

1. Multi-family apartment in Sacramento
2. Price difference: Detached family home vs condo/apartment (in Sacramento)
3. Suburb premium for condos/apartments
4. $\beta_3 + \beta_4$
5. $\beta_2 + \beta_3 + \beta_4$
:::
:::



# Dummy × Continuous Interactions

## Different Slopes for Different Groups {.smaller}

- What if we want the **return to education** to differ by gender?

. . .

$$\ln(wage) = \beta_0 + \beta_1(female) + \beta_2(educ) + \beta_3(female \times educ) + \mu$$

. . .

- $female \times educ$ = 0 for all men, and = `educ` for all women

. . .

- This gives each group a **different intercept AND slope**:

## Two Separate Regression Lines {.smaller}

**Men** ($female = 0$):
$$\ln(wage) = \beta_0 + \beta_2(educ)$$

**Women** ($female = 1$):
$$\ln(wage) = \underbrace{(\beta_0 + \beta_1)}_{\text{different intercept}} + \underbrace{(\beta_2 + \beta_3)}_{\text{different slope}}(educ)$$

. . .

- $\beta_1$ = **intercept** difference (level shift for women at $educ = 0$)
- $\beta_3$ = **slope** difference (how much more/less women gain per year of education)

. . .

::: {.callout-important}
## vs. Dummy-Only Model
Without the interaction, both groups share the **same slope** (parallel lines). The interaction lets the lines have **different slopes** — they can converge, diverge, or even cross.
:::


## Simulated Example {.smaller}

To see this clearly, let's generate data where the difference is obvious:

- Men: 8% return to education; Women: 14% return
- Women start with a lower intercept

. . .

```{r}
#| echo: true

set.seed(42)
n <- 500

sim_data <- tibble(
    female = rep(c(0, 1), each = n/2),
    educ = round(runif(n, min = 8, max = 20)),
    error = rnorm(n, mean = 0, sd = 0.4),
    lwage = 0.8 - 0.90 * female + 0.08 * educ + 0.06 * female * educ + error
) |>
    mutate(female_educ = female * educ)
```

## Simulated Example {.smaller}

```{r}
#| echo: false

sim_data |>
    slice_sample(n = 10) |>
    select(lwage, female, educ, female_educ) |>
    head(8) |>
    mutate(lwage = round(lwage, 3)) |>
    knitr::kable(col.names = c("ln(wage)", "female", "educ", "female × educ"),
                 align = "c")
```

Notice: `female × educ` = 0 for all men, and = `educ` for all women.


## Estimating the Model {.smaller}

```{r}
#| echo: true

lm_slope <- lm(lwage ~ female + educ + female_educ, data = sim_data)
coef(lm_slope) |> round(3)
```

. . .

```{r}
#| echo: false
slope_coefs <- coef(lm_slope)
```

- Return to education for **men**: $\hat{\beta}_2 \approx `r round(slope_coefs[3], 3)`$ → `r round(100*slope_coefs[3], 1)`% per year
- Return to education for **women**: $\hat{\beta}_2 + \hat{\beta}_3 \approx `r round(slope_coefs[3] + slope_coefs[4], 3)`$ → `r round(100*(slope_coefs[3] + slope_coefs[4]), 1)`% per year
- The **interaction** ($\hat{\beta}_3 \approx `r round(slope_coefs[4], 3)`$): women gain `r round(100*slope_coefs[4], 1)` percentage points *more* per year of education


## Visualizing Different Slopes {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

sim_data <- sim_data |>
    mutate(gender_label = ifelse(female == 1, "Women", "Men"))

ggplot(sim_data, aes(x = educ, y = lwage, color = gender_label)) +
    geom_point(alpha = 0.2, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 1.3) +
    scale_color_manual(values = c("Men" = "#0072B2", "Women" = "#D55E00")) +
    labs(x = "Years of Education", y = "ln(wage)",
         title = "Different Returns to Education by Gender",
         color = NULL) +
    theme_minimal(base_size = 16) +
    theme(legend.position = "bottom")
```

- The lines are **not parallel** — women's steeper slope narrows the gap at high education levels


## Side-by-Side: Dummy-Only vs. Interaction {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5

# Dummy only model
lm_dummy_only <- lm(lwage ~ female + educ, data = sim_data)
d_coefs <- coef(lm_dummy_only)

p1_data <- sim_data |>
    mutate(
        pred = d_coefs[1] + d_coefs[2] * female + d_coefs[3] * educ
    )

p1 <- ggplot(p1_data, aes(x = educ, color = gender_label)) +
    geom_point(aes(y = lwage), alpha = 0.15, size = 1) +
    geom_line(aes(y = pred), linewidth = 1.3) +
    scale_color_manual(values = c("Men" = "#0072B2", "Women" = "#D55E00")) +
    labs(x = "Education", y = "ln(wage)", title = "Dummy Only (Parallel Lines)", color = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom")

# Interaction model
p2_data <- sim_data |>
    mutate(
        pred = slope_coefs[1] + slope_coefs[2] * female + slope_coefs[3] * educ + slope_coefs[4] * female_educ
    )

p2 <- ggplot(p2_data, aes(x = educ, color = gender_label)) +
    geom_point(aes(y = lwage), alpha = 0.15, size = 1) +
    geom_line(aes(y = pred), linewidth = 1.3) +
    scale_color_manual(values = c("Men" = "#0072B2", "Women" = "#D55E00")) +
    labs(x = "Education", y = "ln(wage)", title = "Interaction (Different Slopes)", color = NULL) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom")

library(patchwork)
p1 + p2
```


# Computing Total Effects

## The General Rule {.smaller}

When interactions are present, the **total effect** of a variable includes all terms where it appears:

. . .

$$\ln(wage) = \beta_0 + \beta_1(female) + \beta_2(educ) + \beta_3(female \times educ) + \mu$$

. . .

- Effect of education: $\frac{\partial \ln(wage)}{\partial educ} = \beta_2 + \beta_3(female)$
  - For men: $\beta_2$
  - For women: $\beta_2 + \beta_3$

. . .

- Effect of being female (i.e., gender wage gap): $\frac{\partial \ln(wage)}{\partial female} = \beta_1 + \beta_3(educ)$
  - **Depends on education level!**


## The Gender Gap at Different Education Levels {.smaller}

```{r}
#| echo: true

educ_levels <- c(8, 10, 12, 14, 16, 18)

tibble(
    `Education` = educ_levels,
    `Gender gap (log pts)` = round(slope_coefs[2] + slope_coefs[4] * educ_levels, 3),
    `Approx. % gap` = round(100 * (slope_coefs[2] + slope_coefs[4] * educ_levels), 1)
) |> knitr::kable()
```

. . .

::: {.callout-warning}
## Common Mistake
Never report just the main effect $\beta_1$ as "the effect of being female." With an interaction, the true effect is $\beta_1 + \beta_3(educ)$ — it depends on education!
:::


# Why You Need Main Effects

## What Goes Wrong Without Them? {.smaller}

What if we *only* include the interaction — no main effects?

$$\ln(wage) = \beta_0 + \beta_1(female \times married) + \mu$$

. . .

```{r}
#| echo: true

wage_data <- wooldridge::wage1 |>
    mutate(female_married = female * married)

lm_correct <- lm(log(wage) ~ female + married + female_married, data = wage_data)
lm_wrong   <- lm(log(wage) ~ female_married, data = wage_data)
```


## Comparing the Models {.smaller}

```{r}
#| echo: false
#| output: asis

modelsummary(
    fmt = 4,
    list("Correct" = lm_correct, "Interaction Only" = lm_wrong),
    stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
    gof_map = c("nobs", "adj.r.squared"),
    coef_map = c(
        "female" = "Female",
        "married" = "Married",
        "female_married" = "Female × Married"
    )
)
```


## What Can Each Model Predict? {.smaller}

```{r}
#| echo: false

ccoefs <- coef(lm_correct)
wcoefs <- coef(lm_wrong)

tribble(
    ~Group,
    ~`Correct: Formula`, ~`Correct: Value`,
    ~`Wrong: Formula`, ~`Wrong: Value`,
    "Unmarried men",
        "β₀", round(ccoefs[1], 3),
        "β₀", round(wcoefs[1], 3),
    "Unmarried women",
        "β₀ + β₁", round(ccoefs[1] + ccoefs[2], 3),
        "β₀", round(wcoefs[1], 3),
    "Married men",
        "β₀ + β₂", round(ccoefs[1] + ccoefs[3], 3),
        "β₀", round(wcoefs[1], 3),
    "Married women",
        "β₀ + β₁ + β₂ + β₃", round(ccoefs[1] + ccoefs[2] + ccoefs[3] + ccoefs[4], 3),
        "β₀ + β₁", round(wcoefs[1] + wcoefs[2], 3)
) |>
    knitr::kable()
```

. . .

- The interaction-only model gives the **same prediction** to unmarried men, unmarried women, and married men — it can't tell them apart!
- In other words, in lumps all other groups *except married women* into the same baseline group


## The Lesson {.smaller}

::: {.callout-warning}
## Always Include the Main Effects
When you include $X_1 \times X_2$ in a model, **always** also include $X_1$ and $X_2$ separately. Otherwise the interaction absorbs effects it was never meant to capture.
:::

. . .

| Feature | Dummy-only model | Interaction model |
|:--------|:----------------|:-----------------|
| **Intercept** | Different for each group | Different for each group |
| **Slope** | Same for all groups | Different for each group |
| **Assumption** | Effect of $X$ is the same across groups | Effect of $X$ can vary across groups |
| **Flexibility** | Less flexible | More flexible |


## Key Takeaways {.smaller}

::: {.nonincremental}

1. **Interaction terms** = the product of two variables on the RHS
2. **Dummy × Dummy**: the combination of two characteristics may have an effect ≠ sum of individual effects
3. **Dummy × Continuous**: allows different **slopes** (not just different intercepts) for each group
4. **Total effects** must include all terms where a variable appears — don't just report the main effect!
5. **Always include the main effects** alongside the interaction term

:::
