---
title: "Instrumental Variables"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: today
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    embed-resources: true
    fig-align: center
    code-overflow: wrap
    footer: "EC325 | Unit 14 | Instrumental Variables"
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(tidyverse)
library(fixest)
library(modelsummary)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 16) +
          theme(plot.title = element_text(face = "bold", size = 18),
                plot.subtitle = element_text(size = 13, color = "gray40"),
                panel.grid.minor = element_blank()))
```


## Econometrics So Far {.smaller}

- Causal inference toolkit:
  - **RCTs**: gold standard, but rarely available
  - **Multivariate OLS**: control variables for selection on observables
  - **Fixed effects**: control for time-invariant unobservables in panel data
  - **DiD**: control units instead of control variables
- Each of these requires *something* to work:
  - RCTs need randomization
  - FE needs panel data and time-invariant confounding
  - DiD needs a clean treatment vs. control comparison and parallel trends
- What if we have **none** of these? Just one cross-section, treatment is endogenous, no natural experiment in sight?
- Today: a different angle on causal inference --- **instrumental variables (IV)**


## The Promise of IV

- IV asks a *very* different question:

. . .

> *"Can we find some external force that 'shakes up' our endogenous variable in a way that has nothing to do with the things biasing OLS?"*

. . .

- If yes, we can use that external force --- the **instrument** --- to recover the causal effect

- If no, we have to keep looking



# The Endogeneity Problem


## When OLS Fails {.smaller}

- Recall the OLS unbiasedness condition: the **zero conditional mean assumption**

. . .

$$
E[\mu \mid X] = 0
$$

. . .

- When this fails --- when $X$ is correlated with $\mu$ --- $\hat{\beta}_{OLS}$ is biased
- We call the variable $X$ causing the problem an **endogenous** variable

. . .

- Three classic sources of endogeneity:
  1. **Omitted variable bias**
  2. **Selection bias**
  3. **Reverse causality**


## Source 1: Omitted Variable Bias {.smaller}

- We have discussed this a lot!
- Suppose we want to estimate the effect of **education on wages**

. . .

$$
wage_i = \beta_0 + \beta_1 educ_i + \mu_i
$$

. . .

- **Ability** affects both education choices and wages
- We can't measure ability $\rightarrow$ it gets absorbed into $\mu_i$
- Result: $Cov(educ, \mu) \neq 0$, OLS is biased
- This is the classic "ability bias"


## Source 2: Selection Bias {.smaller}

- Want to estimate the effect of **health insurance on health outcomes**

. . .

- Healthier people are *more likely* to purchase insurance
- People with insurance are healthier *because* they were healthier to begin with
- The comparison group is contaminated by selection

. . .

- Result: $Cov(insurance, \mu) \neq 0$, OLS overstates how much insurance "causes" good health


## Source 3: Reverse Causality {.smaller}

- Want to estimate the effect of **police on crime**

. . .

- High-crime areas get more police deployed
- Causation runs both ways: police $\to$ crime, *and* crime $\to$ police
- OLS conflates the two

. . .

- Result: positive correlation between police and crime --- making it look like police *cause* crime!


## Common Thread

- In all three cases, $X$ contains variation correlated with $\mu$
  - $X$ contains ability; $X$ is contaminated with selection bias; or $X$ reverse-causes $Y$
- We need a way to use **only the part of $X$ that is unrelated to $\mu$**

. . .

- Enter: **instrumental variables**



# The Instrumental Variables Idea


## Endogenous vs. Exogenous Variation {.smaller}

- Think of total variation in $X$ as having two types:

. . .

:::: {.columns}
::: {.column width="50%"}
**Bad ("endogenous") variation**

- The part of $X$ correlated with $\mu$
- This is what biases OLS
- We want to *throw it away*
:::

::: {.column width="50%"}
**Good ("exogenous") variation**

- The part of $X$ unrelated to $\mu$
- This is what we want
- We need to *isolate* it
:::
::::

. . .

- IV finds an external variable $Z$ --- the **instrument** --- that "zooms in" to exogenous variation in $X$
- We then use *only* that variation to estimate the causal effect


## What Makes an Instrument Valid?

- For $Z$ to work as an instrument, it must satisfy **two assumptions**

. . .

1. **Relevance**
2. **Exclusion restriction**

. . .

- Both must hold. If either one fails, the instrument does *not* isolate clean variation in $X$
- We'll take them one at a time


## Assumption 1: Relevance {.smaller}

::: {.callout-important}
## Relevance
$$Cov(Z, X) \neq 0$$

The instrument must move the endogenous variable.
:::

. . .

- The instrument must actually do *some* work in shifting $X$
- If $Z$ has no effect on $X$, then there's no exogenous variation to use --- IV gives nothing

. . .

- This is the **testable** IV assumption:
  - Regress $X$ on $Z$
  - Look at the first-stage F-statistic (more on this later in the deck)


## Assumption 2: Exclusion Restriction {.smaller}

::: {.callout-important}
## Exclusion Restriction
$$Cov(Z, \mu) = 0$$

The instrument affects $Y$ **only through** $X$. There is no direct effect of $Z$ on $Y$.
:::

. . .

- This is what turns a correlation into a causal effect
- It says: *the only way $Z$ can move $Y$ is by first moving $X$*

. . .

- This assumption is **not testable** --- there's no statistical test for it
- It has to be defended with **economic reasoning**
- Most of the work in any IV paper is defending this assumption



# A Running Example: Angrist & Evans (1998)


## The Research Question {.smaller}

- What is the causal effect of having an additional child on a mother's labor force participation?
- Important for:
  - Understanding the gender wage gap
  - Designing parental leave and childcare policy

- But fertility is **deeply endogenous**:

. . .

  - **Reverse causality**: women time fertility around their careers
  - **Selection**: less career-oriented women may choose larger families
  - **OVB**: unobserved preferences drive both fertility and work

- OLS regression of employment on number of kids tangles all of this together


## The Clever Instrument {.smaller}

- Angrist and Evans (1998) propose using the **sex composition of the first two children**

. . .

- The idea: parents have a slight preference for **mixed-sex children**
  - Two boys $\to$ slightly more likely to try for a third (hoping for a girl)
  - Two girls $\to$ same logic in reverse
  - One of each $\to$ less reason to have a third

- The sex of a child is essentially biological chance $\to$ same-sex vs. mixed-sex is **as good as random**


## Why This Is a Good Instrument {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Relevance** ($Cov(Z, X) \neq 0$)

- Same-sex parents *are* more likely to have a third child
- We can verify this in the data (and we will!)
- This is testable
:::

::: {.column width="50%"}
**Exclusion** ($Cov(Z, \mu) = 0$)

- Sex composition is essentially random with respect to career orientation, husband's earnings, etc.
- Hard to imagine a *direct* channel from "two boys" to "mom works less"
- Plausible, but **not testable**
:::
::::

. . .

- With the study in hand, we can now draw the diagram



# Graphical Intuition: The IV DAG


```{r}
#| label: dag-helpers
#| include: false

# Reusable function that draws (selected pieces of) the IV DAG.
# Toggle which elements appear with show_z / show_mu / show_excluded.
draw_iv_dag <- function(show_z = FALSE, show_mu = FALSE,
                        show_excluded = FALSE) {

    nodes <- tibble(
        label   = c("Z", "X", "Y", "\u03bc"),
        role    = c("Same-sex first\ntwo children",
                    "Had a\nthird child",
                    "Mother\nworks",
                    "Career orientation\n(unobserved)"),
        x       = c(1, 3, 5, 4),
        y       = c(0, 0, 0, 1.8),
        role_y  = c(-0.70, -0.70, -0.70, 2.65),
        fill_col = c("#0072B2", "#E69F00", "#009E73", "#CC79A7"),
        keep    = c(show_z, TRUE, TRUE, show_mu)
    ) |>
        dplyr::filter(keep)

    p <- ggplot() +
        # Always show X -> Y (the effect we want)
        geom_segment(aes(x = 3.30, xend = 4.70, y = 0, yend = 0),
                     color = "#1F5F8B", linewidth = 1.4,
                     arrow = arrow(length = unit(0.35, "cm"),
                                   type = "closed")) +
        annotate("text", x = 4, y = 0.30,
                 label = "Causal effect\nof interest (\u03b2)",
                 color = "#1F5F8B", size = 4.4, fontface = "bold",
                 lineheight = 0.95)

    if (show_mu) {
        p <- p +
            geom_segment(aes(x = c(3.82, 4.18),
                             xend = c(3.18, 4.82),
                             y = c(1.58, 1.58),
                             yend = c(0.22, 0.22)),
                         color = "#C1272D", linewidth = 1.1,
                         linetype = "dashed",
                         arrow = arrow(length = unit(0.28, "cm"),
                                       type = "closed")) +
            annotate("text", x = 3.28, y = 1.00, label = "Biases X",
                     color = "#C1272D", size = 4, fontface = "italic",
                     angle = 62) +
            annotate("text", x = 4.72, y = 1.00, label = "Biases Y",
                     color = "#C1272D", size = 4, fontface = "italic",
                     angle = -62)
    }

    if (show_z) {
        p <- p +
            geom_segment(aes(x = 1.30, xend = 2.70, y = 0, yend = 0),
                         color = "#1F5F8B", linewidth = 1.4,
                         arrow = arrow(length = unit(0.35, "cm"),
                                       type = "closed")) +
            annotate("text", x = 2, y = 0.30,
                     label = "First stage\nCov(Z, X) \u2260 0",
                     color = "#1F5F8B", size = 4.4, fontface = "bold",
                     lineheight = 0.95)
    }

    if (show_excluded) {
        p <- p +
            geom_segment(aes(x = 1.3, xend = 4.7,
                             y = -1.45, yend = -1.45),
                         color = "gray65", linewidth = 0.7,
                         linetype = "dashed",
                         arrow = arrow(length = unit(0.22, "cm"),
                                       type = "closed")) +
            geom_segment(aes(x = 2.85, xend = 3.15,
                             y = -1.62, yend = -1.28),
                         color = "#C1272D", linewidth = 2) +
            geom_segment(aes(x = 2.85, xend = 3.15,
                             y = -1.28, yend = -1.62),
                         color = "#C1272D", linewidth = 2) +
            annotate("text", x = 3, y = -2.00,
                     label = "Exclusion restriction:\nno direct arrow from Z to Y",
                     color = "gray25", fontface = "italic", size = 4,
                     lineheight = 0.95)
    }

    p +
        geom_point(data = nodes, aes(x = x, y = y, fill = fill_col),
                   shape = 21, size = 24, color = "gray25", stroke = 1.3) +
        geom_text(data = nodes, aes(x = x, y = y, label = label),
                  size = 9, color = "white", fontface = "bold") +
        geom_text(data = nodes, aes(x = x, y = role_y, label = role),
                  size = 4.2, color = "gray20", fontface = "bold",
                  lineheight = 0.95) +
        scale_fill_identity() +
        coord_cartesian(xlim = c(0.2, 5.8),
                        ylim = c(-2.4, 3.1), clip = "off") +
        theme_void() +
        theme(plot.margin = margin(10, 10, 10, 10))
}
```


## Step 1: The Confounding Problem {.smaller}

```{r}
#| label: dag-step1
#| fig-height: 4.8
#| fig-width: 9

draw_iv_dag(show_mu = TRUE)
```

- We *want* the $X \to Y$ arrow --- the parameter $\beta$
- But the unobserved confounder $\mu$ creates a **backdoor path**: $X \leftarrow \mu \to Y$
- Career-minded women have fewer kids *and* are more likely to work $\to$ OLS confuses the two channels


## Step 2: Bring in the Instrument {.smaller}

```{r}
#| label: dag-step2
#| fig-height: 4.8
#| fig-width: 9

draw_iv_dag(show_mu = TRUE, show_z = TRUE)
```

- $Z$ shifts $X$ but is *itself* unaffected by $\mu$
- $Z$ injects clean, exogenous variation into $X$ that we can use


## Step 3: The Crucial *Missing* Arrow {.smaller}

```{r}
#| label: dag-step3
#| fig-height: 5.2
#| fig-width: 9

draw_iv_dag(show_mu = TRUE, show_z = TRUE, show_excluded = TRUE)
```

- The exclusion restriction is what we *don't* see: no direct $Z \to Y$ arrow
- $Z$ can reach $Y$ **only** by going through $X$
- This is what makes IV identify $\beta$


# Simulating Angrist & Evans

```{r}
#| label: sim-data
#| output: false

set.seed(42)
n <- 50000

ae_sim <- tibble(
    # Unobserved "career orientation" --- affects both fertility and work
    career_orientation = rnorm(n, 0, 1),

    # Same-sex instrument (random)
    samesex = rbinom(n, 1, 0.5),

    # More than 2 kids: affected by samesex AND career orientation
    morekids_latent = -0.5 + 0.12 * samesex - 0.3 * career_orientation +
        rnorm(n, 0, 0.5),
    morekids = as.integer(morekids_latent > 0),

    # Labor force participation: affected by morekids AND career orientation
    # True causal effect of morekids is -0.10
    worked_latent = 0.6 - 0.10 * morekids + 0.25 * career_orientation +
        rnorm(n, 0, 0.3),
    mom_worked = as.integer(worked_latent > 0.5)
) |>
    mutate(
        samesex_lab  = if_else(samesex == 1,  "Same sex",   "Mixed sex"),
        morekids_lab = if_else(morekids == 1, "> 2 kids", "\u2264 2 kids")
    )

# Fit OLS up-front so it's available throughout the deck
# (the OLS vs. 2SLS section re-fits it visibly later)
ols <- lm(mom_worked ~ morekids, data = ae_sim)
```


## The Simulated Setup {.smaller}

- $N = 50{,}000$ mothers (synthetic data, but designed to match Angrist-Evans)
- We *know* the truth because we built it:

. . .

| Variable | Description | Effect on... |
|:---------|:------------|:-------------|
| `career_orientation` | Unobserved $\mu$ | $\downarrow$ fertility, $\uparrow$ work |
| `samesex` | Instrument $Z$ | $\uparrow$ fertility |
| `morekids` | Treatment $X$ | $\downarrow$ work (true $\beta = -0.10$) |
| `mom_worked` | Outcome $Y$ | --- |

. . .

- Goal: see whether OLS recovers $\beta = -0.10$. (Spoiler: it doesn't. IV does.)


## Testing Relevance {.smaller}

```{r}
#| echo: true

relevance_check <- ae_sim |>
    group_by(samesex_lab) |>
    summarize(prop_morekids = mean(morekids))

relevance_check
```

. . .

- Same-sex parents are noticeably more likely to have a third child


## Testing Relevance {.smaller}

```{r}
#| label: fig-relevance
#| fig-width: 8
#| fig-height: 4.5

ggplot(relevance_check, aes(x = samesex_lab, y = prop_morekids)) +
    geom_col(fill = "#0072B2", alpha = 0.85, width = 0.55) +
    geom_text(aes(label = scales::percent(prop_morekids, accuracy = 0.1)),
              vjust = -0.5, size = 6, fontface = "bold", color = "#1F5F8B") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.4)) +
    labs(title = "Relevance: Same-Sex Predicts Fertility",
         subtitle = "Same-sex parents about 6 percentage points more likely to have 3+ kids",
         x = "Sex composition of first two children",
         y = "Share with > 2 kids")
```


## "Testing" Excludability {.smaller}

- We **can't** directly test the exclusion restriction
- But we can do a sanity check: is the *reduced-form* effect of $Z$ on $Y$ small?

. . .

```{r}
#| label: fig-exclusion
#| fig-width: 8
#| fig-height: 4

exclusion_check <- ae_sim |>
    group_by(samesex_lab) |>
    summarize(prop_worked = mean(mom_worked))

ggplot(exclusion_check, aes(x = samesex_lab, y = prop_worked)) +
    geom_col(fill = "#E69F00", alpha = 0.85, width = 0.55) +
    geom_text(aes(label = scales::percent(prop_worked, accuracy = 0.1)),
              vjust = -0.5, size = 6, fontface = "bold", color = "#a06700") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.7)) +
    labs(title = "Exclusion check: small reduced-form effect",
         subtitle = "Effect of Z on Y should be small (and only operate through X)",
         x = "Sex composition of first two children",
         y = "Share of mothers working")
```


## Reading the Reduced-Form

- The difference is small (around a percentage point)
- Consistent with the effect operating through fertility, not directly

. . .

- This **doesn't prove** excludability --- we still rely on the economic argument
- But it's at least *consistent* with the assumption



# Two-Stage Least Squares (2SLS)


## The 2SLS Strategy {.smaller}

- IV is typically estimated using **Two-Stage Least Squares (2SLS)**

. . .

**Stage 1 (First Stage):** regress endogenous variable on the instrument

$$
X_i = \pi_0 + \pi_1 Z_i + \nu_i
$$

. . .

Get predicted values: $\hat{X}_i = \hat{\pi}_0 + \hat{\pi}_1 Z_i$

These predicted values contain *only* the variation in $X$ that comes from $Z$

. . .

**Stage 2 (Second Stage):** regress outcome on the predicted values

$$
Y_i = \beta_0 + \beta_1 \hat{X}_i + \varepsilon_i
$$

The coefficient $\hat{\beta}_1^{2SLS}$ is our causal estimate


## Why Does 2SLS Work? {.smaller}

```{r}
#| label: fig-variation
#| fig-width: 9
#| fig-height: 4.5

variation_data <- tibble(
    type = c("Total variation\nin X",
             "Exogenous variation\n(from Z)",
             "Endogenous variation\n(correlated with \u03bc)"),
    value = c(100, 40, 60),
    fill = c("total", "good", "bad")
) |>
    mutate(type = factor(type, levels = type))

ggplot(variation_data, aes(x = type, y = value, fill = fill)) +
    geom_col(width = 0.55) +
    geom_text(aes(label = paste0(value, "%")),
              vjust = -0.5, size = 6, fontface = "bold") +
    scale_fill_manual(values = c("total" = "gray60",
                                  "good"  = "#0072B2",
                                  "bad"   = "#CC79A7")) +
    scale_y_continuous(limits = c(0, 120)) +
    labs(title = "2SLS uses only the 'good' variation",
         subtitle = "OLS uses all variation; IV throws away the part correlated with \u03bc",
         x = NULL, y = "Share of variation (%)") +
    theme(legend.position = "none") +
    annotate("text", x = 2, y = 55, label = "IV\nuses\nthis",
             color = "#0072B2", fontface = "bold", size = 5) +
    annotate("text", x = 3, y = 75, label = "IV\ndiscards\nthis",
             color = "#CC79A7", fontface = "bold", size = 5)
```


## Manual 2SLS: Step 1 {.smaller}

```{r}
#| echo: true

# First stage: regress endogenous variable on instrument
first_stage <- lm(morekids ~ samesex, data = ae_sim)

summary(first_stage)$coefficients
```

. . .

- The first-stage coefficient on `samesex` says: if your first two kids are the same sex, your probability of having a third rises by ~6 percentage points
- The F-statistic on the instrument is well into the hundreds: a strong first stage


## Manual 2SLS: Step 2 {.smaller}

```{r}
#| echo: true

# Get predicted values from first stage
ae_sim <- ae_sim |>
    mutate(morekids_hat = predict(first_stage))

ae_sim |>
    select(samesex, morekids, morekids_hat) |>
    head(8)
```

- `morekids_hat` takes only **two values** --- one for each value of `samesex`
- This is the *exogenous part* of `morekids`


## Manual 2SLS: Step 3 {.smaller}

```{r}
#| echo: true

# Second stage: regress outcome on predicted values
second_stage_manual <- lm(mom_worked ~ morekids_hat, data = ae_sim)
coef(second_stage_manual)["morekids_hat"]
```

. . .

::: {.callout-warning}
## Standard error warning
The manual two-stage approach gives the *correct coefficient*, but the **standard errors are wrong** --- they don't account for first-stage uncertainty.

**Always use a proper 2SLS estimator** (e.g. `feols()`).
:::



# Estimating IV with `feols()`


## The `feols()` 4-Part Formula {.smaller}

- `fixest::feols()` uses a formula with **three vertical bars**:

. . .

```r
outcome ~ exogenous_controls | fixed_effects | endogenous ~ instrument(s)
```

. . .

| Slot | What goes here |
|:-----|:---------------|
| `outcome` | Dependent variable $Y$ |
| `exogenous_controls` | Covariates that are *not* endogenous; use `1` if none |
| `fixed_effects` | Variables to absorb as FEs; use `0` if none |
| `endogenous ~ instrument(s)` | The first-stage formula |


## A Bare-Bones IV {.smaller}

```{r}
#| echo: true

tsls <- feols(mom_worked ~ 1 |  0  | morekids ~ samesex,
              data = ae_sim)

summary(tsls)
```


## What to Look At {.smaller}

- **`fit_morekids`**: the 2SLS estimate of $\beta$ (the `fit_` prefix flags the fitted endogenous variable)
- **Std. error**: the *correct* 2SLS SE (not the broken manual one)
- **F-test (1st stage)**: a built-in instrument-strength diagnostic
- The 2SLS estimate should be close to the true $-0.10$ --- much closer than OLS will get


## Inspecting the First Stage {.smaller}

- `feols()` runs the first stage internally but hides it. Pass `stage = 1` to expose it:

. . .

```{r}
#| echo: true

summary(tsls, stage = 1)
```


## Adding Controls {.smaller}

- Exogenous controls go in the **first slot**
- They are automatically included in *both* stages

. . .

```r
# Illustrative
tsls_ctrl <- feols(mom_worked ~ age + education |
                       0 |
                       morekids ~ samesex,
                   data = ae_sim)
```

. . .

- Use this for variables you want to control for that aren't themselves endogenous (mother's age, education, etc.)


## Adding Fixed Effects {.smaller}

- Fixed effects go in the **second slot**
- Absorbed efficiently --- much faster than creating dummies on big data

. . .

```r
# Illustrative
tsls_fe <- feols(mom_worked ~ age + education |
                     state |
                     morekids ~ samesex,
                 data = ae_sim)
```

. . .

- Equivalent to including state dummies in *both* stages, but vastly faster


## Multiple Instruments {.smaller}

- Adding a second instrument is just a `+` in the first-stage formula

. . .

```r
# Illustrative: add 'twins' as a second instrument for fertility
tsls_multi <- feols(mom_worked ~ 1 |
                        0 |
                        morekids ~ samesex + twins,
                    data = ae_sim)
```

. . .

**Why might you want multiple instruments?**

- More first-stage variation in $X$ $\to$ generally smaller standard errors and tighter inference
- Lets you *cross-check* your instruments: if two supposedly valid instruments produce very different 2SLS estimates, at least one is probably bad

. . .

**What does "overidentified" mean?**

- More instruments than endogenous variables $\to$ the model is **overidentified**
- 2SLS combines the instruments into a single weighted prediction of $X$

. . .

**But more instruments aren't free**

- Each new instrument carries its **own exclusion restriction** to defend --- one bad instrument contaminates the whole estimate
- The "**many weak instruments**" problem: adding marginally-relevant instruments can *amplify* finite-sample bias and drag 2SLS back toward OLS
- A few strong instruments usually beat many weak ones


## Cluster-Robust Standard Errors {.smaller}

- IV standard errors are sensitive to within-cluster correlation
- `feols()` makes this easy:

. . .

```r
# Cluster at estimation time
feols(mom_worked ~ 1 | 0 | morekids ~ samesex,
      cluster = ~ state,
      data = ae_sim)

# Or after the fact
summary(tsls, cluster = ~ state)
```

. . .

- Same advice as for DiD: cluster at the level of treatment assignment


# Weak Instruments


## Why Weak Instruments Are Dangerous {.smaller}

- If the instrument is only *weakly* correlated with $X$:
  - Tiny changes in the data produce wild swings in $\hat{\beta}^{2SLS}$
  - Estimates can become biased *toward* OLS (the very thing we were trying to escape!)
  - Standard errors become unreliable
  - Inference breaks

. . .

::: {.callout-warning}
2SLS with a weak instrument is sometimes *worse* than just running OLS.
:::


## The First-Stage F-Test {.smaller}

- The standard diagnostic for instrument strength is the **first-stage F-statistic**
- It tests $H_0: \pi_1 = 0$ in the first-stage regression

. . .

$$
X_i = \pi_0 + \pi_1 Z_i + \nu_i
$$

. . .

- **Single instrument**: F is just the squared t-statistic on $\pi_1$
- **Multiple instruments**: F is the joint test that *none* of them predict $X$
- A large F means the instruments collectively move $X$ around a lot


## The Traditional "F > 10" Rule {.smaller}

- Long-standing rule of thumb: a first-stage **F above 10** is "strong enough"
- Comes from Stock and Yogo (2005)
- Has been the textbook standard for two decades

. . .

::: {.callout-note}
## The classic rule
$F \geq 10 \implies$ instrument is strong enough to trust 2SLS.
:::


## Recent Evidence: 10 Is Too Lenient {.smaller}

- Lee, McCrary, Moreira, and Porter (2022) show: to justify standard t-test inference on the 2SLS coefficient, you need an F closer to **100** than 10

. . .

- An instrument with $F = 20$ may have fine point estimates, but its confidence intervals are **too narrow**
- Lots of "significant" results in IV papers may not really be

. . .

::: {.callout-important}
## A safer modern default
Treat $F \in [10, 100]$ with caution. Prefer instruments with much higher first-stage F when you can find them.
:::


## What Happens with a Weak Instrument? {.smaller}

```{r}
#| label: weak-iv-setup
#| echo: true

# Replace samesex with random noise --- a useless instrument
set.seed(123)
ae_sim <- ae_sim |>
    mutate(weak_instrument = rnorm(n()))

weak_iv <- feols(mom_worked ~ 1 | 0 | morekids ~ weak_instrument,
                 data = ae_sim)

summary(weak_iv)
```


## Strong vs. Weak: 100 Random Instruments {.smaller}

- Was that one wild fit a fluke? Let's run 2SLS with **100 different random instruments** and see how stable the estimates are.

. . .

```{r}
#| label: fig-weak-iv
#| fig-width: 9.5
#| fig-height: 4.6

set.seed(2026)
n_sims <- 100

# 100 different "instruments", each one pure noise
weak_estimates <- replicate(n_sims, {
    df <- ae_sim |> mutate(weak_z = rnorm(nrow(ae_sim)))
    coef(feols(mom_worked ~ 1 | 0 | morekids ~ weak_z, data = df))["fit_morekids"]
})

mc_df <- bind_rows(
    tibble(Model = "OLS",                                estimate = coef(ols)["morekids"]),
    tibble(Model = "Strong 2SLS\n(same-sex)",            estimate = coef(tsls)["fit_morekids"]),
    tibble(Model = "Weak 2SLS\n(100 random instruments)", estimate = weak_estimates)
) |>
    mutate(Model = factor(Model,
                          levels = c("OLS",
                                     "Strong 2SLS\n(same-sex)",
                                     "Weak 2SLS\n(100 random instruments)")))

# Cap the x-axis at +/-3; note how many weak-IV draws fall outside
x_lo <- -3; x_hi <- 3
n_offscreen <- sum(weak_estimates < x_lo | weak_estimates > x_hi)

ggplot(mc_df, aes(x = estimate, y = Model, color = Model)) +
    geom_vline(xintercept = -0.10, linetype = "dashed",
               color = "darkgreen", linewidth = 1) +
    geom_jitter(height = 0.18, alpha = 0.65, size = 2.8) +
    annotate("text", x = -0.10, y = 3.55,
             label = "True effect = -0.10",
             color = "darkgreen", size = 4, hjust = -0.06) +
    scale_color_manual(values = c(
        "OLS"                                   = "#CC79A7",
        "Strong 2SLS\n(same-sex)"               = "#0072B2",
        "Weak 2SLS\n(100 random instruments)"   = "#D55E00")) +
    coord_cartesian(xlim = c(x_lo, x_hi), clip = "off") +
    labs(title = "Weak instruments are essentially random number generators",
         subtitle = sprintf(
             "OLS and strong 2SLS are stable; weak 2SLS scatters wildly (%d of 100 draws fall outside [−3, 3])",
             n_offscreen),
         x = "2SLS coefficient on `morekids`",
         y = NULL) +
    theme(legend.position = "none",
          plot.margin = margin(10, 30, 10, 10))
```



## The Lesson

- **Always check your first-stage F-statistic**
- A weak instrument can be more misleading than no instrument
- So if you have a weak instrument, you need to stop



# IV Diagnostics


## Computing the Diagnostics {.smaller}

- After estimating, pull out the IV diagnostics with `fitstat()`:

. . .

```{r}
#| echo: true

fitstat(tsls, ~ ivf + ivwald + wh)
```


## What Each One Tests {.smaller}

::: {.callout-note}
## `ivf` --- First-stage F (homoskedastic)
**Null:** the instrument doesn't predict $X$ ($\pi_1 = 0$).
**Reads as:** large F = strong first stage; small F = weak instrument problem.
**Use:** the standard relevance check.
:::

. . .

::: {.callout-note}
## `ivwald` --- Robust first-stage F (Wald form)
**Null:** same as `ivf`, but computed with the robust variance-covariance matrix.
**Reads as:** with clustered or heteroskedasticity-robust SEs, this is the version you should trust --- the homoskedastic `ivf` can mislead.
**Use:** report instead of `ivf` whenever you cluster.
:::

. . .

::: {.callout-note}
## `wh` --- Wu-Hausman endogeneity test
**Null:** $X$ is exogenous (so OLS would be unbiased).
**Reads as:** compares OLS and 2SLS coefficients. Similar $\to$ fail to reject $\to$ OLS may suffice. Very different $\to$ reject $\to$ IV is doing necessary work.
**Use:** sanity-check that you actually *need* IV in the first place.
:::



# OLS vs. 2SLS


## Side-by-Side {.smaller}

```{r}
#| label: ols-fit
#| echo: true

# OLS for comparison (biased)
ols <- lm(mom_worked ~ morekids, data = ae_sim)
```

. . .

```{r}
#| echo: true

modelsummary::modelsummary(
    list("OLS (biased)" = ols, "2SLS (IV)" = tsls),
    coef_map = c("morekids" = "morekids", "fit_morekids" = "morekids"),
    gof_map = c("nobs", "r.squared"),
    output = "markdown"
)
```


## OLS vs. 2SLS: Visual {.smaller}

```{r}
#| label: fig-ols-vs-iv
#| fig-width: 9
#| fig-height: 4.5

coef_comp <- tibble(
    Model = c("OLS (Biased)", "2SLS (IV)"),
    Estimate = c(coef(ols)["morekids"], coef(tsls)["fit_morekids"]),
    SE = c(summary(ols)$coefficients["morekids", "Std. Error"],
           summary(tsls)$se["fit_morekids"])
) |>
    mutate(CI_lower = Estimate - 1.96 * SE,
           CI_upper = Estimate + 1.96 * SE)

ggplot(coef_comp, aes(x = Model, y = Estimate, color = Model)) +
    geom_hline(yintercept = -0.10, linetype = "dashed",
               color = "darkgreen", linewidth = 1) +
    geom_point(size = 5) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                  width = 0.15, linewidth = 1) +
    annotate("text", x = 2.35, y = -0.10, label = "True effect = -0.10",
             color = "darkgreen", hjust = 0, size = 5) +
    scale_color_manual(values = c("OLS (Biased)" = "#CC79A7",
                                   "2SLS (IV)"    = "#0072B2")) +
    coord_flip(clip = "off") +
    labs(title = "OLS overshoots; 2SLS recovers the truth",
         subtitle = "Selection on career orientation makes OLS too negative",
         y = "Effect on labor force participation",
         x = NULL) +
    theme(legend.position = "none",
          plot.margin = margin(10, 80, 10, 10))
```


## What Went Wrong with OLS? {.smaller}

- OLS estimate is *more negative* than the truth
- Because of selection on **career orientation**:
  - Less career-oriented women have more children *and* work less
  - $\rightarrow$ OLS makes children look worse for employment than they really are
- 2SLS uses only the random variation from same-sex composition $\rightarrow$ recovers the truth



# Classic IV Applications


## A Map of the Greatest Hits {.smaller}

| Study | Question | Instrument |
|:------|:---------|:-----------|
| **Card (1995)** | Returns to education | Distance to college |
| **Acemoglu, Johnson & Robinson (2001)** | Do institutions cause growth? | Colonial settler mortality |
| **Levitt (1997)** | Effect of police on crime | Electoral timing |
| **Angrist & Krueger (1991)** | Returns to education | Quarter of birth |

. . .

- Each one uses the same recipe: question $\to$ source of endogeneity $\to$ instrument $\to$ defend exclusion


## Card (1995): Returns to Education {.smaller}

- **Question:** What is the causal effect of education on wages?
- **Problem:** Ability is unobserved; biases OLS upward
- **Instrument:** Distance to the nearest college growing up
- **Why relevant?** Proximity reduces the cost of attending college
- **Why excludable?** Distance to a college shouldn't directly affect wages decades later --- only through education

. . .

- **Finding:** Returns to education are *higher* than OLS suggests
  - Ability bias works in the opposite direction from what many had assumed!


## Acemoglu, Johnson & Robinson (2001): Institutions {.smaller}

- **Question:** Do institutions (rule of law, property rights) cause long-run growth?
- **Problem:** Rich countries can afford better institutions $\rightarrow$ reverse causality
- **Instrument:** Colonial settler mortality rates (17th--19th century)
  - High mortality $\rightarrow$ extractive colonies $\rightarrow$ bad institutions
  - Low mortality $\rightarrow$ settler colonies $\rightarrow$ inclusive institutions
- **Why relevant?** Colonial institutional form persisted to the present
- **Why excludable?** European mortality from disease 200+ years ago shouldn't directly affect today's GDP --- only through the institutions it shaped

. . .

- **Finding:** Institutions have a *large* causal effect on GDP per capita; sparked the "institutions vs. geography" debate


## Levitt (1997): Police and Crime {.smaller}

- **Question:** Does hiring more police cause crime to fall?
- **Problem:** Reverse causality --- high-crime cities hire more police, making OLS look like police *cause* crime
- **Instrument:** Timing of mayoral and gubernatorial elections (police hiring spikes pre-election)
- **Why relevant?** Police staffing rises predictably before elections
- **Why excludable?** The electoral calendar is set by law, not by recent crime trends

. . .

- **Finding:** More police *substantially* reduce crime --- IV estimate flips the (wrong-signed) OLS correlation
- A coding error was later corrected by McCrary (2002); the violent-crime effect survives but is smaller


## Angrist & Krueger (1991): Quarter of Birth {.smaller}

- **Question:** What is the causal return to schooling on earnings?
- **Problem:** Ability bias (same as Card)
- **Instrument:** Quarter of birth, interacted with compulsory schooling laws
  - Q1 babies start school *older*, hit dropout age with *less* education
  - Q4 babies start school *younger*, are forced to stay in *longer*
- **Why relevant?** First-stage relationship is detectable but modest
- **Why excludable?** Quarter of birth is essentially random with respect to ability/family background

. . .

- **Finding:** IV estimates close to OLS $\rightarrow$ ability bias may be smaller than feared
- Bound, Jaeger, and Baker (1995) later showed quarter of birth is a *weak* instrument --- one of the motivating examples for the weak-IV literature



# Wrapping Up


## Key Takeaways {.smaller}

::: {.callout-important}
## What you need for IV to work
**1. Relevance** ($Cov(Z, X) \neq 0$): the instrument predicts $X$. Testable via the first-stage F.

**2. Exclusion** ($Cov(Z, \mu) = 0$): $Z$ affects $Y$ only through $X$. *Not* testable --- defend with economic reasoning.
:::

. . .

- 2SLS uses *only* the exogenous variation in $X$ induced by $Z$
- Weak instruments produce unreliable estimates --- check the first-stage F (and don't trust 10!)
- `feols()` makes 2SLS easy: `outcome ~ controls | FEs | endogenous ~ instrument`

