---
title: "Multivariate Regression and OLS"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 5 | Multivariate Regression and OLS"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


# From Bivariate to Multivariate OLS

## Where We've Been {.smaller}

- Last unit: **bivariate OLS** — one independent variable

. . .

$$y = \beta_0 + \beta_1 x + \mu$$

. . .

- We learned OLS is **unbiased** when GM assumptions 1--4 are met:

1. **Linear in parameters:** $y = \beta_0 + \beta_1 x + \mu$
2. **Random sampling**
3. **Sample variation in $x$**
4. **Zero conditional mean:** $E[\mu | x] = 0$

. . .

- The most important (and hardest to satisfy) is **GM4** — it requires that *everything else* affecting $y$ is uncorrelated with $x$

. . .

- That's a **very strong** assumption — almost never holds in observational data!


## The Problem: A Concrete Example {.smaller}

Consider the effect of education on wages:

$$wage = \beta_0 + \beta_1 \cdot educ + \mu$$

. . .

- The error term $\mu$ contains: ability, family connections, motivation, experience, ...

. . .

- Many of these are **correlated with education**
  - Wealthier families → more education *and* higher wages

. . .

::: {.callout-important}
## The Core Problem
If confounders are in $\mu$ and correlated with $x$, then $E[\mu | x] \neq 0$ and $\hat{\beta}_1$ is **biased**.
:::


## Where We're Going

- **Today:** Multivariate OLS — adding **control variables** to address confounding

. . .

- By including variables that might be confounders, we can make the zero conditional mean assumption more plausible

. . .

::: {.callout-important}
## The Big Question
How does adding control variables help us get closer to **causal** estimates?
:::


# The Multivariate Model

## The Solution: Control for Confounders

- If family income confounds the education-wage relationship, **include it**:

. . .

$$wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$$

. . .

- Now $\beta_1$ is the effect of education **holding family income constant**

- Family income is no longer in $\mu$ — that source of bias is gone!


## The General Model {.smaller}

The multivariate linear regression model with $k$ independent variables:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

. . .

where:

- $y$ is the **dependent variable**
- $x_1, x_2, \ldots, x_k$ are the **independent variables**
- $\beta_j$ is the coefficient on $x_j$
- $\mu$ is the error term — now only contains factors **not included** in the model


## Interpreting Coefficients

- Each coefficient has a **ceteris paribus** interpretation:

. . .

$$\beta_j = \frac{\partial y}{\partial x_j}$$

. . .

- The change in $y$ for a one-unit increase in $x_j$, **holding all other $x$'s constant**

. . .

- This is the key difference from bivariate OLS!
  - Bivariate: "associated with" (everything else floating)
  - Multivariate: "holding constant" (controlling for other factors)


## Interpretation Example {.smaller}

$$wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$$

. . .

- $\beta_1$: Each additional year of education is associated with $\beta_1$ dollars more in wages, **holding family income constant**

. . .

- $\beta_2$: Each additional dollar of family income is associated with $\beta_2$ dollars more in wages, **holding education constant**

## Deriving the Multivariate OLS Estimator {.smaller}

Same logic as bivariate OLS: **minimize the sum of squared residuals**.

. . .

$$SSR = \sum_{i=1}^{n} (y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \hat{\beta}_2 x_{2i} - \cdots - \hat{\beta}_k x_{ki})^2$$

. . .

Taking partial derivatives with respect to each $\hat{\beta}_j$ and setting equal to zero gives $k + 1$ **first-order conditions**:

$$\sum_{i=1}^{n}(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \cdots - \hat{\beta}_k x_{ki}) = 0$$

$$\sum_{i=1}^{n} x_{ji}(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \cdots - \hat{\beta}_k x_{ki}) = 0 \quad \text{for } j = 1, \ldots, k$$

. . .

Solving this system yields the OLS estimates. The algebra is tedious (typically handled with matrix methods), but your computer does this effortlessly with `lm()`.


# Multivariate OLS in R

## Adding Controls is Easy! {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

Same `lm()` function — just use `+` to add variables:

. . .

```{r}
#| echo: true

data(mtcars)

# Bivariate: mpg on horsepower
lm_biv <- lm(mpg ~ hp, data = mtcars)

# Multivariate: add weight as control
lm_multi <- lm(mpg ~ hp + wt, data = mtcars)
```


## Comparing Results {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Bivariate:**
```{r}
#| echo: true
summary(lm_biv)
```
:::

::: {.column width="50%"}
**Multivariate:**
```{r}
#| echo: true
summary(lm_multi)
```
:::

::::

. . .

- Coefficient on `hp` dropped from **-0.068** to **-0.032** — less than half!

. . .

- Why? Horsepower and weight are correlated, and weight independently reduces MPG

- The bivariate estimate was **conflating** the effect of hp with the effect of weight


## Visualizing: The Bivariate View {.smaller}

```{r}
#| echo: false
#| fig-height: 4.5
#| fig-align: center

ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#E63946", linewidth = 1.3) +
  labs(title = "Bivariate: mpg ~ hp",
       subtitle = paste0("β_hp = ", round(coef(lm_biv)["hp"], 3),
                         " — each extra HP associated with ",
                         round(coef(lm_biv)["hp"], 3), " fewer MPG"),
       x = "Horsepower", y = "MPG") +
  theme_minimal(base_size = 16)
```

This is the familiar 2D picture — but it ignores weight entirely. Is the slope **too steep** because heavier cars have more HP *and* worse MPG?


## Visualizing: Adding the Third Dimension {.smaller}

```{r}
#| echo: false

library(plotly)

# Create grid for the regression plane
hp_grid <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 20)
wt_grid <- seq(min(mtcars$wt), max(mtcars$wt), length.out = 20)
grid <- expand.grid(hp = hp_grid, wt = wt_grid)
grid$mpg_pred <- predict(lm_multi, newdata = grid)
plane_z <- matrix(grid$mpg_pred, nrow = 20, ncol = 20)

# Bivariate line at wt = 0 edge for comparison
hp_seq <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 50)
biv_line <- tibble(
  hp = hp_seq,
  wt = 0,
  mpg = coef(lm_biv)[1] + coef(lm_biv)["hp"] * hp_seq
)

plot_ly() |>
  add_markers(
    data = mtcars,
    x = ~hp, y = ~wt, z = ~mpg,
    marker = list(size = 5, color = "#002169", opacity = 0.8),
    text = ~paste0(rownames(mtcars),
                   "<br>HP: ", hp,
                   "<br>Weight: ", wt,
                   "<br>MPG: ", round(mpg, 1)),
    hoverinfo = "text",
    name = "Observations"
  ) |>
  add_trace(
    data = biv_line,
    x = ~hp, y = ~wt, z = ~mpg,
    type = "scatter3d", mode = "lines",
    line = list(color = "#E63946", width = 6, dash = "dash"),
    name = "Bivariate line"
  ) |>
  add_surface(
    x = hp_grid, y = wt_grid, z = ~plane_z,
    colorscale = list(c(0, "#EEB808"), c(1, "#EEB808")),
    opacity = 0.4, showscale = FALSE,
    name = "Multivariate plane"
  ) |>
  layout(
    scene = list(
      xaxis = list(title = "Horsepower"),
      yaxis = list(title = "Weight (1000 lbs)"),
      zaxis = list(title = "MPG"),
      camera = list(eye = list(x = -1.8, y = -1.4, z = 0.8))
    ),
    title = list(
      text = paste0("Multivariate: β_hp = ", round(coef(lm_multi)["hp"], 3),
                    " (was ", round(coef(lm_biv)["hp"], 3), " in bivariate)"),
      font = list(size = 14)),
    margin = list(t = 40)
  )
```

The dashed red line is the old bivariate fit — it sits flat, ignoring weight. The yellow **plane** accounts for both hp and weight. The hp slope flattened from **`r round(coef(lm_biv)["hp"], 3)`** to **`r round(coef(lm_multi)["hp"], 3)`**.


## Try It: Your First Multivariate Regression {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**Using `mtcars`, estimate a regression of `mpg` on `wt`, `hp`, and `drat` (rear axle ratio). Then answer:**

1. What is the coefficient on `wt`? Interpret it in a sentence.
2. How does the `wt` coefficient compare to a bivariate regression of `mpg ~ wt`?

. . .

```{r}
#| output-location: fragment
#| echo: true

lm(mpg ~ wt + hp + drat, data = mtcars) |> coef()
```

. . .

```{r}
#| echo: true

# Compare to bivariate
lm(mpg ~ wt, data = mtcars) |> coef()
```


# Multivariate OLS as Matching

## The Intuition {.smaller}

- How does multivariate OLS actually *control* for something? What is going on *under-the-hood*?

- When we "control for" a variable, we're **comparing like with like**

. . .

- Multivariate OLS is essentially a **matching** estimator with three steps:
  1. **Group** observations by the control variable (e.g., occupation)
  2. **Compute averages within each group** and compare $x$ and $y$ **within** each group
  3. **Take a weighted average** of the within-group comparisons across groups

. . .

- Think of it as making a bunch of separate "apples-to-apples" comparisons and then aggregating them!


## Example: Gender, Wages, and Occupation {.smaller}

- Let's construct an imaginary population where we **know the truth**:

. . .

- **2,000 workers** who are either Male or Female
- Each worker is in one of two occupations: **Office** or **Manual**
- Women are **more likely** to work in office jobs (70% vs. 40% for men)
- The **true** gender wage gap is **\$5,000** (men earn more, holding occupation constant)
- Office jobs pay a **\$20,000 premium** over manual jobs (independent of gender)
- Because gender affects *both* wages and occupation sorting, a naive comparison will **understate** the gender gap

## Example: Gender, Wages, and Occupation {.smaller}

```{r}
#| echo: true
#| code-line-numbers: false

set.seed(42)
n <- 2000

wage_data <- tibble(
  female = sample(c(0, 1), n, replace = TRUE),
  # Women more likely to be in office jobs
  office = rbinom(n, 1, prob = ifelse(female == 1, 0.70, 0.40)),
  # True gender gap: -$5000 for women; office premium: $20000
  wage = 30000 - 5000 * female + 20000 * office + rnorm(n, 0, 6000)
) |>
  mutate(gender = ifelse(female == 1, "Female", "Male"),
         occupation = ifelse(office == 1, "Office", "Manual"))
```


## Visualizing the Data in 3D {.smaller}

```{r}
#| echo: false

library(plotly)

# Jitter the categorical axes so points don't stack
wage_data_3d <- wage_data |>
  mutate(
    x = ifelse(female == 1, 1, 0) + runif(n(), -0.15, 0.15),
    z = ifelse(office == 1, 1, 0) + runif(n(), -0.15, 0.15),
    group = paste(gender, occupation, sep = " / ")
  )

# Cell means for large markers
cell_means_3d <- wage_data |>
  group_by(female, office, gender, occupation) |>
  summarise(avg_wage = mean(wage), .groups = "drop") |>
  mutate(group = paste(gender, occupation, sep = " / "))

plot_ly() |>
  # Individual points (translucent)
  add_markers(
    data = wage_data_3d,
    x = ~x, y = ~wage, z = ~z,
    color = ~group,
    colors = c("Female / Manual" = "#E87D72", "Female / Office" = "#C44E52",
               "Male / Manual" = "#72B5E8", "Male / Office" = "#4878A8"),
    marker = list(size = 2, opacity = 0.15),
    showlegend = FALSE
  ) |>
  # Cell means (large solid markers)
  add_markers(
    data = cell_means_3d,
    x = ~female, y = ~avg_wage, z = ~office,
    color = ~group,
    colors = c("Female / Manual" = "#E87D72", "Female / Office" = "#C44E52",
               "Male / Manual" = "#72B5E8", "Male / Office" = "#4878A8"),
    marker = list(size = 12, opacity = 1,
                  line = list(color = "black", width = 1)),
    text = ~paste0(group, "<br>Avg Wage: $", round(avg_wage, 0)),
    hoverinfo = "text"
  ) |>
  layout(
    scene = list(
      xaxis = list(title = "Gender",
                   tickvals = c(0, 1), ticktext = c("Male", "Female")),
      yaxis = list(title = "Wage ($)"),
      zaxis = list(title = "Occupation",
                   tickvals = c(0, 1), ticktext = c("Manual", "Office")),
      camera = list(eye = list(x = 1.8, y = 1.2, z = 1.0))
    ),
    title = list(text = "Wages by Gender and Occupation",
                 font = list(size = 16)),
    margin = list(t = 40)
  )
```

Each cluster is a **gender × occupation cell**. Large markers show the **cell means**. The matching approach compares means within the same occupation (front-to-back).


## The Naive Estimate is Too Small {.smaller}

```{r}
#| echo: true

# Naive: just gender
lm_naive <- lm(wage ~ female, data = wage_data)

# Controlling for occupation
lm_ctrl <- lm(wage ~ female + office, data = wage_data)
```

. . .

:::: {.columns}

::: {.column width="50%"}
**Without control:**
```{r}
#| echo: false
coef(lm_naive)["female"] |> round(0)
```
:::

::: {.column width="50%"}
**With occupation control:**
```{r}
#| echo: false
coef(lm_ctrl)["female"] |> round(0)
```
:::

::::

. . .

- True gap is **-\$5,000** — the controlled estimate is much closer!

- Naive estimate was **attenuated** because women sort into higher-paying office jobs, which partially offsets the gap


## Step 1: Compute Within-Group Averages {.smaller}

Compute the average wage for **each gender × occupation cell** — this is the 2×2 table:

```{r}
#| echo: true

cell_means <- wage_data |>
  group_by(occupation, gender) |>
  summarise(avg_wage = mean(wage) |> round(0),
            n = n(), .groups = "drop") |>
  arrange(occupation, gender)

cell_means |> knitr::kable(
  col.names = c("Occupation", "Gender", "Avg Wage ($)", "N"))
```

. . .

Now we're comparing men and women **within the same occupation** — apples to apples!


## Step 2: Compute Within-Group Differences {.smaller}

Next, compute the gender wage gap **within each occupation**:

```{r}
#| echo: true

within_diffs <- cell_means |>
  group_by(occupation) |>
  summarise(
    male_wage = avg_wage[gender == "Male"],
    female_wage = avg_wage[gender == "Female"],
    gap = avg_wage[gender == "Female"] - avg_wage[gender == "Male"],
    n_total = sum(n),
    .groups = "drop"
  )

within_diffs |> knitr::kable(
  col.names = c("Occupation", "Male Avg", "Female Avg",
                "Gap (F - M)", "N"),
  digits = 0)
```

. . .

Both within-occupation gaps are **much closer** to the true effect of -$5,000!


## Step 3: Weighted Average Across Groups {.smaller}

Finally, aggregate the within-occupation gaps using a **weighted average**, where the weights reflect each occupation's share of the sample:

```{r}
#| echo: true

within_diffs <- within_diffs |>
  mutate(weight = n_total / sum(n_total))

within_diffs |>
  select(occupation, gap, weight) |>
  knitr::kable(
    col.names = c("Occupation", "Within-Occ Gap ($)", "Weight"),
    digits = 3)
```

. . .

```{r}
#| echo: true

# Weighted average of within-occupation gaps
weighted_estimate <- sum(within_diffs$gap * within_diffs$weight)
weighted_estimate |> round(0)
```

. . .

**Weighted estimate:** `r round(weighted_estimate, 0)` — very close to the true effect of **-\$5,000**!


## The Matching Recipe {.smaller}

::: {.callout-tip}
## What Multivariate OLS Does Under the Hood
:::

1. **Group** observations by the control variable(s) (e.g., same occupation)

. . .

2. **Compare** the outcome across treatment values **within each group** (e.g., the gender wage gap separately for office and manual workers)

. . .

3. **Aggregate** with a weighted average across groups (weights = group sizes)

. . .

This is exactly what the regression $wage = \beta_0 + \beta_1 \cdot female + \beta_2 \cdot office + \mu$ does — it just automates the process!

. . .

```{r}
#| echo: true

# Compare: matching estimate vs regression
c("Matching (weighted avg)" = round(weighted_estimate, 0),
  "Regression" = coef(lm(wage ~ female + office,
                         data = wage_data))["female"] |> round(0),
  "True effect" = -5000)
```


# The Zero Conditional Mean Assumption

## The Multivariate Version {.smaller}

- Recall **Gauss-Markov Assumtion 4** which **must hold for our estimate to be unbiased**

. . .

$E[\mu | x] = 0$

. . .

- All factors contained in $\mu$ must be uncorrelated with $x$

- In multivariate OLS:

. . .

$$E[\mu | x_1, x_2, \ldots, x_k] = 0$$

. . .

- Everything remaining in $\mu$ must be uncorrelated with **all** included variables

. . .

- This is **much more plausible** than the bivariate version — we've pulled confounders out of $\mu$!


## When Does the Assumption Hold? {.smaller}

:::: {.columns}

::: {.column width="50%"}
**More plausible when:**

- We include key confounders
- Good theoretical motivation for controls
- Rich set of observable controls
:::

::: {.column width="50%"}
**Less plausible when:**

- Important confounders are unobservable (ability, motivation)
- Wrong functional form
:::

::::

. . .

::: {.callout-warning}
## A Warning
Adding controls helps but **doesn't guarantee** causality. Unobservable confounders can always lurk in $\mu$.
:::


<!-- ## Try It: Real Data — Boston Housing {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `Boston` dataset from `MASS`, estimate the effect of average room size (`rm`) on median home value (`medv`). Here are some available variables:

| Variable | Description |
|----------|-------------|
| `medv` | Median home value ($1000s) |
| `rm` | Average number of rooms per dwelling |
| `crim` | Per capita crime rate |
| `tax` | Property tax rate per $10,000 |
| `ptratio` | Pupil-teacher ratio by town |
| `lstat` | Percent lower-status population |
| `dis` | Distance to employment centers |
| `nox` | Nitrogen oxide concentration (pollution) |

. . .

**Your task:**

1. Start with the bivariate model: `medv ~ rm`
2. Choose **2--3 controls** you think are confounders. Justify your choices!
3. Compare: how does the coefficient on `rm` change?

. . .

```{r}
#| echo: true
#| output-location: fragment

library(MASS)
data(Boston)

# Bivariate
coef(lm(medv ~ rm, data = Boston))["rm"]

# Example: add your chosen controls
coef(lm(medv ~ rm + crim + tax + ptratio, data = Boston))["rm"]
```

. . .

**Discuss:** Why did the coefficient change? What confounders might *still* be missing? -->


# Omitted Variable Bias

## What is Bias? {.smaller}

- **Bias** is the difference between what our estimator gives us *on average* and the *true* parameter value:

. . .

$$\text{Bias}(\tilde{\beta}_1) = E[\tilde{\beta}_1] - \beta_1$$

. . .

- If $\text{Bias} = 0$, the estimator is **unbiased** — on average, we get the right answer
- If $\text{Bias} \neq 0$, we are **systematically** over- or under-estimating the true effect

. . .

- Bias does **not** shrink with more data — it's a permanent problem!


## The OVB Formula {.smaller}



We want to estimate $\beta_1$. Suppose the **true** model is:

. . .

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \mu$$

. . .

But we omit $x_2$ and estimate:

. . .

$$\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1 x_1$$

. . .

It turns out there is a simple relation between the true value of our parameter, $\beta_1$, 
an *unbiased* estimate of our parameter $\hat{\beta}_1$,
and the *biased* estimate of our parameter $\tilde{\beta}_1$:

. . .

$$\tilde{\beta}_1 = \hat{\beta}_1 + \hat{\beta}_2 \cdot \tilde{\delta}_1$$

. . .

where $\tilde{\delta}_1$ comes from regressing $x_2$ on $x_1$: 

. . .

$$\quad x_2 = \tilde{\delta}_0 + \tilde{\delta}_1 x_1$$


## The OVB Formula {.smaller}

- Taking expectations, we see that:

. . .

$$E[\tilde{\beta}_1] = E[\hat{\beta}_1 + \hat{\beta}_2 \cdot \tilde{\delta}_1]$$

. . .

$$E[\tilde{\beta}_1] = E[\hat{\beta}_1] + E[\hat{\beta}_2] \cdot \tilde{\delta}_1$$

. . .

- Since $E[\hat{\beta}_1] = \beta_1$ and $E[\hat{\beta}_2] = \beta_2$:

. . .

$$E[\tilde{\beta}_1] = \beta_1 + \beta_2 \cdot \tilde{\delta}_1$$

. . .

- Therefore, the bias is:

. . .

$$ Bias(\tilde{\beta}_1) = E[\tilde{\beta}_1] - \beta_1 = \beta_2 \cdot \tilde{\delta}_1$$

. . .

$$\underbrace{Bias(\tilde{\beta}_1)}_{\text{What we get wrong}} = \underbrace{\beta_2}_{\text{Effect of omitted on } y} \times \underbrace{\tilde{\delta}_1}_{\text{Relationship of omitted with } x_1}$$


## Seeing OVB in Action {.smaller}


Let's **simulate** OVB. We know the true DGP, so we can watch bias appear when we omit a variable:

```{r}
#| echo: true

set.seed(123)
n <- 1000

# True DGP: y = 3 + 2*x1 + 4*x2 + 3*x3 + error
x2 <- rnorm(n)
x3 <- rnorm(n)                    # x3 is independent of x1
x1 <- 0.1 * x2 + rnorm(n)        # x1 and x2 are correlated
y <- 3 + 2 * x1 + 4 * x2 + 3 * x3 + rnorm(n)

# Full model (correct)
coef(lm(y ~ x1 + x2 + x3))["x1"]

# Omit x2 (correlated with x1) — biased!
coef(lm(y ~ x1))["x1"]

# Omit x3 (uncorrelated with x1) — no bias!
coef(lm(y ~ x1 + x2))["x1"]
```

## Seeing OVB in Action {.smaller}

Omitting $x_2$ (correlated with $x_1$) biases $\hat{\beta}_1$. Omitting $x_3$ (uncorrelated with $x_1$) does **not** — the estimate barely changes!

. . .

```{r}
#| echo: true

# Verify: bias from omitting x2 = beta2 * delta1
beta2 <- 4
delta1 <- coef(lm(x2 ~ x1))["x1"]
c(
  "beta2" = beta2,
  "delta1" = delta1,
  "Predicted bias" = beta2 * delta1,
  "Actual bias" = coef(lm(y ~ x1))["x1"] - 2)
```

. . .

The OVB formula **works** — the bias is almost exactly $\beta_2 \times \tilde{\delta}_1$! And omitting $x_3$ caused no bias because $\tilde{\delta}_1 \approx 0$.


## Signing the Bias {.smaller}

The bias depends on **two things** — and you can figure out the direction!

. . .

| | $Corr(x_1, x_2) > 0$ | $Corr(x_1, x_2) < 0$ |
|-----|:-----:|:-----:|
| $\beta_2 > 0$ | **Positive** bias ↑ | **Negative** bias ↓ |
| $\beta_2 < 0$ | **Negative** bias ↓ | **Positive** bias ↑ |

. . .

- Positive × Positive = Positive (overestimate)
- Positive × Negative = Negative (underestimate)

. . .

::: {.callout-tip}
## Zero Bias When...
If **either** $\beta_2 = 0$ (omitted variable doesn't affect $y$) **or** $\tilde{\delta}_1 = 0$ (omitted variable uncorrelated with $x_1$), there is **no bias**.
:::


## Example: Positive Bias — Policing and Crime {.smaller}

**True model:** $\quad crime = \beta_0 + \beta_1(police) + \beta_2(poverty) + \mu$

. . .

**What we estimate (omitting poverty):** $\quad \widetilde{crime} = \tilde{\beta}_0 + \tilde{\beta}_1(police)$

. . .

- **Omitted variable:** poverty

. . .

- $\beta_2$ (effect of poverty on crime):
- **Positive** (more poverty → more crime)

. . .

- $\tilde{\delta}_1$ (correlation of poverty with policing):
- **Positive** (high-poverty areas → more police)

. . .

$$\text{Bias} = (+)(+) = \text{Positive}$$

. . .

> The naive estimate makes policing look **less effective** than it actually is (or even that it *increases* crime).


## Example: Negative Bias — Prenatal Care and Birth Weight {.smaller}

**True model:** $\quad birthweight = \beta_0 + \beta_1(prenatal\_visits) + \beta_2(health\_risk) + \mu$

. . .

**What we estimate (omitting health risk):** $\quad \widetilde{birthweight} = \tilde{\beta}_0 + \tilde{\beta}_1(prenatal\_visits)$

. . .

- **Omitted variable:** maternal health risk

. . .

- $\beta_2$ (effect of health risk on birth weight):
- **Negative** (higher risk → lower birth weight)

. . .

- $\tilde{\delta}_1$ (correlation of health risk with prenatal visits): **Positive** (higher-risk mothers get *more* prenatal visits)

. . .

$$\text{Bias} = (-)(+) = \text{Negative}$$

. . .

> The naive estimate **understates** the benefit of prenatal care — high-risk mothers visit more often *and* have worse outcomes, dragging the coefficient toward zero (or even making it negative!).


## Example: Zero Bias — Coffee and Heart Rate {.smaller}

Estimating the effect of coffee consumption on heart rate while omitting height:

$$heart\_rate = \beta_0 + \beta_1(coffee) + \mu$$

. . .

- $\beta_2$ (effect of height on heart rate): 
- **Negative** — taller people tend to have (slightly) lower resting heart rates

. . .

- $\tilde{\delta}_1$ (correlation of height with coffee): 
- **Zero** — how tall you are has nothing to do with how much coffee you drink

. . .

$$\text{Bias} = (-)(0) = 0$$

. . .

> There is **no bias** from omitting height. It fails the second condition: the omitted variable must be correlated with the included regressor. Not every omitted variable is a confounder!


## Class Activity: Sign the Bias! {.smaller}

::: {.callout-note icon=false}
## 🤔 Think-Pair-Share
:::

For each scenario, determine the **direction** of the omitted variable bias:

. . .

1. **Model:** $health = \beta_0 + \beta_1(exercise) + \mu$. Omitted: income.
   - Higher income → more exercise? Higher income → better health?

. . .

2. **Model:** $test\_score = \beta_0 + \beta_1(class\_size) + \mu$. Omitted: school funding.
   - More funding → smaller classes? More funding → higher scores?

. . .

3. **Model:** $crime = \beta_0 + \beta_1(unemployment) + \mu$. Omitted: urban density.
   - Higher density → more unemployment? Higher density → more crime?


# Goodness of Fit: Adjusted R²

## The Problem with R² {.smaller}

- Recall: $R^2$ measures the fraction of variation in $y$ explained by the model

. . .

- **Problem:** $R^2$ can **never decrease** when you add more variables — even irrelevant ones!

. . .

```{r}
#| echo: true

# Add 10 completely irrelevant variables
set.seed(42)
noise_data <- mtcars
for (i in 1:10) noise_data[[paste0("noise", i)]] <- rnorm(nrow(mtcars))
noise_vars <- paste0("noise", 1:10)

c("R² without noise" = summary(lm(mpg ~ hp + wt, data = noise_data))$r.squared,
  "R² with noise"    = summary(lm(mpg ~ hp + wt + ., data = noise_data[c("mpg","hp","wt",noise_vars)]))$r.squared)
```

. . .

- $R^2$ went up even though `random_noise` is meaningless!


## The Fix: Adjusted R² {.smaller}

The **adjusted R-squared** penalizes for the number of variables:

$$\bar{R}^2 = 1 - \frac{(1 - R^2)(n - 1)}{n - k - 1}$$

- where $n$ = number of observations, $k$ = number of regressors, and $(1 - R^2)$ is the share of variance unexplained. 
- As $k$ increases, $n - k - 1$ shrinks, making the penalty larger.

. . .

```{r}
#| echo: true

c("Adj R² without noise" = summary(lm(mpg ~ hp + wt, data = noise_data))$adj.r.squared,
  "Adj R² with noise"    = summary(lm(mpg ~ hp + wt + ., data = noise_data[c("mpg","hp","wt",noise_vars)]))$adj.r.squared)
```

. . .

- Adjusted $R^2$ **decreased** — correctly penalizing the irrelevant variable

. . .

::: {.callout-tip}
## Rule of Thumb
Use **adjusted** $R^2$ when comparing models with different numbers of variables.
:::


# Variance of the Multivariate OLS Estimator

## Extending the Bivariate Result {.smaller}

Recall from the bivariate case:

$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$$

. . .

Precision comes from **variation in $x$** — more variation means more information about the effect of $x$ on $y$.

. . .

In the multivariate case, the same idea applies, but with a twist: **not all variation in $x_j$ is useful**.

. . .

- Some variation in $x_j$ is shared with other regressors — OLS can't tell whether that shared variation is driving $y$ through $x_j$ or through the other controls
- Only the variation in $x_j$ that is **independent of the other regressors** identifies $\hat{\beta}_j$


## How Much Independent Variation Does $x_j$ Have? {.smaller}

Let $R_j^2$ be the R-squared from regressing $x_j$ on all other regressors.

. . .

- $R_j^2$ measures how much of $x_j$'s variation is **shared** with the other controls
- $(1 - R_j^2)$ is the fraction that is **independent** — the part that actually identifies $\hat{\beta}_j$

. . .

So the effective variation available to estimate $\beta_j$ is:

$$SST_j \times (1 - R_j^2)$$

## Variance of $\hat{\beta}_j$: Result {.smaller}

Replacing $SST_x$ with the residual variation $SST_j(1 - R_j^2)$:

. . .

::: {.callout-important}
## Variance of the Multivariate OLS Estimator
$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R_j^2)}$$

where $\sigma^2$ is the error variance, $SST_j = \sum(x_{ji} - \bar{x}_j)^2$ is the total variation in $x_j$, and $R_j^2$ is the R-squared from regressing $x_j$ on all other regressors.
:::

. . .

Three things **increase** variance: **more noise** ($\sigma^2$ ↑), **less variation in $x_j$** ($SST_j$ ↓), **higher collinearity** ($R_j^2$ ↑)

## What Increases Variance? Simulation {.smaller}

```{r}
#| echo: false
#| fig-height: 5.5
#| fig-align: center
#| cache: true

set.seed(325)
n <- 100
n_sims <- 1000
true_beta1 <- 2

# --- Simulation function ---
# DGP: y = 1 + 2*x1 + 1*x2 + error
run_sim <- function(n, sigma, x1_sd, rho, n_sims) {
  estimates <- numeric(n_sims)
  for (i in 1:n_sims) {
    x2 <- rnorm(n)
    x1 <- rho * x2 + sqrt(1 - rho^2) * rnorm(n, sd = x1_sd)
    y <- 1 + true_beta1 * x1 + 1 * x2 + rnorm(n, sd = sigma)
    estimates[i] <- coef(lm(y ~ x1 + x2))["x1"]
  }
  estimates
}

# Baseline: sigma = 1, x1_sd = 1, rho = 0.3
baseline <- run_sim(n, sigma = 1, x1_sd = 1, rho = 0.3, n_sims)

# 1. More noise: sigma = 3
high_noise <- run_sim(n, sigma = 3, x1_sd = 1, rho = 0.3, n_sims)

# 2. Less variation in x1: x1_sd = 0.3
low_var <- run_sim(n, sigma = 1, x1_sd = 0.3, rho = 0.3, n_sims)

# 3. Low collinearity: rho = 0.05
low_collin <- run_sim(n, sigma = 1, x1_sd = 1, rho = 0.05, n_sims)

df_sims <- bind_rows(
  tibble(beta_hat = baseline, scenario = "Baseline (σ=1, sd=1, ρ=0.3)"),
  tibble(beta_hat = high_noise, scenario = "More noise (σ=3)"),
  tibble(beta_hat = low_var, scenario = "Less variation (sd=0.3)"),
  tibble(beta_hat = low_collin, scenario = "Low collinearity (ρ≈0)")
) |>
  mutate(scenario = factor(scenario, levels = c(
    "Baseline (σ=1, sd=1, ρ=0.3)",
    "More noise (σ=3)",
    "Less variation (sd=0.3)",
    "Low collinearity (ρ≈0)"
  )))

ggplot(df_sims, aes(x = beta_hat)) +
  geom_histogram(bins = 40, fill = "#5D8EE5", color = "white", alpha = 0.8) +
  geom_vline(xintercept = true_beta1, color = "#E63946", linewidth = 1, linetype = "dashed") +
  facet_wrap(~ scenario, scales = "free_y", ncol = 2) +
  labs(
    x = expression(hat(beta)[1]),
    y = "Count",
    caption = paste0("1,000 simulations each | n = ", n, " | True β = ", true_beta1, " (dashed red)")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )
```

# Assumptions and Multicollinearity

## Gauss-Markov Assumptions (Extended) {.smaller}

The multivariate versions:

. . .

1. **Linear in parameters:** $y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + \mu$

. . .

2. **Random sampling**

. . .

3. **No perfect multicollinearity** ← UPDATED!

. . .

4. **Zero conditional mean:** $E[\mu | x_1, \ldots, x_k] = 0$

. . .

5. **Homoskedasticity:** $Var(\mu | x_1, \ldots, x_k) = \sigma^2$


## Perfect Multicollinearity {.smaller}

- New assumption 3 includes both **variation in $x_j$** and **no perfect multicolinearity**

- **Perfect multicollinearity**: one variable is an exact linear function of others

. . .

Examples:

- Including `income_dollars` and `income_thousands` (one = other × 1000)
- Including `parent1_income` + `parent2_income` + `family_income` when family = parent1 + parent2

. . .

```{r}
#| echo: true
#| error: true

# Create a perfectly collinear variable
mtcars$hp_doubled <- 2 * mtcars$hp
lm(mpg ~ hp + hp_doubled + wt, data = mtcars) |> coef()
```

. . .

- R drops the collinear variable and gives you `NA` — but you should fix the model!


## Imperfect Multicollinearity {.smaller}

- **Imperfect** multicollinearity (high but not perfect correlation) is *allowed* but problematic

. . .

- It inflates the **variance** of your estimates:

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j (1 - R_j^2)}$$

. . .

- $R_j^2$ = R-squared from regressing $x_j$ on all other $x$'s
- As $R_j^2 \rightarrow 1$, variance $\rightarrow \infty$

. . .

- High correlation between controls makes it hard to **separate their individual effects**
- Your estimates become **imprecise** (large standard errors)


## Visualizing Multicollinearity {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(42)
n_sim <- 500
rho_low <- 0.2
rho_high <- 0.95

sim_mc <- function(rho, n_sim, n = 100) {
  map_dbl(1:n_sim, function(i) {
    x1 <- rnorm(n)
    x2 <- rho * x1 + sqrt(1 - rho^2) * rnorm(n)
    y <- 2 + 3 * x1 + 1 * x2 + rnorm(n)
    coef(lm(y ~ x1 + x2))["x1"]
  })
}

set.seed(42)
low_corr <- tibble(b1 = sim_mc(rho_low, n_sim), label = paste0("Low correlation (r = ", rho_low, ")"))
high_corr <- tibble(b1 = sim_mc(rho_high, n_sim), label = paste0("High correlation (r = ", rho_high, ")"))

bind_rows(low_corr, high_corr) |>
  ggplot(aes(x = b1, fill = label)) +
  geom_histogram(alpha = 0.6, bins = 40, color = "white", position = "identity") +
  geom_vline(xintercept = 3, color = "darkred", linewidth = 1.5, linetype = "dashed") +
  scale_fill_manual(values = c("#5D8EE5", "#EEB808")) +
  labs(x = expression(hat(beta)[1]),
       y = "Count",
       title = "Multicollinearity Inflates Variance",
       subtitle = "Both centered on true value (3), but high correlation → much wider spread",
       fill = "") +
  annotate("text", x = 3, y = 55, label = "True β = 3", color = "darkred",
           hjust = -0.1, fontface = "italic", size = 5) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```


# The Bias-Variance Tradeoff

## Should You Include That Variable? {.smaller}

Adding controls involves a **tradeoff**:

. . .

| Action | Bias | Variance |
|--------|:----:|:--------:|
| Omit a **relevant** variable | ↑ Increases | ↓ Decreases |
| Include an **irrelevant** variable | No change | ↑ Increases |

. . .

- **Too few controls:** risk bias (omitted variable bias)
- **Too many controls:** risk inflated variance (imprecise estimates)

. . .

> Because you can never know *for sure* the true population parameter model,
you can never know *for sure* if a control variable is necessary or not!




## Seeing the Tradeoff {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center
#| cache: true

set.seed(124890)

# Case 1: x2 is irrelevant (beta2 = 0)
pop_irr <- tibble(
  x2 = sample(1:100, 10000, replace = TRUE),
  x1 = 5.68 + 1.89 * x2 + rnorm(10000),
  y = 8.78 + 2.21 * x1 + rnorm(10000)
)

# Case 2: x2 is relevant (beta2 != 0)
pop_rel <- tibble(
  x2 = sample(1:100, 10000, replace = TRUE),
  x1 = 5.68 + 1.89 * x2 + rnorm(10000),
  y = 8.78 + 2.21 * x1 + 0.15 * x2 + rnorm(10000)
)

sim_est <- function(pop, include_x2, n_sims = 500, ss = 100) {
  map_dbl(1:n_sims, function(i) {
    s <- slice_sample(pop, n = ss)
    if (include_x2) coef(lm(y ~ x1 + x2, data = s))["x1"]
    else coef(lm(y ~ x1, data = s))["x1"]
  })
}

set.seed(42)
results <- bind_rows(
  tibble(b1 = sim_est(pop_irr, FALSE), model = "Omit x2", case = "x2 irrelevant (b2 = 0)"),
  tibble(b1 = sim_est(pop_irr, TRUE), model = "Include x2", case = "x2 irrelevant (b2 = 0)"),
  tibble(b1 = sim_est(pop_rel, FALSE), model = "Omit x2", case = "x2 relevant (b2 != 0)"),
  tibble(b1 = sim_est(pop_rel, TRUE), model = "Include x2", case = "x2 relevant (b2 != 0)")
)

ggplot(results, aes(x = b1, fill = model)) +
  geom_histogram(alpha = 0.5, bins = 30, color = "white", position = "identity") +
  geom_vline(xintercept = 2.21, color = "darkred", linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~ case) +
  scale_fill_manual(values = c("Omit x2" = "#5D8EE5", "Include x2" = "#EEB808")) +
  labs(x = expression(hat(beta)[1]), y = "Count",
       title = "The Bias-Variance Tradeoff",
       subtitle = "Dashed line = true b1 = 2.21",
       fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


## Reading the Simulation {.smaller}

**Left panel** ($x_2$ irrelevant):

- Both distributions centered on truth → no bias either way
- Including $x_2$ (yellow) is **wider** → unnecessary variance

. . .

**Right panel** ($x_2$ relevant):

- Omitting $x_2$ (blue) is **shifted** away from truth → biased!
- Including $x_2$ (yellow) is centered on truth → unbiased (but slightly wider)

. . .

::: {.callout-important}
## Key Takeaway
**Bias is usually worse than variance.** A biased estimate is systematically wrong no matter how much data you collect. Extra variance shrinks with larger samples.
:::


## Practical Guidance {.smaller}

::: {.callout-tip}
## Rules of Thumb for Choosing Controls
:::

1. **Include** variables you have strong theoretical reason to believe are confounders

. . .

2. **Be cautious** with highly correlated controls — can inflate variance dramatically

. . .

3. **Larger samples help** — variance decreases with $n$, so the cost of extra controls falls

. . .

4. **Check sensitivity** — if adding a control substantially changes your coefficient of interest, it's probably important


# Summary

## Key Takeaways {.smaller}

- **Multivariate OLS** includes control variables to address confounding:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

. . .

- Coefficients have **ceteris paribus** interpretations: the effect of $x_j$ holding all else constant

. . .

- **OVB formula** tells us the direction of bias: $\text{Bias} = \beta_2 \cdot \tilde{\delta}_1$

. . .

- **Bias-variance tradeoff:** omitting relevant variables → bias; including irrelevant variables → extra variance

. . .

- **Adjusted R²** penalizes for extra variables — use it for model comparison

. . .

- **Perfect multicollinearity** breaks OLS; imperfect multicollinearity inflates variance

## What's Next?

- **Unit 6:** Statistical Inference — hypothesis testing and confidence intervals for regression coefficients

. . .

- We'll finally be able to answer: "Is this coefficient **statistically significant**?"
