---
title: "Multivariate Regression and OLS"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 5 | Multivariate Regression and OLS"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


# From Bivariate to Multivariate OLS

## Where We've Been {.smaller}

- Last unit: **bivariate OLS**, $y = \beta_0 + \beta_1 x + \mu$, is **unbiased** when GM assumptions 1--4 hold
- The hardest to satisfy is **GM4**, $E[\mu | x] = 0$: *everything else* affecting $y$ must be uncorrelated with $x$ --- almost never true in observational data!

. . .

**Example:** $wage = \beta_0 + \beta_1 \cdot educ + \mu$

- The error term $\mu$ contains: ability, family connections, motivation, experience, etc.
- Many of these are **correlated with education** --- wealthier families → more education *and* higher wages, i.e. $E[\mu | X] \neq 0$

. . .

::: {.callout-important}
## The Core Problem
If confounders are in $\mu$ and correlated with $x$, then $E[\mu | x] \neq 0$ and $\hat{\beta}_1$ is **biased**.
:::


# The Multivariate Model

## The Solution: Control for Confounders {.smaller}

- **Today:** how adding **control variables** gets us closer to **causal** estimates

. . .

- If family income confounds the education-wage relationship, **include it**:

. . .

$$wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$$

- Now $\beta_1$ is the effect of education **holding family income constant**
- Family income is no longer in $\mu$ --- that source of bias is gone!

. . .

- In general, with $k$ independent variables:

. . .

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

- $\mu$ now only contains factors **not included** in the model


## Interpreting Coefficients {.smaller}

- Each coefficient has a **ceteris paribus** interpretation: $\beta_j = \dfrac{\partial y}{\partial x_j}$
- The change in $y$ for a one-unit increase in $x_j$, **holding all other $x$'s constant**

. . .

- This is the key difference from bivariate OLS!
  - Bivariate: "associated with" (everything else floating)
  - Multivariate: "holding constant" (the other $x$'s **in the model**)

. . .

**Example:** $wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$

- $\beta_1$: Each additional year of education is associated with $\beta_1$ dollars more in wages, **holding family income constant**
- $\beta_2$: Each additional dollar of family income is associated with $\beta_2$ dollars more in wages, **holding education constant**

. . .

- **But** holding the other $x$'s constant is **not** the same as causal: $\hat{\beta}_1$ is causal only if nothing left in $\mu$ (e.g., ability) is correlated with $educ$ --- that's GM4, coming up


## Deriving the Multivariate OLS Estimator {.smaller}

Same logic as bivariate OLS: choose the $\hat{\beta}$s to **minimize the sum of squared residuals**:

$$SSR = \sum\nolimits_{i=1}^{n} (y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \hat{\beta}_2 x_{2i} - \cdots - \hat{\beta}_k x_{ki})^2$$

. . .

Set the partial derivative with respect to each $\hat{\beta}_j$ equal to zero → $k + 1$ **first-order conditions**. The term in parentheses is the **residual**, $\hat{\mu}_i$:

$$\begin{aligned}
\sum\nolimits_{i=1}^{n}(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \cdots - \hat{\beta}_k x_{ki}) &= \sum\nolimits_{i=1}^{n} \hat{\mu}_i = 0 \\
\sum\nolimits_{i=1}^{n} x_{ji}(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_{1i} - \cdots - \hat{\beta}_k x_{ki}) &= \sum\nolimits_{i=1}^{n} x_{ji} \hat{\mu}_i = 0 \quad \text{for } j = 1, \ldots, k
\end{aligned}$$

. . .

- Solving these $k + 1$ equations gives the OLS estimates: tedious and uninsightful by hand (matrix algebra)


# Multivariate OLS in R

## Adding Controls is Easy! {.smaller}

Same `lm()` function --- just use `+` to add additional $x$, independent variables:

```{r}
#| echo: true

data(mtcars)
lm_biv   <- lm(mpg ~ hp, data = mtcars)        # bivariate
lm_multi <- lm(mpg ~ hp + wt, data = mtcars)   # add weight as a control
summary(lm_biv)
```

## Adding Controls is Easy! {.smaller}

```{r}
#| echo: true

summary(lm_multi)

```

- Coefficient on `hp` dropped from **`r round(coef(lm_biv)["hp"], 3)`** to **`r round(coef(lm_multi)["hp"], 3)`** --- less than half!
- Why? Horsepower and weight are correlated, and weight independently reduces MPG
- The bivariate estimate was **conflating** the effect of hp with the effect of weight


## Visualizing: The Bivariate View {.smaller}

```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 9
#| fig-align: center

ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "#002169", alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "#E63946", linewidth = 1.3) +
  labs(title = "Bivariate: mpg ~ hp",
       subtitle = paste0("Slope = ", round(coef(lm_biv)["hp"], 3),
                         ": each extra horsepower is associated with ",
                         abs(round(coef(lm_biv)["hp"], 3)), " fewer MPG"),
       x = "Horsepower", y = "MPG") +
  theme_minimal(base_size = 16)
```

The familiar 2D picture from Unit 4: the best line through the horsepower--MPG points, **ignoring weight**. But every car also has a weight, so the data really live in **3D**...


## Adding Weight: A Cloud of Points in 3D {.smaller}

```{r}
#| echo: false

library(plotly)

b      <- coef(lm_multi)
b_biv  <- coef(lm_biv)
wt_bar <- mean(mtcars$wt)
wall   <- 5.75   # back wall (heavier than any car), where we draw the 2D "shadow"

cars <- mtcars |>
  mutate(car = rownames(mtcars), fit = fitted(lm_multi), wall_wt = wall)

# Bivariate line and the plane's hp slope at average weight, both drawn on the back wall
hp_line <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 50)
wall_lines <- tibble(
  hp    = hp_line,
  wt    = wall,
  biv   = b_biv[1] + b_biv["hp"] * hp_line,
  plane = b[1] + b["hp"] * hp_line + b["wt"] * wt_bar
)

# Fitted plane over the range of the data
# (plotly wants z as a matrix with rows = y values (wt), columns = x values (hp))
hp_grid <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 25)
wt_grid <- seq(min(mtcars$wt), max(mtcars$wt), length.out = 25)
plane_z <- outer(wt_grid, hp_grid, function(w, h) b[1] + b["hp"] * h + b["wt"] * w)

# Residual sticks: one segment per car from the point to the plane (NA breaks the line)
stick <- function(v_from, v_to) as.vector(rbind(v_from, v_to, NA))
sticks <- tibble(
  hp  = stick(mtcars$hp, mtcars$hp),
  wt  = stick(mtcars$wt, mtcars$wt),
  mpg = stick(mtcars$mpg, fitted(lm_multi))
)

# One figure, built up across slides: same axes and camera every time
fig_3d <- function(show_wall = FALSE, show_plane = FALSE) {
  p <- plot_ly(height = 420)
  if (show_plane) {
    p <- p |>
      add_surface(
        x = hp_grid, y = wt_grid, z = plane_z,
        colorscale = list(c(0, "#EEB808"), c(1, "#EEB808")),
        opacity = 0.5, showscale = FALSE, hoverinfo = "skip",
        name = "Fitted plane", showlegend = TRUE
      ) |>
      add_trace(
        data = sticks, x = ~hp, y = ~wt, z = ~mpg,
        type = "scatter3d", mode = "lines", connectgaps = FALSE,
        line = list(color = "#7F7F7F", width = 3),
        hoverinfo = "skip", name = "Residuals", showlegend = FALSE
      )
  }
  p <- p |>
    add_markers(
      data = cars, x = ~hp, y = ~wt, z = ~mpg,
      marker = list(size = 5, color = "#002169", line = list(color = "white", width = 1)),
      text = ~paste0("<b>", car, "</b><br>HP: ", hp, "  Weight: ", wt, "<br>MPG: ", mpg),
      hoverinfo = "text", name = "Cars"
    )
  if (show_wall) {
    p <- p |>
      add_markers(
        data = cars, x = ~hp, y = ~wall_wt, z = ~mpg,
        marker = list(size = 3.5, color = "#9AA5B8"),
        hoverinfo = "skip", name = "Shadow (ignoring weight)"
      ) |>
      add_trace(
        data = wall_lines, x = ~hp, y = ~wt, z = ~biv,
        type = "scatter3d", mode = "lines", line = list(color = "#E63946", width = 8),
        hoverinfo = "skip", name = paste0("Bivariate fit (slope ", round(b_biv["hp"], 3), ")")
      )
  }
  if (show_plane) {
    p <- p |>
      add_trace(
        data = wall_lines, x = ~hp, y = ~wt, z = ~plane,
        type = "scatter3d", mode = "lines", line = list(color = "#A67C00", width = 8),
        hoverinfo = "skip", name = paste0("Plane at average weight (slope ", round(b["hp"], 3), ")")
      )
  }
  p |>
    layout(
      scene = list(
        xaxis = list(title = "Horsepower", range = c(40, 345)),
        yaxis = list(title = "Weight (1,000 lbs)", range = c(1.3, wall + 0.05),
                     showbackground = TRUE, backgroundcolor = "#EEF1F7"),
        zaxis = list(title = "MPG", range = c(4, 35)),
        aspectmode = "manual", aspectratio = list(x = 1.5, y = 1, z = 0.9),
        camera = list(eye = list(x = 0.4, y = -1.9, z = 0.5),
                      center = list(x = 0, y = 0, z = -0.12))
      ),
      showlegend = TRUE,
      legend = list(orientation = "h", x = 0.5, xanchor = "center", y = 1.0,
                    font = list(size = 14)),
      margin = list(l = 0, r = 0, t = 0, b = 0)
    )
}

fig_3d()
```

- Each car is a point in 3D: **horsepower**, **weight**, and **MPG**. Drag to rotate!
- Heavier cars have **more horsepower** (correlation `r round(cor(mtcars$hp, mtcars$wt), 2)`) and **lower MPG** (correlation `r round(cor(mtcars$wt, mtcars$mpg), 2)`)


## The Bivariate Fit Lives on the Back Wall {.smaller}

```{r}
#| echo: false

fig_3d(show_wall = TRUE)
```

- Flatten every car against the **back wall** (i.e., ignore weight) and you get the 2D scatter from before: the gray **shadow**
- The bivariate regression fits its line (**red**, slope `r round(b_biv["hp"], 3)`) to this shadow, so it can't separate horsepower from weight


## The Multivariate Fit: A Plane {.smaller}

```{r}
#| echo: false

fig_3d(show_wall = TRUE, show_plane = TRUE)
```

- With two regressors, OLS fits a **plane** (yellow), choosing $\hat{\beta}_0, \hat{\beta}_1, \hat{\beta}_2$ to minimize the squared vertical **residuals** (gray sticks)
- The plane's slope along horsepower, `r round(b["hp"], 3)`, is the effect of horsepower **holding weight fixed**. Its trace on the back wall (**gold**) is much flatter than the **red** bivariate line


## Slicing the Plane: Holding Weight Fixed {.smaller}

```{r}
#| echo: false
#| fig-height: 4.5
#| fig-width: 9
#| fig-align: center

b <- coef(lm_multi)
slices <- tibble(wt = c(2.5, 4)) |>
  mutate(intercept = b[1] + b["wt"] * wt,
         label = paste0("Weight held at ", format(wt * 1000, big.mark = ","), " lbs"))

ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_abline(data = slices, aes(intercept = intercept, slope = b["hp"]),
              color = "#A67C00", linewidth = 1.2, linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "#E63946", linewidth = 1.3) +
  geom_point(aes(color = wt), size = 3.2) +
  geom_text(data = slices, aes(x = 340, y = intercept + b["hp"] * 340, label = label),
            hjust = 0, vjust = 0.5, color = "#A67C00", size = 4.3, fontface = "bold") +
  scale_color_gradient(low = "#B9C8E8", high = "#001544", name = "Weight\n(1,000 lbs)") +
  scale_x_continuous(limits = c(50, 470), breaks = seq(50, 350, 50)) +
  labs(x = "Horsepower", y = "MPG") +
  theme_minimal(base_size = 16)
```

- Back in 2D: each gold line is a **slice of the plane** at a fixed weight, and every slice has the same slope (`r round(coef(lm_multi)["hp"], 3)`). This is what "controlling for weight" means
- Heavier cars (**darker**) have more horsepower *and* lower MPG, so the red bivariate line picks up the effect of weight too


# Multivariate OLS as Matching

## Multivariate OLS as Matching: The Population {.smaller}

When we "control for" a variable, we're **comparing like with like**. To see this, take a made-up population where we **know the truth**:

$$wage = 30{,}000 - 5{,}000 \cdot female + 20{,}000 \cdot office + \mu$$

- The **true** gender wage gap is **−\$5,000**, holding occupation constant
- Office jobs pay a **\$20,000 premium** over manual jobs
- Women are **more likely** to work in office jobs: 70% vs. 40% for men

. . .

- We draw a **random sample of 2,000 workers** from this population and see what different approaches estimate

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Code: draw the sample"

set.seed(129)
n <- 2000
wage_data <- tibble(
  female = sample(c(0, 1), n, replace = TRUE),
  office = rbinom(n, 1, prob = ifelse(female == 1, 0.70, 0.40)),
  wage   = 30000 - 5000 * female + 20000 * office + rnorm(n, 0, 6000)
) |>
  mutate(gender = ifelse(female == 1, "Female", "Male"),
         occupation = ifelse(office == 1, "Office", "Manual"))
```

```{r}
#| echo: false
# Number formatting for inline results ({,} keeps LaTeX from spacing the comma)
fmt  <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
fmtm <- function(x, d = 0) gsub(",", "{,}", fmt(x, d))
```


## The Naive Approach: Regress Wage on Female {.smaller .scrollable}

Suppose we ignore occupation and just regress $wage$ on $female$:

$$\widehat{wage} = \hat{\beta}_0 + \hat{\beta}_1 \cdot female$$

::: {.callout-note}
## Recall from Problem Set 1 (Q6d)
When the only regressor is a 0/1 dummy, OLS just compares group averages: $\hat{\beta}_0$ is the average for the 0 group (men), and $\hat{\beta}_1$ is the **difference** in averages (women − men).
:::

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

lm_naive  <- lm(wage ~ female, data = wage_data)
avg_men   <- mean(wage_data$wage[wage_data$female == 0])
avg_women <- mean(wage_data$wage[wage_data$female == 1])

tibble(
  ` `                   = c("Men's average wage", "Women's minus men's average"),
  `From group averages` = fmt(c(avg_men, avg_women - avg_men)),
  `From the regression` = paste(c("$\\hat{\\beta}_0$ =", "$\\hat{\\beta}_1$ ="),
                                fmt(coef(lm_naive)))
) |>
  knitr::kable(align = "lcc")
```

. . .

- Women appear to earn **\$`r fmt(coef(lm_naive)["female"])` more** than men, but the true gap is **−\$5,000**. We got even the **sign** wrong!


## What Are We Missing? Occupation {.smaller}

- That's pretty far off. But remember: there's another variable, $office$, that is correlated with **both** $female$ (our $x$) **and** $wage$ (our $y$)!

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

office_share <- wage_data |> group_by(gender) |> summarise(share = mean(office))

wage_data |>
  summarise(`Corr(office, female)` = sprintf("%.2f", cor(office, female)),
            `Corr(office, wage)`   = sprintf("%.2f", cor(office, wage))) |>
  knitr::kable(align = "cc")
```

. . .

- $office$ and $female$: women are more likely to have office jobs (`r fmt(100 * office_share$share[office_share$gender == "Female"])`% vs. `r fmt(100 * office_share$share[office_share$gender == "Male"])`% of men in our sample)
- $office$ and $wage$: office jobs pay about \$20,000 more
- So the naive $\hat{\beta}_1$ **mixes** the gender gap with the office premium: women look better paid partly because more of them have office jobs


## Matching: Compare Within Occupations {.smaller .scrollable}

- The fix: compare women and men **in the same occupation**, so the office premium drops out of the comparison

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

within_diffs <- wage_data |>
  group_by(occupation) |>
  summarise(male_wage   = mean(wage[gender == "Male"]),
            female_wage = mean(wage[gender == "Female"]),
            gap         = female_wage - male_wage,    # gap within each occupation
            n           = n()) |>
  mutate(weight = n / sum(n))                           # each occupation's share of workers

within_diffs |>
  select(occupation, male_wage, female_wage, gap) |>
  knitr::kable(col.names = c("Occupation", "Men's Avg", "Women's Avg", "Gap (W − M)"),
               digits = 0, format.args = list(big.mark = ","), align = "lccc")
```

. . .

- In **both** occupations, the gap is close to the true **−\$5,000**
- That's what "controlling for" occupation means: comparing like with like


## Matching: Average the Group Estimates {.smaller}

- We now have **two** estimates of the gender gap, one per occupation. To get **one** number, take a **weighted average**, weighting each occupation by its share of workers:

```{r}
#| echo: false
w   <- setNames(within_diffs$weight, within_diffs$occupation)
gap <- setNames(within_diffs$gap,    within_diffs$occupation)
```

. . .

$$
\hat{\beta}_{match} = \underbrace{`r fmtm(w["Manual"], 3)`}_{\text{share manual}} \times (`r fmtm(gap["Manual"])`) + \underbrace{`r fmtm(w["Office"], 3)`}_{\text{share office}} \times (`r fmtm(gap["Office"])`) = `r fmtm(sum(w * gap))`
$$

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

weighted_estimate <- sum(within_diffs$weight * within_diffs$gap)
```

. . .

- Close to the true **−\$5,000**, and nowhere near the naive estimate of **+\$`r fmt(coef(lm_naive)["female"])`**


## The Matching Recipe {.smaller .scrollable}

::: {.callout-tip}
## What Multivariate OLS Does Under the Hood
:::

::: {.nonincremental}
1. **Group** observations by the control variable(s) (e.g., same occupation)
2. **Compare** the outcome **within each group** (the gender gap among office workers, and among manual workers)
3. **Aggregate** with a weighted average across groups
:::

. . .

Regressing $wage$ on $female$ **and** $office$ automates this. (OLS weights each group by its size *and* by how much `female` varies within it; here the two are nearly identical.)

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

lm_ctrl <- lm(wage ~ female + office, data = wage_data)   # control for occupation

tibble(
  `Naive (no control)`    = coef(lm_naive)["female"],
  `Matching by hand`      = weighted_estimate,
  `Regression (+ office)` = coef(lm_ctrl)["female"],
  `Truth`                 = -5000
) |>
  knitr::kable(digits = 0, format.args = list(big.mark = ","), align = "c")
```

. . .

- The naive regression gets even the **sign** wrong; matching and the multivariate regression both land near the truth


# The Gauss-Markov Assumptions, Revisited

## Gauss-Markov Assumptions (Extended) {.smaller}

::: {.nonincremental}
1. **Linear in parameters:** $y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + \mu$
2. **Random sampling**
3. **No perfect multicollinearity** ← UPDATED!
4. **Zero conditional mean:** $E[\mu | x_1, x_2, \ldots, x_k] = 0$ ← UPDATED!
5. **Homoskedasticity:** $Var(\mu | x_1, \ldots, x_k) = \sigma^2$
:::

. . .

- Assumptions 1--4 → OLS is **unbiased**; adding 5 gives us the **variance** formula (later today)

. . .

- Two assumptions change with multiple regressors. We'll take them in order: **assumption 3**, then **assumption 4**


## Assumption 3: No Perfect Multicollinearity {.smaller}

- New assumption 3 includes both **variation in $x_j$** and **no perfect multicollinearity**

- **Perfect multicollinearity**: one variable is an exact linear function of others

. . .

Examples:

- Including `income_dollars` and `income_thousands` (one = other × 1000)
- Including `parent1_income` + `parent2_income` + `family_income` when family = parent1 + parent2

. . .

```{r}
#| echo: true
#| error: true

# Create a perfectly collinear variable
mtcars$hp_doubled <- 2 * mtcars$hp
lm(mpg ~ hp + hp_doubled + wt, data = mtcars) |> coef()
```

. . .

- R drops the collinear variable and gives you `NA` — but you should fix the model!


## Assumption 4: Zero Conditional Mean {.smaller}

$$E[\mu | x_1, x_2, \ldots, x_k] = 0$$

- In words: once we know the $x$'s, the unobserved factors in $\mu$ are **zero on average** --- knowing $x_1, \ldots, x_k$ tells us nothing about $\mu$
- So everything left in $\mu$ must be uncorrelated with **every** included $x$

. . .

- Each control we add **moves a factor out of $\mu$** --- that's why GM4 is **more plausible** with multivariate OLS
- If GM 1--4 hold, OLS is **unbiased**: $E[\hat{\beta}_j] = \beta_j$, so $\hat{\beta}_j$ estimates the **ceteris paribus (causal)** effect of $x_j$

. . .

Two examples: one where a control makes GM4 **hold**, and one where it **helps but isn't enough**.


## Example 1: A Control That Makes GM4 Hold {.smaller}

Suppose crop yield depends on fertilizer, soil quality, rainfall, and pure luck (pests, hail):

:::: {.columns}

::: {.column width="50%"}
**Bivariate:** $yield = \beta_0 + \beta_1 fertilizer + \mu$

$\mu$ contains: **soil quality**, **rainfall**, luck

→ farmers with better soil and more rain also use more fertilizer, so GM4 **fails**
:::

::: {.column width="50%"}
**Multivariate:** $yield = \beta_0 + \beta_1 fertilizer + \beta_2 soil + \beta_3 rain + \mu$

$\mu$ contains: luck

→ luck is unrelated to how much fertilizer a farmer uses, so GM4 **holds**
:::

::::

. . .

- Controlling for soil quality and rainfall **pulls both confounders out of $\mu$**
- What's left in $\mu$ (luck) is uncorrelated with fertilizer, so $\hat{\beta}_1$ is **unbiased**: the ceteris paribus effect of fertilizer


## Example 2: A Control That Isn't Enough {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Bivariate:** $wage = \beta_0 + \beta_1 educ + \mu$

$\mu$ contains: **family income**, ability, motivation, ...

→ family income and ability are correlated with $educ$, so GM4 **fails**
:::

::: {.column width="50%"}
**Multivariate:** $wage = \beta_0 + \beta_1 educ + \beta_2 faminc + \mu$

$\mu$ contains: ability, motivation, ...

→ family income has been **pulled out of $\mu$**; ability is still a concern
:::

::::

. . .

- Unlike the lottery, education is **chosen**, and the reasons people choose it (ability, motivation) are hard to measure
- Controlling for family income makes GM4 **more plausible**, but it probably still **fails**: $\hat{\beta}_1$ is likely still biased by ability


## When Does Zero Conditional Mean Hold? {.smaller}

:::: {.columns}

::: {.column width="50%"}
**More plausible when:**

- We include key confounders
- Good theoretical motivation for controls
- Rich set of observable controls
:::

::: {.column width="50%"}
**Less plausible when:**

- Important confounders are unobservable (ability, motivation)
- Wrong functional form
:::

::::

. . .

::: {.callout-warning}
## A Warning
Adding controls helps but **doesn't guarantee** causality. Unobservable confounders can always lurk in $\mu$.
:::


<!-- ## Try It: Real Data — Boston Housing {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `Boston` dataset from `MASS`, estimate the effect of average room size (`rm`) on median home value (`medv`). Here are some available variables:

| Variable | Description |
|----------|-------------|
| `medv` | Median home value ($1000s) |
| `rm` | Average number of rooms per dwelling |
| `crim` | Per capita crime rate |
| `tax` | Property tax rate per $10,000 |
| `ptratio` | Pupil-teacher ratio by town |
| `lstat` | Percent lower-status population |
| `dis` | Distance to employment centers |
| `nox` | Nitrogen oxide concentration (pollution) |

. . .

**Your task:**

1. Start with the bivariate model: `medv ~ rm`
2. Choose **2--3 controls** you think are confounders. Justify your choices!
3. Compare: how does the coefficient on `rm` change?

. . .

```{r}
#| echo: true
#| output-location: fragment

library(MASS)
data(Boston)

# Bivariate
coef(lm(medv ~ rm, data = Boston))["rm"]

# Example: add your chosen controls
coef(lm(medv ~ rm + crim + tax + ptratio, data = Boston))["rm"]
```

. . .

**Discuss:** Why did the coefficient change? What confounders might *still* be missing? -->



## What is Bias? {.smaller}

- When GM4 fails because we **omit a relevant variable**, our estimate is **biased**. But how should we define bias?

. . .

- Recall that an **unbiased** estimator is right on average: $E[\hat{\beta}_1] = \beta_1$. A **biased** one -- call it $\tilde{\beta}_1$ -- is, on average, the truth **plus something else**:

. . .

$$E[\tilde{\beta}_1] = \beta_1 + \text{something else}$$

. . .

- Call this *something else* the **bias**:

. . .

$$\text{Bias}(\tilde{\beta}_1) = E[\tilde{\beta}_1] - \beta_1$$

. . .

- If $\text{Bias} \neq 0$, we are **systematically** over- or under-estimating the true effect

. . .

- **Question:** when we omit a variable, what exactly *is* the something else?

## The OVB Formula {.smaller}

To find the *something else*, suppose the **true** model is:

. . .

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \mu$$

. . .

But we omit $x_2$ and estimate:

. . .

$$\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1 x_1$$

::: {.callout-note}
## New notation
From here on, **hats** ($\hat{\beta}$) are estimates from the **full** regression that includes $x_2$ (unbiased), and **tildes** ($\tilde{\beta}$) are estimates from the **short** regression that omits $x_2$ (biased).
:::

. . .

It turns out the *biased* and *unbiased* estimates are linked by a simple formula:

. . .

$$\tilde{\beta}_1 = \hat{\beta}_1 + \hat{\beta}_2 \cdot \tilde{\delta}_1$$

. . .

where $\tilde{\delta}_1$ comes from regressing $x_2$ on $x_1$: 

. . .

$$\quad x_2 = \tilde{\delta}_0 + \tilde{\delta}_1 x_1$$


## The OVB Formula {.smaller}

- Taking expectations, we see that:

. . .

$$E[\tilde{\beta}_1] = E[\hat{\beta}_1 + \hat{\beta}_2 \cdot \tilde{\delta}_1]$$

. . .

$$E[\tilde{\beta}_1] = E[\hat{\beta}_1] + E[\hat{\beta}_2] \cdot \tilde{\delta}_1$$

. . .

- Since $E[\hat{\beta}_1] = \beta_1$ and $E[\hat{\beta}_2] = \beta_2$:

. . .

$$E[\tilde{\beta}_1] = \beta_1 + \beta_2 \cdot \tilde{\delta}_1$$

. . .

- So the *something else* -- the bias -- is:

. . .

$$ Bias(\tilde{\beta}_1) = E[\tilde{\beta}_1] - \beta_1 = \beta_2 \cdot \tilde{\delta}_1$$

. . .

$$\underbrace{Bias(\tilde{\beta}_1)}_{\text{What we get wrong}} = \underbrace{\beta_2}_{\text{Relationship of omitted with } y} \times \underbrace{\tilde{\delta}_1}_{\text{Relationship of omitted with } x_1}$$


## Signing the Bias {.smaller}

The bias depends on **two things** — and you can figure out the direction!

. . .

| | $Corr(x_1, x_2) > 0$ | $Corr(x_1, x_2) < 0$ |
|-----|:-----:|:-----:|
| $\beta_2 > 0$ | **Positive** bias ↑ | **Negative** bias ↓ |
| $\beta_2 < 0$ | **Negative** bias ↓ | **Positive** bias ↑ |

. . .

- Positive × Positive = Positive ($E[\tilde{\beta_1}] = \beta_1 + (+)$ - on average you *overestimate* true effect)
- Positive × Negative = Negative ($E[\tilde{\beta_1}] = \beta_1 + (-)$ - on average you *underestimate* true effect)

. . .

::: {.callout-tip}
## Zero Bias When...
If **either** $\beta_2 = 0$ (omitted variable doesn't affect $y$) **or** $\tilde{\delta}_1 = 0$ (omitted variable uncorrelated with $x_1$), there is **no bias**.
:::


## Seeing OVB in Action: The Setup {.smaller}

Let's **simulate** OVB. Because we write the data-generating process (DGP) ourselves, we **know** the true $\beta_1 = 2$:

```{r}
#| echo: true

set.seed(123)
n <- 10000

# True DGP: y = 3 + 2*x1 + 4*x2 + 3*x3 + error
x2 <- rnorm(n)
x3 <- rnorm(n)                     # x3 is independent of x1
x1 <- -0.5 * x2 + rnorm(n)         # x1 and x2 are NEGATIVELY correlated
y  <- 3 + 2 * x1 + 4 * x2 + 3 * x3 + rnorm(n)
```

. . .

**Predict before we run it:** what happens to $\tilde{\beta}_1$ if we leave out $x_2$?

- $\beta_2 = 4 > 0$: $x_2$ raises $y$ → **(+)**
- $Corr(x_1, x_2) < 0$: $\tilde{\delta}_1$ will be negative → **(−)**
- **Bias** $= (+) \times (-) = (-)$: the naive estimate of $\beta_1$ should be **too small**


## Seeing OVB in Action: Naive vs. Full Model {.smaller}

```{r}
#| echo: true

full  <- lm(y ~ x1 + x2 + x3)   # correct model
naive <- lm(y ~ x1)             # omits x2 (and x3)
no_x3 <- lm(y ~ x1 + x2)        # omits only x3, which is uncorrelated with x1

b1 <- function(model) unname(coef(model)["x1"])   # pull out the coefficient on x1
c("True beta1"      = 2,
  "Full model"      = b1(full),
  "Naive (omit x2)" = b1(naive),
  "Omit only x3"    = b1(no_x3)) |> round(3)
```

. . .

- The naive estimate is **far below** the truth: **negative** bias, just as we predicted
- The full model gets it right (≈ 2)

. . .

- Omitting $x_3$ does **not** bias $\hat{\beta}_1$: $x_3$ affects $y$, but it is **uncorrelated** with $x_1$ (condition 2 fails)


## Seeing OVB in Action: Checking the Formula {.smaller}

$\text{Bias}(\tilde{\beta}_1) = \beta_2 \times \tilde{\delta}_1$ --- compute each piece:

```{r}
#| echo: true

beta2          <- 4                           # Step 1: effect of x2 on y (from the DGP)
delta1         <- coef(lm(x2 ~ x1))["x1"]     # Step 2: slope from regressing x2 on x1
predicted_bias <- beta2 * delta1              # Step 3: OVB formula
actual_bias    <- b1(naive) - 2               # Step 4: naive estimate minus the truth
```

```{r}
#| echo: false

tibble(
  Step = c("1", "2", "3", "4"),
  Quantity = c("$\\beta_2$ (effect of $x_2$ on $y$)",
               "$\\tilde{\\delta}_1$ (regress $x_2$ on $x_1$)",
               "Predicted bias $= \\beta_2 \\times \\tilde{\\delta}_1$",
               "Actual bias $= \\tilde{\\beta}_1 - \\beta_1$"),
  Sign = c("(+)", "(−)", "(+) × (−) = (−)", "(−)"),
  Value = c(beta2, delta1, predicted_bias, actual_bias)
) |>
  knitr::kable(digits = 3, align = c("c", "l", "c", "r"))
```

. . .

- The formula works: $\tilde{\beta}_1 \approx \beta_1 + \beta_2 \tilde{\delta}_1 = 2 + (4)(`r round(delta1, 2)`) = `r round(2 + predicted_bias, 2)`$, and the naive regression gave **`r round(b1(naive), 2)`**


## Example: Positive Bias — The Gender Wage Gap {.smaller}

Back to the matching example, where we **know** the truth:

**True model:** $\quad wage = \beta_0 + \beta_1(female) + \beta_2(office) + \mu$, with $\beta_1 = -5{,}000$ and $\beta_2 = 20{,}000$

. . .

**What we estimate (omitting office):** $\quad \widetilde{wage} = \tilde{\beta}_0 + \tilde{\beta}_1(female)$

. . .

- **Omitted variable:** working in an office job

. . .

- $\beta_2$ (effect of an office job on wages): **Positive** (+\$20,000)

. . .

- $\tilde{\delta}_1$ (slope from regressing $office$ on $female$): **Positive**. With a 0/1 regressor, it's the difference in averages: $0.70 - 0.40 = 0.30$

. . .

$$\text{Bias} = \beta_2 \times \tilde{\delta}_1 = 20{,}000 \times 0.30 = +6{,}000$$

. . .

> So $E[\tilde{\beta}_1] = -5{,}000 + 6{,}000 = +1{,}000$, close to the naive **+\$`r fmt(coef(lm_naive)["female"])`** from our sample. Omitting occupation **flips the sign** of the gender gap!


## Example: Negative Bias — Tutoring and Exams {.smaller}

**True model:** $\quad exam = \beta_0 + \beta_1(tutoring\_hours) + \beta_2(prior\_GPA) + \mu$

. . .

**What we estimate (omitting prior GPA):** $\quad \widetilde{exam} = \tilde{\beta}_0 + \tilde{\beta}_1(tutoring\_hours)$

. . .

- **Omitted variable:** prior GPA (how well the student was already doing)

. . .

- $\beta_2$ (effect of prior GPA on exam scores): **Positive** (stronger students score higher)

. . .

- $\tilde{\delta}_1$ (slope from regressing $prior\_GPA$ on $tutoring\_hours$): **Negative** (struggling students are the ones who go to tutoring)

. . .

$$\text{Bias} = (+) \times (-) = \text{Negative}$$

. . .

> The naive estimate **understates** the benefit of tutoring. It can even make tutoring look *harmful*, because the students who use it were already doing worse.


## Class Activity: Sign the Bias! {.smaller}

::: {.callout-note icon=false}
## 🤔 Think-Pair-Share
:::

For each scenario, determine the **direction** of the omitted variable bias:

. . .

1. **Model:** $health = \beta_0 + \beta_1(exercise) + \mu$. Omitted: income.
   - Higher income → more exercise? Higher income → better health?

. . .

2. **Model:** $test\_score = \beta_0 + \beta_1(class\_size) + \mu$. Omitted: school funding.
   - More funding → smaller classes? More funding → higher scores?

. . .

3. **Model:** $crime = \beta_0 + \beta_1(unemployment) + \mu$. Omitted: urban density.
   - Higher density → more unemployment? Higher density → more crime?


## Sign the Bias: Answers {.smaller}

Bias $= \beta_2 \times \tilde{\delta}_1$:

1. **Exercise → health, omitting income:** higher income → better health ($\beta_2 > 0$) and more exercise ($\tilde{\delta}_1 > 0$). Bias $= (+)(+) =$ **positive**: overstates the benefit of exercise
2. **Class size → test scores, omitting funding:** more funding → higher scores ($\beta_2 > 0$) and smaller classes ($\tilde{\delta}_1 < 0$). Bias $= (+)(-) =$ **negative**: $\tilde{\beta}_1$ is too negative, so smaller classes look *more* beneficial than they are
3. **Unemployment → crime, omitting density:** if denser areas have more crime ($\beta_2 > 0$) and more unemployment ($\tilde{\delta}_1 > 0$), then bias $= (+)(+) =$ **positive**: overstates the effect of unemployment on crime


# Goodness of Fit: Adjusted R²

## The Problem with R², and the Fix {.smaller}

- **Problem:** $R^2$ can **never decrease** when you add more variables --- even irrelevant ones!
- **Fix:** the **adjusted** $R^2$ penalizes for the number of regressors $k$: $\quad \bar{R}^2 = 1 - \dfrac{(1 - R^2)(n - 1)}{n - k - 1}$

. . .

```{r}
#| echo: true

set.seed(42)
noise <- matrix(rnorm(nrow(mtcars) * 10), ncol = 10)   # 10 pure-noise variables
small <- summary(lm(mpg ~ hp + wt, data = mtcars))
big   <- summary(lm(mpg ~ hp + wt + noise, data = mtcars))
c("R²"     = small$r.squared,     "R² with noise"     = big$r.squared)
c("Adj R²" = small$adj.r.squared, "Adj R² with noise" = big$adj.r.squared)
```

. . .

- $R^2$ went **up** even though the noise variables are meaningless; adjusted $R^2$ went **down**, correctly penalizing them

. . .

- **Rule of thumb:** use **adjusted** $R^2$ to compare **fit** --- but for **causal** questions, choose controls by whether they're **confounders**, not by fit


# Variance of the Multivariate OLS Estimator

## Step 1: Recall the Bivariate Variance {.smaller}

Under GM 1--5, the bivariate OLS slope has variance

$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}, \qquad SST_x = \sum_{i=1}^{n}(x_i - \bar{x})^2$$

. . .

- **Numerator, $\sigma^2$:** how noisy the unobserved factors in $\mu$ are → more noise, **less** precision
- **Denominator, $SST_x$:** the total variation in $x$ → more spread in $x$ (or a bigger $n$) gives more information about the slope, **more** precision

. . .

In the multivariate case, the same idea applies, but with a twist: **not all of the variation in $x_j$ is useful**.


## Step 2: How Much of $x_j$'s Variation Is Its Own? {.smaller}

- Some variation in $x_j$ is **shared** with the other regressors --- OLS can't tell whether that shared variation moves $y$ through $x_j$ or through the other controls
- Only the variation in $x_j$ that is **independent of the other regressors** identifies $\hat{\beta}_j$

. . .

To measure it, regress $x_j$ on **all the other regressors** and record the R-squared, $R_j^2$:

- $R_j^2$ = share of $x_j$'s variation that is **shared** with the other controls
- $1 - R_j^2$ = share that is **independent**

. . .

**Example:** in `mpg ~ hp + wt`, how much of horsepower's variation is shared with weight?

```{r}
#| echo: true

r2_hp <- summary(lm(hp ~ wt, data = mtcars))$r.squared
r2_hp
```

. . .

About **`r round(100 * r2_hp)`%** of the variation in hp is shared with weight, so only **`r 100 - round(100 * r2_hp)`%** is left to identify the effect of hp.


## Step 3: Swap In the Independent Variation {.smaller}

::: {.nonincremental}
- **Total** variation in $x_j$: $\quad SST_j = \sum_{i=1}^{n}(x_{ji} - \bar{x}_j)^2$
- **Independent** variation in $x_j$: $\quad SST_j \times (1 - R_j^2)$
:::

. . .

Replace $SST_x$ in the bivariate formula with the independent variation:

$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x} \quad \Longrightarrow \quad Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R_j^2)}$$

. . .

- If $x_j$ is uncorrelated with the other regressors, $R_j^2 = 0$ and we are back to the bivariate formula
- The factor $\frac{1}{1 - R_j^2}$ is how much collinearity **inflates** the variance. For hp: $\frac{1}{1 - `r round(r2_hp, 2)`} \approx `r round(1 / (1 - r2_hp), 2)`$, so the variance is about `r round(100 * (1 / (1 - r2_hp) - 1))`% larger than if hp were uncorrelated with weight


## Variance of $\hat{\beta}_j$: Result {.smaller}

Putting it together, under GM 1--5:

::: {.callout-important}
## Variance of the Multivariate OLS Estimator
$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R_j^2)}$$

where $\sigma^2$ is the error variance, $SST_j = \sum(x_{ji} - \bar{x}_j)^2$ is the total variation in $x_j$, and $R_j^2$ is the R-squared from regressing $x_j$ on all other regressors.
:::

. . .

Three things **increase** variance: **more noise** ($\sigma^2$ ↑), **less variation in $x_j$** ($SST_j$ ↓), **higher collinearity** ($R_j^2$ ↑)

. . .

- **Imperfect multicollinearity** (high but not perfect correlation among the $x$'s) is *allowed*, but as $R_j^2 \rightarrow 1$, $Var(\hat{\beta}_j) \rightarrow \infty$
- Highly correlated controls make it hard to **separate their individual effects** → imprecise estimates (large standard errors)


## What Increases Variance? Simulation {.smaller}

```{r}
#| echo: false
#| fig-height: 5.5
#| fig-align: center
#| cache: true

set.seed(325)
n <- 100
n_sims <- 1000
true_beta1 <- 2

# --- Simulation function ---
# DGP: y = 1 + 2*x1 + 1*x2 + error, with cor(x1, x2) = rho and sd(x1) = x1_sd
run_sim <- function(n, sigma, x1_sd, rho, n_sims) {
  estimates <- numeric(n_sims)
  for (i in 1:n_sims) {
    x2 <- rnorm(n)
    x1 <- x1_sd * (rho * x2 + sqrt(1 - rho^2) * rnorm(n))
    y <- 1 + true_beta1 * x1 + 1 * x2 + rnorm(n, sd = sigma)
    estimates[i] <- coef(lm(y ~ x1 + x2))["x1"]
  }
  estimates
}

scenarios <- c(
  "Baseline (σ = 1, sd = 1, ρ = 0.3)",
  "More noise (σ = 3)",
  "Less variation in x1 (sd = 0.3)",
  "High collinearity (ρ = 0.95)"
)

df_sims <- bind_rows(
  tibble(beta_hat = run_sim(n, sigma = 1, x1_sd = 1,   rho = 0.3,  n_sims), scenario = scenarios[1]),
  tibble(beta_hat = run_sim(n, sigma = 3, x1_sd = 1,   rho = 0.3,  n_sims), scenario = scenarios[2]),
  tibble(beta_hat = run_sim(n, sigma = 1, x1_sd = 0.3, rho = 0.3,  n_sims), scenario = scenarios[3]),
  tibble(beta_hat = run_sim(n, sigma = 1, x1_sd = 1,   rho = 0.95, n_sims), scenario = scenarios[4])
) |>
  mutate(scenario = factor(scenario, levels = scenarios))

sd_labels <- df_sims |>
  group_by(scenario) |>
  summarise(label = paste0("SD = ", sprintf("%.2f", sd(beta_hat))))

ggplot(df_sims, aes(x = beta_hat)) +
  geom_histogram(bins = 40, fill = "#5D8EE5", color = "white", alpha = 0.8) +
  geom_vline(xintercept = true_beta1, color = "#E63946", linewidth = 1, linetype = "dashed") +
  geom_text(data = sd_labels, aes(label = label), x = Inf, y = Inf,
            hjust = 1.1, vjust = 1.5, size = 5, fontface = "bold", color = "gray25") +
  facet_wrap(~ scenario, scales = "free_y", ncol = 2) +
  labs(
    x = expression(hat(beta)[1]),
    y = "Count",
    caption = paste0("1,000 simulations each | n = ", n, " | True β = ", true_beta1, " (dashed red)")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )
```


# Estimating $\sigma^2$ and the Standard Error

## The Problem: $\sigma^2$ Is Unknown {.smaller}

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R_j^2)}$$

- We can compute $SST_j$ and $R_j^2$ from our data
- But $\sigma^2 = Var(\mu | x_1, \ldots, x_k)$ is a **population parameter** --- we never observe it

. . .

**We need to estimate $\sigma^2$ from our sample**, in three steps:

::: {.nonincremental}
1. Write $\sigma^2$ as an **average** of something
2. Find a **sample stand-in** for that something
3. **Correct** for the fact that the stand-in is too small
:::


## Step 1: $\sigma^2$ Is the Average Squared Error {.smaller}

From the Gauss-Markov assumptions:

- **GM4:** $E[\mu_i] = 0$ $\qquad$ **GM5:** $Var(\mu_i) = \sigma^2$

. . .

From the stats review, the definition of variance is $Var(X) = E\big[(X - E[X])^2\big]$. Apply it to $\mu_i$:

$$\sigma^2 = Var(\mu_i) = E\big[(\mu_i - E[\mu_i])^2\big] = E\big[(\mu_i - 0)^2\big] = E[\mu_i^2]$$

. . .

- $\sigma^2$ is just the **average squared error**
- If we could see the errors, the natural estimate would be their sample average: $\; \dfrac{1}{n}\sum_{i=1}^{n} \mu_i^2$


## Step 2: Residuals as Proxies for Errors {.smaller}

**Problem:** we can't see $\mu_i$ --- computing it requires the **population** $\beta$'s, which is the whole reason we're running a regression!

. . .

::: {.nonincremental}
- **True error:** $\quad \mu_i = y_i - (\beta_0 + \beta_1 x_{1i} + \dots + \beta_k x_{ki})$ --- uses the **unknown** $\beta$'s
- **Residual:** $\quad \hat{\mu}_i = y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_{1i} + \dots + \hat{\beta}_k x_{ki})$ --- uses our **estimates**
:::

. . .

- **Plug in the residuals:** $\; \dfrac{1}{n}\sum_{i=1}^{n} \hat{\mu}_i^2 = \dfrac{SSR}{n}$

. . .

- But the residuals are **systematically smaller** than the true errors: OLS *chooses* the $\hat{\beta}$'s to make $\sum \hat{\mu}_i^2$ as small as possible --- smaller than $\sum \mu_i^2$ with the true $\beta$'s


## Step 3: The Degrees-of-Freedom Correction {.smaller}

Because the residuals are too small on average, $SSR / n$ **underestimates** $\sigma^2$.

. . .

- Recall the first-order conditions from the OLS derivation: the residuals must satisfy **$k + 1$ equations** ($\sum \hat{\mu}_i = 0$ and $\sum x_{ji} \hat{\mu}_i = 0$ for each $j$)
- So only $n - k - 1$ residuals are free to vary --- each estimated parameter "uses up" one **degree of freedom**

. . .

The fix: divide by $n - k - 1$ instead of $n$:

$$\hat{\sigma}^2 = \frac{\sum_{i=1}^{n}\hat{\mu}_i^2}{n - k - 1} = \frac{SSR}{n - k - 1}, \qquad E[\hat{\sigma}^2] = \sigma^2 \;\; \text{(unbiased)}$$

. . .

::: {.callout-note}
$\hat{\sigma} = \sqrt{\hat{\sigma}^2}$ is what R reports as **"Residual standard error."** In the bivariate case ($k = 1$) the denominator is $n - 2$, the $n - 2$ from Unit 4.
:::


## From $\hat{\sigma}^2$ to the Standard Error {.smaller}

Plug $\hat{\sigma}^2$ into the variance formula and take the square root:

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)} \xrightarrow{\text{estimate } \sigma^2 \text{ with residuals}} se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$$

. . .

- The **standard error** is the estimated standard deviation of the sampling distribution of $\hat{\beta}_j$

. . .

- It tells us: *how much would our estimate typically vary across repeated samples?*

. . .

- Every time you see $se(\hat{\beta}_j)$ in a regression table or R output, this is what's being computed

. . .

> This standard error is the foundation of everything in the next unit: **t-statistics, p-values, and confidence intervals** all depend on it.


## Seeing It in R: Every Piece by Hand {.smaller}

For `lm_multi` ($mpg \sim hp + wt$, so $n = 32$ and $k = 2$), compute $se(\hat{\beta}_{hp})$ step by step:

```{r}
#| echo: true

n_obs     <- nobs(lm_multi); k <- 2
sigma_hat <- sqrt(sum(resid(lm_multi)^2) / (n_obs - k - 1))   # sigma-hat = sqrt(SSR / (n - k - 1))
sst_hp    <- sum((mtcars$hp - mean(mtcars$hp))^2)             # SST for hp
se_hp     <- sigma_hat / sqrt(sst_hp * (1 - r2_hp))           # r2_hp from the variance slides

c("sigma-hat: by hand" = sigma_hat, "sigma-hat: R" = sigma(lm_multi))
c("se(hp): by hand" = se_hp, "se(hp): R" = summary(lm_multi)$coefficients["hp", "Std. Error"])
```

. . .

- Our $\hat{\sigma}$ matches R's **Residual standard error**, and our $se(\hat{\beta}_{hp})$ matches the **Std. Error** column of `summary(lm_multi)`


# The Bias-Variance Tradeoff

## Should You Include That Variable? {.smaller}

Adding controls involves a **tradeoff**:

. . .

| Action | Bias | Variance |
|--------|:----:|:--------:|
| Omit a **relevant** variable | ↑ Increases | ↓ Decreases |
| Include an **irrelevant** variable | No change | ↑ Increases |

. . .

- **Too few controls → bias:** an omitted variable stays in $\mu$. If it affects $y$ ($\beta_2 \neq 0$) **and** moves with $x_1$ ($\tilde{\delta}_1 \neq 0$), then $\text{Bias} = \beta_2 \tilde{\delta}_1 \neq 0$
- **Too many controls → variance:** an unneeded control that moves with $x_1$ raises $R_1^2$, leaving less independent variation in $x_1$, so $Var(\hat{\beta}_1) = \sigma^2 / [SST_1(1 - R_1^2)]$ rises

. . .

> Because you can never know *for sure* the true population parameter model,
you can never know *for sure* if a control variable is necessary or not!


## Seeing the Tradeoff {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center
#| cache: true

# DGP: y = 8.78 + 2.21*x1 + beta2*x2 + error, with cor(x1, x2) = 0.7
# Each simulation draws a fresh sample of n = 100
sim_est <- function(beta2, include_x2, n_sims = 500, n = 100, rho = 0.7) {
  map_dbl(1:n_sims, function(i) {
    x2 <- rnorm(n)
    x1 <- rho * x2 + sqrt(1 - rho^2) * rnorm(n)
    y  <- 8.78 + 2.21 * x1 + beta2 * x2 + rnorm(n)
    if (include_x2) coef(lm(y ~ x1 + x2))["x1"]
    else coef(lm(y ~ x1))["x1"]
  })
}

set.seed(42)
results <- bind_rows(
  tibble(b1 = sim_est(0,   FALSE), model = "Omit x2",    case = "x2 irrelevant (β2 = 0)"),
  tibble(b1 = sim_est(0,   TRUE),  model = "Include x2", case = "x2 irrelevant (β2 = 0)"),
  tibble(b1 = sim_est(0.5, FALSE), model = "Omit x2",    case = "x2 relevant (β2 = 0.5)"),
  tibble(b1 = sim_est(0.5, TRUE),  model = "Include x2", case = "x2 relevant (β2 = 0.5)")
)

ggplot(results, aes(x = b1, fill = model)) +
  geom_histogram(alpha = 0.5, bins = 30, color = "white", position = "identity") +
  geom_vline(xintercept = 2.21, color = "darkred", linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~ case) +
  scale_fill_manual(values = c("Omit x2" = "#5D8EE5", "Include x2" = "#EEB808")) +
  labs(x = expression(hat(beta)[1]), y = "Count",
       title = "The Bias-Variance Tradeoff",
       subtitle = "Dashed line = true β1 = 2.21 | cor(x1, x2) = 0.7 | 500 samples of n = 100",
       fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


## Reading the Simulation {.smaller}

**Left panel** ($x_2$ irrelevant):

- Both distributions centered on truth → no bias either way
- Including $x_2$ (yellow) is **wider** → unnecessary variance, because $x_2$ soaks up some of $x_1$'s variation

. . .

**Right panel** ($x_2$ relevant):

- Omitting $x_2$ (blue) is **shifted** away from truth → biased!
- Including $x_2$ (yellow) is centered on truth → unbiased (but wider)

. . .

::: {.callout-important}
## Key Takeaway
**Bias is usually worse than variance.** A biased estimate is systematically wrong no matter how much data you collect. Extra variance shrinks with larger samples.
:::


## Practical Guidance {.smaller}

::: {.callout-tip}
## Rules of Thumb for Choosing Controls
:::

1. **Include** variables you have strong theoretical reason to believe are confounders

. . .

2. **Be cautious** with highly correlated controls — can inflate variance dramatically

. . .

3. **Larger samples help** — variance decreases with $n$, so the cost of extra controls falls

. . .

4. **Check sensitivity** — if adding a control substantially changes your coefficient of interest, it's probably important


# Summary

## Key Takeaways {.smaller}

::: {.nonincremental}
- **Multivariate OLS** includes control variables to address confounding: $\; y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$
- Coefficients have **ceteris paribus** interpretations: the effect of $x_j$ holding all else constant --- under the hood, OLS is a **matching** estimator
- **OVB formula** tells us the direction of bias: $\text{Bias} = \beta_2 \cdot \tilde{\delta}_1$
- **Variance:** $Var(\hat{\beta}_j) = \sigma^2 / [SST_j(1 - R_j^2)]$; estimate $\sigma^2$ with $SSR/(n - k - 1)$ to get $se(\hat{\beta}_j)$
- **Bias-variance tradeoff:** omitting relevant variables → bias; including irrelevant variables → extra variance
- **Adjusted R²** penalizes for extra variables --- use it to compare fit, not to choose controls
- **Perfect multicollinearity** breaks OLS; imperfect multicollinearity inflates variance
:::

. . .

**Next --- Unit 6: Statistical Inference.** We'll finally be able to answer: "Is this coefficient **statistically significant**?"

