---
title: "Multivariate Regression and OLS"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 5 | Multivariate Regression and OLS"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


# From Bivariate to Multivariate OLS

## Where We've Been

- Last unit: **bivariate OLS** — one independent variable

$$y = \beta_0 + \beta_1 x + \mu$$

. . .

- We learned OLS is **unbiased** when $E[\mu | x] = 0$ (zero conditional mean)

. . .

- But this requires that *everything else* affecting $y$ is uncorrelated with $x$

. . .

- That's a **very strong** assumption — almost never holds in observational data!


## The Problem: A Concrete Example {.smaller}

Consider the effect of education on wages:

$$wage = \beta_0 + \beta_1 \cdot educ + \mu$$

. . .

- The error term $\mu$ contains: ability, family connections, motivation, experience, ...

. . .

- Many of these are **correlated with education**
  - Wealthier families → more education *and* higher wages
  - Higher ability → more education *and* higher wages

. . .

::: {.callout-important}
## The Core Problem
If confounders are in $\mu$ and correlated with $x$, then $E[\mu | x] \neq 0$ and $\hat{\beta}_1$ is **biased**.
:::


## Where We're Going

- **Today:** Multivariate OLS — adding **control variables** to address confounding

. . .

- By including variables that might be confounders, we can make the zero conditional mean assumption more plausible

. . .

::: {.callout-important}
## The Big Question
How does adding control variables help us get closer to **causal** estimates?
:::


# The Multivariate Model

## The Solution: Control for Confounders

- If family income confounds the education-wage relationship, **include it**:

. . .

$$wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$$

. . .

- Now $\beta_1$ is the effect of education **holding family income constant**

- Family income is no longer in $\mu$ — that source of bias is gone!


## The General Model {.smaller}

The multivariate linear regression model with $k$ independent variables:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

. . .

where:

- $y$ is the **dependent variable**
- $x_1, x_2, \ldots, x_k$ are the **independent variables**
- $\beta_j$ is the coefficient on $x_j$
- $\mu$ is the error term — now only contains factors **not included** in the model


## Interpreting Coefficients

- Each coefficient has a **ceteris paribus** interpretation:

. . .

$$\beta_j = \frac{\partial y}{\partial x_j}$$

. . .

- The change in $y$ for a one-unit increase in $x_j$, **holding all other $x$'s constant**

. . .

- This is the key difference from bivariate OLS!
  - Bivariate: "associated with" (everything else floating)
  - Multivariate: "holding constant" (controlling for other factors)


## Interpretation Example {.smaller}

$$wage = \beta_0 + \beta_1 \cdot educ + \beta_2 \cdot family\_income + \mu$$

. . .

- $\beta_1$: Each additional year of education is associated with $\beta_1$ dollars more in wages, **holding family income constant**

. . .

- $\beta_2$: Each additional dollar of family income is associated with $\beta_2$ dollars more in wages, **holding education constant**

. . .

::: {.callout-tip}
## The Mantra
Always include **"holding ___ constant"** (or "controlling for ___") when interpreting multivariate coefficients.
:::


# Multivariate OLS in R

## Adding Controls is Easy! {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

Same `lm()` function — just use `+` to add variables:

. . .

```{r}
#| echo: true

data(mtcars)

# Bivariate: mpg on horsepower
lm_biv <- lm(mpg ~ hp, data = mtcars)

# Multivariate: add weight as control
lm_multi <- lm(mpg ~ hp + wt, data = mtcars)
```


## Comparing Results {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Bivariate:**
```{r}
#| echo: true
coef(lm_biv)
```
:::

::: {.column width="50%"}
**Multivariate:**
```{r}
#| echo: true
coef(lm_multi)
```
:::

::::

. . .

- Coefficient on `hp` dropped from **-0.068** to **-0.032** — less than half!

. . .

- Why? Horsepower and weight are correlated, and weight independently reduces MPG

- The bivariate estimate was **conflating** the effect of hp with the effect of weight


## Visualizing the Difference {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

library(patchwork)

p1 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#002169", linewidth = 1.2) +
  labs(title = "Bivariate: mpg ~ hp",
       subtitle = paste0("β_hp = ", round(coef(lm_biv)["hp"], 3)),
       x = "Horsepower", y = "MPG") +
  theme_minimal(base_size = 14)

# For multivariate, show partial residual plot
mtcars$resid_mpg <- residuals(lm(mpg ~ wt, data = mtcars))
mtcars$resid_hp <- residuals(lm(hp ~ wt, data = mtcars))

p2 <- ggplot(mtcars, aes(x = resid_hp, y = resid_mpg)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#002169", linewidth = 1.2) +
  labs(title = "After controlling for weight",
       subtitle = paste0("β_hp = ", round(coef(lm_multi)["hp"], 3)),
       x = "Horsepower (residualized)", y = "MPG (residualized)") +
  theme_minimal(base_size = 14)

p1 + p2
```

The right panel shows the relationship between hp and mpg **after removing the effect of weight** from both variables.


## Try It: Your First Multivariate Regression {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**Using `mtcars`, estimate a regression of `mpg` on `wt`, `hp`, and `drat` (rear axle ratio). Then answer:**

1. What is the coefficient on `wt`? Interpret it in a sentence.
2. How does the `wt` coefficient compare to a bivariate regression of `mpg ~ wt`?

. . .

```{r}
#| output-location: fragment
#| echo: true

lm(mpg ~ wt + hp + drat, data = mtcars) |> coef()
```

. . .

```{r}
#| echo: true

# Compare to bivariate
lm(mpg ~ wt, data = mtcars) |> coef()
```


# Multivariate OLS as Matching

## The Intuition {.smaller}

- When we "control for" a variable, we're **comparing like with like**

. . .

- Multivariate OLS is essentially a **matching** estimator:
  1. **Group** observations by similar values of the control variables
  2. **Compare** $x$ and $y$ **within** each group
  3. **Take a weighted average** across groups

. . .

- Think of it as making "apples-to-apples" comparisons


## Example: Education, Wages, and Industry {.smaller}

```{r}
#| echo: true
#| code-line-numbers: false

set.seed(42)
n <- 2000

wage_data <- tibble(
  educ_years = sample(c(12, 14, 16), n, replace = TRUE,
                      prob = c(0.4, 0.3, 0.3)),
  # Higher ed → more likely in professional services
  ind_professional = rbinom(n, 1, prob = case_when(
    educ_years == 12 ~ 0.20,
    educ_years == 14 ~ 0.40,
    educ_years == 16 ~ 0.65
  )),
  # True education effect: $3000/year; industry premium: $15000
  wage = 25000 + 3000 * (educ_years - 12) +
    15000 * ind_professional + rnorm(n, 0, 8000)
) |>
  mutate(industry = ifelse(ind_professional == 1,
                           "Professional", "Other"))
```


## The Naive Estimate is Too Large {.smaller}

```{r}
#| echo: true

# Naive: just education
lm_naive <- lm(wage ~ educ_years, data = wage_data)

# Controlling for industry
lm_ctrl <- lm(wage ~ educ_years + ind_professional, data = wage_data)
```

. . .

:::: {.columns}

::: {.column width="50%"}
**Without control:**
```{r}
#| echo: false
coef(lm_naive)["educ_years"] |> round(0)
```
:::

::: {.column width="50%"}
**With industry control:**
```{r}
#| echo: false
coef(lm_ctrl)["educ_years"] |> round(0)
```
:::

::::

. . .

- True effect is **\$3,000/year** — the controlled estimate is much closer!

- Naive estimate was inflated because educated workers sort into high-paying industries


## Visualizing the Matching {.smaller}

```{r}
#| echo: false
#| fig-height: 4.5
#| fig-align: center

cell_means <- wage_data |>
  group_by(industry, educ_years) |>
  summarise(avg_wage = mean(wage), .groups = "drop")

ggplot(wage_data, aes(x = factor(educ_years), y = wage, color = industry)) +
  geom_jitter(alpha = 0.08, width = 0.2, size = 0.8) +
  geom_point(data = cell_means,
             aes(x = factor(educ_years), y = avg_wage, color = industry),
             size = 5, shape = 18) +
  geom_line(data = cell_means,
            aes(x = factor(educ_years), y = avg_wage,
                color = industry, group = industry),
            linewidth = 1.2) +
  scale_color_manual(values = c("Other" = "#5D8EE5", "Professional" = "#EEB808")) +
  labs(x = "Years of Education", y = "Wage ($)",
       title = "Wages by Education and Industry",
       subtitle = "Diamonds show group means — the within-industry slopes reflect the true effect",
       color = "Industry") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

**Controlling for industry** = comparing within the blue dots and within the yellow dots separately.


## Class Activity: Matching by Hand {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Compute the average wage **within each education-industry cell**:

```{r}
#| echo: true
#| output-location: fragment

wage_data |>
  group_by(industry, educ_years) |>
  summarise(avg_wage = mean(wage) |> round(0),
            n = n(), .groups = "drop") |>
  arrange(industry, educ_years)
```

. . .

**Discuss:** Within each industry, how much do wages increase per year of education? Is it close to $3,000?


# The Zero Conditional Mean Assumption

## The Multivariate Version {.smaller}

- In bivariate OLS: $E[\mu | x] = 0$

. . .

- In multivariate OLS:

$$E[\mu | x_1, x_2, \ldots, x_k] = 0$$

. . .

- Everything remaining in $\mu$ must be uncorrelated with **all** included variables

. . .

- This is **more plausible** than the bivariate version — we've pulled confounders out of $\mu$!


## When Does the Assumption Hold? {.smaller}

:::: {.columns}

::: {.column width="50%"}
**More plausible when:**

- We include key confounders
- Good theoretical motivation for controls
- Rich set of observable controls
:::

::: {.column width="50%"}
**Less plausible when:**

- Important confounders are unobservable (ability, motivation)
- Wrong functional form
- Simultaneous causality
:::

::::

. . .

::: {.callout-warning}
## A Warning
Adding controls helps but **doesn't guarantee** causality. Unobservable confounders can always lurk in $\mu$.
:::


# Omitted Variable Bias

## The OVB Formula {.smaller}

Suppose the **true** model is:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \mu$$

But we omit $x_2$ and estimate:

$$\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1 x_1$$

. . .

Then:

$$\tilde{\beta}_1 = \hat{\beta}_1 + \hat{\beta}_2 \cdot \tilde{\delta}_1$$

where $\tilde{\delta}_1$ comes from regressing $x_2$ on $x_1$: $\quad x_2 = \tilde{\delta}_0 + \tilde{\delta}_1 x_1$

. . .

$$\underbrace{\text{Bias}(\tilde{\beta}_1)}_{\text{What we get wrong}} = \underbrace{\beta_2}_{\text{Effect of omitted on } y} \times \underbrace{\tilde{\delta}_1}_{\text{Relationship of omitted with } x_1}$$


## Signing the Bias {.smaller}

The bias depends on **two things** — and you can figure out the direction!

. . .

| | $Corr(x_1, x_2) > 0$ | $Corr(x_1, x_2) < 0$ |
|-----|:-----:|:-----:|
| $\beta_2 > 0$ | **Positive** bias ↑ | **Negative** bias ↓ |
| $\beta_2 < 0$ | **Negative** bias ↓ | **Positive** bias ↑ |

. . .

- Positive × Positive = Positive (overestimate)
- Positive × Negative = Negative (underestimate)

. . .

::: {.callout-tip}
## Zero Bias When...
If **either** $\beta_2 = 0$ (omitted variable doesn't affect $y$) **or** $\tilde{\delta}_1 = 0$ (omitted variable uncorrelated with $x_1$), there is **no bias**.
:::


## Example: Education, Wages, and Ability {.smaller}

Estimating the return to education while omitting ability:

. . .

- $\beta_2$ (effect of ability on wages): **Positive** (more able → higher wages)

. . .

- $\tilde{\delta}_1$ (correlation of ability with education): **Positive** (more able → more education)

. . .

$$\text{Bias} = (+)(+) = \text{Positive}$$

. . .

> Our estimate of the return to education is **too large** — it includes part of the return to ability.


## Example: Policing and Crime {.smaller}

Estimating the effect of police on crime while omitting poverty:

. . .

- $\beta_2$ (effect of poverty on crime): **Positive** (more poverty → more crime)

. . .

- $\tilde{\delta}_1$ (correlation of poverty with policing): **Positive** (high-poverty areas → more police)

. . .

$$\text{Bias} = (+)(+) = \text{Positive}$$

. . .

> The naive estimate makes policing look **less effective** than it actually is (or even that it *increases* crime).


## Class Activity: Sign the Bias! {.smaller}

::: {.callout-note icon=false}
## 🤔 Think-Pair-Share
:::

For each scenario, determine the **direction** of the omitted variable bias:

. . .

1. **Model:** $health = \beta_0 + \beta_1(exercise) + \mu$. Omitted: income.
   - Higher income → more exercise? Higher income → better health?

. . .

2. **Model:** $test\_score = \beta_0 + \beta_1(class\_size) + \mu$. Omitted: school funding.
   - More funding → smaller classes? More funding → higher scores?

. . .

3. **Model:** $crime = \beta_0 + \beta_1(unemployment) + \mu$. Omitted: urban density.
   - Higher density → more unemployment? Higher density → more crime?


## Seeing OVB in Action {.smaller}

```{r}
#| echo: true
#| fig-height: 3
#| fig-align: center

# Simulate data with a confounder
set.seed(123)
n <- 500
sim <- tibble(
  ability = rnorm(n),
  educ = 12 + 2 * ability + rnorm(n),          # ability → education
  wage = 20 + 3 * educ + 5 * ability + rnorm(n, 0, 3)  # ability → wages
)

# Compare estimates
c("Naive (omit ability)" = coef(lm(wage ~ educ, data = sim))["educ"],
  "Controlled" = coef(lm(wage ~ educ + ability, data = sim))["educ"],
  "True effect" = 3)
```

. . .

The naive estimate (~4.5) **overestimates** the true effect (3) because ability is positively correlated with both education and wages.


# Goodness of Fit: Adjusted R²

## The Problem with R² {.smaller}

- Recall: $R^2$ measures the fraction of variation in $y$ explained by the model

. . .

- **Problem:** $R^2$ can **never decrease** when you add more variables — even irrelevant ones!

. . .

```{r}
#| echo: true

# Add a completely irrelevant variable
mtcars$random_noise <- rnorm(nrow(mtcars))

c("R² without noise" = summary(lm(mpg ~ hp + wt, data = mtcars))$r.squared,
  "R² with noise" = summary(lm(mpg ~ hp + wt + random_noise, data = mtcars))$r.squared)
```

. . .

- $R^2$ went up even though `random_noise` is meaningless!


## The Fix: Adjusted R² {.smaller}

The **adjusted R-squared** penalizes for the number of variables:

$$\bar{R}^2 = 1 - \frac{(1 - R^2)(n - 1)}{n - k - 1}$$

. . .

```{r}
#| echo: true

c("Adj R² without noise" = summary(lm(mpg ~ hp + wt, data = mtcars))$adj.r.squared,
  "Adj R² with noise" = summary(lm(mpg ~ hp + wt + random_noise, data = mtcars))$adj.r.squared)
```

. . .

- Adjusted $R^2$ **decreased** — correctly penalizing the irrelevant variable

. . .

::: {.callout-tip}
## Rule of Thumb
Use **adjusted** $R^2$ when comparing models with different numbers of variables.
:::


# Assumptions and Multicollinearity

## Gauss-Markov Assumptions (Extended) {.smaller}

The multivariate versions:

. . .

1. **Linear in parameters:** $y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + \mu$

. . .

2. **Random sampling**

. . .

3. **No perfect multicollinearity** ← NEW!

. . .

4. **Zero conditional mean:** $E[\mu | x_1, \ldots, x_k] = 0$

. . .

5. **Homoskedasticity:** $Var(\mu | x_1, \ldots, x_k) = \sigma^2$


## Perfect Multicollinearity {.smaller}

**Perfect multicollinearity**: one variable is an exact linear function of others

. . .

Examples:

- Including `income_dollars` and `income_thousands` (one = other × 1000)
- Including `parent1_income` + `parent2_income` + `family_income` when family = parent1 + parent2
- Including dummies for **all** categories plus an intercept

. . .

```{r}
#| echo: true
#| error: true

# Create a perfectly collinear variable
mtcars$hp_doubled <- 2 * mtcars$hp
lm(mpg ~ hp + hp_doubled + wt, data = mtcars) |> coef()
```

. . .

R drops the collinear variable and gives you `NA` — but you should fix the model!


## Imperfect Multicollinearity {.smaller}

- **Imperfect** multicollinearity (high but not perfect correlation) is *allowed* but problematic

. . .

- It inflates the **variance** of your estimates:

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j (1 - R_j^2)}$$

. . .

- $R_j^2$ = R-squared from regressing $x_j$ on all other $x$'s
- As $R_j^2 \rightarrow 1$, variance $\rightarrow \infty$

. . .

- High correlation between controls makes it hard to **separate their individual effects**
- Your estimates become **imprecise** (large standard errors)


## Visualizing Multicollinearity {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(42)
n_sim <- 500
rho_low <- 0.2
rho_high <- 0.95

sim_mc <- function(rho, n_sim, n = 100) {
  map_dbl(1:n_sim, function(i) {
    x1 <- rnorm(n)
    x2 <- rho * x1 + sqrt(1 - rho^2) * rnorm(n)
    y <- 2 + 3 * x1 + 1 * x2 + rnorm(n)
    coef(lm(y ~ x1 + x2))["x1"]
  })
}

set.seed(42)
low_corr <- tibble(b1 = sim_mc(rho_low, n_sim), label = paste0("Low correlation (r = ", rho_low, ")"))
high_corr <- tibble(b1 = sim_mc(rho_high, n_sim), label = paste0("High correlation (r = ", rho_high, ")"))

bind_rows(low_corr, high_corr) |>
  ggplot(aes(x = b1, fill = label)) +
  geom_histogram(alpha = 0.6, bins = 40, color = "white", position = "identity") +
  geom_vline(xintercept = 3, color = "darkred", linewidth = 1.5, linetype = "dashed") +
  scale_fill_manual(values = c("#5D8EE5", "#EEB808")) +
  labs(x = expression(hat(beta)[1]),
       y = "Count",
       title = "Multicollinearity Inflates Variance",
       subtitle = "Both centered on true value (3), but high correlation → much wider spread",
       fill = "") +
  annotate("text", x = 3, y = 55, label = "True β₁ = 3", color = "darkred",
           hjust = -0.1, fontface = "italic", size = 5) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```


# The Bias-Variance Tradeoff

## Should You Include That Variable? {.smaller}

Adding controls involves a **tradeoff**:

. . .

| Action | Bias | Variance |
|--------|:----:|:--------:|
| Omit a **relevant** variable | ↑ Increases | ↓ Decreases |
| Include an **irrelevant** variable | No change | ↑ Increases |

. . .

- **Too few controls:** risk bias (omitted variable bias)
- **Too many controls:** risk inflated variance (imprecise estimates)


## Seeing the Tradeoff {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center
#| cache: true

set.seed(124890)

# Case 1: x2 is irrelevant (beta2 = 0)
pop_irr <- tibble(
  x2 = sample(1:100, 10000, replace = TRUE),
  x1 = 5.68 + 1.89 * x2 + rnorm(10000),
  y = 8.78 + 2.21 * x1 + rnorm(10000)
)

# Case 2: x2 is relevant (beta2 != 0)
pop_rel <- tibble(
  x2 = sample(1:100, 10000, replace = TRUE),
  x1 = 5.68 + 1.89 * x2 + rnorm(10000),
  y = 8.78 + 2.21 * x1 + 0.15 * x2 + rnorm(10000)
)

sim_est <- function(pop, include_x2, n_sims = 500, ss = 100) {
  map_dbl(1:n_sims, function(i) {
    s <- slice_sample(pop, n = ss)
    if (include_x2) coef(lm(y ~ x1 + x2, data = s))["x1"]
    else coef(lm(y ~ x1, data = s))["x1"]
  })
}

set.seed(42)
results <- bind_rows(
  tibble(b1 = sim_est(pop_irr, FALSE), model = "Omit x₂", case = "x₂ irrelevant (β₂ = 0)"),
  tibble(b1 = sim_est(pop_irr, TRUE), model = "Include x₂", case = "x₂ irrelevant (β₂ = 0)"),
  tibble(b1 = sim_est(pop_rel, FALSE), model = "Omit x₂", case = "x₂ relevant (β₂ ≠ 0)"),
  tibble(b1 = sim_est(pop_rel, TRUE), model = "Include x₂", case = "x₂ relevant (β₂ ≠ 0)")
)

ggplot(results, aes(x = b1, fill = model)) +
  geom_histogram(alpha = 0.5, bins = 30, color = "white", position = "identity") +
  geom_vline(xintercept = 2.21, color = "darkred", linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~ case) +
  scale_fill_manual(values = c("Omit x₂" = "#5D8EE5", "Include x₂" = "#EEB808")) +
  labs(x = expression(hat(beta)[1]), y = "Count",
       title = "The Bias-Variance Tradeoff",
       subtitle = "Dashed line = true β₁ = 2.21",
       fill = "Model") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


## Reading the Simulation {.smaller}

**Left panel** ($x_2$ irrelevant):

- Both distributions centered on truth → no bias either way
- Including $x_2$ (yellow) is **wider** → unnecessary variance

. . .

**Right panel** ($x_2$ relevant):

- Omitting $x_2$ (blue) is **shifted** away from truth → biased!
- Including $x_2$ (yellow) is centered on truth → unbiased (but slightly wider)

. . .

::: {.callout-important}
## Key Takeaway
**Bias is usually worse than variance.** A biased estimate is systematically wrong no matter how much data you collect. Extra variance shrinks with larger samples.
:::


## Practical Guidance {.smaller}

::: {.callout-tip}
## Rules of Thumb for Choosing Controls
:::

1. **Include** variables you have strong theoretical reason to believe are confounders

. . .

2. **Be cautious** with highly correlated controls — can inflate variance dramatically

. . .

3. **Larger samples help** — variance decreases with $n$, so the cost of extra controls falls

. . .

4. **Check sensitivity** — if adding a control substantially changes your coefficient of interest, it's probably important


# Putting It All Together

## Try It: Full Analysis {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `wage_data` we created, estimate three models and compare:

```{r}
#| echo: true
#| output-location: fragment

# Model 1: No controls
m1 <- lm(wage ~ educ_years, data = wage_data)

# Model 2: Control for industry
m2 <- lm(wage ~ educ_years + ind_professional, data = wage_data)

# Compare
tibble(
  Model = c("No controls", "With industry control"),
  `β_educ` = c(coef(m1)["educ_years"], coef(m2)["educ_years"]),
  `R²` = c(summary(m1)$r.squared, summary(m2)$r.squared),
  `Adj R²` = c(summary(m1)$adj.r.squared, summary(m2)$adj.r.squared)
) |> knitr::kable(digits = 2)
```


## Interpreting Regression Output {.smaller}

::: {.callout-note icon=false}
## 🖥️ Practice
:::

```{r}
#| echo: true

summary(m2)
```


## Class Activity: OVB Detective {.smaller}

::: {.callout-note icon=false}
## 🤔 Think-Pair-Share
:::

A researcher estimates house prices:

| Variable | Coefficient | Std. Error |
|----------|:-----------:|:----------:|
| (Intercept) | 50.23 | 12.45 |
| sqft | 0.15 | 0.02 |
| bedrooms | -8.72 | 3.21 |
| bathrooms | 24.56 | 4.89 |

*Dependent variable: price in $1000s. n = 546, Adj R² = 0.68*

. . .

**Questions:**

1. What happens to price if you add 100 sqft? (Holding bedrooms and bathrooms constant)
2. Why is the bedroom coefficient **negative**? (Hint: we're controlling for sqft!)
3. What omitted variable might bias the sqft coefficient? What direction?


# Summary

## Key Takeaways {.smaller}

- **Multivariate OLS** includes control variables to address confounding:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

. . .

- Coefficients have **ceteris paribus** interpretations: the effect of $x_j$ holding all else constant

. . .

- **OVB formula** tells us the direction of bias: $\text{Bias} = \beta_2 \cdot \tilde{\delta}_1$

. . .

- **Bias-variance tradeoff:** omitting relevant variables → bias; including irrelevant variables → extra variance

. . .

- **Adjusted R²** penalizes for extra variables — use it for model comparison

. . .

- **Perfect multicollinearity** breaks OLS; imperfect multicollinearity inflates variance

## What's Next?

- **Unit 6:** Statistical Inference — hypothesis testing and confidence intervals for regression coefficients

. . .

- We'll finally be able to answer: "Is this coefficient **statistically significant**?"
