---
title: "Good Data Practices with R and RStudio"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
embed-resources: true
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 2 | Good Data Practices"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


# Good Data Practices

## When Good Economists Use Bad Data

- What is the role of the government in times of economic crisis?
- Question received renewed interest in 2008-2009 financial crisis

## When Good Economists Use Bad Data

- In 2010, economists Reinhart and Rogoff published influential research on government debt and economic growth.

- **Key finding:** When a country's debt exceeds 90% of GDP, economic growth drops *dramatically* and becomes negative.

## When Good Economists Use Bad Data

- This strongly influenced policy debates worldwide during the Great Recession.
- It suggested a policy of *austerity*--cutting government expenditures during an economic crisis--was the correct policy
- Went against more established, *Keynesian* policies

```{r}
#| echo: false

cats <- c("0-30%", "30-60%", "60-90%", "> 90%")
rr_calcs <- c(4.1, 2.9, 3.4, -0.1)
hap_calcs <- c(4.2, 3.1, 3.2, 2.2)

rr_data <- tibble(debt_ratio = factor(cats, levels = cats), growth = rr_calcs)
hap_data <- tibble(debt_ratio = factor(cats, levels = cats), growth = hap_calcs)
```

## Reinhart-Rogoff Results

```{r}
#| echo: false
#| fig-height: 4

p1 <- ggplot(rr_data, aes(x = debt_ratio, y = growth)) +
  geom_col(fill = "steelblue") +
  coord_cartesian(ylim = c(-0.5, 5)) +
  labs(
    title = "Reinhart-Rogoff Calculations",
    x = "Debt-to-GDP Ratio",
    y = "Average GDP Growth (%)"
  ) +
  theme_minimal()

print(p1)

```


## When Good Economists Use Bad Data

- However, there was a problem...

- In 2014, graduate student Thomas Herndon tried to replicate the results for a homework assignment.

. . .

He discovered three issues:

::: {.incremental}
- Some observations were mistakenly excluded
- Observations were weighed strangely
- **Excel spreadsheet errors** — some cells weren't included in calculations!
:::

## Corrected Results (Herndon, Ash, Pollin)

```{r}
#| echo: false
#| fig-height: 4

p2 <- ggplot(hap_data, aes(x = debt_ratio, y = growth)) +
  geom_col(fill = "forestgreen") +
  coord_cartesian(ylim = c(-0.5, 5)) +
  labs(
    title = "HAP Corrected Calculations",
    x = "Debt-to-GDP Ratio",
    y = "Average GDP Growth (%)"
  ) +
  theme_minimal()

print((p2))

```

## When Good Economists Use Bad Data

When the data were corrected:

- The dramatic "cliff" in economic growth **disappeared**
- The relationship was much more gradual
- The policy story was **very different**

. . .

::: {.callout-important}
## The Lesson
Good data practices are fundamental to credible research. No sophisticated technique can compensate for errors in the underlying data.
:::

## Good Data Practices

- It is **essential** to first get to know your data before you do any analysis

. . .

Some questions to ask:

- How many observations? What is missing?
- What is the distribution like? Does it make sense?
- Are there any outliers?

## Good Data Practices {.smaller}

- A few good ways to get to know your data:

1. Descriptive statistics
    - **N:** number of observations — any missing?
    - **Mean:** Average values — is it reasonable? 
    - **Min/Max:** Range — any impossible values?

2. Visualizations
    - Can also help identify similar issues
    - Bar graphs, scatterplots, etc.


## Spotting Problems: An Example {.smaller}

- Suppose we're studying education and wages.

- The raw data might look something like this:

. . .

```{r}
#| echo: false

raw_data <- tibble(
  id = 1:100,
  education = sample(c(9:20), size = 100, replace = T),
  wage = sample(c(15:200), size = 100, replace = T),
  age = sample(c(18:65), size = 100, replace = T)
)

knitr::kable(raw_data[1:5,])
```

. . .

- It's hard to tell if something is wrong by staring at a list of numbers...

## Spotting Problems: An Example {.smaller}

- So let's look at some *descriptive statistics* to help:

. . .

```{r}
#| echo: false

# Create problematic data for demonstration
problem_data <- tibble(
  Variable = c("Education (years)", "Wage ($/hr)", "Age"),
  N = c(100, 100, 100),
  Mean = c(14.2, 48.2, 35.4),
  Min = c(-5, 8, 22),
  Max = c(20, 850, 200)
)

knitr::kable(problem_data, align = "lcccc")
```

. . .

**Red Flags:**

- Education = -5 years is **impossible**
- Wage = $850/hr seems **suspiciously high**
- Age = 200 is clearly a **data entry error**

# Getting to Know R

## Starting to work with RStudio

- Before we get into using descriptive statistics to learn about data,
let's first get used to R a little!
- Hopefully, you have already installed R, RStudio, and Git
- Go ahead and open RStudio

## The RStudio Interface {.smaller}

You should see four main panels:

| Panel | Purpose |
|-------|---------|
| **Script Editor** | Write and save your R code |
| **Console** | Execute code, see output |
| **Environment** | View your stored objects |
| **Plots/Help** | View visualizations & documentation |

. . .

- For now, **let's just focus on the console**

## How to Follow Along

Throughout these slides, look for this marker:

. . .

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
Type along in your console as we go
:::

## Basic R Operations {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- R can work as a calculator!
- Type each expression below in your console:

. . .

```{r}
#| echo: true
#| output-location: fragment

4 + 2

```

. . .

```{r}
#| echo: true
#| output-location: fragment

4.65 * 9.3

```

. . .

```{r}
#| echo: true
#| output-location: fragment

(8 + 2.9) * (654 / 98)

```

## Objects and Assignment

- As an **Object-Orientated Language,** the logic of R is built around interacting with **objects**
- An object is a generic name for anything that is "saved" in R
- This includes numbers, vectors, functions, or essentially anything else we will interact with or have to reference

## Objects and Assignment {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- Store things as an object using the assignment operator `<-`
- Try typing these in your console:

. . .

```{r}
#| echo: true
#| output-location: fragment

x <- 3
x

```

. . .

```{r}
#| echo: true
#| output-location: fragment

y <- 5
y

```

. . .

```{r}
#| echo: true
#| output-location: fragment

x + y

```

## Vectors {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- A vector is a sequence of values of the same type
- Use the `c()` function to **combine** values into a vector
- Operations apply to each element:

. . .

```{r}
#| echo: true
#| output-location: fragment

ages <- c(19, 20, 21, 22, 20, 19)
ages

```

. . .

```{r}
#| echo: true
#| output-location: fragment
ages + 10

```

. . .

```{r}
#| echo: true
#| output-location: fragment
primes <- c(1, 3, 5, 7, 9, 11)

ages + primes

```

## Functions {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- Functions are pre-packaged sets of instructions that perform specific tasks
- Every function requires an *argument* to produce output
- Try the `seq()` function, which generates a sequence of numbers:

. . .

```{r}
#| echo: true
#| output-location: fragment

bb <- seq(from = 1, to = 30, by = 3)
bb

```

. . .

- `from`, `to`, and `by` are the arguments of the function
- We set `from = 1` to start at 1, `to = 30` to end at 30, and `by = 3` to increment by 3
- Every functions has a different set of arguments
- We can look up documentation on how to use a function using the `?` function


## Data Frames {.smaller}

- A data frame is the main way datasets are stored as objects
- Basically resembles a spreadsheet: rows = observations, columns = variables
- The `head()` function prints only the first `n` rows:

. . .

```{r}
#| echo: true
#| output-location: fragment

head(mtcars, 4)
```

. . .

- We can pull a specific column with `$` (useful for many functions)

. . .

```{r}
#| echo: true
#| output-location: fragment

head(mtcars$mpg, 10)

```

## Data Types {.smaller}

- Every piece of data in R has a type that determines how it can be used
- The main types are numeric (numbers), character (text), logical (true/false), and factor(categories)
- Depending on the datatype, functions will operate differently or not at all

## Data Types {.smaller}

```{r}
#| echo: true
#| output-location: fragment

# Numeric: numbers for calculations
age <- 25
price <- 19.99
temperature <- -5.2

# Math works naturally
age * price
price * 1.08

```


. . .

```{r}
#| echo: true
#| output-location: fragment

# Character: text strings (always in quotes)
name <- "Boston"
greeting <- 'Hello World'

# But can combine them
paste("Hello", name)  # "Hello Boston"

```

. . .

```{r}
#| echo: true
#| eval: false

# Can't do math on characters!
greeting + "3"  # Error!

```

## R Packages {.smaller}

- R's real power comes from **packages** — collections of functions written by others
- Think of them as "add-ons" that extend what R can do

. . .

**Two-step process:**

. . .

1. Install the package (only do it once!)
2. Load the package using `library()` (every time!)

. . .

```{r}
#| eval: false
#| echo: true

# Step 1: Install ONCE (like downloading an app)
install.packages("tidyverse")

# Step 2: Load EVERY session (like opening the app)
library(tidyverse)
```

#  Data Manipulation in R

## What is the `tidyverse`? {.smaller}

- We will work with the suite of `tidyverse` functions to do data manipulation and exploration
- The `tidyverse` is a collection of R packages designed for data science. 
- These packages share a common philosophy and work seamlessly together


. . .

Includes:

- `dplyr`: the engine for data cleaning and manipulation
- `ggplot2`: the engine for data visualization
- `readr`: the engine for reading a diverse set of data files

. . .

We will work with all of these engines.


## The `mtcars` data

- Let's learn about `tidyverse` functions using the `mtcars` data

. . .

```{r}
#| output-location: fragment

mtcars
```

## The Pipe `|>` {.smaller}

- Takes whatever is to the left of `|>` and passes it as the first argument to the function to the right of `|>`
- Kind of like "and then" in a sentence

. . .

```{r}
#| output-location: fragment

# Take mtcars AND THEN filter to high-mpg cars
mtcars |>
  filter(mpg > 25)
```

## Key Functions for Data Manipulation

- The `tidyverse` gives us a set of intuitive *verbs* for working with data
- Each function does **one thing** well
- They all follow the same pattern: first argument is the data, then the details

## `filter()`: Subset Rows {.smaller}

- **Purpose:** Keep only the rows that meet a condition

. . .

```{r}
#| output-location: fragment

mtcars |>
  filter(mpg > 25)
```

## `select()`: Choose Columns {.smaller}

- **Purpose:** Keep (or drop) specific columns from a data frame

. . .

```{r}
#| output-location: fragment

mtcars |>
  select(mpg, cyl, wt)

```

## `mutate()`: Create New Variables {.smaller}

- **Purpose:** Add new columns or modify existing ones

. . .

```{r}
#| output-location: fragment

mtcars |>
  mutate(kmpl = mpg * 0.425)
```

## Chaining Multiple Verbs Together {.smaller}

- The real power of the pipe: **chain multiple operations** in sequence
- Read it like a sentence: "Take the data, *then* filter, *then* mutate, *then* select..."

. . .

```{r}
#| output-location: fragment

mtcars |>
  filter(cyl == 4) |>
  mutate(kmpl = mpg * 0.425) |>
  select(mpg, kmpl, wt)
```

## Saving Your Work as a New Object {.smaller}

- So far, we've just been **printing** results to the console
- To keep the manipulated data, **save it as a new object** using `<-`

. . .

```{r}
#| output-location: fragment

# Create a new dataset with only efficient 4-cylinder cars
efficient_cars <- mtcars |>
  filter(cyl == 4, mpg > 25) |>
  select(mpg, hp, wt)

efficient_cars
```

. . .

- Now `efficient_cars` is stored in your environment and can be used later!

# Descriptive Statistics in R

## Remember: Know Your Data {.smaller}

Before any analysis, we need to answer three key questions:

. . .

1. **Are there errors?** Is the distribution reasonable?
    - Use descriptive stats: `mean()`, `min()`, `max()`, `summarize()`

. . .

2. **Is there missing data?** How much?
    - Use `is.na()`, `sum(is.na())`

. . .

3. **Are there outliers or patterns?**
    - Use visualizations: histograms, scatterplots

. . .

Let's learn the tools to answer these questions!

## Descriptive Statistics Functions {.smaller}

- R has built-in functions for computing common descriptive statistics
- These operate on **vectors** (i.e., columns of a data frame; reference them using `$`)

. . .

| Function | What it computes |
|----------|-----------------|
| `mean()` | Average |
| `sd()` | Standard deviation |
| `median()` | Median (middle value) |
| `min()` / `max()` | Minimum / Maximum |
| `n()` | Number of observations |

## Descriptive Statistics Functions {.smaller}

```{r}
#| output-location: fragment

mean(mtcars$mpg)
```

. . .

```{r}
#| output-location: fragment

sd(mtcars$mpg)
```

. . .

```{r}
#| output-location: fragment

min(mtcars$mpg)
```


- These are useful on their own, but become **much more powerful** inside `summarize()`

## Handling Missing Data {.smaller}

- Before computing statistics, check for **missing values** (`NA`)
- Many R functions return `NA` if any value is missing

. . .

- `is.na()` checks for missing values; `sum(is.na())` counts them
- The `airquality` dataset (built into R) has some missing values:

. . .

```{r}
#| output-location: fragment

sum(is.na(airquality$Ozone))
```

## Handling Missing Data {.smaller}

- If you try to compute the mean with missing values:

. . .

```{r}
#| output-location: fragment

mean(airquality$Ozone)
```

. . .

- Use `na.rm = TRUE` to ignore missing values:

. . .

```{r}
#| output-location: fragment

mean(airquality$Ozone, na.rm = TRUE)
```

## What Do We Do About Missing Data? {.smaller}

- In this class, we will largely **ignore missing values** using `na.rm = TRUE` (or similar arguments depending on the function)
- This is fine when data are **missing at random**

. . .

- But be careful: sometimes data are **missing systematically**
- When this happens, ignoring missing values can lead to **biased results**

## Systematically Missing Data: Examples {.smaller}

When might data be missing in a non-random way?

. . .

- **Income surveys**: High earners may refuse to report income
  - Dropping them underestimates average income

. . .

- **Health studies**: Sicker patients may drop out of a study
  - Remaining sample looks healthier than it really is

. . .

- **Student evaluations**: Students who dropped the course don't fill them out
  - Evaluations only reflect students who stayed

## Systematically Missing Data {.smaller}

- **Takeaway**: Always ask *why* data might be missing
- If missingness is related to the outcome you're studying, simple fixes like `na.rm = TRUE` can give misleading results

. . .

- Advanced methods exist to handle this (multiple imputation, selection models)
- For now, just be aware of the issue!

## `summarize()`: Compute Summary Statistics {.smaller}

- **Purpose:** Collapse a data frame into a single summary row

. . .

```{r}
#| output-location: fragment

mtcars |>
  summarize(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

## `group_by()` + `summarize()`: Grouped Summaries {.smaller}

- **Purpose:** Split data into groups and summarize each group separately
- Pair `group_by()` with `summarize()` to get statistics *by group*

. . .

```{r}
#| output-location: fragment

mtcars |>
  group_by(cyl) |>
  summarize(
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    n = n()
  )
```

. . .

- When computing descriptive statistics with `summarize()`, you don't need to use `$`!

## Storing Summary Statistics as Objects {.smaller}

- It is sometimes useful to **save** your summary statistics as an object
- This lets you reference them later — for example, in a plot or a report

. . .

```{r}
#| output-location: fragment

mpg_by_cyl <- mtcars |>
  group_by(cyl) |>
  summarize(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg)
  )

mpg_by_cyl
```

. . .

- Now `mpg_by_cyl` is stored in your environment and can be used just like any other data frame

## `count()`: See Unique Values {.smaller}

- **Purpose:** Quickly count the number of rows for each unique value of a variable
- A shortcut for `group_by() |> summarize(n = n())`
- Great for getting a feel for categorical variables in your data

. . .

```{r}
#| output-location: fragment

mtcars |>
  count(cyl)
```

. . .

- You can also count combinations of variables:

. . .

```{r}
#| output-location: fragment

mtcars |>
  count(cyl, am)
```

## `arrange()`: Sort Rows {.smaller}

- **Purpose:** Reorder rows based on one or more columns
- **Key arguments:**
  - `...`: columns to sort by (use `desc()` for descending order)

. . .

```{r}
#| output-location: fragment

mtcars |>
  arrange(desc(mpg)) |>
  select(mpg, cyl, hp)

```

## Quick Summary Tools {.smaller}

- Writing out each statistic with `summarize()` gives you full control, but sometimes you just want a **quick overview**
- R has some handy built-in functions for this

. . .

**`summary()`** — gives you min, quartiles, mean, and max for every column:

. . .

```{r}
#| output-location: fragment

summary(mtcars)
```

## Quick Summary Tools {.smaller}

**`glimpse()`** — shows the structure of your data (variable names, types, and first few values):

```{r}
#| output-location: fragment

glimpse(mtcars)
```

. . .

- `summary()` is great for spotting outliers and missing values at a glance
- `glimpse()` is great for checking variable types and making sure your data loaded correctly

## Practice: Try It Yourself! {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `mtcars` data frame, answer each question using `tidyverse` functions. Try before clicking to reveal the answer!

## Practice 1 {.smaller}

**What is the average horsepower (`hp`) of all cars in the dataset?**

- *Hint:* You only need `summarize()` and `mean()`

. . .

```{r}
#| output-location: fragment

mtcars |>
  summarize(mean_hp = mean(hp))
```

## Practice 2 {.smaller}

**What is the average MPG for cars by the number of gears it has?

- *Hint:* First `group_by()` then `summarize()` and `mean()`

. . .

```{r}
#| output-location: fragment

mtcars |>
  group_by(gear) |>
  summarize(mean_mpg = mean(mpg))
```

## Practice 3 {.smaller}

**Which car has the highest horsepower? Show only its `mpg`, `hp`, and `wt`.**

- *Hint:* `arrange()` in descending order using `desc()`, then `select()` and `head()`

. . .

```{r}
#| output-location: fragment

mtcars |>
  arrange(desc(hp)) |>
  select(mpg, hp, wt) |>
  head(1)
```

## Practice 4 {.smaller}

**Create a summary table showing the number of cars, mean weight (`wt`), and mean MPG for each number of cylinders (`cyl`).**

- *Hint:* `group_by()` + `summarize()` with `n()`, `mean(wt)`, and `mean(mpg)`

. . .

```{r}
#| output-location: fragment

mtcars |>
  group_by(cyl) |>
  summarize(
    n_cars = n(),
    mean_wt = mean(wt),
    mean_mpg = mean(mpg)
  )
```

## Practice 5 {.smaller}

**Among cars that weigh less than 3,000 lbs (`wt < 3`), what is the mean and standard deviation of their quarter-mile time (`qsec`)?**

- *Hint:* `filter()` first, then `summarize()`

. . .

```{r}
#| output-location: fragment

mtcars |>
  filter(wt < 3) |>
  summarize(
    mean_qsec = mean(qsec),
    sd_qsec = sd(qsec),
    n = n()
  )
```


# Working with Scripts

## Why Use Scripts?

- So far, we've been typing directly into the **console**
- But the console doesn't save your work!

. . .

- A **script** is a text file (`.R`) that contains your code
- You can save it, edit it, and run it again later

. . .

- Scripts are essential for **replication** — the foundation of credible research

## The Replication Standard

::: {.callout-important}
## Definition
Research meets the **replication standard** when an independent researcher can exactly duplicate the results using only the information provided.
:::

. . .

- Remember Reinhart-Rogoff: only because they shared their data and code could Thomas Herndon discover the errors
- Without replication materials, the mistakes might never have been found

## Why Replication Matters

1. **Allows others to check our work**
    - Errors happen — replication catches them

. . .

2. **Keeps our work honest**
    - If we know others will check, we're more careful

. . .

3. **Helps our future selves**
    - What seems obvious today is often opaque in 6 months
    - Good documentation means you can return to a project and understand what you did

## Creating a Script in RStudio {.smaller}

- **File → New File → R Script** (or `Ctrl/Cmd + Shift + N`)
- Type your code in the script editor (top-left panel)
- Run code by:
    - Placing cursor on a line and pressing `Ctrl/Cmd + Enter`
    - Selecting multiple lines and pressing `Ctrl/Cmd + Enter`
    - Clicking "Run" button

. . .

- **Save your script** with `Ctrl/Cmd + S` — give it a descriptive name like `unit2_analysis.R`

## Working Directories {.smaller}

- The **working directory** is the folder R looks in when reading or saving files
- If you try to load a file that's not in your working directory, R won't find it!

. . .

- Check your current working directory:

```{r}
#| eval: false
#| echo: true

getwd()
```

. . .

- You *can* set it manually with `setwd()`, but there's a better way...

## R Projects: The Easy Solution

- An **R Project** (`.Rproj` file) automatically sets your working directory to the project folder

. . .

::: {.callout-tip}
## Good News!
If you're using the R Project saved in our course Git repository, your working directory is already set correctly. You don't need to worry about it!
:::

. . .

- Just make sure you open RStudio by clicking on the `.Rproj` file (not just opening RStudio directly)

## Good Coding Practices {.smaller}

**Use comments extensively** — explain what your code does and *why*

```{r}
#| eval: false
#| echo: true

# Calculate average MPG by number of cylinders
# We group by cylinders because engine size affects fuel efficiency
mtcars |>
  group_by(cyl) |>
  summarize(mean_mpg = mean(mpg))
```

. . .

**Use descriptive variable names**

- `x1` ❌ → `years_education` ✅
- `temp` ❌ → `avg_wage_by_state` ✅

## Organizing Your Script {.smaller}

A well-organized script has clear sections:

```{r}
#| eval: false
#| echo: true

# =============================================================
# Project: Unit 2 Analysis
# Author: Your Name
# Date: February 2026
# =============================================================

# Load packages -------------------------------------------
library(tidyverse)

# Load data -----------------------------------------------
my_data <- read_csv("data/my_dataset.csv")

# Clean data ----------------------------------------------
# ... cleaning code here ...

# Analysis ------------------------------------------------
# ... analysis code here ...
```

## The Bottom Line

- **Always work in scripts**, not the console
- **Comment your code** so others (and future you) can understand it
- **Save and organize** your work logically

. . .

::: {.callout-tip}
## Remember
If you can't replicate your own results, neither can anyone else!
:::



#  Visualizations in R

## Why Visualize Data?

- Numbers are useful, but graphs often reveal patterns that statistics miss
- Effective visualization is both a **diagnostic tool** and a way to **communicate results**

. . .

::: {.callout-tip}
## Key Insight
A good visualization translates complex data into a clear visual format, revealing patterns, trends, and relationships that would be difficult to see in tables of numbers.
:::

## The Grammar of Graphics: `ggplot2`

- We use `ggplot2` (part of the `tidyverse`) for creating visualizations
- The "gg" stands for **grammar of graphics**

. . .

Every ggplot has two main components:

1. **The scaffolding**: titles, axis labels, legends, sources
2. **The content**: the actual data encoded as visual elements (points, lines, bars)

. . .

- Key insight: we build plots **layer by layer**

## Installing a New Package {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- For visualization practice, we'll use the `palmerpenguins` dataset
- First, install and load the package:

```{r}
#| eval: false
#| echo: true

# Install the package (only need to do this once!)
install.packages("palmerpenguins")

# Load it
library(palmerpenguins)
```

```{r}
#| echo: false

library(palmerpenguins)
```

## The Palmer Penguins Data {.smaller}

- The `palmerpenguins` dataset contains measurements of 344 penguins from three species
- Let's take a look:

. . .

```{r}
#| output-location: fragment

glimpse(penguins)
```

## Building a Plot: Start with Data

- Start by creating an empty plot object with `ggplot()`

. . .

```{r}
#| output-location: fragment
#| fig-height: 3.5

ggplot(data = penguins)
```

- This creates the plotting area but doesn't show any data yet!

## Building a Plot: Add a Layer {.smaller}

- Add a **layer** with points using `geom_point()`
- Specify an **aesthetic mapping** with `aes()`: which variables go where

. . .

```{r}
#| output-location: fragment
#| fig-height: 3.5

ggplot(data = penguins) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g))
```

## What Do We See?

- Clear **positive relationship**: penguins with longer flippers tend to be heavier
- But could this be driven by **different species**?

. . .

- Let's encode species information using **color**

## Adding Color by Species {.smaller}

```{r}
#| output-location: fragment
#| fig-height: 4

ggplot(data = penguins) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, color = species))
```

## What Do We Learn?

- Gentoo penguins (blue) are generally larger
- The positive relationship exists **within each species**
- This suggests the relationship isn't just due to species differences

## What is a Line of Best Fit?

- A **line of best fit** is a straight line that best summarizes the relationship between two variables
- It is the line that minimizes the total distance between itself and all the data points

. . .

- Useful for:
  - Visualizing the **direction** of a relationship (positive or negative)
  - Summarizing the **overall trend** in the data
  - Making **predictions** for values we haven't observed

## Adding a Line of Best Fit {.smaller}

- Add a line of best fit with `geom_smooth()`
- This is a new **layer**, not a new aesthetic

. . .

```{r}
#| output-location: fragment
#| fig-height: 3.5

ggplot(data = penguins) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_smooth(aes(x = flipper_length_mm, y = body_mass_g), method = "lm")
```

## Proper Labeling {.smaller}

- Good labels are **essential** for clear communication
- Use `labs()` to add titles, axis labels, and legend titles

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(data = penguins) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_smooth(aes(x = flipper_length_mm, y = body_mass_g), method = "lm") +
  labs(
    title = "Body Mass and Flipper Length",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

## Histograms: Visualizing Distributions {.smaller}

- Histograms show the distribution of a **single continuous variable**
- Use `geom_histogram()` to create them

. . .

```{r}
#| output-location: fragment
#| fig-height: 3.5

ggplot(penguins) +
  geom_histogram(aes(x = body_mass_g), binwidth = 200,
                 fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Penguin Body Mass",
       x = "Body Mass (g)", y = "Count") +
  theme_minimal()
```

## The `binwidth` Argument {.smaller}

- `binwidth` controls how wide each bar is
- Smaller binwidths = more bars = more detail

```{r}
#| echo: false
#| layout-ncol: 3
#| fig-height: 3

ggplot(penguins) +

  geom_histogram(aes(x = body_mass_g), binwidth = 100, fill = "steelblue", alpha = 0.7) +
  labs(title = "binwidth = 100", x = "Body Mass (g)", y = "Count") +
  theme_minimal()

ggplot(penguins) +
  geom_histogram(aes(x = body_mass_g), binwidth = 200, fill = "steelblue", alpha = 0.7) +
  labs(title = "binwidth = 200", x = "Body Mass (g)", y = "Count") +
  theme_minimal()

ggplot(penguins) +
  geom_histogram(aes(x = body_mass_g), binwidth = 500, fill = "steelblue", alpha = 0.7) +
  labs(title = "binwidth = 500", x = "Body Mass (g)", y = "Count") +
  theme_minimal()
```

## Quick Reference: Key ggplot2 Functions {.smaller}

| Function | What it does |
|----------|--------------|
| `ggplot(data)` | Initialize plot with data |
| `aes(x, y, color, ...)` | Map variables to aesthetics |
| `geom_point()` | Add points (scatterplot) |
| `geom_smooth()` | Add trend line |
| `geom_histogram()` | Add histogram |
| `geom_col()` | Add bar chart |
| `labs()` | Add labels and titles |
| `theme_minimal()` | Clean, minimal theme |

## Building Intuition

The layered approach of ggplot2 mirrors how we think about data analysis:

. . .

1. Start with the **data**
2. Add **visual representations** (geoms)
3. Refine with **labels and formatting**

. . .

Experiment with different geometries, aesthetics, and layers to find the visualization that best tells your data's story!

## Practice: Visualization Exercises {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `penguins` data, try to create each of the following visualizations. Give each one a try before clicking to reveal the answer!

## Visualization Practice 1 {.smaller}

**Create a histogram of penguin flipper length (`flipper_length_mm`).**

- *Hint:* Use `geom_histogram()` with `aes(x = ...)`

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm), binwidth = 5,
                 fill = "steelblue", alpha = 0.7) +
  labs(x = "Flipper Length (mm)", y = "Count") +
  theme_minimal()
```

## Visualization Practice 2 {.smaller}

**Create a scatterplot of bill length (`bill_length_mm`) vs. bill depth (`bill_depth_mm`).**

- *Hint:* Use `geom_point()` with `aes(x = ..., y = ...)`

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm)) +
  labs(x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  theme_minimal()
```

## Visualization Practice 3 {.smaller}

**Modify the previous scatterplot to color the points by species.**

- *Hint:* Add `color = species` inside `aes()`

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(penguins) +
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  labs(x = "Bill Length (mm)", y = "Bill Depth (mm)", color = "Species") +
  theme_minimal()
```

## Visualization Practice 4 {.smaller}

**Add a line of best fit to the scatterplot of flipper length vs. body mass, and add proper axis labels and a title.**

- *Hint:* Combine `geom_point()`, `geom_smooth(method = "lm")`, and `labs()`

. . .

```{r}
#| output-location: fragment
#| fig-height: 2.5

ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g), alpha = 0.6) +
  geom_smooth(aes(x = flipper_length_mm, y = body_mass_g), method = "lm") +
  labs(title = "Penguin Body Mass vs. Flipper Length",
       x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()
```

