---
title: "Good Data Practices with R and RStudio"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 2 | Good Data Practices"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(palmerpenguins)
```


# Good Data Practices

## When Good Economists Use Bad Data

- What is the role of the government in times of economic crisis?
- Question received renewed interest in 2008-2009 financial crisis

## When Good Economists Use Bad Data

- In 2010, economists Reinhart and Rogoff published influential research on government debt and economic growth.

- **Key finding:** When a country's debt exceeds 90% of GDP, economic growth drops *dramatically* and becomes negative.

## When Good Economists Use Bad Data

- This strongly influenced policy debates worldwide during the Great Recession.
- It suggested a policy of *austerity*--cutting government expenditures during an economic crisis--was the correct policy
- Went against more established, *Keynesian* policies

```{r}
#| echo: false

cats <- c("0-30%", "30-60%", "60-90%", "> 90%")
rr_calcs <- c(4.1, 2.9, 3.4, -0.1)
hap_calcs <- c(4.2, 3.1, 3.2, 2.2)

rr_data <- tibble(debt_ratio = factor(cats, levels = cats), growth = rr_calcs)
hap_data <- tibble(debt_ratio = factor(cats, levels = cats), growth = hap_calcs)
```

## Reinhart-Rogoff Results

```{r}
#| echo: false
#| fig-height: 4

p1 <- ggplot(rr_data, aes(x = debt_ratio, y = growth)) +
  geom_col(fill = "steelblue") +
  coord_cartesian(ylim = c(-0.5, 5)) +
  labs(
    title = "Reinhart-Rogoff Calculations",
    x = "Debt-to-GDP Ratio",
    y = "Average GDP Growth (%)"
  ) +
  theme_minimal()

print(p1)

```


## When Good Economists Use Bad Data

- However, there was a problem...

- In 2014, graduate student Thomas Herndon tried to replicate the results for a homework assignment.

. . .

He discovered three issues:

::: {.incremental}
- Some observations were mistakenly excluded
- Observations were weighed strangely
- **Excel spreadsheet errors** — some cells weren't included in calculations!
:::

## Corrected Results (Herndon, Ash, Pollin)

```{r}
#| echo: false
#| fig-height: 4

p2 <- ggplot(hap_data, aes(x = debt_ratio, y = growth)) +
  geom_col(fill = "forestgreen") +
  coord_cartesian(ylim = c(-0.5, 5)) +
  labs(
    title = "HAP Corrected Calculations",
    x = "Debt-to-GDP Ratio",
    y = "Average GDP Growth (%)"
  ) +
  theme_minimal()

print((p2))

```

## When Good Economists Use Bad Data

When the data were corrected:

- The dramatic "cliff" in economic growth **disappeared**
- The relationship was much more gradual
- The policy story was **very different**

. . .

::: {.callout-important}
## The Lesson
Good data practices are fundamental to credible research. No sophisticated technique can compensate for errors in the underlying data.
:::

## Good Data Practices

- It is **essential** to first get to know your data before you do any analysis

. . .

Some questions to ask:

- How many observations? What is missing?
- What is the distribution like? Does it make sense?
- Are there any outliers?

## Good Data Practices {.smaller}

- A few good ways to get to know your data:

1. Descriptive statistics
    - **N:** number of observations — any missing?
    - **Mean:** Average values — is it reasonable? 
    - **Min/Max:** Range — any impossible values?

2. Visualizations
    - Can also help identify similar issues
    - Bar graphs, scatterplots, etc.


## Spotting Problems: An Example {.smaller}

- Suppose we're studying education and wages.

- The raw data might look something like this:

. . .

```{r}
#| echo: false

raw_data <- tibble(
  id = 1:100,
  education = sample(c(9:20), size = 100, replace = T),
  wage = sample(c(15:200), size = 100, replace = T),
  age = sample(c(18:65), size = 100, replace = T)
)

knitr::kable(raw_data[1:5,])
```

. . .

- It's hard to tell if something is wrong by staring at a list of numbers...

## Spotting Problems: An Example {.smaller}

- So let's look at some *descriptive statistics* to help:

. . .

```{r}
#| echo: false

# Create problematic data for demonstration
problem_data <- tibble(
  Variable = c("Education (years)", "Wage ($/hr)", "Age"),
  N = c(100, 100, 100),
  Mean = c(14.2, 48.2, 35.4),
  Min = c(-5, 8, 22),
  Max = c(20, 850, 200)
)

knitr::kable(problem_data, align = "lcccc")
```

. . .

**Red Flags:**

- Education = -5 years is **impossible**
- Wage = $850/hr seems **suspiciously high**
- Age = 200 is clearly a **data entry error**

# Getting to Know R

## Starting to work with RStudio

- Before we get into using descriptive statistics to learn about data,
let's first get used to R a little!
- Hopefully, you have already installed R, RStudio, and Git
- Go ahead and open RStudio

## The RStudio Interface {.smaller}

You should see four main panels:

| Panel | Purpose |
|-------|---------|
| **Script Editor** | Write and save your R code |
| **Console** | Execute code, see output |
| **Environment** | View your stored objects |
| **Plots/Help** | View visualizations & documentation |

. . .

- For now, **let's just focus on the console**

- Feel free to follow along!

## Basic R Operations {.smaller}

- R can work as a calculator!
- Type each one of the expressions below in your console and let me know what you get

. . .

```{r}
#| echo: true
#| output-location: fragment

4 + 2

```

. . .

```{r}
#| echo: true
#| output-location: fragment

4.65 * 9.3

```

. . .

```{r}
#| echo: true
#| output-location: fragment

(8 + 2.9) * (654 / 98)

```

## Objects and Assignment

- As an **Object-Orientated Language,** the logic of R is built around interacting with **objects**
- An object is a generic name for anything that is "saved" in R
- This includes numbers, vectors, functions, or essentially anything else we will interact with or have to reference

## Objects and Assignment {.smaller}

- Store things as an object using the assignment operator `<-`
- We can then just recall the object by name to access whatever is saved to it!
- We can then use it in calculations, functions, etc.

. . .

```{r}
#| echo: true
#| output-location: fragment

x <- 3
x

```

. . .

```{r}
#| echo: true
#| output-location: fragment

y <- 5
y

```

. . .

```{r}
#| echo: true
#| output-location: fragment

x + y

```

## Vectors

- A vector is a sequence of values of the same type
- We can perform operations on vectors, and it will apply it to each element of the vector

. . .

```{r}
#| echo: true
#| output-location: fragment

ages <- c(19, 20, 21, 22, 20, 19)
ages

```

. . .

```{r}
#| echo: true
#| output-location: fragment
ages + 10

```

. . .

```{r}
#| echo: true
#| output-location: fragment
primes <- c(1, 3, 5, 7, 9, 11)

ages + primes

```

## Functions {.smaller}

- Functions are pre-packaged sets of instructions that perform specific tasks
- Every function requires we input an *argument*, which it uses to produce some output

- The `seq()` function generates a sequence of numbers:

. . .

```{r}
#| echo: true
#| output-location: fragment

bb <- seq(from = 1, to = 30, by = 3)
bb

```

. . .

- `from`, `to`, and `by` are the arguments of the function
- We set `from = 1` to start at 1, `to = 30` to end at 30, and `by = 3` to increment by 3
- Every functions has a different set of arguments
- We can look up documentation on how to use a function using the `?` function


## Data Frames {.smaller}

- A data frame is the main way datasets are stored as objects
- Basically resembles a spreadsheet
- rows = observations**, **columns = variables
- (the `head()` function will print only the first `n` rows to the console)

. . .

```{r}
#| echo: true
#| output-location: fragment
#| 
head(penguins, 4)
```

. . .

- We can Pull a specific column with `$` (useful for many functions)

. . .

```{r}
#| echo: true
#| output-location: fragment

head(penguins$body_mass_g, 10)

```

## Data Types {.smaller}

- Every piece of data in R has a type that determines how it can be used
- The main types are numeric (numbers), character (text), logical (true/false), and factor(categories)
- Depending on the datatype, functions will operate differently or not at all

## Data Types {.smaller}

```{r}
#| echo: true
#| output-location: fragment

# Numeric: numbers for calculations
age <- 25
price <- 19.99
temperature <- -5.2

# Math works naturally
age * price
price * 1.08

```


. . .

```{r}
#| echo: true
#| output-location: fragment

# Character: text strings (always in quotes)
name <- "Boston"
greeting <- 'Hello World'

# But can combine them
paste("Hello", name)  # "Hello Boston"

```

. . .

```{r}
#| echo: true
#| eval: false

# Can't do math on characters!
greeting + "3"  # Error!

```

## R Packages {.smaller}

- R's real power comes from **packages** — collections of functions written by others
- Think of them as "add-ons" that extend what R can do

. . .

**Two-step process:**

. . .

1. Install the package (only do it once!)
2. Load the package using `library()` (every time!)

```{r}
#| eval: false
#| echo: true

# Step 1: Install ONCE (like downloading an app)
install.packages("tidyverse")

# Step 2: Load EVERY session (like opening the app)
library(tidyverse)
```


#  Data Analysis in R

## SLIDES IN PROGRESS

#  Visualizations in R




