---
title: "Dummy Variables and Non-Linear Models"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 7 | Dummy Variables and Non-Linear Models"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
library(palmerpenguins)
options(scipen = 999)
```


# Dummy Variables

## What Are Dummy Variables? {.smaller}

- So far we've used **quantitative** regressors: wages, education, experience...

- But many important variables are **qualitative**: race, gender, region, program participation


- A **dummy variable** (aka indicator or binary variable) translates **qualitative info into numbers**:

. . .

$$
D = \begin{cases} 1 & \text{if the characteristic is present} \\ 0 & \text{otherwise} \end{cases}
$$

. . .

- Examples: $female = 1$ if female, $south = 1$ if in the South, $treated = 1$ if in treatment group


## What Does a Dummy Look Like? {.smaller}

- Suppose we want to know the effect of completing calculus in high school on college Economics test grades

- We might have data that looks like this:

. . .

```{r}
#| echo: false

econmath_data <- wooldridge::econmath |>
  mutate(hs_math_level = case_when(calculus == 1 ~ "calculus",
    calculus == 0 ~ "algebra")) |>
  select(score, hsgpa, hs_math_level)

head(econmath_data, 6)


```

- We can't directly estimate a model using the `hs_math_level` variable
- But we *can* recode this qualitative information into quantitative form using a dummy variable

## Creating a Dummy Variable {.smaller}

- Define a new dummy variable called $calculus$ where:

. . .

$$
Calculus_i = \begin{cases} 1 & \text{if student $i$ took calculus in high school} \\ 0 & \text{otherwise} \end{cases}
$$


## Creating a Dummy Variable with `case_when()` {.smaller}

- In R, `case_when()` is useful for creating new variables (like dummies) based on conditions from existing variables

- The syntax is as follows:

```r
case_when(
  condition1 ~ value1,
  condition2 ~ value2,
  TRUE       ~ default_value   # catch-all for everything else
)
```

- Each line is a **condition ~ value** pair — R checks conditions top-to-bottom and assigns the first match
- `TRUE ~ ...` acts as an "else" — it catches anything not matched above

## Creating a Dummy Variable with `case_when()` {.smaller}


```{r}
#| echo: true

econmath_data <- econmath_data |>
  mutate(calculus = case_when(
      hs_math_level == "calculus" ~ 1,
      hs_math_level == "algebra" ~ 0,
      TRUE ~ 0
  ))

head(econmath_data, 6)
```


## Interpreting Dummy Coefficients {.smaller}

- Say we want to estimate the following regression:

$$score = \beta_0 + \beta_1(hsgpa) + \beta_2(calculus) + \mu$$

- We simply include the dummy variable on the RHS!

. . .

```{r}
#| echo: true

lm1 <- lm(score ~ hsgpa + calculus, data = wooldridge::econmath)
summary(lm1)
```

. . .

- But how do we interpret this?

## Interpreting Dummy Coefficients {.smaller}

- $\hat{\beta}_2 \approx 5.8$: Students who took calculus score **5.8 points higher** on average *relative to those who didn't*, *holding GPA constant*

- But what is actually going on "under-the-hood" when we estimate a regression with dummy variables?

## What Dummies Do: Two Regression Lines {.smaller}

- Start with our model: $\widehat{score} = \hat{\beta}_0 + \hat{\beta}_1(hsgpa) + \hat{\beta}_2(calculus)$

- **Plug in** $calculus = 0$ (no calculus):

. . .

$$\widehat{score} = \hat{\beta}_0 + \hat{\beta}_1(hsgpa) + \hat{\beta}_2(0) = \hat{\beta}_0 + \hat{\beta}_1(hsgpa)$$

. . .

- **Plug in** $calculus = 1$ (took calculus):

. . .

$$\widehat{score} = \hat{\beta}_0 + \hat{\beta}_1(hsgpa) + \hat{\beta}_2(1) = \underbrace{(\hat{\beta}_0 + \hat{\beta}_2)}_{\text{new intercept}} + \hat{\beta}_1(hsgpa)$$

. . .

::: {.callout-important}
## The Key
Both groups have the **same slope** ($\hat{\beta}_1$) — the effect of GPA is the same for both. But the calculus group's line is shifted **up** by exactly $\hat{\beta}_2$.
:::

## What Dummies Do: A Picture {.smaller}

- Dummy variables essentially allow for a *separate intercept* for each value of your qualitative variables (i.e., each group!)

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

b <- coef(lm1)

ggplot(wooldridge::econmath, aes(x = hsgpa, y = score)) +
  geom_point(aes(color = factor(calculus)), alpha = 0.3, size = 1.5) +
  geom_abline(intercept = b[1], slope = b[2],
              color = "#D55E00", linewidth = 1.2) +
  geom_abline(intercept = b[1] + b[3], slope = b[2],
              color = "#0072B2", linewidth = 1.2) +
  annotate("segment", x = 4.2, xend = 4.2,
           y = b[1] + b[2]*4.2,
           yend = b[1] + b[3] + b[2]*4.2,
           arrow = arrow(ends = "both", length = unit(0.15, "inches")),
           linewidth = 0.8) +
  annotate("text", x = 4.35,
           y = b[1] + b[3]/2 + b[2]*4.2,
           label = paste0("β₂ ≈ ", round(b[3], 1)),
           fontface = "bold", size = 5) +
  scale_color_manual(values = c("0" = "#D55E00", "1" = "#0072B2"),
                     labels = c("No Calculus", "Took Calculus")) +
  coord_cartesian(xlim = c(2, 4.5), ylim = c(30, 95)) +
  labs(x = "High School GPA", y = "Econ Score", color = NULL) +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```


## The Dummy Variable Trap {.smaller}

Why not include dummies for **both** groups?

$$score = \beta_0 + \beta_2(calculus) + \beta_3(no\_calculus) + \mu$$

. . .

- For every observation: $calculus + no\_calculus = 1$

. . .

- This is **perfect multicollinearity** — OLS cannot solve it!

. . .

::: {.callout-important}
## The Rule
With $g$ categories, include $g - 1$ dummy variables. The omitted category is the **baseline** (reference group).
:::


## Multiple Categories {.smaller}

::: columns
::: {.column width="55%"}
Three penguin species → **two** dummies:

```{r}
#| echo: true

lm_pen <- lm(flipper_length_mm ~ body_mass_g +
              species, data = penguins)
coef(lm_pen) |> round(3)
```

:::
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5

ggplot(penguins |> drop_na(),
       aes(x = body_mass_g, y = flipper_length_mm, color = species)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("#D55E00", "#009E73", "#0072B2")) +
  labs(x = "Body Mass (g)", y = "Flipper Length (mm)", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

:::
:::

. . .

- Baseline = Adelie. Chinstrap flippers are ~5.6 mm longer; Gentoo ~15.7 mm longer (same body mass)


## But How Do We Encode Multiple Categories? {.smaller}

R's `factor()` handles this behind the scenes — but what is it actually doing?

With $g$ categories, we create $g - 1$ dummies using `case_when()`:

```{r}
#| echo: true

set.seed(42)
pen_sample <- penguins |>
  drop_na() |>
  select(species, flipper_length_mm) |>
  slice_sample(n = 6) |>
  mutate(
    chinstrap = case_when(species == "Chinstrap" ~ 1, TRUE ~ 0),
    gentoo    = case_when(species == "Gentoo"    ~ 1, TRUE ~ 0)
  )

knitr::kable(pen_sample)
```

. . .

- Adelie → both dummies = 0 (the **baseline**)
- Each observation gets exactly **one** dummy = 1 (or none, if baseline)


## Interpreting: Everything Is Relative to the Baseline {.smaller}

Back to our penguin regression (baseline = **Adelie**):

```{r}
#| echo: false

coef(lm_pen) |> round(3)
```

. . .

- $\hat{\beta}_{Chinstrap} \approx 5.59$: Chinstraps have flippers **5.6 mm longer than Adelie**, holding body mass constant

. . .

- $\hat{\beta}_{Gentoo} \approx 15.68$: Gentoos have flippers **15.7 mm longer than Adelie**, holding body mass constant

. . .

- What about Chinstrap vs. Gentoo? We can **subtract**: $15.68 - 5.59 \approx 10.1$ mm

. . .

::: {.callout-important}
## Key Insight
Every dummy coefficient is a **comparison to the omitted baseline**. To compare two non-baseline groups, subtract their coefficients.
:::


## What Happens When We Change the Baseline? {.smaller}

::: columns
::: {.column width="50%"}
**Baseline = Adelie**
```{r}
#| echo: true

lm_a <- lm(flipper_length_mm ~
  body_mass_g + species,
  data = penguins)
coef(lm_a) |> round(3)
```
:::
::: {.column width="50%"}
**Baseline = Gentoo**
```{r}
#| echo: true

pen2 <- penguins |>
  mutate(species = relevel(
    species, ref = "Gentoo"))
lm_g <- lm(flipper_length_mm ~
  body_mass_g + species,
  data = pen2)
coef(lm_g) |> round(3)
```
:::
:::

. . .

- The **body mass** coefficient is identical in both → changing the baseline doesn't affect other variables!
- The dummy coefficients change sign/magnitude because the reference point changed


## When Does the Baseline Matter? {.smaller}

**It doesn't matter** when you only care about the coefficient on another variable (e.g., body mass):

- The estimate, standard error, and p-value are all **identical** regardless of baseline

. . .

**It matters** when you want to directly interpret or test a dummy coefficient:

- "Are Chinstrap flippers significantly different from **Adelie**?" → make Adelie the baseline
- "Are Chinstrap flippers significantly different from **Gentoo**?" → make Gentoo the baseline

. . .

::: {.callout-tip}
## Practical Advice
Choose a baseline that makes your comparisons natural. For studying the gender wage gap, use male as the baseline so the female coefficient directly measures the gap.
:::


## Factor Variables in R

- Creating dummies manually is tedious — R does it automatically with **factors**:

```{r}
#| echo: true
#| eval: false

# R picks first alphabetical category as baseline:
lm(y ~ x + factor(region), data = df)

# Use relevel() to choose your own baseline:
lm(y ~ x + relevel(factor(region), ref = "South"), data = df)
```

. . .

- R omits the first alphabetical category by default

- Use `relevel()` to choose a different baseline


# Non-Linear Models: Quadratics

## Why Non-Linear? {.smaller}

- Linear models assume a **constant** marginal effect — every extra year of experience raises wages by the same amount

. . .

- But many relationships curve:
  - Experience → wages: rises fast early, flattens later
  - Pollution → health: gets worse at an increasing rate
  - Advertising → sales: diminishing returns

. . .

- We can capture this with a **quadratic** (polynomial) model:

$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \mu$$


## Quadratic: The Shape {.smaller}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5.5

tibble(x = seq(0, 50, 0.5)) |>
  mutate(
    `β₂ < 0 (inverted U)` = 3.7 + 0.30*x - 0.006*x^2,
    `β₂ > 0 (U-shape)` = 15 - 0.50*x + 0.01*x^2
  ) |>
  pivot_longer(-x, names_to = "Shape", values_to = "Y") |>
  ggplot(aes(x = x, y = Y, color = Shape)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("#D55E00", "#0072B2")) +
  labs(x = "X", y = "Y", color = NULL,
       title = "The Sign of β₂ Determines the Curve") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```

- $\beta_2 < 0$: inverted-U (wages & experience)
- $\beta_2 > 0$: U-shape (cost curves)


## Estimating a Quadratic {.smaller}

$$wage = \beta_0 + \beta_1(exper) + \beta_2(exper^2) + \mu$$

```{r}
#| echo: true

wage_data <- wage1 |> mutate(exper2 = exper^2)
lm_quad <- lm(wage ~ exper + exper2, data = wage_data)
coef(lm_quad) |> round(4)
```

. . .

- $\hat{\beta}_1 > 0$ and $\hat{\beta}_2 < 0$ → inverted-U confirmed!


## Visualizing the Quadratic Fit {.smaller}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5.5

ggplot(wage1, aes(x = exper, y = wage)) +
  geom_point(alpha = 0.3, color = "gray50", size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x,
              aes(color = "Linear"), se = FALSE, linewidth = 1.2) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2),
              aes(color = "Quadratic"), se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("Linear" = "#0072B2", "Quadratic" = "#D55E00")) +
  labs(x = "Experience (years)", y = "Hourly Wage ($)", color = "Model",
       title = "Quadratic Captures the Curvature") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "bottom")
```


## Interpreting Quadratics {.smaller}

The marginal effect of $X$ on $Y$ is **not constant** — it depends on $X$:

$$\frac{\partial Y}{\partial X} = \beta_1 + 2\beta_2 X$$

. . .

With $\hat{\beta}_1 \approx 0.298$ and $\hat{\beta}_2 \approx -0.006$:

| Experience | Marginal Effect | Interpretation |
|:----------:|:---------------:|:---------------|
| 5 years | $0.298 + 2(-0.006)(5) = 0.24$ | +\$0.24/hr per extra year |
| 25 years | $0.298 + 2(-0.006)(25) ≈ 0.00$ | Near the peak! |
| 40 years | $0.298 + 2(-0.006)(40) = -0.18$ | −\$0.18/hr per extra year |

. . .

**Peak:** $x^* = -\frac{\beta_1}{2\beta_2} = -\frac{0.298}{2(-0.006)} \approx 25$ years


## Marginal Effect Across Experience {.smaller}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5

b <- coef(lm_quad)

tibble(exper = 0:50) |>
  mutate(marginal_effect = b["exper"] + 2 * b["exper2"] * exper) |>
  ggplot(aes(x = exper, y = marginal_effect)) +
  geom_line(color = "#0072B2", linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = -b["exper"] / (2 * b["exper2"]),
             linetype = "dotted", color = "#D55E00", linewidth = 1) +
  annotate("text", x = -b["exper"]/(2*b["exper2"]) + 3, y = 0.05,
           label = paste0("Peak ≈ ", round(-b["exper"]/(2*b["exper2"]), 0), " yrs"),
           color = "#D55E00", fontface = "bold", size = 5) +
  labs(x = "Experience (years)",
       y = "Marginal Effect on Wage ($/hr)",
       title = "The Marginal Effect Declines with Experience") +
  theme_minimal(base_size = 16)
```


# Logarithmic Transformations

## Why Use Logs? {.smaller}

- Many economic variables (wages, income, prices) are **right-skewed**

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4

p1 <- ggplot(wage1, aes(x = wage)) +
  geom_histogram(fill = "#0072B2", color = "white", bins = 40) +
  labs(x = "Wage ($/hr)", y = "Count", title = "Wages (in Levels)") +
  theme_minimal(base_size = 14)

p2 <- ggplot(wage1, aes(x = log(wage))) +
  geom_histogram(fill = "#D55E00", color = "white", bins = 40) +
  labs(x = "ln(Wage)", y = "Count", title = "Wages (Logged)") +
  theme_minimal(base_size = 14)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

. . .

- Logging compresses the right tail → more symmetric distribution

- More importantly: logs let us talk about **percentage changes**


## The Key Property of Logs {.smaller}

The reason logs work for percentages:

$$\ln(Y_1) - \ln(Y_0) = \ln\!\left(\frac{Y_1}{Y_0}\right) \approx \frac{Y_1 - Y_0}{Y_0}$$

. . .

A **difference in logs** ≈ the **proportional change** (for small changes)

. . .

| Wage Change | Proportional Change | Difference in Logs |
|:----------:|:-------------------:|:-----------------:|
| \$20 → \$21 | $1/20 = 0.050$ | $\ln(21)-\ln(20) = 0.049$ |
| \$20 → \$22 | $2/20 = 0.100$ | $\ln(22)-\ln(20) = 0.095$ |
| \$20 → \$25 | $5/20 = 0.250$ | $\ln(25)-\ln(20) = 0.223$ |

. . .

Works great for small changes; for larger ones, use exact formula: $100 \times (e^{\beta} - 1)$%


## Log-Linear: The Derivation {.smaller}

The most common log model in applied economics:

$$\ln(Y) = \beta_0 + \beta_1 X + \mu$$

. . .

**Step 1:** Increase $X$ by one unit (from $X_0$ to $X_0 + 1$):

$$\ln(Y_1) = \beta_0 + \beta_1(X_0 + 1)$$
$$\ln(Y_0) = \beta_0 + \beta_1 X_0$$

. . .

**Step 2:** Subtract:

$$\ln(Y_1) - \ln(Y_0) = \beta_1$$


## Log-Linear: The Derivation (cont.) {.smaller}

**Step 3:** Use our approximation $\ln(Y_1) - \ln(Y_0) \approx \frac{Y_1 - Y_0}{Y_0}$:

$$\frac{Y_1 - Y_0}{Y_0} \approx \beta_1$$

. . .

**Step 4:** Multiply by 100 to get percent:

$$\text{% change in } Y \approx 100 \times \beta_1$$

. . .

::: {.callout-important}
## Log-Linear Interpretation
A **one-unit** increase in $X$ → approximately $100 \times \beta_1$ **percent** change in $Y$
:::


## Log-Linear: Example {.smaller}

```{r}
#| echo: true

lm_loglin <- lm(log(wage) ~ educ, data = wage1)
coef(lm_loglin) |> round(4)
```

. . .

- $\hat{\beta}_1 \approx 0.083$ → one more year of education is associated with ≈ **8.3% higher wages**

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4

ggplot(wage1, aes(x = educ, y = wage)) +
  geom_point(alpha = 0.3, color = "gray50", size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x,
              aes(color = "Level-Level"), se = FALSE, linewidth = 1.2) +
  stat_function(fun = function(x) exp(coef(lm_loglin)[1] + coef(lm_loglin)[2]*x),
                aes(color = "Log-Level"), linewidth = 1.2) +
  scale_color_manual(values = c("Level-Level" = "#0072B2", "Log-Level" = "#D55E00")) +
  labs(x = "Education (years)", y = "Hourly Wage ($)", color = "Model") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```


## Other Log Models {.smaller}

| Model | Equation | $X$ changes by... | $Y$ changes by... |
|:------|:---------|:------------------:|:------------------:|
| **Level-Level** | $Y = \beta_0 + \beta_1 X$ | 1 unit | $\beta_1$ units |
| **Log-Level** | $\ln(Y) = \beta_0 + \beta_1 X$ | 1 unit | $(100\beta_1)$% |
| **Level-Log** | $Y = \beta_0 + \beta_1\ln(X)$ | 1% | $\beta_1/100$ units |
| **Log-Log** | $\ln(Y) = \beta_0 + \beta_1\ln(X)$ | 1% | $\beta_1$% |

. . .

::: {.callout-tip}
## Quick Rule
- **Logged** variable → think in **percents**
- **Level** variable → think in **units**
:::


## Log Models: Examples {.smaller}

::: columns
::: {.column width="50%"}

**Level-Log**

$bwght = 116.1 + 2.47 \cdot \ln(faminc)$

A 1% increase in family income → $2.47/100 = 0.025$ ounce increase in birthweight

:::
::: {.column width="50%"}

**Log-Log** (Elasticity)

$\ln(wage) = -0.02 + 0.83 \cdot \ln(educ)$

A 1% increase in education → ≈ 0.83% increase in wages

$\beta_1$ is the **elasticity** — a unit-free measure of responsiveness

:::
:::

. . .

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.5

bw_data <- bwght |> filter(faminc > 0)
lm_ll <- lm(bwght ~ log(faminc), data = bw_data)

p1 <- ggplot(bw_data, aes(x = faminc, y = bwght)) +
  geom_point(alpha = 0.2, color = "gray50", size = 1) +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#009E73", se = FALSE, linewidth = 1.2) +
  labs(x = "Family Income ($1000s)", y = "Birthweight (oz)",
       title = "Level-Log") +
  theme_minimal(base_size = 12)

we_data <- wage1 |> filter(educ > 0)
lm_loglog <- lm(log(wage) ~ log(educ), data = we_data)

p2 <- ggplot(we_data, aes(x = educ, y = wage)) +
  geom_point(alpha = 0.2, color = "gray50", size = 1) +
  stat_function(fun = function(x) exp(coef(lm_loglog)[1] + coef(lm_loglog)[2]*log(x)),
                color = "#D55E00", linewidth = 1.2) +
  labs(x = "Education (years)", y = "Hourly Wage ($)",
       title = "Log-Log") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```


## When to Use Logs? {.smaller}

**Commonly logged:**

- Wages, income, prices, population, firm revenue, expenditures
- Positive, right-skewed, where % changes are more meaningful

. . .

**Typically NOT logged:**

- Education (years), age, rates/percentages, dummy variables
- Variables that can be zero or negative ($\ln(0)$ is undefined!)

. . .

::: {.callout-warning}
## Watch Out
The "$100 \times \beta_1$%" approximation works for small $\beta_1$ (roughly $|\beta_1| < 0.20$).

For larger coefficients, use the **exact** formula: $100 \times (e^{\beta_1} - 1)$%
:::


# Putting It All Together


## The Big Reference Table {.smaller}

| Model | Equation | Interpretation of $\hat{\beta}_1$ |
|:------|:---------|:----------------------------------|
| **Level-Level** | $Y = \beta_0 + \beta_1 X$ | 1-unit ↑ $X$ → $\beta_1$-unit Δ $Y$ |
| **Log-Level** | $\ln(Y) = \beta_0 + \beta_1 X$ | 1-unit ↑ $X$ → $(100\beta_1)$% Δ $Y$ |
| **Level-Log** | $Y = \beta_0 + \beta_1 \ln(X)$ | 1% ↑ $X$ → $\beta_1/100$ unit Δ $Y$ |
| **Log-Log** | $\ln(Y) = \beta_0 + \beta_1 \ln(X)$ | 1% ↑ $X$ → $\beta_1$% Δ $Y$ |
| **Quadratic** | $Y = \beta_0 + \beta_1 X + \beta_2 X^2$ | Marginal effect = $\beta_1 + 2\beta_2 X$ |
| **Dummy** | $Y = \beta_0 + \beta_1 D$ | Diff. between $D=1$ and baseline |
| **Dummy (log)** | $\ln(Y) = \beta_0 + \beta_1 D$ | ≈ $(100\beta_1)$% diff. from baseline |


## Key Takeaways

1. **Dummy variables** let us include qualitative information — the coefficient is the difference in intercepts between groups

. . .

2. **Quadratic models** capture non-linear relationships — the marginal effect depends on $X$

. . .

3. **Logging** a variable means we interpret its changes in **percent** terms

. . .

4. Always ask: **is the dependent variable logged? Is $X$ logged?** That tells you the interpretation.
