# Heteroskedasticity {#sec-heteroskedasticity}

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
library(lmtest)
library(sandwich)
library(fixest)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 12) +
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(size = 11, color = "gray40"),
                panel.grid.minor = element_blank()))

```

```{=html}

<style>
.problem {
    background-color: #f8f9fa;
    border: 1px solid #dee2e6;
    border-radius: 8px;
    padding: 20px;
    margin: 20px 0;
}
.context {
    background-color: #e8f4f8;
    padding: 12px;
    border-radius: 5px;
    margin-bottom: 15px;
    border-left: 4px solid #2874a6;
}
.question {
    margin: 15px 0;
    padding: 10px;
    background-color: #fff3cd;
    border-radius: 5px;
}
.answer-row {
    display: flex;
    align-items: center;
    margin: 10px 0;
    flex-wrap: wrap;
    gap: 10px;
}
.answer-row label {
    min-width: 320px;
    font-weight: 500;
}
select {
    padding: 8px 12px;
    font-size: 14px;
    border: 2px solid #ccc;
    border-radius: 5px;
    min-width: 300px;
    cursor: pointer;
}
select:focus {
    border-color: #2874a6;
    outline: none;
}
.correct {
    border-color: #28a745 !important;
    background-color: #d4edda !important;
}
.incorrect {
    border-color: #dc3545 !important;
    background-color: #f8d7da !important;
}
.explanation {
    margin-top: 10px;
    padding: 15px;
    background-color: #e7f3ff;
    border-radius: 5px;
    border-left: 4px solid #2874a6;
}
.note {
    font-size: 0.85em;
    color: #666;
    font-style: italic;
}
</style>

<script>
function checkAnswer(selectElement, correctValue) {
    if (selectElement.value === '') {
        selectElement.classList.remove('correct', 'incorrect');
    } else if (selectElement.value === correctValue) {
        selectElement.classList.remove('incorrect');
        selectElement.classList.add('correct');
    } else {
        selectElement.classList.remove('correct');
        selectElement.classList.add('incorrect');
    }
}
</script>

```


::: {.callout-tip collapse="true"}
## **Key Questions**
- What is homoskedasticity and why is it a Gauss-Markov assumption?
- What happens to OLS estimates when heteroskedasticity is present?
- How does heteroskedasticity affect hypothesis tests?
- What are heteroskedasticity-robust standard errors and when should we use them?
- What is the cost of using robust standard errors when they are not needed?
:::

::: {.callout-note collapse="true"}
## **Suggested Readings**
- @wooldridge_introductory_2019, Ch. 8
:::


One of the Gauss-Markov assumptions is that the variance of the error term is constant across all values of the explanatory variables.
When this assumption fails,
we have **heteroskedasticity**.
Our coefficient estimates remain unbiased,
but our standard errors become unreliable---and
unreliable standard errors mean unreliable inference.


## The Homoskedasticity Assumption

### Gauss-Markov Assumption 5

Recall the Gauss-Markov assumptions that make OLS the Best Linear Unbiased Estimator (BLUE).
The fifth assumption is **homoskedasticity**:

::: {.callout-note}
## Homoskedasticity
The error term $\mu$ has the **same variance** given any value of the explanatory variables:
$$Var(\mu | x_1, x_2, ..., x_k) = \sigma^2$$
The variance is constant---it doesn't depend on $x$.
:::

This assumption is crucial because it allows us to derive the correct formula for the variance of $\hat{\beta}$,
which in turn gives us valid standard errors,
t-statistics,
and p-values.

### What Does Homoskedasticity Look Like?

In a homoskedastic regression,
the spread of observations around the regression line is **constant** across all values of $x$.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-homoskedastic
#| fig-cap: "Homoskedastic data: the spread of points around the regression line is constant across all values of x."

set.seed(123)
n <- 200

# Generate homoskedastic data
homo_data <- tibble(
    x = runif(n, 0, 100),
    error = rnorm(n, 0, 15),  # Constant variance
    y = 10 + 0.5 * x + error
)

ggplot(homo_data, aes(x = x, y = y)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    # Add bands to show constant variance
    geom_ribbon(
        data = tibble(x = seq(0, 100, length.out = 100)) |>
            mutate(yhat = 10 + 0.5 * x),
        aes(x = x, ymin = yhat - 30, ymax = yhat + 30),
        alpha = 0.15, fill = "darkred",
        inherit.aes = FALSE
    ) +
    labs(
        title = "Homoskedastic Data",
        subtitle = "The spread (variance) of errors is constant across all x values",
        x = "X",
        y = "Y"
    ) +
    annotate("text", x = 80, y = 15, label = "Constant\nspread", 
             color = "darkred", size = 4, fontface = "bold")
```

The "cone" of data points has roughly the same width whether $x$ is small or large.


## Heteroskedasticity: When Variance Changes

### What Is Heteroskedasticity?

**Heteroskedasticity** occurs when the variance of the error term changes with the level of the explanatory variable:
$$Var(\mu_i | x_i) = \sigma_i^2$$

The subscript $i$ indicates that variance depends on observation $i$'s value of $x$.

### What Does Heteroskedasticity Look Like?

The classic pattern is a "fan" or "cone" shape,
where the spread of data increases (or decreases) as $x$ changes:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-heteroskedastic
#| fig-cap: "Heteroskedastic data: the spread of points around the regression line increases with x, creating a 'fan' or 'cone' pattern."

set.seed(123)

# Generate heteroskedastic data (variance increases with x)
hetero_data <- tibble(
    x = runif(n, 0, 100),
    # Variance increases with x
    error = rnorm(n, 0, 0.3 * x),
    y = 10 + 0.5 * x + error
)

ggplot(hetero_data, aes(x = x, y = y)) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    # Add expanding bands to show increasing variance
    geom_ribbon(
        data = tibble(x = seq(0, 100, length.out = 100)) |>
            mutate(yhat = 10 + 0.5 * x, spread = 0.6 * x),
        aes(x = x, ymin = yhat - spread, ymax = yhat + spread),
        alpha = 0.15, fill = "darkred",
        inherit.aes = FALSE
    ) +
    labs(
        title = "Heteroskedastic Data",
        subtitle = "The spread (variance) of errors increases with x",
        x = "X",
        y = "Y"
    ) +
    annotate("text", x = 85, y = 20, label = "Spread\nincreases", 
             color = "darkred", size = 4, fontface = "bold")
```

This pattern is common in economic data.
For example:

- **Income and consumption:** Higher-income households have more variable consumption patterns
- **Firm size and profits:** Larger firms have more variable profit margins
- **Experience and wages:** More experienced workers have more variable wages

### Side-by-Side Comparison

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-comparison
#| fig-cap: "Comparing homoskedastic (left) and heteroskedastic (right) data. The key difference is whether the spread of points changes with x."
#| fig-width: 10
#| fig-height: 4

# Combine data for faceted plot
combined_data <- bind_rows(
    homo_data |> mutate(type = "Homoskedastic"),
    hetero_data |> mutate(type = "Heteroskedastic")
)

ggplot(combined_data, aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    facet_wrap(~type, scales = "free_y") +
    labs(
        title = "Homoskedastic vs. Heteroskedastic Data",
        x = "X",
        y = "Y"
    )
```


## Why Does Heteroskedasticity Matter?

### Coefficients Are Still Unbiased

**Heteroskedasticity does NOT bias our coefficient estimates.**
Unbiasedness of OLS depends only on Gauss-Markov assumptions 1--4,
not on homoskedasticity,
so $E[\hat{\beta}] = \beta$ still holds.

### Standard Errors Are Wrong

The standard error formulas we have been using assume homoskedasticity.
When that assumption is violated:

- The formula $SE(\hat{\beta}_1) = \frac{\hat{\sigma}}{\sqrt{SST_x}}$ is **invalid**
- Computed standard errors will typically be **too small**
- So t-statistics are **too large**
- And p-values are **too small**

In other words,
**we reject the null hypothesis too often**,
even when it is true.
This means we will be too confident in stating there is a "true" effect of $x$ on $y$ when there is in fact no statistically significant relationship.

### Type I Error: A Quick Review

::: {.callout-important}
## Type I Error
A **Type I error** (false positive) occurs when we reject a true null hypothesis.

When we set $\alpha = 0.05$,
we're accepting a 5% probability of making this error.
If our test is working correctly,
we should reject true null hypotheses exactly 5% of the time.
:::

Another way to think about "rejecting the null too often" is in terms of Type I error.
With heteroskedasticity and standard OLS standard errors,
the actual Type I error rate can be much higher than 5%---sometimes 10%, 15%, or even higher.


## Simulation: Seeing the Problem

### Setup: Simulating Hypothesis Tests

To see the problem concretely,
we simulate data where the **true effect is zero** ($\beta_1 = 0$)
and check how often we incorrectly reject the null.

First, let's look at the baseline distribution of estimates under **homoskedasticity** (where everything works correctly):

```{r}
#| echo: true
#| message: false
#| warning: false
#| cache: true

# Simulation parameters
n_sims <- 2000       # Number of simulations
n_obs <- 100         # Observations per sample
beta0_true <- 5      # True intercept
beta1_true <- 0      # True slope (NULL HYPOTHESIS IS TRUE!)

set.seed(42)

# Function for ONE simulation trial under HOMOSKEDASTICITY
run_homo_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    error <- rnorm(n_obs, mean = 0, sd = 10)  # Constant variance
    y <- beta0_true + beta1_true * x + error

    model <- lm(y ~ x)

    beta1_hat <- coef(model)["x"]
    se_ols <- summary(model)$coefficients["x", "Std. Error"]
    pval_ols <- summary(model)$coefficients["x", "Pr(>|t|)"]

    return(data.frame(beta1_hat = beta1_hat, se_ols = se_ols, pval_ols = pval_ols))
}

# Run simulations
homo_results <- map_dfr(1:n_sims, run_homo_trial)

# Calculate rejection rate and average SE
homo_reject_rate <- mean(homo_results$pval_ols < 0.05)
homo_avg_se <- mean(homo_results$se_ols)
homo_true_sd <- sd(homo_results$beta1_hat)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-homo-sampling
#| fig-cap: "Under homoskedasticity, the actual sampling distribution of the slope estimate (blue) closely matches what OLS standard errors predict (red dashed curve). The standard errors are getting it right."

ggplot(homo_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30,
                   fill = "steelblue", alpha = 0.5, color = "white") +
    stat_function(
        fun = dnorm,
        args = list(mean = 0, sd = homo_avg_se),
        color = "red",
        linewidth = 1.2,
        linetype = "dashed"
    ) +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 1) +
    labs(
        title = "Sampling Distribution Under Homoskedasticity",
        subtitle = "Blue = actual distribution of estimates | Red dashed = what OLS SEs predict",
        x = expression(hat(beta)[1]),
        y = "Density"
    ) +
    annotate("text", x = max(homo_results$beta1_hat) * 0.6, y = 0.4,
             label = paste0("True SD of estimates: ", round(homo_true_sd, 3),
                            "\nAvg OLS SE: ", round(homo_avg_se, 3)),
             color = "gray30", size = 3.5, hjust = 0)
```

Under homoskedasticity,
the actual spread of the estimates (blue histogram) closely matches
what the standard error formula predicts (red dashed curve).
Confidence intervals and hypothesis tests work as intended:
the Type I error rate is `r scales::percent(homo_reject_rate, accuracy = 0.1)`,
close to the nominal 5%,
exactly the pre-defined level we are comfortable with.

### Now With Heteroskedasticity

Now we run the same simulation
with heteroskedastic errors---the variance of the error term increases with $x$.
Before looking at sampling distributions,
here is what one draw of heteroskedastic data looks like next to homoskedastic data:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-sim-data-comparison
#| fig-cap: "One simulated sample under each scenario. In the homoskedastic case (left), the spread of points around the line is constant. In the heteroskedastic case (right), the spread fans out as x increases."
#| fig-width: 10
#| fig-height: 4

set.seed(42)

# One draw of each type
sim_homo <- tibble(
    x = runif(n_obs, 1, 50),
    y = beta0_true + beta1_true * x + rnorm(n_obs, 0, 10),
    type = "Homoskedastic"
)

sim_hetero <- tibble(
    x = runif(n_obs, 1, 50),
    y = beta0_true + beta1_true * x + rnorm(n_obs, 0, x^3 * 0.001),
    type = "Heteroskedastic"
)

bind_rows(sim_homo, sim_hetero) |>
    mutate(type = factor(type, levels = c("Homoskedastic", "Heteroskedastic"))) |>
    ggplot(aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    facet_wrap(~type, scales = "free_y") +
    labs(
        title = "Simulated Data: Homoskedastic vs. Heteroskedastic",
        subtitle = "True effect = 0 in both cases; notice the 'fan' shape on the right",
        x = "X",
        y = "Y"
    )
```

In both panels the true slope is zero,
so the regression line should be flat.
The difference is in the **spread**:
on the left it is constant,
while on the right it fans out as $x$ increases.

Repeating this 2,000 times under each scenario
gives us the sampling distribution of $\hat{\beta}_1$:

```{r}
#| echo: true
#| message: false
#| warning: false
#| cache: true

# Function for ONE simulation trial under HETEROSKEDASTICITY
run_hetero_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    # Variance INCREASES with x (heteroskedasticity!)
    error <- rnorm(n_obs, mean = 0, sd = x^3 * 0.001)
    y <- beta0_true + beta1_true * x + error

    model <- lm(y ~ x)

    beta1_hat <- coef(model)["x"]
    res_ols <- coeftest(model)
    res_robust <- coeftest(model, vcov. = vcovHC(model, type = "HC1"))

    return(data.frame(
        beta1_hat = beta1_hat,
        se_ols = res_ols["x", "Std. Error"],
        se_robust = res_robust["x", "Std. Error"],
        pval_ols = res_ols["x", "Pr(>|t|)"],
        pval_robust = res_robust["x", "Pr(>|t|)"]
    ))
}

# Run simulations
hetero_results <- map_dfr(1:n_sims, run_hetero_trial)

# Calculate rejection rates and average SE
ols_reject_rate <- mean(hetero_results$pval_ols < 0.05)
robust_reject_rate <- mean(hetero_results$pval_robust < 0.05)
hetero_avg_se <- mean(hetero_results$se_ols)
hetero_true_sd <- sd(hetero_results$beta1_hat)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-sampling-comparison
#| fig-cap: "Sampling distributions of the slope estimate under homoskedasticity (top) and heteroskedasticity (bottom). The red dashed curve shows what OLS standard errors predict. Under homoskedasticity the curve matches reality; under heteroskedasticity the actual estimates are far more spread out than OLS expects."
#| fig-width: 8
#| fig-height: 7

# --- Helper: build one panel as a standalone ggplot ---
make_panel <- function(results, avg_se, true_sd, reject_rate, panel_title) {

    # Normal curve data spanning the range of actual estimates
    xlim <- c(-1, 1) * max(abs(results$beta1_hat)) * 1.1
    x_seq <- seq(xlim[1], xlim[2], length.out = 300)
    curve_df <- tibble(x = x_seq, density = dnorm(x_seq, 0, avg_se))

    # Annotation
    label_text <- paste0(
        "True SD of estimates: ", round(true_sd, 3),
        "\nAvg OLS SE: ", round(avg_se, 3),
        "\nType I error rate: ", scales::percent(reject_rate, accuracy = 0.1)
    )

    ggplot(results, aes(x = beta1_hat)) +
        geom_histogram(aes(y = after_stat(density)), bins = 35,
                       fill = "steelblue", alpha = 0.5, color = "white") +
        geom_line(data = curve_df, aes(x = x, y = density),
                  color = "red", linewidth = 1.2, linetype = "dashed") +
        geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
        annotate("label", x = xlim[2] * 0.95,
                 y = max(curve_df$density) * 0.9,
                 label = label_text,
                 hjust = 1, vjust = 1, size = 3,
                 color = "gray20", fill = "white", label.size = 0.3) +
        labs(title = panel_title,
             x = expression(hat(beta)[1]),
             y = "Density") +
        theme(plot.title = element_text(size = 12))
}

p_homo  <- make_panel(homo_results, homo_avg_se, homo_true_sd,
                       homo_reject_rate, "Homoskedastic")
p_hetero <- make_panel(hetero_results, hetero_avg_se, hetero_true_sd,
                        ols_reject_rate, "Heteroskedastic")

p_homo / p_hetero +
    plot_annotation(
        title = expression(
            paste("Sampling Distributions of ", hat(beta)[1])
        ),
        subtitle = "Blue = actual distribution of estimates | Red dashed = what OLS standard errors predict"
    )
```

In the top panel (homoskedasticity),
the red dashed curve tracks the blue histogram closely:
the true SD of the estimates (`r round(homo_true_sd, 3)`) nearly perfectly matches
the OLS standard error (`r round(homo_avg_se, 3)`) formula,
and the Type I error rate is `r scales::percent(homo_reject_rate, accuracy = 0.1)`.

Now look at the **x-axis scale** in the bottom panel.
The estimates range much further from zero---the
true SD is `r round(hetero_true_sd, 3)`,
roughly `r round(hetero_true_sd / homo_true_sd)` times larger than under homoskedasticity.
As can be seen, the OLS standard error formula (red line) is quite far off from the actually distribution of estimates,
with the mass of estimates being much more spread out and shorter than the case with homoskedastic errors.
The red curve is far too narrow for the actual distribution,
and the resulting Type I error rate is `r scales::percent(ols_reject_rate, accuracy = 0.1)`---more
than **twice** the nominal 5%.


## Heteroskedasticity-Robust Standard Errors

### The Fix

Luckily, there is a relatively simple solution for this problem.
Instead of using the standard error formula we have worked with so far,
use a different formula for the standard errors---one that accounts for heteroskedasticity.
These are called **heteroskedasticity-robust standard errors** (also known as White or Huber-White standard errors).

The standard OLS variance formula is:
$$Var(\hat{\beta}_1) = \frac{\sum_{i=1}^{n}\hat{u}_i^2 / (n-2)}{SST_x^2}$$

The heteroskedasticity-robust formula is:
$$Var(\hat{\beta}_1) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2\hat{u}_i^2}{SST_x^2}$$

The key, intuitive difference is that robust standard errors **weight each squared residual by how far its $x_i$ is from the mean**.
This accounts for the fact that residuals may have different variances at different values of $x$.

### A Small Numerical Example

To make these formulas concrete,
consider a small example with $n = 8$ observations.
The data have a clear heteroskedastic pattern---small
residuals at low $x$ and large residuals at high $x$:

```{r}
#| echo: true
#| message: false
#| warning: false

# A small dataset with obvious heteroskedasticity
# x-values are unevenly spaced (like income in thousands)
tiny <- tibble(
    x = c(5, 10, 15, 20, 30, 50, 75, 100),
    y = c(3.7, 7.5, 6.6, 9.2, 11.7, 24.3, 31.5, 78.6)
)

# Fit OLS
tiny_model <- lm(y ~ x, data = tiny)

# Pieces we need
tiny <- tiny |>
    mutate(
        uhat    = residuals(tiny_model),       # residuals
        uhat_sq = uhat^2,                      # squared residuals
        x_dev   = x - mean(x),                 # (x_i - x-bar)
        x_dev_sq = x_dev^2                     # (x_i - x-bar)^2
    )

tiny
```

The squared residuals (`uhat_sq`) at $x = 75$ and $x = 100$ are much larger
than those at $x = 10$ or $x = 20$---more
variability at higher values of $x$.

From here we can compute $SST_x = \sum (x_i - \bar{x})^2$, the sum of squared deviations of $x$:

```{r}
#| echo: true

SST_x <- sum(tiny$x_dev_sq)
SST_x
```

**Standard OLS variance** pools all the squared residuals into a single estimate of $\sigma^2$,
then divides by $SST_x$:

```{r}
#| echo: true

# Pooled variance estimate: sum of squared residuals / (n - 2)
sigma2_hat <- sum(tiny$uhat_sq) / (nrow(tiny) - 2)

# Standard OLS variance of beta1_hat
var_ols <- sigma2_hat / SST_x

# Standard OLS SE
se_ols <- sqrt(var_ols)
se_ols
```

This formula treats every residual the same---it averages them into one $\hat{\sigma}^2$.
When the error variance is constant,
that is fine.
But here the moderate residuals at low $x$ are averaged with the large residuals at high $x$,
diluting the signal about how variable the estimate is.

**Robust variance** instead weights each squared residual by $(x_i - \bar{x})^2$,
so observations far from $\bar{x}$ get more say:

```{r}
#| echo: true

# Robust variance: sum of (x_i - x-bar)^2 * uhat_i^2, divided by SST_x^2
var_robust <- sum(tiny$x_dev_sq * tiny$uhat_sq) / SST_x^2

# Robust SE
se_robust <- sqrt(var_robust)
se_robust
```

The robust SE (`r round(se_robust, 3)`) is **larger** than the standard OLS SE (`r round(se_ols, 3)`).
The observations farthest from $\bar{x}$---at $x = 75$ and $x = 100$---receive
the biggest $(x_i - \bar{x})^2$ weights,
and those same observations have the largest squared residuals.
The robust formula therefore gives more influence to the high-variance observations,
while the standard formula averages them away.

### Visualizing the Fix

Now, let's plot the actual distribution of heteroskedastic estimates,
the distribution as implied by the normal standard error formula (in red),
and the distribution implied by the robust standard error forumula (in green):

```{r}
#| echo: false
#| message: false
#| warning: false

# Use results from earlier simulation
avg_se_robust <- mean(hetero_results$se_robust)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-sampling-dist
#| fig-cap: "The actual sampling distribution (blue) compared to what standard OLS SEs predict (red dashed) vs. robust SEs (green). OLS SEs are too narrow; robust SEs match reality."

ggplot(hetero_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30,
                   fill = "steelblue", alpha = 0.5, color = "white") +
    stat_function(
        fun = dnorm,
        args = list(mean = 0, sd = hetero_avg_se),
        color = "red",
        linewidth = 1.2,
        linetype = "dashed"
    ) +
    stat_function(
        fun = dnorm,
        args = list(mean = 0, sd = avg_se_robust),
        color = "darkgreen",
        linewidth = 1.2
    ) +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 1) +
    labs(
        title = "The Robust SE Fix",
        subtitle = "Blue = actual distribution | Red dashed = OLS SE (too narrow) | Green = Robust SE (correct)",
        x = expression(hat(beta)[1]),
        y = "Density"
    ) +
    annotate("text", x = hetero_true_sd * 1.2, y = 0.35, label = "OLS SE\n(too narrow)",
             color = "red", size = 3.5, fontface = "bold") +
    annotate("text", x = -hetero_true_sd * 1.2, y = 0.25, label = "Robust SE\n(correct)",
             color = "darkgreen", size = 3.5, fontface = "bold")
```

The red dashed curve (standard OLS SEs) is too narrow.
The green curve (robust SEs) matches the actual spread of the estimates.

### Comparing Type I Error Rates

As a final check, let's confirm that the robust SEs give us an acceptable Type 1 error rate (about 5%).

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-type1-comparison
#| fig-cap: "Type I error rates across different scenarios. Robust standard errors maintain the correct 5% rate even with heteroskedasticity."

type1_data <- tibble(
    Scenario = c("Homoskedastic\n(OLS SE)",
                 "Heteroskedastic\n(OLS SE)",
                 "Heteroskedastic\n(Robust SE)"),
    Rate = c(homo_reject_rate, ols_reject_rate, robust_reject_rate),
    Status = c("Correct", "Inflated!", "Correct")
)

type1_data$Scenario <- factor(type1_data$Scenario, levels = type1_data$Scenario)

ggplot(type1_data, aes(x = Scenario, y = Rate, fill = Status)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
    geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
              vjust = -0.5, size = 5, fontface = "bold") +
    scale_fill_manual(values = c("Correct" = "steelblue", "Inflated!" = "coral")) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.15)) +
    labs(
        title = "Type I Error Rates: The Heteroskedasticity Problem",
        subtitle = "Dashed line shows the target 5% rate",
        x = "",
        y = "Type I Error Rate (False Positive Rate)"
    ) +
    theme(legend.position = "none") +
    annotate("text", x = 2, y = 0.13, label = "Over-rejection!",
             color = "coral", fontface = "bold", size = 4)
```

The first bar is the baseline:
under homoskedasticity with standard OLS standard errors,
the rejection rate is `r scales::percent(homo_reject_rate, accuracy = 0.1)`.

The middle bar shows the problem.
Under heteroskedasticity with the same OLS standard errors,
the rejection rate jumps to `r scales::percent(ols_reject_rate, accuracy = 0.1)`.
Roughly 1 in `r round(1/ols_reject_rate)` tests will turn up a
significant result **that does not exist**.

The third bar shows the fix.
Switching to robust standard errors brings the
rejection rate back to `r scales::percent(robust_reject_rate, accuracy = 0.1)`.


## Implementing Robust Standard Errors in R

### Using `fixest`

The `fixest` package makes this straightforward:

```{r}
#| echo: true
#| message: false
#| warning: false

# Generate some heteroskedastic data
set.seed(123)
example_data <- tibble(
    x = runif(200, 1, 50),
    y = 5 + 0.5 * x + rnorm(200, 0, sqrt(0.1 * x^3))
)

# Standard OLS (wrong SEs with heteroskedasticity)
model_ols <- feols(y ~ x, data = example_data)

# With heteroskedasticity-robust SEs
model_robust <- feols(y ~ x, vcov = "HC1", data = example_data)
```

```{r}
#| echo: true
#| message: false
#| warning: false

# Compare the results
modelsummary(
    list("Standard SE" = model_ols, "Robust SE" = model_robust),
    stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
    gof_map = c("nobs", "r.squared")
)
```

The coefficients are identical---unbiasedness is unaffected.
The standard errors differ (robust SEs are larger here),
which changes the t-statistics and p-values.

Importantly,
when using normal OLS standard errors,
the estimate *was* statistically significant at the 10% level,
since we had a p-value less than 0.1.
However,
with robust SEs, this is no longer the case: we can no longer reject the null hypothesis at any conventional threshold.
*This is an important analytical difference!*
We can see clearly that we may falsely conclude there is a "true" effect of our x variable when there is in fact not if we do not use the appropriate standard errors.


## What If There's No Heteroskedasticity?

If robust standard errors fix the problem when heteroskedasticity is present,
what happens when we use them on **homoskedastic** data
where the standard OLS formula is already correct?

Let's take a look at the distribution of the different standard errors for the same homoskedastic data.

```{r}
#| echo: true
#| message: false
#| warning: false
#| cache: true

set.seed(321)

# Run simulation under HOMOSKEDASTICITY using robust SEs
run_homo_robust_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    error <- rnorm(n_obs, mean = 0, sd = 10)  # Constant variance
    y <- beta0_true + beta1_true * x + error

    model <- lm(y ~ x)

    res_ols <- coeftest(model)
    res_robust <- coeftest(model, vcov. = vcovHC(model, type = "HC1"))

    return(data.frame(
        beta1_hat = coef(model)["x"],
        se_ols = res_ols["x", "Std. Error"],
        se_robust = res_robust["x", "Std. Error"],
        pval_ols = res_ols["x", "Pr(>|t|)"],
        pval_robust = res_robust["x", "Pr(>|t|)"]
    ))
}

homo_robust_results <- map_dfr(1:n_sims, run_homo_robust_trial)

# Rejection rates
homo_ols_reject <- mean(homo_robust_results$pval_ols < 0.05)
homo_robust_reject <- mean(homo_robust_results$pval_robust < 0.05)

# Average SEs
homo_ols_avg_se <- mean(homo_robust_results$se_ols)
homo_robust_avg_se <- mean(homo_robust_results$se_robust)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-homo-robust-comparison
#| fig-cap: "Under homoskedasticity, robust standard errors are slightly larger than standard OLS SEs, but the Type I error rate remains close to 5%."
#| fig-width: 10
#| fig-height: 4

homo_se_comparison <- homo_robust_results |>
    pivot_longer(cols = c(se_ols, se_robust),
                 names_to = "method", values_to = "se") |>
    mutate(method = case_when(
        method == "se_ols" ~ paste0("Standard OLS SE\n(Avg: ",
                                     round(homo_ols_avg_se, 4), ")"),
        method == "se_robust" ~ paste0("Robust SE\n(Avg: ",
                                        round(homo_robust_avg_se, 4), ")")
    ))

ggplot(homo_se_comparison, aes(x = se, fill = method)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30,
                   alpha = 0.6, color = "white", position = "identity") +
    scale_fill_manual(values = c("steelblue", "darkgreen")) +
    labs(
        title = "Standard Errors Under Homoskedasticity: OLS vs. Robust",
        subtitle = "Robust SEs are slightly larger on average, but not by much",
        x = "Standard Error",
        y = "Density",
        fill = ""
    ) +
    theme(legend.position = "bottom")
```

Each simulation produces both a standard OLS standard error and a robust standard error for the same regression.
The figure plots the distribution of each across 2,000 simulations.
If robust SEs impose a large cost under homoskedasticity,
the green distribution would sit noticeably to the right of the blue one,
meaning wider confidence intervals and lower power.
In practice the two distributions nearly overlap.

The two SE estimates track each other closely.
The average robust SE (`r round(homo_robust_avg_se, 4)`) is slightly
larger than the average OLS SE (`r round(homo_ols_avg_se, 4)`)---a
difference of about `r scales::percent((homo_robust_avg_se - homo_ols_avg_se) / homo_ols_avg_se, accuracy = 0.1)`.
The Type I error rate is
`r scales::percent(homo_robust_reject, accuracy = 0.1)` with robust SEs
versus `r scales::percent(homo_ols_reject, accuracy = 0.1)` with standard SEs---both
on target.

Why is the cost so small?
Recall the two variance formulas.
The standard formula uses $\hat{\sigma}^2 / SST_x$,
which pools all squared residuals into a single variance estimate.
The robust formula uses $\sum (x_i - \bar{x})^2 \hat{u}_i^2 / SST_x^2$,
which weights each squared residual by $(x_i - \bar{x})^2$.
When homoskedasticity holds,
the squared residuals $\hat{u}_i^2$ are roughly the same size everywhere,
so the $(x_i - \bar{x})^2$ weights do not change much---the
weighted average ends up close to the unweighted average.
The two formulas converge toward the same answer,
and the robust SE is only slightly noisier because it estimates observation-level variances
rather than a single pooled $\sigma^2$.

The cost of robust standard errors when homoskedasticity holds is therefore a small loss of efficiency
(slightly wider confidence intervals).
The cost of *not* using them when heteroskedasticity is present is
a Type I error rate of `r scales::percent(ols_reject_rate, accuracy = 0.1)`
instead of 5%.
The tradeoff is one-sided.
For this reason, it is generally considered good practice to *always* report heteroskedasticity robust SEs when doing regression analysis.
From this point forward,
this is a habit you should get used to.


## Summary

Heteroskedasticity is a common issue in economic data where the variance of errors changes with the explanatory variables.

**Key Points:**

1. **Heteroskedasticity does NOT bias coefficients** - $E[\hat{\beta}] = \beta$ still holds
2. **Heteroskedasticity DOES invalidate standard errors** - leading to incorrect inference
3. **The main symptom is inflated Type I error rates** - we reject true nulls too often
4. **The solution is simple: use robust standard errors** - specify `vcov = "HC1"` in `feols()`

**Best Practice:**
Always use heteroskedasticity-robust standard errors.
They're valid whether or not heteroskedasticity is present,
so there's no downside to using them.


## Check Your Understanding

:::: {.content-visible when-format="html"}

For each question below, 
select the best answer from the dropdown menu. 

```{=html}
<div class="question">
<div class="answer-row">
    <label>i. What is heteroskedasticity? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">When the error term is correlated with x</option>
        <option value="b">When the variance of the error term changes with x</option>
        <option value="c">When the mean of the error term is not zero</option>
        <option value="d">When observations are not independent</option>
    </select>
</div>

<div class="answer-row">
    <label>ii. Does heteroskedasticity bias OLS coefficient estimates? </label>
    <select onchange="checkAnswer(this, 'a')">
        <option value="">-- Select Answer --</option>
        <option value="a">No, coefficients remain unbiased</option>
        <option value="b">Yes, coefficients become biased upward</option>
        <option value="c">Yes, coefficients become biased downward</option>
        <option value="d">It depends on the form of heteroskedasticity</option>
    </select>
</div>

<div class="answer-row">
    <label>iii. What goes wrong when we have heteroskedasticity but use standard OLS standard errors? </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">Coefficients become inconsistent</option>
        <option value="b">R² becomes meaningless</option>
        <option value="c">Standard errors are typically too small, leading to too many false positives</option>
        <option value="d">The regression line no longer minimizes squared residuals</option>
    </select>
</div>

<div class="answer-row">
    <label>iv. In a residual plot, what pattern suggests heteroskedasticity? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">Residuals forming a horizontal line at zero</option>
        <option value="b">Residuals forming a fan or cone shape</option>
        <option value="c">Residuals forming a U-shape</option>
        <option value="d">Residuals clustering in the center</option>
    </select>
</div>

<div class="answer-row">
    <label>v. Why is it considered best practice to always use robust standard errors? </label>
    <select onchange="checkAnswer(this, 'd')">
        <option value="">-- Select Answer --</option>
        <option value="a">They make coefficients larger</option>
        <option value="b">They increase R²</option>
        <option value="c">They are required by academic journals</option>
        <option value="d">They are valid whether or not heteroskedasticity is present</option>
    </select>
</div>

<div class="answer-row">
    <label>vi. <strong>(Challenge)</strong> If standard OLS standard errors are 0.05 and robust standard errors are 0.08 for the same coefficient, what does this suggest? </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">The coefficient is biased</option>
        <option value="b">The model has omitted variable bias</option>
        <option value="c">Heteroskedasticity is likely present, and OLS SEs are underestimating uncertainty</option>
        <option value="d">The sample size is too small</option>
    </select>
</div>
</div>

```

::: {.callout-tip collapse="true"}
## Show Explanation

i. Heteroskedasticity means the variance of the error term is not constant—it changes depending on the value of x. For example, variance might increase with x, creating a "fan" pattern in the data.

ii. No, heteroskedasticity does NOT cause bias in coefficient estimates. Unbiasedness depends only on Gauss-Markov assumptions 1-4 (linearity, random sampling, no perfect collinearity, zero conditional mean). The homoskedasticity assumption (GM5) affects efficiency and inference, not unbiasedness.

iii. When heteroskedasticity is present but ignored, the standard error formula is wrong. Typically, standard errors are underestimated, making t-statistics too large and p-values too small. This leads to rejecting true null hypotheses too often (inflated Type I error rate).

iv. The classic sign of heteroskedasticity is a "fan" or "cone" shape in the residual plot, where the spread of residuals increases (or decreases) as the fitted values change. Constant spread would indicate homoskedasticity.

v. Robust standard errors are valid whether or not heteroskedasticity exists. Without heteroskedasticity, they are slightly less efficient but still correct. With heteroskedasticity, they are correct while standard SEs are wrong.

vi. Robust SEs larger than standard OLS SEs suggest heteroskedasticity is present. The standard formula is underestimating the true variability in β̂ because it pools residual variance across all observations rather than weighting by leverage.

:::

::::

:::: {.content-hidden when-format="html"}

*Note: This section contains interactive content only available in the HTML version.*

::::
