# Instrumental Variables {#sec-iv}

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(tidyverse)
library(fixest)
library(wooldridge)
library(modelsummary)

options(scipen = 999)

theme_set(theme_minimal(base_size = 12) +
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(size = 11, color = "gray40"),
                panel.grid.minor = element_blank()))

```

```{=html}

<style>
.problem {
    background-color: #f8f9fa;
    border: 1px solid #dee2e6;
    border-radius: 8px;
    padding: 20px;
    margin: 20px 0;
}
.context {
    background-color: #e8f4f8;
    padding: 12px;
    border-radius: 5px;
    margin-bottom: 15px;
    border-left: 4px solid #2874a6;
}
.question {
    margin: 15px 0;
    padding: 10px;
    background-color: #fff3cd;
    border-radius: 5px;
}
.answer-row {
    display: flex;
    align-items: center;
    margin: 10px 0;
    flex-wrap: wrap;
    gap: 10px;
}
.answer-row label {
    min-width: 320px;
    font-weight: 500;
}
select {
    padding: 8px 12px;
    font-size: 14px;
    border: 2px solid #ccc;
    border-radius: 5px;
    min-width: 300px;
    cursor: pointer;
}
select:focus {
    border-color: #2874a6;
    outline: none;
}
.correct {
    border-color: #28a745 !important;
    background-color: #d4edda !important;
}
.incorrect {
    border-color: #dc3545 !important;
    background-color: #f8d7da !important;
}
.explanation {
    margin-top: 10px;
    padding: 15px;
    background-color: #e7f3ff;
    border-radius: 5px;
    border-left: 4px solid #2874a6;
}
.note {
    font-size: 0.85em;
    color: #666;
    font-style: italic;
}
</style>

<script>
function checkAnswer(selectElement, correctValue) {
    if (selectElement.value === '') {
        selectElement.classList.remove('correct', 'incorrect');
    } else if (selectElement.value === correctValue) {
        selectElement.classList.remove('incorrect');
        selectElement.classList.add('correct');
    } else {
        selectElement.classList.remove('correct');
        selectElement.classList.add('incorrect');
    }
}
</script>

```


::: {.callout-tip collapse="true"}
## **Key Questions**
- What is endogeneity and why does it cause OLS to fail?
- What are instrumental variables and what makes an instrument valid?
- How does Two-Stage Least Squares (2SLS) work?
- What is the Local Average Treatment Effect (LATE)?
- How do we test whether an instrument is strong enough?
:::

::: {.callout-note collapse="true"}
## **Suggested Readings**
- TBD
:::


We've built up a powerful toolkit for causal inference:
randomized experiments,
multivariate regression,
fixed effects,
and difference-in-differences.
But what happens when none of these methods are available?

Sometimes we have only cross-sectional data,
treatment is clearly endogenous,
we cannot randomize,
and there's no natural experiment to exploit.
In these situations,
**instrumental variables (IV)** offers a path forward---if we can find the right instrument.


## The Fundamental Problem: Endogeneity

### When OLS Fails

Recall that OLS provides unbiased estimates when the zero conditional mean assumption holds:
$E[\mu | X] = 0$.
When this assumption fails---when $X$ is correlated with the error term---we say $X$ is **endogenous**,
and OLS is biased.

Endogeneity arises from three main sources:

**1. Omitted Variable Bias**

We want to estimate the effect of education on wages,
but ability affects both education choices and wages.
Since we can't measure ability,
it's absorbed into the error term,
biasing our estimate.

**2. Selection Bias**

We want to estimate the effect of health insurance on health outcomes,
but healthier people are more likely to purchase insurance.
Health status drives insurance decisions,
contaminating our comparison.

**3. Reverse Causality**

We want to estimate the effect of police on crime,
but high-crime areas get more police deployed.
The causation runs both ways,
making it impossible to isolate the effect of policing.

### Formalizing the Problem

Suppose we want to estimate:

$$
y = \beta_0 + \beta_1 x + \mu
$$

When $E[\mu | x] \neq 0$,
the variable $x$ is **endogenous**.
OLS will be biased and inconsistent---no matter how large our sample,
we won't converge to the true $\beta_1$.

We cannot interpret the OLS coefficient as a causal effect.


## The Instrumental Variables Solution

### The Key Insight

The IV approach finds a variable $Z$ that "shakes up" $X$ in a way that's unrelated to the error term.
Think of it as finding a source of **exogenous variation** in $X$---variation that comes from outside the system and isn't contaminated by the factors causing endogeneity.

### Two Requirements for a Valid Instrument

For $Z$ to be a valid instrument,
it must satisfy two conditions:

::: {.callout-important}
## The Two IV Assumptions

**1. Relevance:** $Cov(Z, X) \neq 0$

The instrument must be correlated with the endogenous variable.
This is **testable**---we can check whether $Z$ predicts $X$.

**2. Exclusion Restriction:** $Cov(Z, \mu) = 0$

The instrument affects $Y$ **only through** $X$.
There is no direct effect of $Z$ on $Y$.
This is **not directly testable**---it requires economic reasoning and assumption.
:::

### Graphical Intuition

The IV strategy can be visualized as a causal diagram:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-iv-dag
#| fig-cap: "The instrumental variables strategy. Z affects Y only through X (exclusion restriction), while unobservables (μ) affect both X and Y, creating endogeneity."

# Create simple DAG visualization
dag_data <- tibble(
    x = c(1, 2, 3, 2.5),
    y = c(0, 0, 0, 1),
    label = c("Instrument\n(Z)", "Treatment\n(X)", "Outcome\n(Y)", "Unobservables\n(μ)")
)

arrows <- tibble(
    x_start = c(1, 2, 2.5, 2.5),
    y_start = c(0, 0, 1, 1),
    x_end = c(2, 3, 2, 3),
    y_end = c(0, 0, 0, 0),
    type = c("good", "good", "bad", "bad")
)

ggplot() +
    # Draw arrows
    geom_segment(data = arrows, 
                 aes(x = x_start, y = y_start, xend = x_end, yend = y_end, color = type),
                 arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
                 linewidth = 1.2) +
    # Draw nodes
    geom_point(data = dag_data, aes(x = x, y = y), size = 20, color = "white") +
    geom_point(data = dag_data, aes(x = x, y = y), size = 18, 
               color = c("#0072B2", "#E69F00", "#009E73", "#CC79A7")) +
    # Add labels
    geom_text(data = dag_data, aes(x = x, y = y, label = label), 
              size = 3.5, fontface = "bold", color = "white") +
    # Add annotation for exclusion
    annotate("text", x = 1.5, y = -0.3, 
             label = "Relevance:\nZ predicts X", size = 3, color = "#0072B2") +
    annotate("text", x = 2.5, y = -0.3, 
             label = "Causal effect\nwe want", size = 3, color = "#E69F00") +
    annotate("text", x = 2.75, y = 0.6, 
             label = "Endogeneity\nproblem", size = 3, color = "#CC79A7") +
    scale_color_manual(values = c("good" = "#0072B2", "bad" = "#CC79A7")) +
    coord_cartesian(xlim = c(0.5, 3.5), ylim = c(-0.5, 1.3)) +
    theme_void() +
    theme(legend.position = "none") +
    labs(title = "The IV Strategy: Using Exogenous Variation")
```

The key insight:
$X$ has two types of variation---"good" variation unrelated to $\mu$,
and "bad" variation correlated with $\mu$.
OLS uses both and is therefore biased.
IV uses only the "good" variation induced by $Z$,
giving us an unbiased estimate.


## Application: Fertility and Labor Supply

### The Research Question

What is the causal effect of having children on mothers' labor force participation?
This question is crucial for understanding gender wage gaps,
designing parental leave policies,
and evaluating childcare subsidies.

But we cannot randomize fertility!
And the decision to have children is deeply endogenous:

- **Reverse causality:** Women may time children based on career factors
- **Selection:** Less career-oriented women may choose larger families
- **Omitted variables:** Unobserved preferences affect both fertility and work

### The Angrist and Evans (1998) Instrument

In a landmark study,
Angrist and Evans (1998) found a clever instrument for fertility:
the **sex composition of the first two children**.

The instrument works because parents prefer mixed-sex children.
If your first two children are the same sex (both boys or both girls),
you're more likely to try for a third child to "get" the other sex.

**Why is this relevant?** ($Cov(Z, X) \neq 0$)

- Parents with same-sex children are more likely to have a third child
- This is an empirical fact we can verify

**Why is this excludable?** ($Cov(Z, \mu) = 0$)

- Sex of children is essentially random (about 50% boys, 50% girls)
- Hard to imagine why same-sex vs. mixed-sex would *directly* affect labor supply
- The only plausible channel is through fertility decisions

### Simulating the Angrist-Evans Setup

Let's create simulated data that mirrors the Angrist-Evans setting to illustrate how IV works:

```{r}
#| echo: false
#| message: false
#| warning: false

# Simulate Angrist-Evans style data
set.seed(42)
n <- 50000

# Generate data
ae_sim <- tibble(
    # Unobserved "career orientation" - affects both fertility and work
    career_orientation = rnorm(n, 0, 1),
    
    # Same-sex instrument (random)
    samesex = rbinom(n, 1, 0.5),
    
    # More than 2 kids: affected by samesex AND career orientation
    morekids_latent = -0.5 + 0.07 * samesex - 0.3 * career_orientation + rnorm(n, 0, 0.5),
    morekids = as.integer(morekids_latent > 0),
    
    # Labor force participation: affected by morekids AND career orientation
    # True causal effect of morekids is -0.10
    worked_latent = 0.6 - 0.10 * morekids + 0.25 * career_orientation + rnorm(n, 0, 0.3),
    mom_worked = as.integer(worked_latent > 0.5)
) |>
    mutate(
        samesex_lab = if_else(samesex == 1, "Same sex", "Mixed sex"),
        morekids_lab = if_else(morekids == 1, "> 2 kids", "≤ 2 kids")
    )
```

### Testing the Relevance Condition

First,
let's verify that same-sex siblings predict having more than two children:

```{r}
#| echo: true
#| message: false
#| warning: false

# Check relevance: does samesex predict morekids?
relevance_check <- ae_sim |>
    group_by(samesex_lab) |>
    summarize(prop_morekids = mean(morekids))

relevance_check
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-relevance
#| fig-cap: "The relevance condition: parents with same-sex children are more likely to have a third child."

ggplot(relevance_check, aes(x = samesex_lab, y = prop_morekids)) +
    geom_col(fill = "steelblue", alpha = 0.8, width = 0.6) +
    geom_text(aes(label = scales::percent(prop_morekids, accuracy = 0.1)),
              vjust = -0.5, size = 5, fontface = "bold") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.5)) +
    labs(
        title = "Relevance Check: Same-Sex Predicts Fertility",
        subtitle = "Parents with same-sex children are more likely to have 3+ kids",
        x = "Sex Composition of First Two Children",
        y = "Proportion with More Than 2 Kids"
    )
```

The difference is substantial---same-sex parents are about 7 percentage points more likely to have a third child.
The instrument is relevant.

### Testing Excludability (Sort of)

We cannot directly test the exclusion restriction,
but we can check whether the instrument has a small *reduced-form* relationship with the outcome:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-exclusion
#| fig-cap: "Checking excludability: the direct relationship between same-sex and labor supply is small, consistent with the effect operating through fertility."

exclusion_check <- ae_sim |>
    group_by(samesex_lab) |>
    summarize(prop_worked = mean(mom_worked))

ggplot(exclusion_check, aes(x = samesex_lab, y = prop_worked)) +
    geom_col(fill = "coral", alpha = 0.8, width = 0.6) +
    geom_text(aes(label = scales::percent(prop_worked, accuracy = 0.1)),
              vjust = -0.5, size = 5, fontface = "bold") +
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.7)) +
    labs(
        title = "Exclusion Check: Direct Effect Should Be Small",
        subtitle = "Any effect of same-sex on labor supply should operate through fertility",
        x = "Sex Composition of First Two Children",
        y = "Proportion of Mothers Working"
    )
```

The difference is small (about 1-2 percentage points),
consistent with the effect operating through fertility rather than directly.
Of course,
this doesn't *prove* excludability---we must rely on economic reasoning to argue there's no direct channel.


## Two-Stage Least Squares (2SLS)

### The Estimation Strategy

IV models are typically estimated using **Two-Stage Least Squares (2SLS)**:

**Stage 1 (First Stage):** Regress the endogenous variable on the instrument:
$$
X_i = \pi_0 + \pi_1 Z_i + \nu_i
$$

Get predicted values: $\hat{X}_i = \hat{\pi}_0 + \hat{\pi}_1 Z_i$

These predicted values contain only the variation in $X$ that comes from $Z$---the exogenous variation.

**Stage 2 (Second Stage):** Regress the outcome on the predicted values:
$$
Y_i = \beta_0 + \beta_1 \hat{X}_i + \mu_i
$$

The coefficient $\hat{\beta}_1^{2SLS}$ is our causal estimate.

### Why Does This Work?

The first stage "purges" $X$ of its problematic variation.
Since $Z$ is exogenous (by assumption),
the predicted values $\hat{X}$ contain only exogenous variation.
When we regress $Y$ on $\hat{X}$,
we're using only this "clean" variation,
giving us an unbiased estimate.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2sls-intuition
#| fig-cap: "2SLS uses only the variation in X that comes from the instrument Z, discarding the endogenous variation."

# Visualize the 2SLS intuition
variation_data <- tibble(
    type = c("Total variation\nin X", "Exogenous variation\n(from Z)", "Endogenous variation\n(correlated with μ)"),
    value = c(100, 40, 60),
    fill = c("total", "good", "bad")
)

variation_data$type <- factor(variation_data$type, 
                               levels = c("Total variation\nin X", 
                                         "Exogenous variation\n(from Z)", 
                                         "Endogenous variation\n(correlated with μ)"))

ggplot(variation_data, aes(x = type, y = value, fill = fill)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = paste0(value, "%")), vjust = -0.5, size = 5, fontface = "bold") +
    scale_fill_manual(values = c("total" = "gray60", "good" = "#0072B2", "bad" = "#CC79A7")) +
    scale_y_continuous(limits = c(0, 120)) +
    labs(
        title = "2SLS Uses Only the 'Good' Variation",
        subtitle = "OLS uses all variation; IV uses only variation induced by the instrument",
        x = "",
        y = "Share of Variation (%)"
    ) +
    theme(legend.position = "none") +
    annotate("text", x = 2, y = 55, label = "IV uses\nthis", color = "#0072B2", fontface = "bold") +
    annotate("text", x = 3, y = 75, label = "IV discards\nthis", color = "#CC79A7", fontface = "bold")
```

### Implementing 2SLS

Let's estimate the model step by step:

**Step 1: The First Stage**

```{r}
#| echo: true
#| message: false
#| warning: false

# First stage: regress endogenous variable on instrument
first_stage <- lm(morekids ~ samesex, data = ae_sim)
summary(first_stage)
```

The F-statistic on `samesex` is very large (over 100),
indicating a strong instrument.

**Step 2: Get Predicted Values**

```{r}
#| echo: true
#| message: false
#| warning: false

# Get predicted values from first stage
ae_sim <- ae_sim |>
    mutate(morekids_hat = predict(first_stage))

# Look at the predictions
ae_sim |>
    select(samesex, morekids, morekids_hat) |>
    head(10)
```

**Step 3: The Second Stage**

```{r}
#| echo: true
#| message: false
#| warning: false

# Second stage: regress outcome on predicted values
# NOTE: Standard errors from this approach are WRONG
second_stage_manual <- lm(mom_worked ~ morekids_hat, data = ae_sim)
coef(second_stage_manual)["morekids_hat"]
```

::: {.callout-warning}
## Standard Error Warning
The manual two-stage approach gives the correct coefficient,
but the standard errors are **wrong** because they don't account for uncertainty in the first stage.
Always use a proper 2SLS estimator that computes correct standard errors.
:::

### Proper 2SLS with `feols()`

The `fixest` package provides proper 2SLS estimation with correct standard errors:

```{r}
#| echo: true
#| message: false
#| warning: false

# Proper 2SLS estimation
# Syntax: outcome ~ controls | fixed effects | endogenous ~ instrument
tsls <- feols(mom_worked ~ 1 |  # Just intercept (no controls)
                           0 |  # No fixed effects
                           morekids ~ samesex,  # First stage
              data = ae_sim)

summary(tsls)
```

### Comparing OLS and 2SLS

```{r}
#| echo: true
#| message: false
#| warning: false

# OLS for comparison (biased)
ols <- lm(mom_worked ~ morekids, data = ae_sim)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ols-vs-iv
#| fig-cap: "Comparing OLS (biased) and 2SLS (unbiased) estimates. OLS overstates the negative effect because career orientation confounds the relationship."

# Extract coefficients
coef_comparison <- tibble(
    Model = c("OLS (Biased)", "2SLS (IV)"),
    Estimate = c(coef(ols)["morekids"], coef(tsls)["fit_morekids"]),
    SE = c(summary(ols)$coefficients["morekids", "Std. Error"],
           summary(tsls)$se["fit_morekids"])
) |>
    mutate(
        CI_lower = Estimate - 1.96 * SE,
        CI_upper = Estimate + 1.96 * SE
    )

ggplot(coef_comparison, aes(x = Model, y = Estimate)) +
    geom_hline(yintercept = -0.10, linetype = "dashed", color = "darkgreen", linewidth = 1) +
    geom_point(size = 4, color = "steelblue") +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                  width = 0.15, color = "steelblue", linewidth = 1) +
    annotate("text", x = 2.3, y = -0.10, label = "True effect = -0.10", 
             color = "darkgreen", hjust = 0, size = 4) +
    coord_flip() +
    labs(
        title = "OLS vs. 2SLS Estimates",
        subtitle = "OLS is biased; 2SLS recovers the true causal effect",
        x = "",
        y = "Effect on Labor Force Participation"
    )
```

The OLS estimate is more negative than the true effect (-0.10) because it's contaminated by selection:
women with lower career orientation both have more children *and* are less likely to work,
making the correlation more negative than the causal effect.

The 2SLS estimate is much closer to the true effect because it uses only the exogenous variation from the same-sex instrument.


## Testing for Weak Instruments

### The Weak Instrument Problem

If the instrument is only weakly correlated with $X$,
2SLS performs poorly:

- Estimates become biased (toward OLS!)
- Standard errors become unreliable
- Inference is invalid

### The F-Test Rule of Thumb

**Stock and Yogo (2005)** established a rule of thumb:

::: {.callout-note}
## Weak Instrument Test
The first-stage F-statistic should exceed **10** for the instrument to be considered strong.
With F < 10, 2SLS estimates may be severely biased.
:::

In our example,
the first-stage F-statistic is well over 100---a very strong instrument.

### What Happens with a Weak Instrument?

Let's see what happens when we use a weak (essentially random) instrument:

```{r}
#| echo: true
#| message: false
#| warning: false

# Create a weak instrument (random noise)
set.seed(123)
ae_sim <- ae_sim |>
    mutate(weak_instrument = rnorm(n()))

# Try 2SLS with weak instrument
weak_iv <- feols(mom_worked ~ 1 | 0 | morekids ~ weak_instrument,
                 data = ae_sim)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-weak-iv
#| fig-cap: "With a weak instrument (random noise), the 2SLS estimate becomes unreliable with huge standard errors."

# Compare strong and weak IV
weak_comparison <- tibble(
    Model = c("OLS", "2SLS (Strong IV)", "2SLS (Weak IV)"),
    Estimate = c(coef(ols)["morekids"], 
                 coef(tsls)["fit_morekids"],
                 coef(weak_iv)["fit_morekids"]),
    SE = c(summary(ols)$coefficients["morekids", "Std. Error"],
           summary(tsls)$se["fit_morekids"],
           summary(weak_iv)$se["fit_morekids"])
) |>
    mutate(
        CI_lower = Estimate - 1.96 * SE,
        CI_upper = Estimate + 1.96 * SE,
        # Cap for visualization
        CI_lower_plot = pmax(CI_lower, -1.5),
        CI_upper_plot = pmin(CI_upper, 1.5)
    )

ggplot(weak_comparison, aes(x = Model, y = Estimate)) +
    geom_hline(yintercept = 0, linetype = "solid", color = "gray60") +
    geom_hline(yintercept = -0.10, linetype = "dashed", color = "darkgreen", linewidth = 1) +
    geom_point(size = 4, color = c("coral", "steelblue", "purple")) +
    geom_errorbar(aes(ymin = CI_lower_plot, ymax = CI_upper_plot), 
                  width = 0.15, 
                  color = c("coral", "steelblue", "purple"),
                  linewidth = 1) +
    coord_flip(ylim = c(-1.5, 1.5)) +
    labs(
        title = "Strong vs. Weak Instruments",
        subtitle = "Weak instruments produce unreliable estimates with huge confidence intervals",
        x = "",
        y = "Effect on Labor Force Participation"
    ) +
    annotate("text", x = 0.7, y = -0.10, label = "True effect", 
             color = "darkgreen", size = 3.5)
```

With a weak instrument,
the estimate is wildly off and the confidence interval is enormous.
**Always check your first-stage F-statistic!**


## What Does IV Estimate? The LATE

### The Local Average Treatment Effect

An important subtlety:
IV does not estimate the same quantity as OLS or DiD.
It estimates the **Local Average Treatment Effect (LATE)**---the causal effect for a specific subpopulation called **compliers**.

### Four Types of People

In the fertility example,
we can categorize people by how their fertility responds to the instrument:

| Type | Description |
|:-----|:------------|
| **Always-takers** | Would have 3+ kids regardless of sex composition |
| **Never-takers** | Would never have 3+ kids regardless of sex composition |
| **Compliers** | Have 3+ kids *because* first two were same-sex |
| **Defiers** | Have 3+ kids *only if* first two were different-sex |

**IV identifies the causal effect only for compliers**---those whose treatment status was actually changed by the instrument.

### Who Are the Compliers?

In the Angrist-Evans context,
compliers are parents who:

- Wanted mixed-sex children strongly enough to try for a third
- Would have stopped at two kids if the first two were different sexes

This is **not**:

- All parents with 3+ children
- Parents who wanted large families regardless
- Parents who strongly preferred small families

### Trade-offs of IV Estimation

| Advantages | Disadvantages |
|:-----------|:--------------|
| Credible causal identification | Estimates LATE, not ATE |
| Handles endogeneity | Less generalizable |
| No need for RCT or natural experiment | Requires strong assumptions |
| Works with cross-sectional data | Often larger standard errors |


## Classic IV Applications

Instrumental variables have been used to study many important questions:

| Study | Question | Instrument |
|:------|:---------|:-----------|
| **Card (1995)** | Returns to education | Distance to college |
| **Acemoglu et al. (2001)** | Do institutions cause growth? | Colonial settler mortality |
| **Levitt (1997)** | Effect of police on crime | Electoral timing |
| **Angrist & Krueger (1991)** | Returns to education | Quarter of birth |

### Example: Card (1995) - Returns to Education

**Question:** What is the causal effect of education on wages?

**Problem:** Ability is unobserved but affects both education and wages,
biasing OLS upward.

**Instrument:** Distance to the nearest college.
People who grew up closer to a college got more education.

**Why relevant?** Proximity reduces the cost of attending college.

**Why excludable?** Distance to a college shouldn't directly affect wages decades later
(only through education).

**Finding:** Returns to education are actually *higher* than OLS suggests---the ability bias works in the opposite direction from what many expected.


## Summary

Instrumental variables provide a powerful tool for causal inference when other methods fail.
The key is finding a variable that affects treatment but has no direct effect on the outcome.

**The two requirements:**

1. **Relevance:** The instrument predicts the endogenous variable (testable via first-stage F-statistic > 10)
2. **Exclusion:** The instrument affects the outcome only through the treatment (requires economic reasoning---not directly testable)

**Key insights:**

- 2SLS uses only the exogenous variation in $X$ induced by $Z$
- IV estimates the Local Average Treatment Effect (LATE) for compliers
- Weak instruments produce unreliable estimates---always check the first-stage F-statistic
- The exclusion restriction is crucial but untestable---think carefully about whether it holds

IV requires strong assumptions,
but when those assumptions are credible,
it provides a path to causal inference even in challenging observational settings.


## Check Your Understanding

:::: {.content-visible when-format="html"}

For each question below, 
select the best answer from the dropdown menu. 

```{=html}
<div class="question">
<div class="answer-row">
    <label>i. What does it mean for a variable to be "endogenous"? </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">The variable is measured with error</option>
        <option value="b">The variable has a large variance</option>
        <option value="c">The variable is correlated with the error term</option>
        <option value="d">The variable is binary</option>
    </select>
</div>

<div class="answer-row">
    <label>ii. Which IV assumption can be directly tested? </label>
    <select onchange="checkAnswer(this, 'a')">
        <option value="">-- Select Answer --</option>
        <option value="a">Relevance (Z predicts X)</option>
        <option value="b">Exclusion restriction (Z affects Y only through X)</option>
        <option value="c">Both can be tested</option>
        <option value="d">Neither can be tested</option>
    </select>
</div>

<div class="answer-row">
    <label>iii. In the Angrist-Evans study, why is same-sex a valid instrument for fertility? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">Same-sex children are more expensive to raise</option>
        <option value="b">Parents prefer mixed-sex children and try for another if they have same-sex</option>
        <option value="c">Same-sex children make mothers more likely to work</option>
        <option value="d">Same-sex is correlated with education</option>
    </select>
</div>

<div class="answer-row">
    <label>iv. What is the rule of thumb for detecting weak instruments? </label>
    <select onchange="checkAnswer(this, 'd')">
        <option value="">-- Select Answer --</option>
        <option value="a">R² > 0.5 in the first stage</option>
        <option value="b">p-value < 0.05 on the instrument</option>
        <option value="c">First-stage F-statistic > 100</option>
        <option value="d">First-stage F-statistic > 10</option>
    </select>
</div>

<div class="answer-row">
    <label>v. IV estimates the effect for which subpopulation? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">Always-takers</option>
        <option value="b">Compliers</option>
        <option value="c">Never-takers</option>
        <option value="d">The entire population</option>
    </select>
</div>

<div class="answer-row">
    <label>vi. <strong>(Challenge)</strong> If OLS shows a negative relationship between X and Y, but 2SLS shows no relationship, what might this suggest? </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">The instrument is weak</option>
        <option value="b">The true effect is negative but larger than OLS suggests</option>
        <option value="c">Selection bias was creating a spurious negative correlation; the true causal effect may be zero</option>
        <option value="d">The exclusion restriction is violated</option>
    </select>
</div>
</div>

```

::: {.callout-tip collapse="true"}
## Show Explanation

i. Endogeneity means the explanatory variable is correlated with the error term: Cov(X, μ) ≠ 0. This violates the key OLS assumption and causes bias. Common sources include omitted variables, selection, and reverse causality.

ii. Relevance (Cov(Z, X) ≠ 0) can be tested by regressing X on Z and checking the F-statistic. The exclusion restriction (Cov(Z, μ) = 0) cannot be directly tested because we don't observe μ. It must be justified with economic reasoning.

iii. Parents tend to prefer having children of both sexes. If their first two children are the same sex, they're more likely to try for a third to "get" the other sex. Since biological sex is essentially random, this creates exogenous variation in fertility.

iv. Stock and Yogo (2005) established that a first-stage F-statistic below 10 indicates a weak instrument. With weak instruments, 2SLS estimates are biased (toward OLS) and inference is unreliable.

v. IV estimates the Local Average Treatment Effect (LATE) for compliers—those whose treatment status was actually changed by the instrument. It doesn't identify effects for always-takers or never-takers, whose treatment status is unaffected by the instrument.

vi. This pattern suggests that the OLS relationship was driven by selection or omitted variable bias rather than a true causal effect. For example, if people who choose X also have characteristics that lead to lower Y, OLS would show a negative correlation even if X has no causal effect. 2SLS, by using only exogenous variation, reveals that the true causal effect is zero.

:::

::::

:::: {.content-hidden when-format="html"}

*Note: This section contains interactive content only available in the HTML version.*

::::
