# R Functions and Packages Reference {#sec-r-reference}

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(tidyverse)
```

This appendix provides a quick reference for all the R functions and packages used throughout this book.
Use this as a handy guide when you need to remember the syntax or arguments for a particular function.


## Packages Used in This Book

### tidyverse

The `tidyverse` is a collection of R packages designed for data science.
When you load `tidyverse`, it automatically loads several packages including `ggplot2`, `dplyr`, `tidyr`, `readr`, and others.

**What it's used for:**

- Data manipulation and transformation (`dplyr`)
- Data visualization (`ggplot2`)
- Reshaping data (`tidyr`)
- Reading data files (`readr`)

**Loading:**
```{r}
#| eval: false
library(tidyverse)
```

---

### modelsummary

The `modelsummary` package creates publication-quality tables for regression results and descriptive statistics.

**What it's used for:**

- Creating formatted regression tables comparing multiple models
- Generating descriptive statistics tables
- Exporting tables to Word, LaTeX, or HTML

**Loading:**
```{r}
#| eval: false
library(modelsummary)
```

---

### palmerpenguins

A dataset package containing measurements of penguins from Palmer Station, Antarctica.
Great for learning data visualization and exploration.

**What it's used for:**

- Practice dataset for learning R
- Data visualization examples
- Regression examples

**Loading:**
```{r}
#| eval: false
library(palmerpenguins)
data(penguins)
```

---

### wooldridge

Contains all datasets from Wooldridge's *Introductory Econometrics* textbook.

**What it's used for:**

- Real-world econometric datasets
- Regression examples (wages, crime, etc.)

**Loading:**
```{r}
#| eval: false
library(wooldridge)
data(wage1)  # Example: load the wage1 dataset
```

---

### fixest

A fast and powerful package for fixed effects and panel data regression.

**What it's used for:**

- Panel data regression with fixed effects
- Clustered standard errors
- Very large datasets (faster than `lm()`)

**Loading:**
```{r}
#| eval: false
library(fixest)
```

---

### lmtest

Provides diagnostic tests for linear regression models.

**What it's used for:**

- Breusch-Pagan test for heteroskedasticity
- Other specification tests

**Loading:**
```{r}
#| eval: false
library(lmtest)
```

---

### sandwich

Provides robust covariance matrix estimators (heteroskedasticity-consistent standard errors).

**What it's used for:**

- Robust standard errors
- Heteroskedasticity-consistent (HC) standard errors

**Loading:**
```{r}
#| eval: false
library(sandwich)
```

---

### patchwork

Easily combine multiple ggplot2 plots into a single figure.

**What it's used for:**

- Arranging multiple plots side by side
- Creating multi-panel figures

**Loading:**
```{r}
#| eval: false
library(patchwork)
```

---


## Data Inspection Functions

### `head()`

Display the first few rows of a data frame.

**Arguments:**

- `x`: A data frame or vector
- `n`: Number of rows to display (default: 6)

**Example:**
```{r}
head(mtcars, n = 3)
```

---

### `glimpse()`

Get a compact overview of a data frame's structure and column types.

**Arguments:**

- `x`: A data frame

**Example:**
```{r}
glimpse(mtcars)
```

---

### `summary()`

Generate summary statistics for each variable.

**Arguments:**

- `object`: A data frame, vector, or model object

**Example:**
```{r}
summary(mtcars$mpg)
```

---

### `nrow()` and `ncol()`

Return the number of rows or columns in a data frame.

**Arguments:**

- `x`: A data frame or matrix

**Example:**
```{r}
nrow(mtcars)
ncol(mtcars)
```

---

### `colnames()`

Return or set the column names of a data frame.

**Arguments:**

- `x`: A data frame or matrix

**Example:**
```{r}
colnames(mtcars)
```

---

### `class()` and `typeof()`

Return the class or underlying type of an object.

**Arguments:**

- `x`: Any R object

**Example:**
```{r}
class(mtcars$mpg)
typeof(mtcars$mpg)
```

---


## Descriptive Statistics Functions

### `mean()`

Calculate the arithmetic mean.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
x <- c(10, 20, 30, NA, 50)
mean(x, na.rm = TRUE)
```

---

### `sd()`

Calculate the standard deviation.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
sd(mtcars$mpg)
```

---

### `median()`

Calculate the median (middle value).

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
median(mtcars$mpg)
```

---

### `min()` and `max()`

Find the minimum or maximum value.

**Arguments:**

- `...`: Numeric vectors
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
min(mtcars$mpg)
max(mtcars$mpg)
```

---

### `sum()`

Calculate the sum of all values.

**Arguments:**

- `...`: Numeric vectors
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
sum(mtcars$mpg)
```

---

### `var()`

Calculate the variance.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
var(mtcars$mpg)
```

---

### `cor()`

Calculate the correlation between two variables.

**Arguments:**

- `x`, `y`: Numeric vectors
- `use`: How to handle missing values (e.g., `"complete.obs"`)

**Example:**
```{r}
cor(mtcars$mpg, mtcars$hp)
```

---


## Data Manipulation Functions (dplyr)

### `select()`

Choose which columns to keep.

**Arguments:**

- `.data`: A data frame
- `...`: Column names or selection helpers

**Example:**
```{r}
mtcars |>
  select(mpg, cyl, hp) |>
  head(3)
```

---

### `filter()`

Keep rows that meet a condition.

**Arguments:**

- `.data`: A data frame
- `...`: Logical conditions

**Example:**
```{r}
mtcars |>
  filter(mpg > 25) |>
  head(3)
```

---

### `mutate()`

Create new variables or modify existing ones.

**Arguments:**

- `.data`: A data frame
- `...`: Name-value pairs of expressions

**Example:**
```{r}
mtcars |>
  mutate(kpl = mpg * 0.425) |>  # Convert to km per liter
  select(mpg, kpl) |>
  head(3)
```

---

### `group_by()`

Group data by one or more variables (usually followed by `summarize()`).

**Arguments:**

- `.data`: A data frame
- `...`: Variables to group by

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(avg_mpg = mean(mpg))
```

---

### `summarize()` / `summarise()`

Calculate summary statistics for each group.

**Arguments:**

- `.data`: A data frame (usually grouped)
- `...`: Name-value pairs of summary functions

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    avg_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

---

### `count()`

Count observations by group.

**Arguments:**

- `x`: A data frame
- `...`: Variables to count by
- `sort`: Sort by count? (default: `FALSE`)

**Example:**
```{r}
mtcars |>
  count(cyl)
```

---

### `arrange()`

Sort rows by one or more variables.

**Arguments:**

- `.data`: A data frame
- `...`: Variables to sort by (use `desc()` for descending)

**Example:**
```{r}
mtcars |>
  arrange(desc(mpg)) |>
  head(3)
```

---

### `case_when()`

Vectorized if-else for creating categorical variables.

**Arguments:**

- `...`: A sequence of two-sided formulas: `condition ~ value`

**Example:**
```{r}
mtcars |>
  mutate(
    mpg_category = case_when(
      mpg < 15 ~ "Low",
      mpg < 25 ~ "Medium",
      TRUE ~ "High"
    )
  ) |>
  count(mpg_category)
```

---

### `ifelse()`

Simple conditional: if condition is TRUE, return one value; otherwise, another.

**Arguments:**

- `test`: A logical condition
- `yes`: Value if TRUE
- `no`: Value if FALSE

**Example:**
```{r}
mtcars |>
  mutate(efficient = ifelse(mpg > 20, 1, 0)) |>
  count(efficient)
```

---

### `n()`

Count the number of observations in the current group (used inside `summarize()`).

**Arguments:** None

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(count = n())
```

---


## Regression Functions

### `lm()`

Fit a linear model using Ordinary Least Squares (OLS).

**Arguments:**

- `formula`: A formula like `y ~ x1 + x2`
- `data`: A data frame

**Example:**
```{r}
reg <- lm(mpg ~ hp + wt, data = mtcars)
summary(reg)
```

---

### `summary()` (for regression)

Display detailed regression results including coefficients, standard errors, t-statistics, and p-values.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
#| eval: false
reg <- lm(mpg ~ hp, data = mtcars)
summary(reg)
```

---

### `coef()`

Extract the estimated coefficients from a model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
coef(reg)
```

---

### `confint()`

Calculate confidence intervals for model coefficients.

**Arguments:**

- `object`: A fitted model object
- `level`: Confidence level (default: 0.95)

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
confint(reg, level = 0.95)
```

---

### `predict()`

Generate predicted values from a fitted model.

**Arguments:**

- `object`: A fitted model object
- `newdata`: Optional data frame with new predictor values

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
# Predict for existing data
head(predict(reg))

# Predict for new data
predict(reg, newdata = data.frame(hp = c(100, 150, 200)))
```

---

### `residuals()`

Extract residuals (prediction errors) from a fitted model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
head(residuals(reg))
```

---

### `feols()` (fixest package)

Fit linear models with fixed effects and robust standard errors.

**Arguments:**

- `fml`: A formula (use `|` for fixed effects)
- `data`: A data frame
- `vcov`: Type of standard errors (e.g., `"HC1"` for robust)

**Example:**
```{r}
#| eval: false
library(fixest)
reg <- feols(mpg ~ hp + wt, data = mtcars, vcov = "HC1")
summary(reg)
```

---

### `modelsummary()` (modelsummary package)

Create publication-quality regression tables.

**Arguments:**

- `models`: A model or list of models
- `stars`: Show significance stars?
- `gof_map`: Which goodness-of-fit statistics to show

**Example:**
```{r}
#| eval: false
library(modelsummary)
reg1 <- lm(mpg ~ hp, data = mtcars)
reg2 <- lm(mpg ~ hp + wt, data = mtcars)
modelsummary(list(reg1, reg2), stars = TRUE)
```

---


## Visualization Functions (ggplot2)

### `ggplot()`

Initialize a ggplot object.

**Arguments:**

- `data`: A data frame
- `mapping`: Aesthetic mappings created by `aes()`

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point()
```

---

### `aes()`

Define aesthetic mappings (which variables map to x, y, color, etc.).

**Arguments:**

- `x`, `y`: Variables for axes
- `color`, `fill`: Variables for color
- `size`, `shape`: Variables for size and shape

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point()
```

---

### `geom_point()`

Add points to a plot (scatterplot).

**Arguments:**

- `size`: Point size
- `color`: Point color
- `alpha`: Transparency (0 to 1)

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(size = 3, color = "steelblue", alpha = 0.7)
```

---

### `geom_line()`

Add lines connecting points.

**Arguments:**

- `linewidth`: Line thickness
- `color`: Line color
- `linetype`: Line type (e.g., "dashed")

**Example:**
```{r}
#| fig-height: 3
# Line plot (useful for time series)
df <- data.frame(x = 1:10, y = cumsum(rnorm(10)))
ggplot(df, aes(x = x, y = y)) +
  geom_line(color = "steelblue", linewidth = 1)
```

---

### `geom_histogram()`

Create a histogram.

**Arguments:**

- `binwidth`: Width of each bin
- `bins`: Number of bins
- `fill`: Bar fill color
- `color`: Bar outline color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "white")
```

---

### `geom_boxplot()`

Create a boxplot.

**Arguments:**

- `fill`: Box fill color
- `outlier.color`: Color of outlier points

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "lightblue")
```

---

### `geom_bar()` and `geom_col()`

Create bar charts. `geom_bar()` counts observations; `geom_col()` uses values directly.

**Arguments:**

- `fill`: Bar fill color
- `stat`: For `geom_bar()`, use `"identity"` to plot values directly

**Example:**
```{r}
#| fig-height: 3
# geom_bar counts automatically
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue")
```

---

### `geom_smooth()`

Add a smoothed conditional mean (often a regression line).

**Arguments:**

- `method`: Smoothing method (e.g., `"lm"` for linear)
- `se`: Show confidence interval? (default: `TRUE`)
- `color`: Line color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

---

### `geom_hline()` and `geom_vline()`

Add horizontal or vertical reference lines.

**Arguments:**

- `yintercept` / `xintercept`: Where to draw the line
- `linetype`: Line type (e.g., `"dashed"`)
- `color`: Line color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_hline(yintercept = mean(mtcars$mpg), linetype = "dashed", color = "red")
```

---

### `labs()`

Add labels to the plot (title, axis labels, etc.).

**Arguments:**

- `title`: Plot title
- `x`, `y`: Axis labels
- `color`, `fill`: Legend titles

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs. Horsepower",
    x = "Horsepower",
    y = "Miles per Gallon"
  )
```

---

### `facet_wrap()`

Create small multiples (separate panels for each level of a variable).

**Arguments:**

- `facets`: A formula like `~ variable`
- `nrow`, `ncol`: Number of rows or columns
- `scales`: Should scales be fixed or free?

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl, nrow = 1)
```

---

### `theme_minimal()`

Apply a clean, minimal theme to the plot.

**Arguments:** None (or see `theme()` for customization)

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  theme_minimal()
```

---


## Utility Functions

### `c()`

Combine values into a vector.

**Arguments:**

- `...`: Values to combine

**Example:**
```{r}
x <- c(1, 2, 3, 4, 5)
x
```

---

### `seq()`

Generate a sequence of numbers.

**Arguments:**

- `from`: Starting value
- `to`: Ending value
- `by`: Increment
- `length.out`: Desired length of sequence

**Example:**
```{r}
seq(from = 0, to = 10, by = 2)
seq(from = 0, to = 1, length.out = 5)
```

---

### `rep()`

Repeat values.

**Arguments:**

- `x`: Value(s) to repeat
- `times`: Number of times to repeat

**Example:**
```{r}
rep(c("A", "B"), times = 3)
```

---

### `sample()`

Take a random sample.

**Arguments:**

- `x`: Vector to sample from
- `size`: Number of items to sample
- `replace`: Sample with replacement?

**Example:**
```{r}
set.seed(123)
sample(1:10, size = 5, replace = FALSE)
```

---

### `set.seed()`

Set the random number seed for reproducibility.

**Arguments:**

- `seed`: An integer

**Example:**
```{r}
set.seed(42)
rnorm(3)  # Will always produce the same values
```

---

### `rnorm()`

Generate random numbers from a normal distribution.

**Arguments:**

- `n`: Number of values to generate
- `mean`: Mean of the distribution (default: 0)
- `sd`: Standard deviation (default: 1)

**Example:**
```{r}
set.seed(123)
rnorm(5, mean = 100, sd = 15)
```

---

### `factor()`

Create a factor (categorical variable).

**Arguments:**

- `x`: A vector
- `levels`: The allowed levels (in order)
- `labels`: Labels for the levels

**Example:**
```{r}
education <- c("HS", "College", "HS", "Graduate")
factor(education, levels = c("HS", "College", "Graduate"))
```

---

### `as.factor()`, `as.numeric()`, `as.character()`

Convert objects to a different type.

**Arguments:**

- `x`: Object to convert

**Example:**
```{r}
x <- c("1", "2", "3")
as.numeric(x)
```

---

### `is.na()`

Check for missing values.

**Arguments:**

- `x`: A vector or data frame

**Example:**
```{r}
x <- c(1, 2, NA, 4)
is.na(x)
sum(is.na(x))  # Count missing values
```

---

### `library()`

Load an installed package.

**Arguments:**

- `package`: Name of the package (unquoted)

**Example:**
```{r}
#| eval: false
library(tidyverse)
```

---

### `install.packages()`

Install a package from CRAN (run once, not in scripts).

**Arguments:**

- `pkgs`: Package name (quoted)

**Example:**
```{r}
#| eval: false
install.packages("tidyverse")
```

---


## The Pipe Operator: `|>`

The pipe operator takes the output from the left side and passes it as the first argument to the function on the right.
This allows you to chain operations together in a readable way.

**Example without pipe:**
```{r}
#| eval: false
# Nested functions are hard to read
summary(select(filter(mtcars, mpg > 20), mpg, hp))
```

**Example with pipe:**
```{r}
# Piped version is much clearer
mtcars |>
  filter(mpg > 20) |>
  select(mpg, hp) |>
  summary()
```

::: {.callout-tip}
## Keyboard Shortcut
In RStudio, type **Cmd/Ctrl + Shift + M** to insert the pipe operator.
:::

---


## Quick Reference Tables

### Descriptive Statistics Functions

| Function | Purpose | Key Argument |
|----------|---------|--------------|
| `mean()` | Average | `na.rm = TRUE` |
| `sd()` | Standard deviation | `na.rm = TRUE` |
| `median()` | Middle value | `na.rm = TRUE` |
| `min()`, `max()` | Extremes | `na.rm = TRUE` |
| `sum()` | Total | `na.rm = TRUE` |
| `var()` | Variance | `na.rm = TRUE` |
| `cor()` | Correlation | `use = "complete.obs"` |
| `summary()` | Multiple stats | — |

: Descriptive statistics functions {#tbl-desc-stats-ref}

### Data Manipulation Functions (dplyr)

| Function | Purpose | Example |
|----------|---------|---------|
| `select()` | Choose columns | `select(data, col1, col2)` |
| `filter()` | Choose rows | `filter(data, x > 5)` |
| `mutate()` | Create/modify variables | `mutate(data, new = x * 2)` |
| `group_by()` | Group data | `group_by(data, category)` |
| `summarize()` | Aggregate | `summarize(data, avg = mean(x))` |
| `count()` | Count rows | `count(data, category)` |
| `arrange()` | Sort rows | `arrange(data, desc(x))` |

: Data manipulation functions {#tbl-dplyr-ref}

### Regression Functions

| Function | Purpose | Package |
|----------|---------|---------|
| `lm()` | OLS regression | base R |
| `summary()` | Model results | base R |
| `coef()` | Coefficients | base R |
| `confint()` | Confidence intervals | base R |
| `predict()` | Fitted values | base R |
| `residuals()` | Residuals | base R |
| `feols()` | Fixed effects + robust SE | fixest |
| `modelsummary()` | Formatted tables | modelsummary |

: Regression functions {#tbl-reg-ref}

### ggplot2 Geometries

| Function | Plot Type |
|----------|-----------|
| `geom_point()` | Scatterplot |
| `geom_line()` | Line plot |
| `geom_histogram()` | Histogram |
| `geom_boxplot()` | Boxplot |
| `geom_bar()` | Bar chart (counts) |
| `geom_col()` | Bar chart (values) |
| `geom_smooth()` | Smoothed line/regression |
| `geom_hline()` | Horizontal line |
| `geom_vline()` | Vertical line |

: ggplot2 geometries {#tbl-ggplot-ref}
