# R Functions and Packages Reference {#sec-r-reference}

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(tidyverse)
```

This appendix provides a quick reference for all the R functions and packages used throughout this book.
Use this as a handy guide when you need to remember the syntax or arguments for a particular function.


## Packages Used in This Book

### tidyverse

The `tidyverse` is a collection of R packages designed for data science.
When you load `tidyverse`, it automatically loads several packages including `ggplot2`, `dplyr`, `tidyr`, `readr`, and others.

**What it's used for:**

- Data manipulation and transformation (`dplyr`)
- Data visualization (`ggplot2`)
- Reshaping data (`tidyr`)
- Reading data files (`readr`)

**Loading:**
```{r}
#| eval: false
library(tidyverse)
```

---

### modelsummary

The `modelsummary` package creates publication-quality tables for regression results and descriptive statistics.

**What it's used for:**

- Creating formatted regression tables comparing multiple models
- Generating descriptive statistics tables
- Exporting tables to Word, LaTeX, or HTML

**Loading:**
```{r}
#| eval: false
library(modelsummary)
```

---

### palmerpenguins

A dataset package containing measurements of penguins from Palmer Station, Antarctica.
Great for learning data visualization and exploration.

**What it's used for:**

- Practice dataset for learning R
- Data visualization examples
- Regression examples

**Loading:**
```{r}
#| eval: false
library(palmerpenguins)
data(penguins)
```

---

### wooldridge

Contains all datasets from Wooldridge's *Introductory Econometrics* textbook.

**What it's used for:**

- Real-world econometric datasets
- Regression examples (wages, crime, etc.)

**Loading:**
```{r}
#| eval: false
library(wooldridge)
data(wage1)  # Example: load the wage1 dataset
```

---

### fixest

A fast and powerful package for fixed effects and panel data regression.

**What it's used for:**

- Panel data regression with fixed effects
- Clustered standard errors
- Very large datasets (faster than `lm()`)

**Loading:**
```{r}
#| eval: false
library(fixest)
```

---

### lmtest

Provides diagnostic tests for linear regression models.

**What it's used for:**

- Breusch-Pagan test for heteroskedasticity
- Other specification tests

**Loading:**
```{r}
#| eval: false
library(lmtest)
```

---

### sandwich

Provides robust covariance matrix estimators (heteroskedasticity-consistent standard errors).

**What it's used for:**

- Robust standard errors
- Heteroskedasticity-consistent (HC) standard errors

**Loading:**
```{r}
#| eval: false
library(sandwich)
```

---

### patchwork

Easily combine multiple ggplot2 plots into a single figure.

**What it's used for:**

- Arranging multiple plots side by side
- Creating multi-panel figures

**Loading:**
```{r}
#| eval: false
library(patchwork)
```

---


## Data Inspection Functions

### `head()`

Display the first few rows of a data frame.

**Arguments:**

- `x`: A data frame or vector
- `n`: Number of rows to display (default: 6)

**Example:**
```{r}
head(mtcars, n = 3)
```

---

### `glimpse()`

Get a compact overview of a data frame's structure and column types.

**Arguments:**

- `x`: A data frame

**Example:**
```{r}
glimpse(mtcars)
```

---

### `summary()`

Generate summary statistics for each variable.

**Arguments:**

- `object`: A data frame, vector, or model object

**Example:**
```{r}
summary(mtcars$mpg)
```

---

### `nrow()` and `ncol()`

Return the number of rows or columns in a data frame.

**Arguments:**

- `x`: A data frame or matrix

**Example:**
```{r}
nrow(mtcars)
ncol(mtcars)
```

---

### `colnames()`

Return or set the column names of a data frame.

**Arguments:**

- `x`: A data frame or matrix

**Example:**
```{r}
colnames(mtcars)
```

---

### `class()` and `typeof()`

Return the class or underlying type of an object.

**Arguments:**

- `x`: Any R object

**Example:**
```{r}
class(mtcars$mpg)
typeof(mtcars$mpg)
```

---


## Descriptive Statistics Functions

### `mean()`

Calculate the arithmetic mean.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
x <- c(10, 20, 30, NA, 50)
mean(x, na.rm = TRUE)
```

---

### `sd()`

Calculate the standard deviation.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
sd(mtcars$mpg)
```

---

### `median()`

Calculate the median (middle value).

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
median(mtcars$mpg)
```

---

### `min()` and `max()`

Find the minimum or maximum value.

**Arguments:**

- `...`: Numeric vectors
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
min(mtcars$mpg)
max(mtcars$mpg)
```

---

### `sum()`

Calculate the sum of all values.

**Arguments:**

- `...`: Numeric vectors
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
sum(mtcars$mpg)
```

---

### `var()`

Calculate the variance.

**Arguments:**

- `x`: A numeric vector
- `na.rm`: Remove missing values? (default: `FALSE`)

**Example:**
```{r}
var(mtcars$mpg)
```

---

### `cor()`

Calculate the correlation between two variables.

**Arguments:**

- `x`, `y`: Numeric vectors
- `use`: How to handle missing values (e.g., `"complete.obs"`)

**Example:**
```{r}
cor(mtcars$mpg, mtcars$hp)
```

---


## Data Manipulation Functions (dplyr)

### `select()`

Choose which columns to keep.

**Arguments:**

- `.data`: A data frame
- `...`: Column names or selection helpers

**Example:**
```{r}
mtcars |>
  select(mpg, cyl, hp) |>
  head(3)
```

---

### `filter()`

Keep rows that meet a condition.

**Arguments:**

- `.data`: A data frame
- `...`: Logical conditions

**Example:**
```{r}
mtcars |>
  filter(mpg > 25) |>
  head(3)
```

---

### `mutate()`

Create new variables or modify existing ones.

**Arguments:**

- `.data`: A data frame
- `...`: Name-value pairs of expressions

**Example:**
```{r}
mtcars |>
  mutate(kpl = mpg * 0.425) |>  # Convert to km per liter
  select(mpg, kpl) |>
  head(3)
```

---

### `group_by()`

Group data by one or more variables (usually followed by `summarize()`).

**Arguments:**

- `.data`: A data frame
- `...`: Variables to group by

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(avg_mpg = mean(mpg))
```

---

### `summarize()` / `summarise()`

Calculate summary statistics for each group.

**Arguments:**

- `.data`: A data frame (usually grouped)
- `...`: Name-value pairs of summary functions

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    avg_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

---

### `count()`

Count observations by group.

**Arguments:**

- `x`: A data frame
- `...`: Variables to count by
- `sort`: Sort by count? (default: `FALSE`)

**Example:**
```{r}
mtcars |>
  count(cyl)
```

---

### `arrange()`

Sort rows by one or more variables.

**Arguments:**

- `.data`: A data frame
- `...`: Variables to sort by (use `desc()` for descending)

**Example:**
```{r}
mtcars |>
  arrange(desc(mpg)) |>
  head(3)
```

---

### `case_when()`

Vectorized if-else for creating categorical variables.

**Arguments:**

- `...`: A sequence of two-sided formulas: `condition ~ value`

**Example:**
```{r}
mtcars |>
  mutate(
    mpg_category = case_when(
      mpg < 15 ~ "Low",
      mpg < 25 ~ "Medium",
      TRUE ~ "High"
    )
  ) |>
  count(mpg_category)
```

---

### `ifelse()`

Simple conditional: if condition is TRUE, return one value; otherwise, another.

**Arguments:**

- `test`: A logical condition
- `yes`: Value if TRUE
- `no`: Value if FALSE

**Example:**
```{r}
mtcars |>
  mutate(efficient = ifelse(mpg > 20, 1, 0)) |>
  count(efficient)
```

---

### `n()`

Count the number of observations in the current group (used inside `summarize()`).

**Arguments:** None

**Example:**
```{r}
mtcars |>
  group_by(cyl) |>
  summarize(count = n())
```

---


## Regression Functions

### `lm()`

Fit a linear model using Ordinary Least Squares (OLS).

**Arguments:**

- `formula`: A formula like `y ~ x1 + x2`
- `data`: A data frame

**Example:**
```{r}
reg <- lm(mpg ~ hp + wt, data = mtcars)
summary(reg)
```

---

### `summary()` (for regression)

Display detailed regression results including coefficients, standard errors, t-statistics, and p-values.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
#| eval: false
reg <- lm(mpg ~ hp, data = mtcars)
summary(reg)
```

---

### `coef()`

Extract the estimated coefficients from a model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
coef(reg)
```

---

### `confint()`

Calculate confidence intervals for model coefficients.

**Arguments:**

- `object`: A fitted model object
- `level`: Confidence level (default: 0.95)

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
confint(reg, level = 0.95)
```

---

### `predict()`

Generate predicted values from a fitted model.

**Arguments:**

- `object`: A fitted model object
- `newdata`: Optional data frame with new predictor values

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
# Predict for existing data
head(predict(reg))

# Predict for new data
predict(reg, newdata = data.frame(hp = c(100, 150, 200)))
```

---

### `residuals()`

Extract residuals (prediction errors) from a fitted model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
head(residuals(reg))
```

---

### `feols()` (fixest package)

Fit linear models with fixed effects and robust standard errors.

**Arguments:**

- `fml`: A formula (use `|` for fixed effects)
- `data`: A data frame
- `vcov`: Type of standard errors (e.g., `"HC1"` for robust)

**Example:**
```{r}
#| eval: false
library(fixest)
reg <- feols(mpg ~ hp + wt, data = mtcars, vcov = "HC1")
summary(reg)
```

---

### `modelsummary()` (modelsummary package)

Create publication-quality regression tables.

**Arguments:**

- `models`: A model or list of models
- `stars`: Show significance stars?
- `gof_map`: Which goodness-of-fit statistics to show

**Example:**
```{r}
#| eval: false
library(modelsummary)
reg1 <- lm(mpg ~ hp, data = mtcars)
reg2 <- lm(mpg ~ hp + wt, data = mtcars)
modelsummary(list(reg1, reg2), stars = TRUE)
```

---


## Visualization Functions (ggplot2)

### `ggplot()`

Initialize a ggplot object.

**Arguments:**

- `data`: A data frame
- `mapping`: Aesthetic mappings created by `aes()`

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point()
```

---

### `aes()`

Define aesthetic mappings (which variables map to x, y, color, etc.).

**Arguments:**

- `x`, `y`: Variables for axes
- `color`, `fill`: Variables for color
- `size`, `shape`: Variables for size and shape

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point()
```

---

### `geom_point()`

Add points to a plot (scatterplot).

**Arguments:**

- `size`: Point size
- `color`: Point color
- `alpha`: Transparency (0 to 1)

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(size = 3, color = "steelblue", alpha = 0.7)
```

---

### `geom_line()`

Add lines connecting points.

**Arguments:**

- `linewidth`: Line thickness
- `color`: Line color
- `linetype`: Line type (e.g., "dashed")

**Example:**
```{r}
#| fig-height: 3
# Line plot (useful for time series)
df <- data.frame(x = 1:10, y = cumsum(rnorm(10)))
ggplot(df, aes(x = x, y = y)) +
  geom_line(color = "steelblue", linewidth = 1)
```

---

### `geom_histogram()`

Create a histogram.

**Arguments:**

- `binwidth`: Width of each bin
- `bins`: Number of bins
- `fill`: Bar fill color
- `color`: Bar outline color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "white")
```

---

### `geom_boxplot()`

Create a boxplot.

**Arguments:**

- `fill`: Box fill color
- `outlier.color`: Color of outlier points

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(fill = "lightblue")
```

---

### `geom_bar()` and `geom_col()`

Create bar charts. `geom_bar()` counts observations; `geom_col()` uses values directly.

**Arguments:**

- `fill`: Bar fill color
- `stat`: For `geom_bar()`, use `"identity"` to plot values directly

**Example:**
```{r}
#| fig-height: 3
# geom_bar counts automatically
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue")
```

---

### `geom_smooth()`

Add a smoothed conditional mean (often a regression line).

**Arguments:**

- `method`: Smoothing method (e.g., `"lm"` for linear)
- `se`: Show confidence interval? (default: `TRUE`)
- `color`: Line color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

---

### `geom_hline()` and `geom_vline()`

Add horizontal or vertical reference lines.

**Arguments:**

- `yintercept` / `xintercept`: Where to draw the line
- `linetype`: Line type (e.g., `"dashed"`)
- `color`: Line color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_hline(yintercept = mean(mtcars$mpg), linetype = "dashed", color = "red")
```

---

### `labs()`

Add labels to the plot (title, axis labels, etc.).

**Arguments:**

- `title`: Plot title
- `x`, `y`: Axis labels
- `color`, `fill`: Legend titles

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs. Horsepower",
    x = "Horsepower",
    y = "Miles per Gallon"
  )
```

---

### `facet_wrap()`

Create small multiples (separate panels for each level of a variable).

**Arguments:**

- `facets`: A formula like `~ variable`
- `nrow`, `ncol`: Number of rows or columns
- `scales`: Should scales be fixed or free?

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl, nrow = 1)
```

---

### `theme_minimal()`

Apply a clean, minimal theme to the plot.

**Arguments:** None (or see `theme()` for customization)

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  theme_minimal()
```

---


## Utility Functions

### `c()`

Combine values into a vector.

**Arguments:**

- `...`: Values to combine

**Example:**
```{r}
x <- c(1, 2, 3, 4, 5)
x
```

---

### `seq()`

Generate a sequence of numbers.

**Arguments:**

- `from`: Starting value
- `to`: Ending value
- `by`: Increment
- `length.out`: Desired length of sequence

**Example:**
```{r}
seq(from = 0, to = 10, by = 2)
seq(from = 0, to = 1, length.out = 5)
```

---

### `rep()`

Repeat values.

**Arguments:**

- `x`: Value(s) to repeat
- `times`: Number of times to repeat

**Example:**
```{r}
rep(c("A", "B"), times = 3)
```

---

### `sample()`

Take a random sample.

**Arguments:**

- `x`: Vector to sample from
- `size`: Number of items to sample
- `replace`: Sample with replacement?

**Example:**
```{r}
set.seed(123)
sample(1:10, size = 5, replace = FALSE)
```

---

### `set.seed()`

Set the random number seed for reproducibility.

**Arguments:**

- `seed`: An integer

**Example:**
```{r}
set.seed(42)
rnorm(3)  # Will always produce the same values
```

---

### `rnorm()`

Generate random numbers from a normal distribution.

**Arguments:**

- `n`: Number of values to generate
- `mean`: Mean of the distribution (default: 0)
- `sd`: Standard deviation (default: 1)

**Example:**
```{r}
set.seed(123)
rnorm(5, mean = 100, sd = 15)
```

---

### `factor()`

Create a factor (categorical variable).

**Arguments:**

- `x`: A vector
- `levels`: The allowed levels (in order)
- `labels`: Labels for the levels

**Example:**
```{r}
education <- c("HS", "College", "HS", "Graduate")
factor(education, levels = c("HS", "College", "Graduate"))
```

---

### `as.factor()`, `as.numeric()`, `as.character()`

Convert objects to a different type.

**Arguments:**

- `x`: Object to convert

**Example:**
```{r}
x <- c("1", "2", "3")
as.numeric(x)
```

---

### `is.na()`

Check for missing values.

**Arguments:**

- `x`: A vector or data frame

**Example:**
```{r}
x <- c(1, 2, NA, 4)
is.na(x)
sum(is.na(x))  # Count missing values
```

---

### `library()`

Load an installed package.

**Arguments:**

- `package`: Name of the package (unquoted)

**Example:**
```{r}
#| eval: false
library(tidyverse)
```

---

### `install.packages()`

Install a package from CRAN (run once, not in scripts).

**Arguments:**

- `pkgs`: Package name (quoted)

**Example:**
```{r}
#| eval: false
install.packages("tidyverse")
```

---


## The Pipe Operator: `|>`

The pipe operator takes the output from the left side and passes it as the first argument to the function on the right.
This allows you to chain operations together in a readable way.

**Example without pipe:**
```{r}
#| eval: false
# Nested functions are hard to read
summary(select(filter(mtcars, mpg > 20), mpg, hp))
```

**Example with pipe:**
```{r}
# Piped version is much clearer
mtcars |>
  filter(mpg > 20) |>
  select(mpg, hp) |>
  summary()
```

::: {.callout-tip}
## Keyboard Shortcut
In RStudio, type **Cmd/Ctrl + Shift + M** to insert the pipe operator.
:::

---


## Statistical Distribution Functions

R has a consistent naming convention for distribution functions. Each distribution has four functions, prefixed by a letter:

- `d` = **density** (the height of the PDF at a given value)
- `p` = **probability** (the CDF---cumulative probability up to a given value)
- `q` = **quantile** (the inverse of the CDF---find the value for a given probability)
- `r` = **random** (generate random draws from the distribution)

For example, for the normal distribution: `dnorm()`, `pnorm()`, `qnorm()`, `rnorm()`.

### `dnorm()` and `dt()`

Compute the probability density function (PDF) for the normal or t-distribution. Useful for plotting distribution curves.

**Arguments:**

- `x`: Value(s) at which to evaluate the density
- `mean`, `sd`: Parameters of the normal distribution (for `dnorm()`)
- `df`: Degrees of freedom (for `dt()`)

**Example:**
```{r}
# Height of the standard normal PDF at x = 0
dnorm(0, mean = 0, sd = 1)

# Height of the t-distribution PDF at x = 2 with 30 df
dt(2, df = 30)
```

---

### `pt()`

Compute the cumulative distribution function (CDF) for the t-distribution. This gives you the probability that a t-distributed random variable is less than or equal to a given value. Essential for computing p-values.

**Arguments:**

- `q`: The t-value(s) to evaluate
- `df`: Degrees of freedom
- `lower.tail`: If `TRUE` (default), returns $P(T \leq q)$; if `FALSE`, returns $P(T > q)$

**Example:**
```{r}
# P-value for a two-sided test with t = 2.3 and 100 df
2 * pt(abs(2.3), df = 100, lower.tail = FALSE)
```

---

### `qt()`

Compute the quantile function (inverse CDF) for the t-distribution. Given a probability, it returns the corresponding t-value. Used to find critical values for hypothesis tests.

**Arguments:**

- `p`: Probability (between 0 and 1)
- `df`: Degrees of freedom
- `lower.tail`: If `TRUE` (default), finds the value where $P(T \leq q) = p$; if `FALSE`, finds the value where $P(T > q) = p$

**Example:**
```{r}
# Critical value for a two-sided 5% test with 100 df
# We want the value that leaves 2.5% in the upper tail
qt(0.025, df = 100, lower.tail = FALSE)
```

---

### `qf()`

Compute the quantile function for the F-distribution. Used to find critical values for F-tests.

**Arguments:**

- `p`: Probability
- `df1`: Numerator degrees of freedom (number of restrictions)
- `df2`: Denominator degrees of freedom (from the unrestricted model)
- `lower.tail`: If `FALSE`, returns the value where $P(F > q) = p$

**Example:**
```{r}
# Critical value for an F-test with 2 and 494 degrees of freedom at 5%
qf(0.05, df1 = 2, df2 = 494, lower.tail = FALSE)
```

---

### `runif()`

Generate random draws from a uniform distribution.

**Arguments:**

- `n`: Number of values to generate
- `min`: Minimum value (default: 0)
- `max`: Maximum value (default: 1)

**Example:**
```{r}
set.seed(123)
runif(5, min = 1, max = 10)
```

---

### `rexp()`

Generate random draws from an exponential distribution.

**Arguments:**

- `n`: Number of values to generate
- `rate`: Rate parameter $\lambda$ (default: 1). The mean of the distribution is $1/\lambda$.

**Example:**
```{r}
set.seed(123)
rexp(5, rate = 0.5)  # Mean = 1/0.5 = 2
```

---

### `rbinom()`

Generate random draws from a binomial distribution. Useful for simulating binary outcomes (e.g., treatment assignment).

**Arguments:**

- `n`: Number of values to generate
- `size`: Number of trials
- `prob`: Probability of success on each trial

**Example:**
```{r}
set.seed(123)
# Simulate 10 coin flips (1 = heads, 0 = tails)
rbinom(10, size = 1, prob = 0.5)
```

---

### `rt()`

Generate random draws from a t-distribution.

**Arguments:**

- `n`: Number of values to generate
- `df`: Degrees of freedom

**Example:**
```{r}
set.seed(123)
rt(5, df = 30)
```

---


## Model Diagnostic Functions

### `anova()`

Compare nested models using an F-test. Pass the restricted (smaller) model first, then the unrestricted (larger) model. Tests whether the additional variables in the unrestricted model are jointly significant.

**Arguments:**

- `object`: One or more fitted model objects

**Example:**
```{r}
reg_small <- lm(mpg ~ hp, data = mtcars)
reg_large <- lm(mpg ~ hp + wt + cyl, data = mtcars)
anova(reg_small, reg_large)
```

---

### `nobs()`

Return the number of observations used to fit a model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
nobs(reg)
```

---

### `df.residual()`

Return the residual degrees of freedom ($n - k - 1$) from a fitted model.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp + wt, data = mtcars)
df.residual(reg)  # 32 - 2 - 1 = 29
```

---

### `resid()`

Extract residuals from a fitted model. Equivalent to `residuals()`.

**Arguments:**

- `object`: A fitted model object

**Example:**
```{r}
reg <- lm(mpg ~ hp, data = mtcars)
head(resid(reg))
```

---


## Additional Data Manipulation Functions

### `tibble()`

Create a data frame (modern version). Works like `data.frame()` but with better defaults: it doesn't convert strings to factors and prints more cleanly.

**Arguments:**

- `...`: Name-value pairs of columns

**Example:**
```{r}
tibble(
  name = c("Alice", "Bob", "Carol"),
  age = c(25, 30, 35),
  income = c(50000, 60000, 70000)
)
```

---

### `bind_rows()`

Stack data frames on top of each other (by rows).

**Arguments:**

- `...`: Data frames to bind together

**Example:**
```{r}
df1 <- tibble(x = 1:3, y = c("a", "b", "c"))
df2 <- tibble(x = 4:6, y = c("d", "e", "f"))
bind_rows(df1, df2)
```

---

### `pivot_longer()`

Reshape data from wide format to long format. Useful for making data "tidy" for ggplot2.

**Arguments:**

- `data`: A data frame
- `cols`: Columns to pivot into longer format
- `names_to`: Name of the new column that will contain the old column names
- `values_to`: Name of the new column that will contain the values

**Example:**
```{r}
wide_data <- tibble(
  id = 1:3,
  score_2020 = c(80, 90, 85),
  score_2021 = c(85, 92, 88)
)

wide_data |>
  pivot_longer(
    cols = starts_with("score"),
    names_to = "year",
    values_to = "score"
  )
```

---

### `slice_sample()`

Randomly sample rows from a data frame.

**Arguments:**

- `.data`: A data frame
- `n`: Number of rows to sample
- `replace`: Sample with replacement? (default: `FALSE`)

**Example:**
```{r}
set.seed(123)
mtcars |>
  slice_sample(n = 5)
```

---

### `map_dfr()` (purrr)

Apply a function to each element of a list or vector and combine the results into a single data frame by binding rows. Useful for running simulations.

**Arguments:**

- `.x`: A list or vector to iterate over
- `.f`: A function to apply to each element

**Example:**
```{r}
# Run 5 simulations and collect results
map_dfr(1:5, function(i) {
  x <- rnorm(50)
  tibble(sim = i, mean_x = mean(x))
})
```

---


## Additional ggplot2 Functions

### `annotate()`

Add text, labels, or shapes to a plot at specific coordinates. Unlike `geom_text()`, `annotate()` is for adding single annotations rather than mapping data to text.

**Arguments:**

- `geom`: Type of annotation (e.g., `"text"`, `"rect"`, `"segment"`)
- `x`, `y`: Position of the annotation
- `label`: Text to display (for `"text"` geom)
- `parse`: If `TRUE`, interpret the label as a plotmath expression (default: `FALSE`)
- `color`, `size`: Styling options

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  annotate("text", x = 300, y = 30,
           label = "hat(beta)[1] == -0.068",
           parse = TRUE, color = "red", size = 5)
```

---

### `stat_function()`

Overlay a mathematical function on a ggplot. Useful for plotting theoretical distributions on top of histograms.

**Arguments:**

- `fun`: The function to plot (e.g., `dnorm`, `dt`)
- `args`: A list of additional arguments to pass to the function
- `color`, `linewidth`: Styling options

**Example:**
```{r}
#| fig-height: 3
ggplot(data.frame(x = rnorm(500)), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "lightblue", color = "black") +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1),
                color = "red", linewidth = 1.2)
```

---

### `geom_segment()`

Draw line segments between specified start and end points. Useful for adding arrows, error bars, or connecting points.

**Arguments:**

- `aes(x, y, xend, yend)`: Start and end coordinates
- `arrow`: Add arrowheads with `arrow()`
- `color`, `linewidth`, `linetype`: Styling options

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_segment(aes(x = 150, y = 25, xend = 150, yend = 15),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "red", linewidth = 1)
```

---

### `geom_ribbon()` and `geom_area()`

Shade a region between a `ymin` and `ymax`. `geom_area()` is a special case where `ymin = 0`. Useful for shading regions under distribution curves (e.g., p-values, rejection regions).

**Arguments:**

- `aes(x, ymin, ymax)`: Boundaries of the shaded region
- `fill`: Fill color
- `alpha`: Transparency

**Example:**
```{r}
#| fig-height: 3
shade_data <- tibble(
  x = seq(-3, 3, length.out = 200),
  y = dnorm(x)
)

ggplot(shade_data, aes(x = x)) +
  geom_line(aes(y = y)) +
  geom_ribbon(data = shade_data |> filter(x >= 1.96),
              aes(ymin = 0, ymax = y),
              fill = "red", alpha = 0.5)
```

---

### `geom_errorbar()`

Add error bars to a plot. Can be vertical (default) or horizontal (with `orientation = "y"`). Commonly used for confidence interval plots.

**Arguments:**

- `aes(ymin, ymax)`: Lower and upper bounds (vertical), or `aes(xmin, xmax)` with `orientation = "y"` (horizontal)
- `width`: Width of the error bar caps
- `orientation`: Set to `"y"` for horizontal error bars

**Example:**
```{r}
#| fig-height: 3
ci_data <- tibble(
  variable = c("hp", "wt", "cyl"),
  estimate = c(-0.03, -3.8, -1.5),
  lower = c(-0.05, -5.1, -2.8),
  upper = c(-0.01, -2.5, -0.2)
)

ggplot(ci_data, aes(y = variable, x = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = lower, xmax = upper),
                width = 0.2, orientation = "y") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal()
```

---

### `geom_density()`

Plot a smoothed density estimate. An alternative to histograms for visualizing distributions.

**Arguments:**

- `fill`: Fill color
- `alpha`: Transparency
- `color`: Line color

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "steelblue", alpha = 0.5)
```

---

### `scale_color_manual()` and `scale_fill_manual()`

Manually set colors for categorical variables mapped to `color` or `fill`.

**Arguments:**

- `values`: A named vector of colors
- `labels`: Optional labels for the legend

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point(size = 2) +
  scale_color_manual(values = c("4" = "steelblue", "6" = "orange", "8" = "red"))
```

---

### `coord_cartesian()`

Zoom into a region of the plot without dropping data points (unlike `xlim()`/`ylim()`, which remove data outside the range).

**Arguments:**

- `xlim`: Range for x-axis as `c(min, max)`
- `ylim`: Range for y-axis as `c(min, max)`

**Example:**
```{r}
#| fig-height: 3
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  coord_cartesian(xlim = c(50, 200), ylim = c(15, 35))
```

---


## Quick Reference Tables

### Descriptive Statistics Functions

| Function | Purpose | Key Argument |
|----------|---------|--------------|
| `mean()` | Average | `na.rm = TRUE` |
| `sd()` | Standard deviation | `na.rm = TRUE` |
| `median()` | Middle value | `na.rm = TRUE` |
| `min()`, `max()` | Extremes | `na.rm = TRUE` |
| `sum()` | Total | `na.rm = TRUE` |
| `var()` | Variance | `na.rm = TRUE` |
| `cor()` | Correlation | `use = "complete.obs"` |
| `summary()` | Multiple stats | — |

: Descriptive statistics functions {#tbl-desc-stats-ref}

### Data Manipulation Functions (dplyr)

| Function | Purpose | Example |
|----------|---------|---------|
| `select()` | Choose columns | `select(data, col1, col2)` |
| `filter()` | Choose rows | `filter(data, x > 5)` |
| `mutate()` | Create/modify variables | `mutate(data, new = x * 2)` |
| `group_by()` | Group data | `group_by(data, category)` |
| `summarize()` | Aggregate | `summarize(data, avg = mean(x))` |
| `count()` | Count rows | `count(data, category)` |
| `arrange()` | Sort rows | `arrange(data, desc(x))` |

: Data manipulation functions {#tbl-dplyr-ref}

### Regression Functions

| Function | Purpose | Package |
|----------|---------|---------|
| `lm()` | OLS regression | base R |
| `summary()` | Model results | base R |
| `coef()` | Coefficients | base R |
| `confint()` | Confidence intervals | base R |
| `predict()` | Fitted values | base R |
| `residuals()` / `resid()` | Residuals | base R |
| `anova()` | F-test (compare models) | base R |
| `nobs()` | Number of observations | base R |
| `df.residual()` | Residual degrees of freedom | base R |
| `feols()` | Fixed effects + robust SE | fixest |
| `modelsummary()` | Formatted tables | modelsummary |

: Regression functions {#tbl-reg-ref}

### Statistical Distribution Functions

| Function | Purpose | Example |
|----------|---------|---------|
| `dnorm()`, `dt()` | Density (PDF height) | `dnorm(0)`, `dt(2, df=30)` |
| `pt()` | CDF for t-distribution | `pt(2.3, df=100)` |
| `qt()` | Critical values (t-dist) | `qt(0.025, df=100)` |
| `qf()` | Critical values (F-dist) | `qf(0.05, df1=2, df2=494)` |
| `rnorm()` | Random normal draws | `rnorm(100, mean=0, sd=1)` |
| `runif()` | Random uniform draws | `runif(100, min=0, max=1)` |
| `rbinom()` | Random binomial draws | `rbinom(100, size=1, prob=0.5)` |
| `rexp()` | Random exponential draws | `rexp(100, rate=0.5)` |
| `rt()` | Random t-dist draws | `rt(100, df=30)` |

: Statistical distribution functions {#tbl-dist-ref}

### ggplot2 Geometries

| Function | Plot Type |
|----------|-----------|
| `geom_point()` | Scatterplot |
| `geom_line()` | Line plot |
| `geom_histogram()` | Histogram |
| `geom_density()` | Smoothed density |
| `geom_boxplot()` | Boxplot |
| `geom_bar()` | Bar chart (counts) |
| `geom_col()` | Bar chart (values) |
| `geom_smooth()` | Smoothed line/regression |
| `geom_segment()` | Line segments/arrows |
| `geom_ribbon()` / `geom_area()` | Shaded regions |
| `geom_errorbar()` | Error bars / CIs |
| `geom_hline()` | Horizontal line |
| `geom_vline()` | Vertical line |
| `annotate()` | Text/shape annotations |
| `stat_function()` | Overlay math functions |

: ggplot2 geometries {#tbl-ggplot-ref}
