---
title: "Concept Check 2: Good Data Practices"
subtitle: "EC325 - Introduction to Econometrics"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
---

# Instructions

**Total Points:** 60 points

**Due Date:** [INSERT DUE DATE]

**Submission Requirements:**

You will submit **two files**:

1. **R Script** (`lastname_firstname_cc2.R`) - Contains all your code
2. **Write-up** (PDF or Word) - Contains your written answers and any tables/figures

**Tips for Success:**

- Your R script should be well-organized and include comments explaining what each section does
- Run your code line-by-line to make sure each part works before moving on
- If you get an error, read the error message carefully—it often tells you what went wrong
- Copy any R output or figures into your write-up document

---

# Part A: Understanding Your Data (20 points)

For this assignment, you will use the `penguins` dataset from the `palmerpenguins` package. This dataset contains measurements of 344 penguins from three species collected near Palmer Station, Antarctica.

## Setup

**Start your R script with the following code.** Copy this exactly into a new R script file:

```{r}
#| eval: false
#| echo: true

# =============================================================================
# Concept Check 2: Good Data Practices
# Your Name Here
# Date Here
# =============================================================================

# Load required packages
# (If you get an error, you may need to install these first - see note below)
library(tidyverse)
library(palmerpenguins)

# Load the penguins data into your environment
data(penguins)
```

::: {.callout-note}
## Installing Packages (If Needed)

If you get an error saying a function or package is not found, you need to install it first. Run these commands ONE TIME in your console (not in your script):

```{r}
#| eval: false
install.packages("tidyverse")
install.packages("palmerpenguins")
```

After installing, the `library()` commands in your script should work.
:::

## A.1: First Look at the Data (4 points)

The `glimpse()` function gives us a quick overview of a dataset. Run the following code:

```{r}
#| eval: false
#| echo: true

# Get an overview of the penguins dataset
glimpse(penguins)
```

**Using the output from `glimpse()`, answer the following questions in your write-up:**

a) How many observations (rows) are in the dataset? *(Hint: Look for "Rows:" in the output)* (1 point)

b) How many variables (columns) are in the dataset? *(Hint: Look for "Columns:" in the output)* (1 point)

c) List all the variable names. (1 point)

d) What type of variable is `species`? What type is `body_mass_g`? *(Hint: The abbreviation after each variable name tells you the type. `fct` means factor/categorical, `int` means integer/whole number, `dbl` means double/decimal number)* (1 point)

## A.2: Missing Data (4 points)

Missing data is shown as `NA` in R. The `summary()` function tells us how many missing values each variable has.

**Run this code:**

```{r}
#| eval: false
#| echo: true

# Get summary statistics for all variables
summary(penguins)
```

**Answer in your write-up:**

a) Look at the output for the `sex` variable. How many NA's (missing values) does it have? (2 points)

b) In 2-3 sentences, explain why it's important to know about missing data *before* conducting an analysis. Give one example of how ignoring missing data could lead to misleading conclusions. (2 points)

## A.3: Descriptive Statistics (6 points)

Now let's calculate specific statistics. The `mean()` and `sd()` functions calculate the mean and standard deviation.

**Important:** When data has missing values (NA), we need to tell R to ignore them using `na.rm = TRUE`. Otherwise, R will return NA as the answer.

**Run this code to calculate statistics for flipper length:**

```{r}
#| eval: false
#| echo: true

# Calculate mean flipper length (na.rm = TRUE tells R to ignore missing values)
mean(penguins$flipper_length_mm, na.rm = TRUE)

# Calculate standard deviation of flipper length
sd(penguins$flipper_length_mm, na.rm = TRUE)
```

a) What is the mean flipper length? What is the standard deviation? Round to 2 decimal places. (2 points)

**Now you try! Write code to calculate the mean and standard deviation of `body_mass_g`:**

```{r}
#| eval: false
#| echo: true

# Calculate mean body mass
# YOUR CODE HERE: Replace the blank with the correct variable name
mean(penguins$__________, na.rm = TRUE)

# Calculate standard deviation of body mass
# YOUR CODE HERE: Replace the blank with the correct variable name
sd(penguins$__________, na.rm = TRUE)
```

b) What is the mean body mass? What is the standard deviation? (2 points)

c) In your write-up, interpret the standard deviation of body mass in plain English. What does it tell us about how much penguin weights vary? *(Hint: The SD tells us roughly how far a typical observation is from the mean)* (2 points)

## A.4: Frequency Tables (6 points)

When working with categorical variables (like species), we want to count how many observations fall into each category.

**Run this code to count penguins by species:**

```{r}
#| eval: false
#| echo: true

# Count the number of penguins in each species
penguins |>
  count(species)
```

a) Which species has the most observations? How many? (2 points)

**Now let's create a cross-tabulation—counting by TWO variables at once:**

```{r}
#| eval: false
#| echo: true

# Count penguins by species AND island
penguins |>
  count(species, island)
```

b) Copy this output into your write-up. (2 points)

c) Looking at this table, are all three species found on all three islands? Which species are found on which islands? What does this tell us about the data? (2 points)

---

# Part B: Data Visualization (24 points)

Good visualizations reveal patterns that numbers alone might miss. We'll use `ggplot2` to create graphs.

## B.1: Histogram (5 points)

A histogram shows the distribution of a single numeric variable.

**Run this code to create a histogram of body mass:**

```{r}
#| eval: false
#| echo: true

# Create a histogram of penguin body mass
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Penguin Body Mass",
    x = "Body Mass (grams)",
    y = "Number of Penguins"
  ) +
  theme_minimal()
```

a) Copy this figure into your write-up. (2 points)

b) Describe the shape of the distribution in 2-3 sentences. Consider: (3 points)

   - Is it symmetric (roughly equal on both sides) or skewed (tail on one side)?
   - Is it unimodal (one peak) or bimodal (two peaks)?
   - What might explain the shape you see?

## B.2: Scatterplot (7 points)

A scatterplot shows the relationship between two numeric variables.

**Step 1: Run this code to create a basic scatterplot:**

```{r}
#| eval: false
#| echo: true

# Basic scatterplot: flipper length vs. body mass
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  labs(
    title = "Flipper Length vs. Body Mass",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

a) What pattern do you see? Is there a positive relationship (both increase together), negative relationship (one increases as the other decreases), or no clear relationship? (2 points)

**Step 2: Now let's add color to show the different species. Run this code:**

```{r}
#| eval: false
#| echo: true

# Scatterplot with points colored by species
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(
    title = "Flipper Length vs. Body Mass by Species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

b) Copy this colored scatterplot into your write-up. (2 points)

c) In 2-3 sentences, describe what you observe. Does the positive relationship hold within each species, or is it just an artifact of comparing different species? Which species tends to be largest? (3 points)

## B.3: Boxplot (4 points)

Boxplots are great for comparing distributions across groups.

**Run this code:**

```{r}
#| eval: false
#| echo: true

# Boxplot comparing body mass across species
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Body Mass by Penguin Species",
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Reading a Boxplot

- The **thick line in the middle** is the median (middle value)
- The **box** contains the middle 50% of the data
- The **whiskers** extend to show most of the remaining data
- **Dots beyond the whiskers** are potential outliers (unusual values)
:::

a) Copy this boxplot into your write-up. (1 point)

b) Which species has the highest median body mass? (1 point)

c) Do you see any outliers (dots beyond the whiskers)? If so, for which species? (2 points)

## B.4: Spotting Data Problems with a "Messy" Dataset (8 points)

One key purpose of visualization is to detect potential data errors. In this section, you'll work with a dataset that has some problems—just like real-world data often does!

**Run this code to create a dataset on student grades with some data entry errors:**

```{r}
#| eval: false
#| echo: true

# Create a "messy" dataset with some data problems
# This simulates what real data often looks like before cleaning

student_grades <- tibble(
  student_id = 1:25,
  hours_studied = c(2, 5, 3, 8, 1, 6, 4, 7, 2, 5,
                    3, 9, 4, 6, 50, 3, 7, 2, 8, 4,   # Note: 50 hours is suspicious!
                    5, 6, 3, 7, 4),
  exam_score = c(65, 78, 70, 92, 55, 85, 72, 88, 60, 80,
                 68, 95, 75, 82, 73, 71, 890, 62, 90, 74,  # Note: 890 is suspicious!
                 79, 83, 69, 86, 76),
  grade_level = c("Junior", "Senior", "Junior", "Senior", "Sophomore",
                  "Junior", "Senior", "Junior", "Sophomore", "Senior",
                  "Junior", "Senior", "Junior", "Senior", "Junior",
                  "Sophomore", "Senior", "Junior", "Senior", "Junior",
                  "Senior", "Junior", "Senior", "Junior", "17")  # Note: "17" should be a grade level!
)
```

**Step 1: Look at the summary statistics.**

```{r}
#| eval: false
#| echo: true

# Check summary statistics - do you notice anything unusual?
summary(student_grades)
```

a) Looking at the summary output, identify TWO variables that have suspicious values. What are the suspicious values, and why do they seem wrong? (2 points)

**Step 2: Create a histogram to visualize one of the problems.**

Now create a histogram of `hours_studied`. Use the code below as a template:

```{r}
#| eval: false
#| echo: true

# Create a histogram of hours studied
# YOUR CODE: Fill in the blanks

ggplot(student_grades, aes(x = __________)) +
  geom_histogram(binwidth = 5, fill = "coral", alpha = 0.7) +
  labs(
    title = "__________",      # Add a descriptive title
    x = "__________",          # Label the x-axis
    y = "Number of Students"
  ) +
  theme_minimal()
```

b) Copy your histogram into your write-up. Does the visualization help you spot the outlier? Where is it on the graph? (2 points)

**Step 3: Create a scatterplot to see the relationship (and the problem).**

```{r}
#| eval: false
#| echo: true

# Create a scatterplot of hours studied vs exam score
# YOUR CODE: Fill in the blanks

ggplot(student_grades, aes(x = hours_studied, y = __________)) +
  geom_point(size = 3, color = "darkblue") +
  labs(
    title = "__________",
    x = "Hours Studied",
    y = "__________"
  ) +
  theme_minimal()
```

c) Copy your scatterplot into your write-up. You should be able to see TWO obvious outliers—one on each axis. Circle or describe where they are. (2 points)

d) In 2-3 sentences, explain what you would do with these outliers before running a regression analysis. Should you automatically delete them? What questions would you want to answer first? (2 points)

::: {.callout-tip}
## Hints for the Fill-in-the-Blanks

- For the histogram, you want to visualize `hours_studied`
- For the scatterplot, the y-axis should be `exam_score`
- Titles should describe what the graph shows (e.g., "Distribution of Hours Studied")
- Axis labels should include units if applicable (e.g., "Exam Score (points)")
:::

---

# Part C: Documentation and Replication (10 points)

## C.1: Code Comments (4 points)

Good code includes comments that explain what the code does and *why*. This helps others (and your future self!) understand your work.

**Your task:** Add comments to each line of the following code block. Write the commented version in your R script.

```{r}
#| eval: false
#| echo: true

# Add a comment above EACH line explaining what it does
# Example: "# Remove rows where sex is missing"

penguins_clean <- penguins |>
  filter(!is.na(sex)) |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  select(species, island, body_mass_kg, sex)
```

::: {.callout-tip}
## Hints for Understanding This Code

- `filter()` keeps only rows that meet a condition
- `!is.na(sex)` means "sex is NOT missing"
- `mutate()` creates new variables
- `select()` chooses which variables to keep
- The `|>` symbol (called a "pipe") sends the result to the next line
:::

## C.2: Short Answer Questions (6 points)

**Answer the following questions in your write-up (2-3 sentences each):**

a) What is the "replication standard" in research? Why is it important for scientific credibility? (2 points)

b) Give two reasons why examining descriptive statistics and visualizations is important *before* running regression analysis. (2 points)

c) The chapter discussed the Reinhart-Rogoff case, where a graduate student found errors in influential economic research. What lesson does this teach us about research practices? (2 points)

---

# Submission Checklist

Before submitting, verify:

**R Script (`lastname_firstname_cc2.R`):**

- [ ] Header with your name, date, and assignment title
- [ ] All code runs without errors (try running the whole script from top to bottom)
- [ ] Code is organized with comments
- [ ] All required analyses are included

**Write-up (`lastname_firstname_cc2.pdf`):**

- [ ] All questions are answered with clear question numbers (A.1a, A.1b, etc.)
- [ ] All figures are included and labeled
- [ ] Written explanations are complete sentences
- [ ] Professional formatting

---

::: {.callout-tip}
## Final Tips

- **Run your code frequently!** Don't write 50 lines and then try to run them all. Run each piece as you go.
- **Read error messages carefully.** They often tell you exactly what's wrong.
- **Google is your friend.** If you're stuck, try searching for the error message.
- **Start early.** This gives you time to get help if you need it.
:::

*Good luck! Remember: The goal is not just to get the "right" answer, but to develop good habits for working with data that will serve you throughout your career.*
