---
title: "Binary Outcomes"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: today
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    embed-resources: true
    fig-align: center
    code-overflow: wrap
    footer: "EC325 | Unit 10 | Binary Outcomes"
    scrollable: true
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(tidyverse)
library(wooldridge)
library(fixest)
library(modelsummary)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 16) +
          theme(plot.title = element_text(face = "bold", size = 18),
                plot.subtitle = element_text(size = 13, color = "gray40"),
                panel.grid.minor = element_blank()))
```


## Where We Are

- We've been building a regression toolkit all semester
- Started with bivariate OLS, added multiple regression, statistical inference
- Incorporated **qualitative information on the right-hand side**: dummy variables for categories, interaction effects for differential slopes
- Made our models more flexible: polynomials, logs, robust SEs

. . .

Today: what happens when the **left-hand side** --- the dependent variable --- is qualitative?


## Binary Outcomes Are Everywhere

| Field | Outcome | $y = 1$ | $y = 0$ |
|:------|:--------|:--------|:--------|
| Labor | Employed? | Working | Not working |
| Health | Low birthweight? | < 2,500g | $\geq$ 2,500g |
| Education | Graduated? | Yes | No |
| Finance | Defaulted? | Yes | No |
| Policy | Enrolled in program? | Yes | No |

. . .

These are all **binary** (or **dummy**) dependent variables: $y$ can only be 0 or 1.


## The Key Insight (1/2)

When $y$ is 0 or 1, what does $E(y \mid \mathbf{x})$ mean?

. . .

Recall the formula for expected value. When $y$ takes only two values:

$$E(y \mid \mathbf{x}) = \sum_{\text{all values}} [\text{value} \times P(\text{value})]$$

. . .

$$= 1 \times P(y = 1 \mid \mathbf{x}) \;+\; 0 \times P(y = 0 \mid \mathbf{x})$$

. . .

$$= P(y = 1 \mid \mathbf{x})$$


## The Key Insight (2/2)

So we have:

$$E(y \mid \mathbf{x}) = P(y = 1 \mid \mathbf{x})$$

. . .

Why does this matter? Recall that OLS estimates the conditional expectation $E(y \mid \mathbf{x})$.

. . .

So when $y$ is binary, OLS is estimating a **probability**.

. . .

::: {.highlight-box}
With a binary dependent variable, $\hat{y}$ is the **predicted probability** that $y = 1$, given the values of $\mathbf{x}$.
:::


## The Linear Probability Model

Just run OLS with a binary $y$:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + \mu$$

. . .

This is called the **linear probability model** (LPM).

. . .

Since $\hat{y} = \hat{P}(y = 1 \mid \mathbf{x})$, the coefficients have a special interpretation:

::: {.highlight-box}
$\hat{\beta}_j$ = the change in the **probability** of $y = 1$ for a one-unit increase in $x_j$, holding all else constant.
:::


## Interpretation: Percentage Points!

::: {.callout-warning}
## Be Careful with Language
If $\hat{\beta}_1 = -0.15$, we say:

"A one-unit increase in $x_1$ **decreases the probability** of $y = 1$ by **15 percentage points**."

**Not** "15 percent"! Percentage points and percent are different things.
:::

. . .

- Suppose $P(y = 1) = 0.60$ and $\hat{\beta}_1 = -0.15$:
  - New probability: $0.60 - 0.15 = 0.45$
  - That's a 15 **percentage point** decrease
  - But it's a 25 **percent** decrease (since $0.15 / 0.60 = 0.25$)


## Example: Women's Labor Force Participation

```{r}
#| label: data-setup

lf_data <- wooldridge::mroz
```

The `mroz` dataset from `wooldridge`: **753 married women** in 1975.

- `inlf` = 1 if in the labor force, 0 otherwise
- `kidslt6` = number of children under age 6
- `kidsge6` = number of children aged 6--18
- `nwifeinc` = husband's income (thousands of \$)
- `educ` = woman's years of education

. . .

**Question:** what factors affect whether a married woman participates in the labor force?


## The Raw Data

```{r}
#| label: fig-inlf-bar
#| fig-width: 7
#| fig-height: 4

ggplot(lf_data, aes(x = factor(inlf), fill = factor(inlf))) +
    geom_bar(width = 0.5) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 5) +
    scale_fill_manual(values = c("0" = "#D55E00", "1" = "steelblue")) +
    scale_x_discrete(labels = c("0" = "Not in LF (y = 0)", "1" = "In LF (y = 1)")) +
    labs(x = NULL, y = "Count",
         title = "Labor Force Participation Among Married Women") +
    theme(legend.position = "none")
```

About 57% of women in the sample are in the labor force. Unlike continuous outcomes, there are no in-between values.


## Raw Relationship: Young Children and LFP

```{r}
#| label: fig-kids-lfp
#| fig-width: 8
#| fig-height: 3.5

lf_data |>
    group_by(kidslt6) |>
    summarise(pct_inlf = mean(inlf), n = n()) |>
    ggplot(aes(x = factor(kidslt6), y = pct_inlf)) +
    geom_col(fill = "steelblue", width = 0.5) +
    geom_text(aes(label = paste0(round(pct_inlf * 100, 1), "%\n(n=", n, ")")),
              vjust = -0.3, size = 4) +
    scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.8)) +
    labs(x = "Number of Children Under Age 6",
         y = "Share in Labor Force",
         title = "Labor Force Participation by Number of Young Children") +
    theme(plot.title = element_text(hjust = 0.5))
```

Clear pattern: more young children $\rightarrow$ lower LFP. Let's see what the LPM tells us.


## Estimating the LPM

```{r}
#| echo: true

reg_lpm <- lm(inlf ~ kidslt6 + kidsge6 + nwifeinc + educ,
              data = lf_data)
summary(reg_lpm)
```


## Interpreting the Coefficients {.smaller}

::: {.columns}
::: {.column width="50%"}
### `kidslt6`: $\hat{\beta}_1 \approx -0.19$

Each additional child under 6 **decreases** the probability of working by about 19 percentage points.

### `educ`: $\hat{\beta}_4 \approx 0.06$

Each additional year of education **increases** the probability of working by about 4 percentage points.
:::

::: {.column width="50%"}
### `kidsge6`: $\hat{\beta}_2 \approx 0.01$

Older children have little effect --- small and not significant.

### `nwifeinc`: $\hat{\beta}_3 \approx -0.008$

Each additional \$1,000 of husband's income reduces probability by about 0.5 pp.
:::
:::


## Two Problems with the LPM

The LPM is simple and intuitive, but it has two important drawbacks:

. . .

1. **Built-in heteroskedasticity**: when $y$ is binary, the error variance *always* depends on $\mathbf{x}$, violating Gauss-Markov assumption 5. (We already know how to fix this!)

. . .

2. **Out-of-bounds predictions**: the LPM is a straight line, so it can produce predicted probabilities **less than 0** or **greater than 1** --- which are impossible.

. . .

Let's look at each one.


## Problem 1: Built-In Heteroskedasticity {.smaller}

When $y$ is binary, the error $\mu = y - E(y \mid \mathbf{x})$ can only take **two possible values** for a given $\mathbf{x}$:

. . .

- If $y = 1$: $\;\mu = 1 - p(\mathbf{x})$, which happens with probability $p(\mathbf{x})$
- If $y = 0$: $\;\mu = 0 - p(\mathbf{x}) = -p(\mathbf{x})$, which happens with probability $1 - p(\mathbf{x})$

. . .

where $p(\mathbf{x}) = E(y \mid \mathbf{x})$ is the predicted probability.

. . .

The variance of this error is:

$$\text{Var}(\mu \mid \mathbf{x}) = p(\mathbf{x})\big[1 - p(\mathbf{x})\big]^2 + \big[1 - p(\mathbf{x})\big]\big[p(\mathbf{x})\big]^2 = p(\mathbf{x})\big(1 - p(\mathbf{x})\big)$$

. . .

This depends on $p(\mathbf{x})$, which depends on $\mathbf{x}$. **The error variance changes with $\mathbf{x}$** --- that's heteroskedasticity, by definition!


## Visualizing LPM Heteroskedasticity

```{r}
#| label: fig-het-residuals
#| fig-width: 9
#| fig-height: 4

set.seed(42)
n_sim <- 200
x_sim <- runif(n_sim, -3, 3)
p_true <- plogis(-1 + 1.5 * x_sim)
y_sim <- rbinom(n_sim, 1, p_true)
sim_df <- tibble(x = x_sim, y = y_sim)
lpm_sim <- lm(y ~ x, data = sim_df)
sim_df$fitted <- fitted(lpm_sim)

ggplot(sim_df, aes(x = x, y = y)) +
    geom_segment(aes(xend = x, yend = fitted), color = "gray60", alpha = 0.4) +
    geom_point(aes(color = factor(y)), alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1, linetype = "dashed") +
    scale_color_manual(values = c("0" = "steelblue", "1" = "#D55E00"),
                       labels = c("y = 0", "y = 1"), name = NULL) +
    annotate("label", x = -2.2, y = 0.6, size = 3.5, fill = "white", label = "Low p̂ → small residuals") +
    annotate("label", x = 0.8, y = -0.18, size = 3.5, fill = "white", label = "p̂ ≈ 0.5 → large residuals") +
    annotate("label", x = 2.5, y = 0.4, size = 3.5, fill = "white", label = "High p̂ → small residuals") +
    labs(x = "x", y = "y (0 or 1)",
         title = "Residuals Fan Out in the Middle",
         subtitle = "Residual spread is largest where p̂ ≈ 0.5 and smallest near 0 or 1") +
    theme(legend.position = "bottom")
```


## The Fix: Robust Standard Errors

We already know how to handle this from our heteroskedasticity unit!

. . .

Just use `feols()` with `vcov = "HC1"`:

```{r}
#| echo: true

# Standard OLS SEs (wrong under heteroskedasticity)
lpm_default <- feols(inlf ~ kidslt6 + kidsge6 + nwifeinc + educ,
                     data = lf_data)

# Robust SEs (correct)
lpm_robust <- feols(inlf ~ kidslt6 + kidsge6 + nwifeinc + educ,
                    vcov = "HC1", data = lf_data)
```

## The LPM in Practice

```{r}
#| label: fig-het-variance
#| fig-width: 7
#| fig-height: 4

p_seq <- seq(0.01, 0.99, by = 0.01)
ggplot(tibble(p = p_seq, v = p_seq * (1 - p_seq)), aes(x = p, y = v)) +
    geom_line(color = "steelblue", linewidth = 1.2) +
    geom_point(data = tibble(p = 0.5, v = 0.25), size = 3, color = "#D55E00") +
    annotate("text", x = 0.58, y = 0.255, label = "Max at p = 0.5",
             color = "#D55E00", size = 4) +
    labs(x = "Predicted Probability p(x)", y = "Var(error) = p(1 - p)",
         title = "Error Variance in the LPM",
         subtitle = "Variance is largest at p = 0.5 and shrinks toward 0 and 1")
```

- Problem 1 (heteroskedasticity): ✅ solved with robust SEs
- Problem 2 (out-of-bounds): still a concern --- let's look at it now


## Problem 2: Out-of-Bounds Predictions

```{r}
#| label: fig-oob
#| fig-width: 9
#| fig-height: 4

lf_data <- lf_data |>
    mutate(pred_lpm = predict(reg_lpm))

ggplot(lf_data, aes(x = pred_lpm)) +
    geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 0.8) +
    annotate("rect", xmin = -0.35, xmax = 0, ymin = 0, ymax = Inf,
             alpha = 0.15, fill = "red") +
    labs(x = "Predicted Probability (LPM)", y = "Count",
         title = "LPM Predicted Probabilities",
         subtitle = "Red region: impossible probabilities below zero!")
```

The LPM is a **straight line** --- for extreme enough values of $\mathbf{x}$, it will cross the 0 or 1 boundary. No easy fix for this one --- we need a different model.


## An Alternative: Logistic/Logit Regression

What if instead of a straight line, we used a function that **keeps predictions between 0 and 1**?

$$P(y = 1 \mid \mathbf{x}) = G(\beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k)$$

. . .

where $G(z) = \frac{\exp(z)}{1 + \exp(z)}$ is the **logistic function**.

. . .

The idea: wrap the linear index $\beta_0 + \beta_1 x_1 + \cdots$ inside $G(\cdot)$, which maps any real number into the interval $(0, 1)$.


## The Logistic Function

```{r}
#| label: fig-logistic
#| fig-width: 9
#| fig-height: 4

z_seq <- seq(-6, 6, by = 0.1)
ggplot(tibble(z = z_seq, G = plogis(z_seq)), aes(x = z, y = G)) +
    geom_line(color = "steelblue", linewidth = 1.4) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray70") +
    geom_point(data = tibble(z = 0, G = 0.5), size = 3, color = "#D55E00") +
    annotate("text", x = 1, y = 0.54, label = "G(0) = 0.5", color = "#D55E00", size = 5) +
    annotate("text", x = -4.5, y = 0.15, label = "Approaches 0\nbut never reaches it",
             color = "gray40", size = 3.5, fontface = "italic") +
    annotate("text", x = 4.5, y = 0.85, label = "Approaches 1\nbut never reaches it",
             color = "gray40", size = 3.5, fontface = "italic") +
    annotate("segment", x = -1.5, xend = 1.5, y = 0.25, yend = 0.75,
             color = "#009E73", linewidth = 0.8, linetype = "dashed") +
    annotate("text", x = 2.2, y = 0.4, label = "Nearly linear\nin the middle",
             color = "#009E73", size = 3.5, fontface = "italic") +
    labs(x = "z (linear index)", y = "G(z)",
         title = "The Logistic (Sigmoid) Function",
         subtitle = "Always between 0 and 1 --- no impossible predictions!")
```


## Key Properties {.smaller}

- **Always between 0 and 1** --- no out-of-bounds predictions
- **S-shaped**: nearly linear in the middle, curves at extremes
- **Symmetric** around $z = 0$ (where $G = 0.5$)

. . .

Because $G(\cdot)$ is nonlinear, we can't use OLS. Instead: **Maximum Likelihood Estimation** (MLE).

. . .

::: {.highlight-box}
**OLS** minimizes $\sum(y_i - \hat{y}_i)^2$.

**MLE** finds coefficients that make the observed data **most probable**: high $\hat{p}_i$ when $y_i = 1$, low $\hat{p}_i$ when $y_i = 0$.
:::


## Estimating Logit in R {.smaller}

Just like switching from `lm()` to `feols()`, we use `feglm()` from `fixest`:

```{r}
#| echo: true

logit_reg <- feglm(inlf ~ kidslt6 + kidsge6 + nwifeinc + educ,
                   family = binomial(link = "logit"),
                   data = lf_data)
summary(logit_reg)
```

. . .

- The key argument is `family = binomial(link = "logit")`
- This tells `feglm()` to use the logistic formula in the regression

## Side-by-Side Comparison {.smaller}

::: {.columns}
::: {.column width="55%"}
```{r}
#| echo: true

modelsummary(
  list("LPM" = reg_lpm,
       "Logit" = logit_reg),
  stars = c('*' = 0.1,
            '**' = 0.05,
            '***' = 0.01),
  gof_map = c("nobs",
              "r.squared",
              "logLik"))
```
:::

::: {.column width="45%"}
- **Signs** are the same across both models --- they agree on directions
- But **magnitudes** differ: logit coefficients are on a **different scale** (log-odds)
- How do we interpret them?
:::
:::


## Interpreting Logit: The Problem

In the LPM, $\hat{\beta}_j$ is the change in **probability** --- simple and direct.

In logit, $\hat{\beta}_j$ is the change in the **log-odds** --- not directly intuitive.

. . .

Three ways to make sense of logit coefficients:

1. **Log-odds** (direct, but abstract)
2. **Odds ratios** (exponentiate the coefficient)
3. **Marginal effects** (most practical --- directly comparable to LPM)

. . .

Let's walk through all three.


## Interpretation 1: Log-Odds {.smaller}

The **odds** of an event are: $\;\text{Odds} = \frac{P(y = 1)}{1 - P(y = 1)}$

. . .

- If $P = 0.75$: odds = $0.75 / 0.25 = 3$ ("3 to 1 in favor")
- If $P = 0.50$: odds = $0.50 / 0.50 = 1$ ("even odds")
- If $P = 0.20$: odds = $0.20 / 0.80 = 0.25$ ("4 to 1 against")

. . .

The logit model is linear in the **log-odds**:

$$\ln\left(\frac{P(y = 1 \mid \mathbf{x})}{1 - P(y = 1 \mid \mathbf{x})}\right) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$

. . .

So $\hat{\beta}_j$ is the change in **log-odds** for a one-unit increase in $x_j$.

. . .

::: {.caution-box}
This is mathematically clean, but not very intuitive --- most people don't think in log-odds. The main practical use: **positive** coefficients increase the probability; **negative** coefficients decrease it.
:::


## Interpretation 2: Odds Ratios {.smaller}

Exponentiating the logit coefficient gives the **odds ratio**: $\exp(\hat{\beta}_j)$.

```{r}
#| label: fig-odds-ratios
#| fig-width: 8
#| fig-height: 3.5

logit_coefs <- coef(logit_reg)[-1]
logit_ses <- se(logit_reg)[-1]

or_data <- tibble(
    variable = c("Kids < 6", "Kids 6-18", "Non-wife Income", "Education"),
    odds_ratio = exp(logit_coefs),
    lower = exp(logit_coefs - 1.96 * logit_ses),
    upper = exp(logit_coefs + 1.96 * logit_ses)
)

ggplot(or_data, aes(x = odds_ratio, y = reorder(variable, odds_ratio))) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
    geom_pointrange(aes(xmin = lower, xmax = upper), color = "steelblue", size = 0.8) +
    labs(x = "Odds Ratio (with 95% CI)", y = NULL,
         title = "Odds Ratios for Labor Force Participation",
         subtitle = "Values < 1 reduce the odds; values > 1 increase the odds")
```

- **Kids < 6**: OR $\approx$ 0.42 --- each young child **cuts** odds of working about 0.40
- **Education**: OR $\approx$ 1.22 --- each year **increases** odds by about a quarter


## Interpretation 3: Marginal Effects {.smaller}

Unlike the LPM, the logit's effect on probability **depends on where you start**.

. . .

To compute the marginal effect of a variable:

1. **Pick specific values** for all the $x$ variables
2. Compute the **predicted probability** at those values
3. **Increase the variable of interest by 1**, keeping everything else the same
4. Compute the **new predicted probability**
5. **Subtract**: the difference is the marginal effect

. . .

::: {.highlight-box}
Marginal effect = $\hat{P}(y = 1 \mid x_j + 1) \;-\; \hat{P}(y = 1 \mid x_j)$
:::

. . .

Let's walk through a concrete example.


## A Quick Note on `predict()` {.smaller}

R's `predict()` function lets us compute $\hat{y}$ automatically for any values of $\mathbf{x}$.


Instead of manually plugging numbers into the estimated equation, we pass `predict()` a **fitted model** and a `newdata` data frame with the $x$ values we want:

```{r}
#| echo: true
#| eval: false

predict(my_model, newdata = data.frame(x1 = 5, x2 = 10), type = "response")
```

. . .

- The column names in `newdata` must **exactly match** the variable names in the model
- For logit models, `type = "response"` tells R to return a **predicted probability** (not the log-odds)
- Without `type = "response"`, `predict()` returns the raw linear index $\hat{\beta}_0 + \hat{\beta}_1 x_1 + \cdots$


## Marginal Effect: A Specific Woman {.smaller}

Consider a specific woman: **0 kids under 6**, 1 older kid, husband's income \$20k, 12 years of education.

What is the effect of her having **one young child**?

. . .

```{r}
#| echo: true

# Step 1: Predicted probability with 0 kids under 6
prob_0kids <- predict(logit_reg, type = "response",
                      newdata = data.frame(kidslt6 = 0, kidsge6 = 1,
                                           nwifeinc = 20, educ = 12))

# Step 2: Predicted probability with 1 kid under 6 (everything else the same)
prob_1kid <- predict(logit_reg, type = "response",
                     newdata = data.frame(kidslt6 = 1, kidsge6 = 1,
                                          nwifeinc = 20, educ = 12))

cat("P(working | 0 kids < 6):", round(prob_0kids, 4), "\n")
cat("P(working | 1 kid < 6): ", round(prob_1kid, 4), "\n")
cat("Marginal effect:        ", round(prob_1kid - prob_0kids, 4), "\n")
```

. . .

For this particular woman, one young child decreases the probability of working by about **26 percentage points**.


## A Different Woman, A Different ME {.smaller}

Now consider a woman with the same characteristics but **husband's income of \$50k**:

```{r}
#| echo: true

prob_0kids_rich <- predict(logit_reg, type = "response",
                           newdata = data.frame(kidslt6 = 0, kidsge6 = 1,
                                                nwifeinc = 50, educ = 12))
prob_1kid_rich <- predict(logit_reg, type = "response",
                          newdata = data.frame(kidslt6 = 1, kidsge6 = 1,
                                               nwifeinc = 50, educ = 12))

cat("P(working | 0 kids < 6):", round(prob_0kids_rich, 4), "\n")
cat("P(working | 1 kid < 6): ", round(prob_1kid_rich, 4), "\n")
cat("Marginal effect:        ", round(prob_1kid_rich - prob_0kids_rich, 4), "\n")
```

. . .

- The marginal effect is **different** because this woman starts at a different predicted probability. 
- Just like all non-linear models, the ME depends on the values of $x_j$

## Marginal Effect at the Mean {.smaller}

- How do we know which $x$ values to report ME at?
- The standard approach: evaluate at the **mean** of all explanatory variables.

. . .

```{r}
#| echo: true

# Compute means of each variable
mean_data <- lf_data |>
    summarize(kidslt6 = mean(kidslt6), kidsge6 = mean(kidsge6),
              nwifeinc = mean(nwifeinc), educ = mean(educ))
mean_data
```

. . .

```{r}
#| echo: true

# Probability at the mean
prob_at_mean <- predict(logit_reg, type = "response", newdata = mean_data)

# Probability with one more kid under 6
mean_data_plus1 <- mean_data |> mutate(kidslt6 = kidslt6 + 1)
prob_plus1 <- predict(logit_reg, type = "response", newdata = mean_data_plus1)

marginal_effect <- prob_plus1 - prob_at_mean
cat("Marginal effect at the mean:", round(marginal_effect, 4), "\n")
```

## How Does the Logit ME Compare to the LPM? {.smaller}

```{r}
#| echo: true

cat("Logit ME at mean:", round(marginal_effect, 4), "\n")
cat("LPM coefficient:  ", round(coef(reg_lpm)["kidslt6"], 4), "\n")
```

. . .

They're **very similar**!

. . .

This is not a coincidence. Look back at the logistic function: it is nearly **linear** in the middle of its range. For observations near $p = 0.5$, the straight line (LPM) and the S-curve (logit) give approximately the same slope.

::: {.highlight-box}
Near the center of the probability distribution, the LPM and logit produce **similar marginal effects**. They diverge mainly at the tails.
:::


## LPM vs. Logit: Predictions Compared

```{r}
#| label: fig-lpm-vs-logit
#| fig-width: 9
#| fig-height: 4

lf_data <- lf_data |>
    mutate(
        logit_index = predict(logit_reg, type = "link"),
        logit_pred = predict(logit_reg, type = "response")
    )

lpm_on_index <- lm(pred_lpm ~ logit_index, data = lf_data)

index_grid <- tibble(logit_index = seq(min(lf_data$logit_index) - 1,
                                        max(lf_data$logit_index) + 1, by = 0.05)) |>
    mutate(
        logit_pred = plogis(logit_index),
        lpm_pred = predict(lpm_on_index, newdata = pick(everything()))
    )

pred_long <- index_grid |>
    pivot_longer(cols = c(lpm_pred, logit_pred), names_to = "model", values_to = "predicted") |>
    mutate(model = ifelse(model == "lpm_pred", "LPM", "Logit"))

ggplot() +
    geom_jitter(data = lf_data, aes(x = logit_index, y = inlf),
                alpha = 0.1, height = 0.03, color = "gray50", size = 0.8) +
    geom_line(data = pred_long |> filter(model == "LPM"),
              aes(x = logit_index, y = predicted, color = "LPM"),
              linewidth = 1, linetype = "dashed") +
    geom_line(data = pred_long |> filter(model == "Logit"),
              aes(x = logit_index, y = predicted, color = "Logit"),
              linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "red", alpha = 0.5) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "red", alpha = 0.5) +
    scale_color_manual(values = c("LPM" = "steelblue", "Logit" = "#D55E00")) +
    coord_cartesian(ylim = c(-0.3, 1.3)) +
    labs(x = "Linear Index", y = "Predicted Probability",
         color = "Model",
         title = "LPM vs. Logit Predictions",
         subtitle = "They agree closely in the middle but diverge at the tails") +
    theme(legend.position = "bottom")
```


## Rounding Up: LPM vs. Logit {.smaller}

::: {.columns}
::: {.column width="50%"}
### Linear Probability Model

- Models probability as a **straight line**: $P(y = 1) = \mathbf{x}\boldsymbol{\beta}$
- Estimated by **OLS**
- Coefficients = direct change in probability
- Can predict outside [0, 1]
- Always heteroskedastic (use robust SEs)
:::

::: {.column width="50%"}
### Logistic Regression

- Models probability through the **S-curve**: $P(y = 1) = G(\mathbf{x}\boldsymbol{\beta})$
- Estimated by **MLE**
- Coefficients = change in log-odds (convert via OR or ME)
- Predictions always in (0, 1)
- Heteroskedasticity handled by design
:::
:::

. . .

::: {.highlight-box}
**The punchline:** despite these differences, the two models typically produce **very similar estimated effects** across most of the distribution of $\mathbf{x}$. They only diverge meaningfully at the tails, where predicted probabilities are near 0 or 1. For the vast majority of observations in a typical dataset, you will get essentially the same answer from either approach.
:::


## Practical Advice {.smaller}

For causal analysis...

. . .

- **In economics**: the LPM is the workhorse. Most applied micro papers use it as the primary specification, sometimes reporting logit as a robustness check.

- **In health sciences/comp sci/other fields**: logistic regression is the standard. Odds ratios are commonly references lingua franca.

. . .

But for Prediction...

- logit wins across fields --- its predictions are always valid probabilities.


## When to Use Which? {.smaller}

| | LPM | Logistic Regression |
|:---|:---|:---|
| **Estimation** | OLS | Maximum Likelihood |
| **Coefficients** | Change in probability | Change in log-odds |
| **Interpretation** | Direct and intuitive | Requires conversion (OR or ME) |
| **Predictions** | Can fall outside [0, 1] | Always in (0, 1) |
| **Heteroskedasticity** | Always present (use robust SEs) | Handled by MLE |
| **Common in** | Economics | Health sciences, epidemiology |
| **Best for** | Causal inference | Prediction |


## Summary {.smaller}

1. When $y$ is binary, $E(y \mid \mathbf{x}) = P(y = 1 \mid \mathbf{x})$
2. **LPM**: run OLS; $\hat{\beta}_j$ = change in probability (percentage points)
3. LPM problems: out-of-bounds predictions and always heteroskedastic (use robust SEs!)
4. **Logit**: wraps predictions in the logistic function; estimated by MLE
5. Logit coefficients = log-odds; convert via odds ratios or marginal effects
6. **Marginal effects at the mean** are the most practical way to compare logit to LPM
7. Near the center of the distribution, the two models typically **agree closely**


# Try It Yourself

## Try It Yourself: Setup {.smaller}

The `bwght` dataset in `wooldridge` has data on infant birthweight. Create a binary variable `low_bwght` equal to 1 if `bwght < 88` (below 2,500 grams).

. . .

```{r}
#| echo: true
#| eval: true

bw_data <- wooldridge::bwght |>
  mutate(low_bwght = ifelse(bwght < 88, 1, 0))
```

- We'll use this dataset for the next three questions.

## Question 1: Estimate Two Models {.smaller}

Estimate an LPM (with robust SEs) and a logit model of the form:

$$low\_bwght = \beta_0 + \beta_1 \cdot cigs + \beta_2 \cdot faminc + \beta_3 \cdot motheduc + \mu$$

. . .

```{r}
#| echo: true
#| eval: true

lpm_bw  <- feols(low_bwght ~ cigs + faminc + motheduc,
                 vcov = "HC1", data = bw_data)

logit_bw <- feglm(low_bwght ~ cigs + faminc + motheduc,
                  family = binomial(link = "logit"), data = bw_data)
```

## Question 2: Interpret the LPM {.smaller}

Interpret the LPM coefficient on `cigs` in a sentence.

```{r}
#| echo: true
#| eval: true

summary(lpm_bw)

```

. . .

**Answer (template):**

> "Holding family income and mother's education constant, smoking one additional cigarette per day during pregnancy is associated with a $0.3$ percentage-point change in the probability that the infant is born with low birthweight."

- Remember: in the LPM, coefficients are interpreted in **percentage points** of probability.

## Question 3: Marginal Effect from Logit {.smaller}

Using `predict()`, compute the marginal effect of 1 additional cigarette for a woman at the **mean of all variables** using the logistic regression estimates. How does it compare to the LPM coefficient?

. . .

```{r}
#| echo: true
#| eval: true

# Average woman in the sample
avg_woman <- bw_data |>
  summarize(cigs = mean(cigs, na.rm = TRUE),
            faminc = mean(faminc, na.rm = TRUE),
            motheduc = mean(motheduc, na.rm = TRUE))

# Predicted probability at the mean, and at +1 cigarette
p0 <- predict(logit_bw, newdata = avg_woman)
p1 <- predict(logit_bw, newdata = avg_woman |> mutate(cigs = cigs + 1))

# Marginal effect of one more cigarette at the mean
p1 - p0
```

- Compare this number to $\hat{\beta_1}$ from the LPM — near the center of the distribution, the two should be very close.


