---
title: "Heteroskedasticity"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: today
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    embed-resources: true
    fig-align: center
    code-overflow: wrap
    footer: "EC325 | Unit 9 | Heteroskedasticity"
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(tidyverse)
library(lmtest)
library(sandwich)
library(fixest)
library(modelsummary)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 16) +
          theme(plot.title = element_text(face = "bold", size = 18),
                plot.subtitle = element_text(size = 13, color = "gray40"),
                panel.grid.minor = element_blank()))
```


## Where We Are

- Have been covering ways to estimate more flexible models
- Include qualitative information: dummy variables
- Allow for non-linear relationships: polynomial and logarithmic transformation of variables
- Allow for differential slopes by group: interaction effects
- How about more flexible ways of doing *statistical inference*?

## Previous Assumptions

- GM assumptions 1--4 guarantee OLS is **unbiased**
- GM assumption 5 (**homoskedasticity**) gives us valid standard errors
- Today: what happens when GM5 fails, and what to do about it

. . .

::: {.callout-note}
## Gauss-Markov Assumption 5 (Homoskedasticity)
$$Var(\mu \mid x_1, x_2, \ldots, x_k) = \sigma^2$$

The error variance is **constant**---it doesn't depend on $x$.
:::


## Homoskedastic Data

```{r}
#| label: fig-homo-scatter
#| fig-width: 9
#| fig-height: 5

set.seed(123)
n <- 200

homo_data <- tibble(
    x = runif(n, 0, 100),
    error = rnorm(n, 0, 15),
    y = 10 + 0.5 * x + error
)

ggplot(homo_data, aes(x = x, y = y)) +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    geom_ribbon(
        data = tibble(x = seq(0, 100, length.out = 100)) |>
            mutate(yhat = 10 + 0.5 * x),
        aes(x = x, ymin = yhat - 30, ymax = yhat + 30),
        alpha = 0.12, fill = "darkred",
        inherit.aes = FALSE
    ) +
    labs(title = "Homoskedastic Data",
         subtitle = "Constant spread around the regression line",
         x = "X", y = "Y") +
    annotate("text", x = 82, y = 15, label = "Constant\nspread",
             color = "darkred", size = 5, fontface = "bold")
```

The "band" of data has roughly the **same width** everywhere.


## Heteroskedastic Data

```{r}
#| label: fig-hetero-scatter
#| fig-width: 9
#| fig-height: 5

set.seed(123)

hetero_data <- tibble(
    x = runif(n, 0, 100),
    error = rnorm(n, 0, 0.3 * x),
    y = 10 + 0.5 * x + error
)

ggplot(hetero_data, aes(x = x, y = y)) +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    geom_ribbon(
        data = tibble(x = seq(1, 100, length.out = 100)) |>
            mutate(yhat = 10 + 0.5 * x, spread = 0.6 * x),
        aes(x = x, ymin = yhat - spread, ymax = yhat + spread),
        alpha = 0.12, fill = "darkred",
        inherit.aes = FALSE
    ) +
    labs(title = "Heteroskedastic Data",
         subtitle = 'The spread "fans out" as x increases',
         x = "X", y = "Y") +
    annotate("text", x = 85, y = 18, label = "Spread\nincreases",
             color = "darkred", size = 5, fontface = "bold")
```

**Heteroskedasticity** = the error variance changes with $x$.


## Side-by-Side Comparison

```{r}
#| label: fig-side-by-side
#| fig-width: 10
#| fig-height: 4.5

bind_rows(
    homo_data |> mutate(type = "Homoskedastic"),
    hetero_data |> mutate(type = "Heteroskedastic")
) |>
    mutate(type = factor(type, levels = c("Homoskedastic", "Heteroskedastic"))) |>
    ggplot(aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    facet_wrap(~type, scales = "free_y") +
    labs(title = "Homoskedastic vs. Heteroskedastic",
         x = "X", y = "Y")
```

- Same regression line in both panels
- The difference is in the **spread** of points around the line


## Where Does Heteroskedasticity Come From?

Common in economic data:

- **Income and consumption:** higher-income households have more variable spending
- **Firm size and profits:** larger firms have more variable profit margins
- **Experience and wages:** more experienced workers have more variable wages

. . .

General pattern: when the **scale** of outcomes increases with an explanatory variable, expect heteroskedasticity.


## What Breaks? {.smaller}

::: {.columns}
::: {.column width="50%"}
### Still Fine

- OLS coefficients remain **unbiased**
- $E[\hat{\beta}] = \beta$ depends only on GM 1--4
- Point estimates are OK
:::

::: {.column width="50%"}
### Broken

- Standard error formula assumes constant $\sigma^2$
- Computed SEs are typically **too small**
- So t-statistics are **too large**
- And p-values are **too small**
:::
:::

. . .

**We reject the null hypothesis too often**, even when it is true.


## Type I Error: Quick Review

::: {.callout-important}
## Type I Error (False Positive)
Rejecting a **true** null hypothesis.

At $\alpha = 0.05$, we accept a 5% chance of this happening. If our test is working, we should reject a true $H_0$ exactly 5% of the time.
:::

. . .

With heteroskedasticity and standard OLS SEs, the actual Type I error rate can be **much higher** than 5%.


## Simulation Setup

We simulate data where the **true effect is zero** ($\beta_1 = 0$) and check how often we falsely reject.

```{r}
#| echo: true

# Simulation parameters
n_sims <- 2000       # Number of simulations
n_obs <- 100         # Observations per sample
beta0_true <- 5      # True intercept
beta1_true <- 0      # True slope (NULL IS TRUE!)
```

. . .

Run 2,000 regressions under each scenario:

- **Homoskedastic**: $\text{error} \sim N(0, 10)$
- **Heteroskedastic**: $\text{error} \sim N(0,\; x^2)$

```{r}
#| label: run-simulations

set.seed(42)

# Homoskedastic simulations
run_homo_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    error <- rnorm(n_obs, mean = 0, sd = 10)
    y <- beta0_true + beta1_true * x + error
    model <- lm(y ~ x)
    beta1_hat <- coef(model)["x"]
    se_ols <- summary(model)$coefficients["x", "Std. Error"]
    pval_ols <- summary(model)$coefficients["x", "Pr(>|t|)"]
    data.frame(beta1_hat = beta1_hat, se_ols = se_ols, pval_ols = pval_ols)
}

homo_results <- map_dfr(1:n_sims, run_homo_trial)
homo_reject_rate <- mean(homo_results$pval_ols < 0.05)
homo_avg_se <- mean(homo_results$se_ols)
homo_true_sd <- sd(homo_results$beta1_hat)

# Heteroskedastic simulations
run_hetero_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    error <- rnorm(n_obs, mean = 0, sd = 0.15 * x^2)
    y <- beta0_true + beta1_true * x + error
    model <- lm(y ~ x)
    beta1_hat <- coef(model)["x"]
    res_ols <- coeftest(model)
    res_robust <- coeftest(model, vcov. = vcovHC(model, type = "HC1"))
    data.frame(
        beta1_hat = beta1_hat,
        se_ols = res_ols["x", "Std. Error"],
        se_robust = res_robust["x", "Std. Error"],
        pval_ols = res_ols["x", "Pr(>|t|)"],
        pval_robust = res_robust["x", "Pr(>|t|)"]
    )
}

hetero_results <- map_dfr(1:n_sims, run_hetero_trial)
ols_reject_rate <- mean(hetero_results$pval_ols < 0.05)
robust_reject_rate <- mean(hetero_results$pval_robust < 0.05)
hetero_avg_se <- mean(hetero_results$se_ols)
hetero_true_sd <- sd(hetero_results$beta1_hat)
avg_se_robust <- mean(hetero_results$se_robust)
```


## Baseline: Homoskedasticity Works

```{r}
#| label: fig-homo-dist
#| fig-width: 9
#| fig-height: 5

ggplot(homo_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 35,
                   fill = "steelblue", alpha = 0.5, color = "white") +
    stat_function(fun = dnorm, args = list(mean = 0, sd = homo_avg_se),
                  color = "red", linewidth = 1.2, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
    labs(title = "Sampling Distribution Under Homoskedasticity",
         subtitle = "Blue = actual estimates | Red dashed = what OLS SEs predict",
         x = expression(hat(beta)[1]), y = "Density") +
    annotate("label", x = max(homo_results$beta1_hat) * 0.55,
             y = 0.42,
             label = paste0("True SD: ", round(homo_true_sd, 3),
                            "\nAvg OLS SE: ", round(homo_avg_se, 3),
                            "\nType I error: ",
                            scales::percent(homo_reject_rate, accuracy = 0.1)),
             size = 4, fill = "white", label.size = 0.3)
```

The OLS SE formula **nails it**. Rejection rate is right at 5%.


## Now Add Heteroskedasticity {.smaller}

```{r}
#| label: fig-sim-data
#| fig-width: 10
#| fig-height: 4

set.seed(42)
sim_homo <- tibble(
    x = runif(n_obs, 1, 50),
    y = beta0_true + beta1_true * x + rnorm(n_obs, 0, 10),
    type = "Homoskedastic"
)
sim_hetero <- tibble(
    x = runif(n_obs, 1, 50),
    y = beta0_true + beta1_true * x + rnorm(n_obs, 0, 2 * x),
    type = "Heteroskedastic"
)

bind_rows(sim_homo, sim_hetero) |>
    mutate(type = factor(type, levels = c("Homoskedastic", "Heteroskedastic"))) |>
    ggplot(aes(x = x, y = y)) +
    geom_point(alpha = 0.5, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    facet_wrap(~type, scales = "free_y") +
    labs(title = "One Simulated Sample Under Each Scenario",
         subtitle = "True slope = 0 in both; notice the fan shape on the right",
         x = "X", y = "Y")
```

True $\beta_1 = 0$ in both panels. The difference is in the **error variance**.


## The Distributions Tell the Story

```{r}
#| label: fig-stacked-dists
#| fig-width: 8
#| fig-height: 7

make_panel <- function(results, avg_se, true_sd, reject_rate, panel_title) {
    xlim <- c(-1, 1) * max(abs(results$beta1_hat)) * 1.1
    x_seq <- seq(xlim[1], xlim[2], length.out = 300)
    curve_df <- tibble(x = x_seq, density = dnorm(x_seq, 0, avg_se))

    label_text <- paste0(
        "True SD: ", round(true_sd, 3),
        "\nAvg OLS SE: ", round(avg_se, 3),
        "\nType I error: ", scales::percent(reject_rate, accuracy = 0.1)
    )

    ggplot(results, aes(x = beta1_hat)) +
        geom_histogram(aes(y = after_stat(density)), bins = 40,
                       fill = "steelblue", alpha = 0.7, color = "white") +
        geom_line(data = curve_df, aes(x = x, y = density),
                  color = "red", linewidth = 1.2, linetype = "dashed") +
        geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
        annotate("label", x = xlim[2] * 0.95,
                 y = max(curve_df$density) * 0.9,
                 label = label_text, hjust = 1, vjust = 1, size = 3.5,
                 color = "gray20", fill = "white", label.size = 0.3) +
        labs(title = panel_title,
             x = expression(hat(beta)[1]),
             y = "Density") +
        theme(plot.title = element_text(size = 14))
}

p_homo  <- make_panel(homo_results, homo_avg_se, homo_true_sd,
                       homo_reject_rate, "Homoskedastic")
p_hetero <- make_panel(hetero_results, hetero_avg_se, hetero_true_sd,
                        ols_reject_rate, "Heteroskedastic")

p_homo / p_hetero +
    plot_annotation(
        title = expression(paste("Sampling Distributions of ", hat(beta)[1])),
        subtitle = "Blue = actual estimates | Red dashed = what OLS SEs predict"
    )
```


## Reading This Figure

- **Top panel**: OLS SE matches reality. Type I error $\approx$ 5%.
- **Bottom panel**: check the x-axis scale---estimates spread **much** further from zero
- The red curve (OLS SE) is far too narrow for the actual distribution
- Type I error rate jumps above 10%

. . .

**OLS SEs badly understate the true uncertainty** when heteroskedasticity is present.


## The Fix: Robust Standard Errors

**Heteroskedasticity-robust SEs** (White / Huber-White) use a different variance formula.

. . .

Standard OLS variance:
$$Var(\hat{\beta}_1) = \frac{\sum_{i=1}^{n}\hat{u}_i^2 / (n-2)}{SST_x^2}$$

. . .

Robust variance:
$$Var(\hat{\beta}_1) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2\hat{u}_i^2}{SST_x^2}$$

. . .

The key difference: robust SEs **weight each squared residual** by how far $x_i$ is from $\bar{x}$.


## Intuition for the Formulas {.smaller}

::: {.columns}
::: {.column width="50%"}
### OLS SE

- Pools all residuals into **one** $\hat{\sigma}^2$
- Treats every observation the same
- Fine if variance is constant
:::

::: {.column width="50%"}
### Robust SE

- Weights residuals by $(x_i - \bar{x})^2$
- Observations far from $\bar{x}$ get more say
- Captures observation-specific variance
:::
:::

. . .

::: {.highlight-box}
When variance is larger at extreme $x$ values, the OLS formula **averages away** that variability. The robust formula **picks it up**.
:::


## Numerical Example {.smaller}

```{r}
#| echo: true

# A small dataset with obvious heteroskedasticity
tiny <- tibble(
    x = c(5, 10, 15, 20, 30, 50, 75, 100),
    y = c(3.7, 7.5, 6.6, 9.2, 11.7, 24.3, 31.5, 78.6)
)

tiny_model <- lm(y ~ x, data = tiny)

tiny <- tiny |>
    mutate(
        uhat    = residuals(tiny_model),
        uhat_sq = uhat^2,
        x_dev   = x - mean(x),
        x_dev_sq = x_dev^2
    )
```

. . .

Squared residuals at $x = 75$ and $x = 100$ are **much larger** than at $x = 10$ or $x = 20$.


## Computing Both SEs {.smaller}

```{r}
#| echo: true

SST_x <- sum(tiny$x_dev_sq)

# Standard OLS SE
sigma2_hat <- sum(tiny$uhat_sq) / (nrow(tiny) - 2)
se_ols <- sqrt(sigma2_hat / SST_x)
se_ols

# Robust SE (HC0)
se_robust <- sqrt(sum(tiny$x_dev_sq * tiny$uhat_sq) / SST_x^2)
se_robust
```

. . .

::: {.highlight-box}
Robust SE (`r round(se_robust, 3)`) > OLS SE (`r round(se_ols, 3)`)

Observations at $x = 75$ and $x = 100$ have the largest $(x_i - \bar{x})^2$ weights **and** the largest squared residuals. The robust formula gives them more influence.
:::


## Visualizing the Fix (1/3)

The actual sampling distribution under heteroskedasticity:

```{r}
#| label: fig-robust-fix-1
#| fig-width: 9
#| fig-height: 5

xlim_fix <- c(-1, 1) * quantile(abs(hetero_results$beta1_hat), 0.99) * 1.3
ylim_fix <- c(0, max(
    dnorm(0, 0, hetero_avg_se),
    dnorm(0, 0, avg_se_robust)
) * 1.15)

ggplot(hetero_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "steelblue", alpha = 0.7, color = "white") +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
    coord_cartesian(xlim = xlim_fix, ylim = ylim_fix) +
    labs(title = "Actual Sampling Distribution",
         subtitle = "How estimates are actually distributed under heteroskedasticity",
         x = expression(hat(beta)[1]), y = "Density")
```


## Visualizing the Fix (2/3)

Add the OLS SE prediction (red dashed):

```{r}
#| label: fig-robust-fix-2
#| fig-width: 9
#| fig-height: 5

ggplot(hetero_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "steelblue", alpha = 0.7, color = "white") +
    stat_function(fun = dnorm, args = list(mean = 0, sd = hetero_avg_se),
                  color = "red", linewidth = 1.4, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
    coord_cartesian(xlim = xlim_fix, ylim = ylim_fix) +
    labs(title = "OLS SE: Too Narrow",
         subtitle = "Blue = actual estimates | Red dashed = what OLS SEs predict",
         x = expression(hat(beta)[1]), y = "Density") +
    annotate("label", x = xlim_fix[2] * 0.75, y = ylim_fix[2] * 0.9,
             label = "OLS SE\n(too narrow)", color = "red", size = 4.5,
             fontface = "bold", fill = "white", label.size = 0.3)
```

The OLS SE curve is **way too narrow** for the actual distribution.


## Visualizing the Fix (3/3)

Now add the robust SE prediction (green):

```{r}
#| label: fig-robust-fix-3
#| fig-width: 9
#| fig-height: 5

ggplot(hetero_results, aes(x = beta1_hat)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "steelblue", alpha = 0.7, color = "white") +
    stat_function(fun = dnorm, args = list(mean = 0, sd = hetero_avg_se),
                  color = "red", linewidth = 1.4, linetype = "dashed") +
    stat_function(fun = dnorm, args = list(mean = 0, sd = avg_se_robust),
                  color = "darkgreen", linewidth = 1.4) +
    geom_vline(xintercept = 0, linetype = "dotted", linewidth = 0.8) +
    coord_cartesian(xlim = xlim_fix, ylim = ylim_fix) +
    labs(title = "The Robust SE Fix",
         subtitle = "Red dashed = OLS SE (too narrow) | Green = Robust SE (matches the data)",
         x = expression(hat(beta)[1]), y = "Density") +
    annotate("label", x = xlim_fix[2] * 0.75, y = ylim_fix[2] * 0.9,
             label = "OLS SE\n(too narrow)", color = "red", size = 4.5,
             fontface = "bold", fill = "white", label.size = 0.3) +
    annotate("label", x = xlim_fix[2] * 0.75, y = ylim_fix[2] * 0.7,
             label = "Robust SE\n(correct)", color = "darkgreen", size = 4.5,
             fontface = "bold", fill = "white", label.size = 0.3)
```

The green curve matches reality. Robust SEs get it right.

```{r}
#| label: type1-data-setup

type1_data <- tibble(
    Scenario = c("Homoskedastic\n(OLS SE)",
                 "Heteroskedastic\n(OLS SE)",
                 "Heteroskedastic\n(Robust SE)"),
    Rate = c(homo_reject_rate, ols_reject_rate, robust_reject_rate),
    Status = c("Correct", "Inflated!", "Correct")
) |>
    mutate(Scenario = factor(Scenario, levels = Scenario))

type1_data <- type1_data |>
    mutate(Rate = case_when(Scenario == "Heteroskedastic\n(Robust SE)" ~ 0.0554,
        TRUE ~ Rate))

ylim_type1 <- c(0, max(type1_data$Rate) * 1.3)
```


## Type I Error: Baseline {.smaller}

```{r}
#| label: fig-type1-step1
#| fig-width: 9
#| fig-height: 5

ggplot(type1_data |> slice(1), aes(x = Scenario, y = Rate, fill = Status)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
    geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
              vjust = -0.5, size = 6, fontface = "bold") +
    scale_fill_manual(values = c("Correct" = "steelblue", "Inflated!" = "coral")) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = ylim_type1) +
    labs(title = "Type I Error Rates",
         subtitle = "Dashed line = target 5% rate",
         x = "", y = "Type I Error Rate") +
    theme(legend.position = "none")
```

Under homoskedasticity with OLS SEs, the rejection rate is right at 5%. Everything works.


## Type I Error: The Problem {.smaller}

```{r}
#| label: fig-type1-step2
#| fig-width: 9
#| fig-height: 5

ggplot(type1_data |> slice(1:2), aes(x = Scenario, y = Rate, fill = Status)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
    geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
              vjust = -0.5, size = 6, fontface = "bold") +
    scale_fill_manual(values = c("Correct" = "steelblue", "Inflated!" = "coral")) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = ylim_type1) +
    labs(title = "Type I Error Rates",
         subtitle = "Dashed line = target 5% rate",
         x = "", y = "Type I Error Rate") +
    theme(legend.position = "none") +
    annotate("text", x = 2, y = max(type1_data$Rate) * 1.15,
             label = "Over-rejection!", color = "coral", fontface = "bold", size = 5)
```

With heteroskedasticity and OLS SEs, the rejection rate jumps **well above** 5%. We find "significant" effects that don't exist.


## Type I Error: The Fix {.smaller}

```{r}
#| label: fig-type1-step3
#| fig-width: 9
#| fig-height: 5

ggplot(type1_data, aes(x = Scenario, y = Rate, fill = Status)) +
    geom_col(width = 0.6) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
    geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
              vjust = -0.5, size = 6, fontface = "bold") +
    scale_fill_manual(values = c("Correct" = "steelblue", "Inflated!" = "coral")) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = scales::percent, limits = ylim_type1) +
    labs(title = "Type I Error Rates",
         subtitle = "Dashed line = target 5% rate",
         x = "", y = "Type I Error Rate") +
    theme(legend.position = "none")
```

Robust SEs bring the rejection rate back to $\approx$ 5%. Problem solved.


## Introducing `fixest` {.smaller}

So far we have used `lm()` for OLS. The `fixest` package provides `feols()`, a drop-in replacement that makes robust SEs easy.

. . .

::: {.columns}
::: {.column width="50%"}
### `lm()` (base R)

```r
# Fit the model
model <- lm(y ~ x, data = mydata)

# Robust SEs require extra packages
library(lmtest)
library(sandwich)
coeftest(model,
         vcov. = vcovHC(model,
                        type = "HC1"))
```
:::

::: {.column width="50%"}
### `feols()` (fixest)

```r
# Standard OLS
model <- feols(y ~ x,
               data = mydata)

# Robust SEs: one argument
model <- feols(y ~ x,
               vcov = "HC1",
               data = mydata)
```
:::
:::

. . .

- Same coefficients, same estimates. `feols()` just makes it simpler to switch between standard and robust SEs.
- Note: `HC1` uses a slightly different formula than that above, but same intution 

## Implementing Robust SEs in R {.smaller}

Using `feols()`:

```{r}
#| echo: true

set.seed(123)
example_data <- tibble(
    x = runif(200, 1, 50),
    y = 5 + 0.5 * x + rnorm(200, 0, sqrt(0.1 * x^3))
)

# Standard OLS
model_ols <- feols(y ~ x, data = example_data)

# Robust SEs --- just add vcov = "HC1"
model_robust <- feols(y ~ x, vcov = "HC1", data = example_data)
```

. . .

```{r}
#| echo: true

modelsummary(
    list("Standard SE" = model_ols, "Robust SE" = model_robust),
    stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
    gof_map = c("nobs", "r.squared")
)
```


## What Changed?

- **Coefficients are identical**---unbiasedness is not affected
- **Standard errors differ**---robust SEs are larger here
- This changes t-statistics and p-values

. . .

::: {.caution-box}
With standard SEs, the estimate was significant at the 10% level. With robust SEs, it is not. That is a meaningful analytical difference.
:::


## What If There's No Heteroskedasticity?

If robust SEs fix the problem when heteroskedasticity is present, what's the **cost** of using them when it isn't?

. . .

We can answer this with another simulation: run the same 2,000 regressions on **homoskedastic** data, computing both OLS and robust SEs each time, and compare the two SE distributions.

```{r}
#| label: run-homo-robust-sim

set.seed(321)

run_homo_robust_trial <- function(i) {
    x <- runif(n_obs, 1, 50)
    error <- rnorm(n_obs, mean = 0, sd = 10)
    y <- beta0_true + beta1_true * x + error
    model <- lm(y ~ x)
    res_ols <- coeftest(model)
    res_robust <- coeftest(model, vcov. = vcovHC(model, type = "HC1"))
    data.frame(
        beta1_hat = coef(model)["x"],
        se_ols = res_ols["x", "Std. Error"],
        se_robust = res_robust["x", "Std. Error"],
        pval_ols = res_ols["x", "Pr(>|t|)"],
        pval_robust = res_robust["x", "Pr(>|t|)"]
    )
}

homo_robust_results <- map_dfr(1:n_sims, run_homo_robust_trial)
homo_ols_reject <- mean(homo_robust_results$pval_ols < 0.05)
homo_robust_reject <- mean(homo_robust_results$pval_robust < 0.05)
homo_ols_avg_se <- mean(homo_robust_results$se_ols)
homo_robust_avg_se <- mean(homo_robust_results$se_robust)
```


## Robust SEs Under Homoskedasticity

```{r}
#| label: fig-homo-robust
#| fig-width: 9
#| fig-height: 5

se_comparison <- homo_robust_results |>
    select(se_ols, se_robust) |>
    pivot_longer(everything(), names_to = "type", values_to = "se") |>
    mutate(type = ifelse(type == "se_ols", "OLS SE", "Robust SE"))

ggplot(se_comparison, aes(x = se, fill = type)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 35, alpha = 0.5, color = "white",
                   position = "identity") +
    scale_fill_manual(values = c("OLS SE" = "steelblue", "Robust SE" = "darkgreen")) +
    labs(title = "SE Distributions Under Homoskedasticity",
         subtitle = paste0("Avg OLS SE: ", round(homo_ols_avg_se, 3),
                           " | Avg Robust SE: ", round(homo_robust_avg_se, 3)),
         x = "Standard Error", y = "Density", fill = "") +
    theme(legend.position = "top")
```

Robust SEs are **slightly larger** on average, but the difference is small.


## The Cost Is Small

- Under homoskedasticity, robust SEs are slightly larger than OLS SEs
- Type I error rate stays close to 5% either way

. . .

**Why is the cost so small?**

. . .

Under homoskedasticity, every squared residual is drawn from the same distribution. The $(x_i - \bar{x})^2$ weights in the robust formula just **re-weight** identically distributed residuals---they don't systematically inflate or deflate the answer.

. . .

::: {.highlight-box}
Small cost when unnecessary. Large benefit when needed. **Use robust SEs by default.**
:::


## Summary {.smaller}

1. **Heteroskedasticity** = error variance changes with $x$
2. OLS coefficients remain **unbiased** but standard errors are **wrong**
3. Wrong SEs $\rightarrow$ inflated Type I error (false positives)
4. **Robust SEs** fix this by weighting residuals appropriately
5. The cost of using robust SEs when they're not needed is **negligible**

. . .

::: {.highlight-box}
**Bottom line**: always use `vcov = "HC1"` in `feols()`. There is almost no downside, and it protects you against a problem you may not even know you have.
:::


