---
title: "Panel Data and Fixed Effects"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: today
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    embed-resources: true
    fig-align: center
    code-overflow: wrap
    footer: "EC325 | Unit 11 | Panel Data & Fixed Effects"
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(tidyverse)
library(wooldridge)
library(fixest)
library(modelsummary)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 16) +
          theme(plot.title = element_text(face = "bold", size = 18),
                plot.subtitle = element_text(size = 13, color = "gray40"),
                panel.grid.minor = element_blank()))
```


## Where We Are

- Started with the "gold standard" of causal inference: **RCTs**
- Next best alternative: **multivariate OLS** --- control for observable confounders
- Last unit: **binary outcomes** --- more flexible models for real-life data
- Central threat throughout: **omitted variable bias**
- We can control for what we can observe... but what about **unobservable** confounders? Ability, motivation, culture, history?

. . .

Today: a new data structure --- **panel data** --- that lets us control for unobservables.


## Cross-Sectional Data

Most datasets we've looked at so far have been **cross-sectional**: units observed at a single point in time.

- Housing prices in Ames, Iowa in 2010
- Wages and education data from the CPS
- Health outcomes in the Oregon Health Insurance Experiment
- Women's labor force participation (`mroz`)

## What We Can Do with Cross-Sectional Data

With a snapshot from a single point in time, we can:

- **Describe a population**: averages, medians, distributions
- **Identify correlations**: do two variables move together?
- **Control for observable confounders**: multivariate OLS, holding other measured factors constant

. . .

But there's a big limitation...

## What Cross-Sectional Data Struggles With {.smaller}

```{r}
#| label: fig-cross-section
#| fig-width: 8
#| fig-height: 3.5

cs_data <- tibble(
    City = c("Boston", "Denver", "Miami"),
    Year = 2020,
    Unemployment = c(6.2, 5.1, 7.8),
    Crime = c(312, 285, 410)
)

ggplot(cs_data, aes(x = Unemployment, y = Crime, label = City)) +
    geom_point(size = 4, color = "steelblue") +
    geom_text(vjust = -1, size = 5) +
    coord_cartesian(xlim = c(4.5, 8.5), ylim = c(250, 450)) +
    labs(x = "Unemployment Rate (%)", y = "Crime Rate",
         title = "Cross-Sectional Data: Three Cities in 2020",
         subtitle = "Each city observed once --- we can only compare across cities")
```

Miami has high unemployment **and** high crime. But is that because unemployment *causes* crime, or because Miami differs in many unobserved ways (geography, culture, history)?

. . .

We can't control for what we can't measure!


## Panel Data {.smaller}

This is in contrast to **panel data**: the same units observed at *multiple points in time*.

- NBA players observed across multiple seasons
- Employment and minimum wages in states across multiple quarters
- Pre-period and post-period data in an experiment
- Stock prices for multiple companies tracked daily for a year

. . .

| Example | Unit ($i$) | Time ($t$) |
|:--------|:-----|:-----|
| Workers tracked annually | Person | Year |
| States observed quarterly | State | Quarter |
| Firms reported monthly | Firm | Month |
| Countries measured yearly | Country | Year |


## What If We Followed Cities Over Time? {.smaller}

```{r}
#| label: fig-panel-intro
#| fig-width: 9
#| fig-height: 4

panel_data <- tibble(
    City = rep(c("Boston", "Denver", "Miami"), each = 3),
    Year = rep(2018:2020, 3),
    Unemployment = c(4.8, 5.5, 6.2, 4.0, 4.5, 5.1, 6.5, 7.0, 7.8),
    Crime = c(298, 305, 312, 270, 278, 285, 390, 401, 410)
)

ggplot(panel_data, aes(x = Unemployment, y = Crime, color = City, group = City)) +
    geom_point(size = 3) +
    geom_line(linewidth = 1, arrow = arrow(length = unit(0.15, "inches"), type = "closed")) +
    geom_text(data = panel_data |> filter(Year == 2020),
              aes(label = City), vjust = -1.2, size = 4.5, show.legend = FALSE) +
    scale_color_manual(values = c("Boston" = "steelblue", "Denver" = "#D55E00", "Miami" = "#009E73")) +
    coord_cartesian(ylim = c(260, 430)) +
    labs(x = "Unemployment Rate (%)", y = "Crime Rate",
         title = "Panel Data: Three Cities Over Three Years",
         subtitle = "Arrows show each city's trajectory over time") +
    theme(legend.position = "none")
```

Now we can ask: when unemployment rises **within a given city**, does crime rise too?


## Benefits of Panel Data

1. **Controlling for Unobserved Heterogeneity**
    - Observing units over time lets us control for all *time-invariant* characteristics --- both **observed and unobserved**!
    - Geography, genetics, a firm's core industry, "culture" --- as long as they're stable, panel methods can remove their influence
    - This is the **big one** and the focus of today's lecture

## Benefits of Panel Data (cont.)

2. **Analyzing Dynamics of Change**
    - We can study how variables change over time and why
    - Analyze the impact of events, policies, or treatments by comparing units before and after
    - Examine lags and leads (does a change in $x$ today affect $y$ tomorrow?)

. . .

3. **Increased Statistical Power**
    - Pooling data across units *and* time increases total sample size ($N \times T$)
    - More data $\rightarrow$ more precise estimates (smaller standard errors)


## The Key Insight

::: {.highlight-box}
Panel data lets us control for **all time-invariant characteristics** of each unit --- whether observed or unobserved --- by using only **within-unit** variation.
:::

. . .

- **Between** variation: comparing *different* cities to each other (confounded by all ways cities differ)
- **Within** variation: comparing *each city to itself* over time (city-level confounders cancel out)

. . .

The question shifts from "do cities with more unemployment have more crime?" to "when unemployment rises *within a city*, does crime rise too?"


## The Fixed Effects Model

We formalize this with the **fixed effects model**:

$$y_{it} = \beta_0 + \beta_1 x_{it} + a_i + \mu_{it}$$

. . .

- $y_{it}$: outcome for unit $i$ at time $t$
- $x_{it}$: explanatory variable (varies across units **and** time)
- $a_i$: **unit fixed effect** --- captures all time-invariant characteristics of unit $i$
- $\mu_{it}$: idiosyncratic error (the leftover)



## What Does $a_i$ Capture? {.smaller}

The unit fixed effect $a_i$ absorbs **everything** about unit $i$ that doesn't change over time:

. . .

- A city's geographic location
- A city's historical legacy and founding conditions
- Stable institutional features (legal system, political structure)
- Baseline demographics (to the extent they're stable)
- Measurement differences (how crime is recorded)
- "Culture" and other hard-to-measure factors

. . .

We don't need to **observe** these things. We don't need to **measure** them. We just need them to be **stable** over the time periods we observe.


## How Does It Work? The Within Transformation (1/3) {.smaller}

We can't observe $a_i$ directly. But we can **eliminate** it.

. . .

**Step 1:** For each unit $i$, compute the time average of every variable:

$$\bar{y}_i = \beta_0 + \beta_1 \bar{x}_i + a_i + \bar{\mu}_i$$


## How Does It Work? The Within Transformation (2/3) {.smaller}

**Step 2:** Subtract the time-average equation from the original:

\begin{align}
y_{it} - \bar{y}_i &= (\beta_0 - \beta_0) + \beta_1(x_{it} - \bar{x}_i) + (a_i - a_i) + (\mu_{it} - \bar{\mu}_i)
\end{align}

. . .

What happened?

- $\beta_0 - \beta_0 = 0$ --- the intercept drops out
- $a_i - a_i = 0$ --- **the fixed effect drops out!** Since $a_i$ is constant over time, its time average is just itself

. . .

We're left with only **demeaned** variables:

$$\ddot{y}_{it} = \beta_1 \ddot{x}_{it} + \ddot{\mu}_{it}$$

where $\ddot{y}_{it} = y_{it} - \bar{y}_i$ (deviations from each unit's own average), and so on.


## How Does It Work? The Within Transformation (3/3)

::: {.highlight-box}
**Why "within" transformation?**

After demeaning, all cross-sectional differences between units are gone. The only variation left is how each unit **differs from its own average** over time.

$\beta_1$ is identified purely from **within-unit changes**.
:::

. . .

Let's see this in action with a concrete example...


## Within Transformation: A Small Example (1/3) {.smaller}

Suppose we have 3 cities observed over 2 years. Here's our **raw data**:

```{r}
#| echo: true

example_df <- tibble(
  city = rep(c("A", "B", "C"), each = 2),
  year = rep(c(1, 2), times = 3),
  unem = c(5, 7, 3, 5, 8, 6),
  crime = c(30, 34, 20, 26, 40, 38)
)
example_df
```

. . .

City A: unemployment went up (5 to 7), crime went up (30 to 34). City C: unemployment went *down* (8 to 6), crime went *down* (40 to 38). That's the within-unit variation we want to use.


## Within Transformation: A Small Example (2/3) {.smaller}

**Step 1:** Compute each city's time average, then subtract:

```{r}
#| echo: true

example_df <- example_df |>
  group_by(city) |>
  mutate(
    mean_unem = mean(unem),
    mean_crime = mean(crime),
    unem_demean = unem - mean_unem,
    crime_demean = crime - mean_crime
  ) |>
  ungroup()

example_df |> select(city, year, unem, mean_unem, unem_demean, crime, mean_crime, crime_demean)
```

. . .

Notice: the **demeaned** values only reflect how each city differs from **its own** average --- all level differences between cities are gone.


## Within Transformation: A Small Example (3/3) {.smaller}

**Step 2:** Regress the demeaned variables:

```{r}
#| echo: true

lm(crime_demean ~ unem_demean, data = example_df) |> coef()
```

. . .

Now compare to a regression using *between* variation:

```{r}
#| echo: true

lm(crime ~ unem, data = example_df) |> coef()
```



## Going Further: Time Fixed Effects {.smaller}

- We removed **unit-level** confounders by demeaning across time. But what about shocks that hit **all** cities in a given year (e.g., a recession, a national policy change)? 
- These are called **time-fixed effects**

. . .

We can go a step further: take our **unit-demeaned** data and demean it again --- this time across **units within each year**:

```{r}
#| echo: true

example_df <- example_df |>
  group_by(year) |>
  mutate(
    yr_mean_unem_dm = mean(unem_demean),
    yr_mean_crime_dm = mean(crime_demean),
    unem_dm2 = unem_demean - yr_mean_unem_dm,
    crime_dm2 = crime_demean - yr_mean_crime_dm
  ) |>
  ungroup()

example_df |> select(city, year, unem_demean, yr_mean_unem_dm, unem_dm2, crime_demean, yr_mean_crime_dm, crime_dm2)
```


## Time Fixed Effects: The Result {.smaller}

Now regress the **doubly-demeaned** variables (no intercept needed):

```{r}
#| echo: true

lm(crime_dm2 ~ unem_dm2 - 1, data = example_df) |> coef()
```

. . .

Compare to a regression with dummies for **both** city and year:

```{r}
#| echo: true

lm(crime ~ unem + factor(city) + factor(year), data = example_df) |> coef()
```

. . .

::: {.highlight-box}
Same coefficient on `unem`! Sequential demeaning --- first by unit, then by time --- is equivalent to including **both** unit and time fixed effects. The first demeaning removes time-invariant unit confounders; the second removes unit-invariant time shocks.
:::


## Equivalently: Dummy Variable Regression {.smaller}

Instead of manually demeaning, we can estimate the same model using **dummy variable regression**:

$$y_{it} = \beta_0 + \beta_1 x_{it} + a_i + \tau_t + \mu_{it}$$

. . .

Include a **dummy variable for each unit** $i$ (minus one omitted category) and a **dummy variable for each time** $t$ (minus one omitted period).

. . .

- This produces the **same** $\hat{\beta}_1$ as the within transformation
- But also gives us estimates of each unit's and time period's dummy coefficient
- Don't forget the *dummy variable trap*!

. . .

In R, we can use `factor()` to automatically create dummies, or --- much more efficiently --- use `feols()` from `fixest`.

. . .

::: {.callout-note}
## A Variation You May See
Some implementations **drop the intercept** and include a dummy for **every** unit (no omitted category). This yields a different $\hat{\beta}_0$ interpretation (each dummy captures the unit's own intercept), but $\hat{\beta}_1$ is **identical** either way.
:::


## Example: Unemployment and Crime {.smaller}

```{r}
#| label: crime-data-setup

crime_data <- wooldridge::crime2 |>
    mutate(city_id = rep(1:46, each = 2))
```

The `crime2` dataset: **46 cities** observed in **1982 and 1987**.

. . .

Let's start with a **cross-sectional** regression using just 1987:

```{r}
#| echo: true

reg_cross <- lm(crmrte ~ unem, data = filter(crime_data, year == 87))
coef(reg_cross)
```

. . .

The coefficient on `unem` is **negative**: higher unemployment $\rightarrow$ *less* crime?

This makes no intuitive sense. Almost certainly **OVB** --- cities with low unemployment also have other "good" characteristics (wealth, strong institutions) that reduce crime.


## Now Use the Panel: Add Fixed Effects {.smaller}

```{r}
#| echo: true

reg_fe <- feols(crmrte ~ unem | city_id + d87, data = crime_data)
coef(reg_fe)
```

. . .

Now the coefficient is **positive**: when unemployment rises *within a city*, crime rises too.

. . .

::: {.highlight-box}
Fixed effects removed the bias from unobserved city characteristics, revealing the true positive relationship between unemployment and crime.
:::


## Side-by-Side Results {.smaller}

```{r}
#| echo: true

modelsummary(list("Cross-Section" = reg_cross,
                  "Fixed Effects" = reg_fe),
             gof_map = c("adj.r.squared", "nobs"),
             add_rows = tribble(
                 ~term, ~`Cross-Section`, ~`Fixed Effects`,
                 "City FE", "No", "Yes",
                 "Year FE", "No", "Yes"
             ))
```

. . .

Notice the bottom of the table: researchers report which fixed effects are included as "Yes/No" rows rather than reporting every dummy coefficient.


## Using `feols()` in Practice {.smaller}

The `feols()` function from `fixest` puts fixed effects after a vertical bar `|`:

```{r}
#| echo: true
#| eval: false

feols(y ~ x1 + x2 | unit + time, data = mydata)
```

. . .

**Left of `|`:** standard regression formula --- variables you want coefficients for.

**Right of `|`:** fixed effects (absorbed) --- dummies created automatically, coefficients **not** reported.

. . .

This is equivalent to including a dummy for every unit and time period, but `feols()` is **much faster** and keeps the output clean.


## A Worked Example: Police and Crime {.smaller}

```{r}
#| label: small-panel-setup

city_panel <- tibble(
    city = rep(c("Austin", "Baltimore", "Denver", "Portland", "Raleigh"), each = 4),
    year = rep(2017:2020, times = 5),
    police_spend = c(3.8, 4.0, 4.3, 4.1, 6.2, 6.5, 6.4, 6.0,
                     4.5, 4.7, 5.0, 4.6, 3.5, 3.4, 3.2, 3.0,
                     3.9, 4.2, 4.5, 4.3),
    property_crime = c(35.2, 34.1, 32.8, 33.5, 48.6, 47.3, 47.9, 49.2,
                       38.4, 37.1, 35.6, 37.8, 42.1, 43.5, 44.8, 46.2,
                       30.1, 29.2, 27.8, 28.9)
)
```

5 cities, 4 years. First, let's ignore the panel structure and look at the **cross-sectional** (between) relationship:

```{r}
#| label: fig-between-scatter
#| fig-width: 8
#| fig-height: 3.5

ggplot(city_panel, aes(x = police_spend, y = property_crime)) +
    geom_point(size = 3, color = "gray50") +
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
    labs(x = "Police Spending (per capita, $100s)",
         y = "Property Crime Rate (per 1,000)",
         title = "Cross-Sectional: Pooling All Observations",
         subtitle = "Positive slope: cities that spend more on police have MORE crime?")
```

. . .

This is the **between** comparison --- it looks like police spending *increases* crime.


## Now Layer in the Panel Structure {.smaller}

Same data, but now we connect observations **within** each city over time:

```{r}
#| label: fig-within-scatter
#| fig-width: 8
#| fig-height: 3.5

ggplot(city_panel, aes(x = police_spend, y = property_crime, color = city)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, alpha = 0.7) +
    scale_color_brewer(palette = "Set1") +
    labs(x = "Police Spending (per capita, $100s)",
         y = "Property Crime Rate (per 1,000)",
         title = "Within Cities: A Different Story",
         subtitle = "Within each city, more spending is associated with LESS crime",
         color = NULL) +
    theme(legend.position = "bottom")
```

. . .

The **within-city** slopes tell the opposite story from the cross-sectional one!


## Why the Misleading Cross-Sectional Pattern? {.smaller}

Baltimore spends the most on police **and** has the highest crime.

. . .

But this doesn't mean police spending *causes* crime. Cities with high crime invest more in policing precisely **because** they have high crime.

. . .

Those same cities also have persistent characteristics (poverty, density, historical disinvestment) that keep crime elevated regardless of spending.

. . .

::: {.caution-box}
The cross-sectional comparison conflates the **effect of police spending** with all the other ways these cities differ. We need fixed effects to separate them.
:::


## Estimating with `feols()` {.smaller}

```{r}
#| echo: true

fe_reg <- feols(property_crime ~ police_spend | city + year,
                data = city_panel)
summary(fe_reg)
```

. . .

After absorbing city and year fixed effects, more police spending is associated with **lower** crime --- the opposite of what the cross-sectional data suggested.


## Extracting the Fixed Effects {.smaller}

- We can retrieve the estimated unit and time fixed effects with `fixef()`:

```{r}
#| echo: true

fe_estimates <- fixef(fe_reg)
fe_estimates$city
fe_estimates$year
```

. . .

- **City FE**: Baltimore has the highest baseline crime; Raleigh the lowest. These capture everything about each city that is stable over time.
- **Year FE**: common year-to-year shocks affecting all cities.
- We usually don't care about these coefficients *per se*, but they can give useful information:
  - "Accounting for police spending, Baltimore has the highest baseline crime (relative to the omitted city)"
  - "2020 had the highest crime across all cities (relative to the omitted year)"
- The main point: they absorb confounders so $\hat{\beta}_1$ is unbiased.


## What Do FE Control For? {.smaller}

**Unit FE** control for anything specific to a unit that is **constant over time**: geography, historical legacy, stable institutions, baseline demographics, measurement differences, culture.

. . .

**Time FE** control for anything that hits **all units equally** in a given period: macroeconomic shocks, national policy changes, seasonal patterns, technology trends.

. . .

**Neither** controls for factors that vary **both across units and over time** --- for example, a city-specific economic boom or a policy change that only affects some cities. These are **unit-specific, time-varying confounders** and must be controlled for explicitly (or remain a source of potential bias).


## A Classification Exercise {.smaller}

::: {.nonincremental}
For each confounder, is it absorbed by unit FE, time FE, or neither?

- A city's geographic terrain?
:::

. . .

**Unit FE** --- constant across time for a given city

::: {.nonincremental}
- National recession in 2020?
:::

. . .

**Time FE** --- hits all cities equally in that year

::: {.nonincremental}
- City-specific economic boom in 2019?
:::

. . .

**Neither** --- varies across cities *and* over time

::: {.nonincremental}
- Nationwide sentencing reform?
:::

. . .

**Time FE** --- affects all cities in the same period

::: {.nonincremental}
- A city's founding history?
:::

. . .

**Unit FE** --- permanent characteristic of that city


## Classification Exercise (cont.) {.smaller}

::: {.nonincremental}
- A new mayor elected in one city in 2018?
:::

. . .

**Neither** --- specific to one city at one point in time

::: {.nonincremental}
- A state's distance from the border?
:::

. . .

**Unit FE** --- doesn't change over time

::: {.nonincremental}
- Nationwide COVID-19 lockdowns in 2020?
:::

. . .

**Time FE** --- common shock across all units

::: {.nonincremental}
- A city-specific policing strategy introduced in 2019?
:::

. . .

**Neither** --- unit-specific and time-varying

. . .

The "neither" cases are **unit-specific, time-varying confounders**. These require explicit control variables.


## Adding Time-Varying Controls {.smaller}

- One way to address the fact that FEs don'y account for **unit-specific, time-varying confounders** is to directly control for them!

- Can do so by putting them in the RHS of our regression, like any other variable

. . .

$$y_{it} = \beta_1 x_{it} + \gamma z_{it} + a_i + \tau_t + \mu_{it}$$

- Of course, this requires that you are directly able to observe these variables, so only a partial solution

## Adding Time-Varying Controls {.smaller}

In `feols()`, time-varying controls go on the **left** side of `|`:

. . .

```{r}
#| echo: true

city_panel <- city_panel |>
    mutate(unem_rate = c(3.1, 2.8, 2.9, 5.0, 5.8, 5.5, 5.3, 7.6,
                         3.0, 2.7, 2.8, 5.4, 3.8, 3.6, 3.9, 6.2,
                         3.4, 3.1, 3.0, 5.1))

fe_with_control <- feols(property_crime ~ police_spend + unem_rate | city + year,
                         data = city_panel)
```

. . .

```{r}
#| echo: true

modelsummary(list("No Control" = fe_reg, "With Unemployment" = fe_with_control),
             gof_map = c("adj.r.squared"),
             add_rows = tribble(
                 ~term, ~`No Control`, ~`With Unemployment`,
                 "City FE", "Yes", "Yes",
                 "Year FE", "Yes", "Yes"
             ))
```



## When Are Fixed Effects Most Useful? {.smaller}

**Good fit for FE:**

- Genuine panel data (same units, multiple periods)
- Meaningful within-unit variation in $x$
- Time-invariant confounders are a primary concern
- Treatment effects are relatively immediate

. . .

**Less useful when:**

- Little within-unit variation (everyone's $x$ is stable over time)
- Treatment effects take a long time to materialize
- Time-varying confounders are the main concern
- Very few time periods (less precise estimates)


## Summary {.smaller}

1. **Panel data**: same units observed over multiple time periods
2. The **within transformation** eliminates all time-invariant confounders by demeaning
3. **Unit FE** absorb permanent unit characteristics (observed and unobserved)
4. **Time FE** absorb shocks common to all units in a given period
5. In R: `feols(y ~ x | unit + time, data = mydata)`
6. Can add **time-varying controls** on the left side of `|`
7. **Cannot** include time-invariant variables --- they're collinear with unit FE

. . .

::: {.highlight-box}
Fixed effects $\neq$ no OVB. Unit-specific, time-varying confounders can still bias results. But FE is a **major improvement** over cross-sectional analysis when time-invariant confounders are the primary concern.
:::


## Try It Yourself {.smaller}

The `wagepan` dataset in `wooldridge` is a balanced panel of 545 men observed annually from 1980--1987.

```{r}
#| echo: true
#| eval: false

wage_data <- wooldridge::wagepan
```

. . .

**1.** How many workers? How many years? Confirm it's balanced. (*Hint:* `n_distinct()`)

. . .

**2.** Using only 1987 data, estimate: $\ln(wage) = \beta_0 + \beta_1 \cdot union + \mu$. Interpret $\hat{\beta}_1$.

. . .

**3.** Now estimate with worker and year FE: `feols(lwage ~ union | nr + year)`. How does $\hat{\beta}_1$ change? Why?

. . .

**4.** Try adding `educ` as a control in the FE model. What happens? Why?


