---
title: "Regression Discontinuity"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: today
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    embed-resources: true
    fig-align: center
    code-overflow: wrap
    footer: "EC325 | Unit 15 | Regression Discontinuity"
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
#| label: setup

library(tidyverse)
library(fixest)
library(modelsummary)
library(patchwork)

options(scipen = 999)

theme_set(theme_minimal(base_size = 16) +
          theme(plot.title = element_text(face = "bold", size = 18),
                plot.subtitle = element_text(size = 13, color = "gray40"),
                panel.grid.minor = element_blank()))
```


## Where We've Been {.smaller}

- Our causal inference toolkit so far:
  - **RCTs**: random assignment by design
  - **Multivariate OLS**: control for selection on observables
  - **Fixed effects**: absorb time-invariant unobservables
  - **DiD**: control units net out common trends
  - **IV**: exogenous variation from an instrument
- All require finding variation in $X$ that is **as good as random** or **totally exogenous**
- Today: one more tool --- one of the cleanest --- **regression discontinuity (RD)**


## The Big Idea {.smaller}

- Modern life is full of **rigid rules and cutoffs**:

. . .

  - Turn 21 $\to$ legal drinking
  - Score 70+ on an exam $\to$ admitted to a selective school
  - Income below threshold $\to$ qualify for Medicaid
  - 50.01% vote share $\to$ win the election

- At the cutoff, treatment flips on (or off) **abruptly**
- Units just above and just below are nearly identical *except* for treatment
- $\to$ rigid rules generate **natural experiments**



# The RD Intuition


## A Jump at the Cutoff

```{r}
#| label: fig-intuition
#| fig-width: 9
#| fig-height: 4.6

set.seed(1)
n <- 300
x <- runif(n, 0, 100)
d <- as.numeric(x >= 50)
y <- 30 + 0.5 * x + 8 * d + rnorm(n, 0, 4)
intuition <- tibble(x = x, y = y, d = factor(d, labels = c("Untreated", "Treated")))

ggplot(intuition, aes(x = x, y = y, color = d)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(aes(group = d), method = "lm", se = FALSE, linewidth = 1.3) +
    geom_vline(xintercept = 50, linetype = "dashed", color = "black") +
    annotate("segment", x = 50, xend = 50, y = 55, yend = 63,
             arrow = arrow(type = "closed", length = unit(0.3, "cm"),
                           ends = "both"),
             color = "#e7298a", linewidth = 1.2) +
    annotate("text", x = 53, y = 59, label = "Causal\neffect",
             hjust = 0, color = "#e7298a", size = 5, fontface = "bold") +
    annotate("text", x = 50, y = 95, label = "Cutoff",
             hjust = -0.1, size = 4.5) +
    scale_color_manual(values = c("Untreated" = "grey50",
                                  "Treated"   = "#0072B2")) +
    labs(x = "Running Variable (X)", y = "Outcome (Y)", color = NULL) +
    theme(legend.position = "bottom")
```


## Key Vocabulary {.smaller}

::: {.callout-note}
## Three terms you need to know
- **Running variable** (or *forcing variable*): the continuous variable whose value determines treatment (e.g., age, test score, vote share).
- **Cutoff** (or *threshold*): the value of the running variable at which treatment flips (e.g., age 21, 70%, 50%).
- **Sharp RD**: treatment is a *deterministic* function of the running variable --- 0 below the cutoff, 1 above (or vice versa).
:::

. . .

- Today we focus on **sharp** RD
- A *fuzzy* variant exists (cutoff shifts the *probability* of treatment) --- handled with IV



# Running Example: The MLDA


## Birthdays and Funerals {.smaller}

- Carpenter & Dobkin (2009): does the **minimum legal drinking age** of 21 save lives?

. . .

- Hard to answer with regular tools:
  - Can't randomize legal drinking age
  - Before/after comparisons confound aging with drinking
- But the MLDA itself runs the experiment for us:
  - On your 21st birthday, legal access flips on
  - Yesterday's you and today's you are essentially identical, *except for the fact that you are now 21*
- Compare death rates **just below** and **just above** age 21


## Simulating Carpenter & Dobkin {.smaller}

```{r}
#| echo: true

set.seed(234)

mlda_data <- tibble(months = seq(-24, 23, by = 1)) |>
    mutate(
        over21 = as.numeric(months >= 0),
        # True jump = 7.7 deaths per 100,000 (AP Table 4.1)
        death_rate = 90 - 0.15 * months + 7.7 * over21 +
                     rnorm(n(), 0, 1)
    )
```

. . .

```{r}
#| label: fig-raw
#| fig-width: 9
#| fig-height: 3.6

ggplot(mlda_data, aes(x = months, y = death_rate, color = factor(over21))) +
    geom_point(size = 3, alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
    scale_color_manual(values = c("0" = "grey50", "1" = "#0072B2"),
                       labels = c("Under 21", "Over 21")) +
    annotate("text", x = 0, y = 104, label = "21st birthday",
             hjust = -0.05, size = 4.5) +
    labs(x = "Age in months (0 = 21st birthday)",
         y = "Death rate per 100,000",
         color = NULL) +
    theme(legend.position = "bottom")
```



# The Sharp RD Regression


## A Linear Model {.smaller}

- Simplest sharp RD specification:

. . .

$$
Y_a = \alpha + \rho D_a + \gamma a + e_a
$$

. . .

- $D_a = 1$ if $a \geq a_0$, else 0
- $\gamma a$ controls for the **smooth trend** in the running variable
- $\rho$ = the **vertical jump** at the cutoff $\to$ our causal estimate

. . .

::: {.callout-tip}
## What makes RD special
Treatment is a deterministic function of the running variable. Control for the running variable correctly $\to$ no classical OVB. The *whole* identification argument lives in modeling that one trend.
:::


## Estimating It {.smaller}

```{r}
#| echo: true

simple_rd <- feols(death_rate ~ over21 + months,
                   vcov = "HC1",
                   data = mlda_data)

summary(simple_rd)
```

- Coefficient on `over21` $\to$ estimate of the MLDA jump
- Should land near the true $\rho = 7.7$


## Visualizing the Fit

```{r}
#| label: fig-simple-fit
#| fig-width: 9
#| fig-height: 4.6

mlda_data <- mlda_data |>
    mutate(yhat_linear = predict(simple_rd))

ggplot(mlda_data, aes(x = months, y = death_rate, color = factor(over21))) +
    geom_point(size = 2.5, alpha = 0.7) +
    geom_line(aes(y = yhat_linear), linewidth = 1.3) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_color_manual(values = c("0" = "grey50", "1" = "#0072B2"),
                       labels = c("Under 21", "Over 21")) +
    annotate("segment",
             x = 0.1, xend = 0.1,
             y = 90, yend = 90 + coef(simple_rd)["over21"],
             arrow = arrow(type = "closed", length = unit(0.25, "cm"),
                           ends = "both"),
             color = "#e7298a", linewidth = 1.1) +
    annotate("text", x = 1.5,
             y = 95 + coef(simple_rd)["over21"] / 2,
             label = paste0("ρ ≈ ",
                            round(coef(simple_rd)["over21"], 1)),
             color = "#e7298a", hjust = 0, size = 5.5,
             fontface = "bold") +
    labs(title = "The RD jump *is* the causal effect",
         x = "Age in months (0 = 21st birthday)",
         y = "Death rate per 100,000",
         color = NULL) +
    theme(legend.position = "bottom") +
    coord_cartesian(ylim = c(75, 105))
```



## Linear Isn't Always Enough {.smaller}

```{r}
#| label: fig-three-panels
#| fig-width: 10
#| fig-height: 4.4

set.seed(101)
make_panel <- function(type) {
    x <- seq(0.01, 0.99, length.out = 80)
    d <- as.numeric(x >= 0.5)
    y <- switch(type,
        "A" = 0.3 + 0.8 * x + 0.6 * d + rnorm(length(x), 0, 0.05),
        "B" = 0.0 + 2.2 * sqrt(x) - 0.9 * x + 0.5 * d +
              rnorm(length(x), 0, 0.05),
        "C" = 0.4 + 0.4 * x + 6 * (x - 0.5)^2 * sign(x - 0.5) +
              rnorm(length(x), 0, 0.04)
    )
    tibble(x = x, y = y, d = d, panel = paste("Panel", type))
}

three_panels <- bind_rows(make_panel("A"), make_panel("B"), make_panel("C"))

three_labels <- tibble(
    panel = c("Panel A", "Panel B", "Panel C"),
    label = c("True jump,\nlinear trend",
              "True jump,\nnonlinear trend",
              "No jump,\nbut nonlinear trend")
)

ggplot(three_panels, aes(x = x, y = y)) +
    geom_point(aes(color = factor(d)), size = 1.6, alpha = 0.7) +
    geom_smooth(aes(group = d), method = "loess", span = 0.9,
                se = FALSE, linewidth = 0.6, color = "gray40",
                linetype = "dotted") +
    geom_smooth(aes(group = d, color = factor(d)),
                method = "lm", se = FALSE, linewidth = 1.1) +
    geom_vline(xintercept = 0.5, linetype = "dashed") +
    geom_text(data = three_labels,
              aes(x = 0.05, y = 2.55, label = label),
              hjust = 0, size = 4, inherit.aes = FALSE) +
    facet_wrap(~ panel) +
    scale_color_manual(values = c("0" = "grey50", "1" = "#0072B2"),
                       guide = "none") +
    labs(x = "Running variable (X)", y = "Outcome (Y)",
         caption = "Solid lines = naive linear fit on each side. Dotted = true smooth shape.") +
    theme(strip.text = element_text(face = "bold"),
          plot.caption = element_text(size = 10, color = "gray40"))
```

- Panel C is the danger: a **smooth curve** with no real jump can *look* like a jump under a naive linear fit



## Two Ways to Handle Nonlinearity {.smaller}

:::: {.columns}
::: {.column width="50%"}
**1. Parametric**

- Add polynomials & interactions
- Let the slope (and curvature) bend differently on each side of the cutoff
- Uses *all* the data
:::

::: {.column width="50%"}
**2. Nonparametric**

- Restrict to a narrow window around the cutoff
- Over a small window, any smooth curve looks linear
- Uses only data *near* the cutoff
:::
::::

. . .

- Both are useful. Best practice: report **both** and show the answer is stable.


## Parametric: Three Specifications {.smaller}

```{r}
#| echo: true

rd_linear     <- feols(death_rate ~ over21 + months,
                       vcov = "HC1", data = mlda_data)

rd_linear_int <- feols(death_rate ~ over21 + months + months:over21,
                       vcov = "HC1", data = mlda_data)

rd_quad       <- feols(death_rate ~ over21 + months + I(months^2) +
                           months:over21 + I(months^2):over21,
                       vcov = "HC1", data = mlda_data)
```

. . .

- **Linear**: same slope on each side
- **Linear, interacted**: different slopes on each side
- **Quadratic, interacted**: different curvatures, too


## Parametric: Visual Comparison

```{r}
#| label: fig-parametric
#| fig-width: 10
#| fig-height: 4.4

mlda_fits <- mlda_data |>
    mutate(
        fit_linear = predict(rd_linear),
        fit_lint   = predict(rd_linear_int),
        fit_quad   = predict(rd_quad)
    ) |>
    select(months, over21, death_rate,
           fit_linear, fit_lint, fit_quad) |>
    pivot_longer(starts_with("fit_"), names_to = "model", values_to = "yhat") |>
    mutate(model = recode(model,
        fit_linear = "Linear",
        fit_lint   = "Linear, interacted",
        fit_quad   = "Quadratic, interacted"
    ),
    model = factor(model, levels = c("Linear", "Linear, interacted",
                                      "Quadratic, interacted")))

ggplot(mlda_fits, aes(x = months)) +
    geom_point(aes(y = death_rate, color = factor(over21)),
               size = 1.7, alpha = 0.55) +
    geom_line(aes(y = yhat, group = interaction(model, over21)),
              linewidth = 1.1, color = "#d95f02") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    facet_wrap(~ model) +
    scale_color_manual(values = c("0" = "grey50", "1" = "#0072B2"),
                       guide = "none") +
    labs(x = "Age in months (0 = 21st birthday)",
         y = "Death rate per 100,000") +
    theme(strip.text = element_text(face = "bold"))
```

- Estimates are stable across specs $\to$ **a hallmark of credible RD**

. . .

::: {.callout-warning}
## Don't go higher than quadratic
4th- or 5th-order polynomials chase noise near the boundaries and *invent* jumps. Gelman & Imbens (2019) recommend stopping at quadratic and using local linear regression instead.
:::



# Nonparametric RD


## Bandwidth: A Bias-Variance Tradeoff {.smaller}

- Restrict to $|a - a_0| \leq b$ for some bandwidth $b$
- Then run a simple linear RD inside the window

. . .

:::: {.columns}
::: {.column width="50%"}
**Narrow bandwidth**

- Less data $\to$ bigger SE
- Less room for nonlinearity to bite
- $\to$ **lower bias, higher variance**
:::

::: {.column width="50%"}
**Wide bandwidth**

- More data $\to$ smaller SE
- More chance to mistake curvature for a jump
- $\to$ **higher bias, lower variance**
:::
::::


## Trying Three Bandwidths {.smaller}

```{r}
#| echo: true

rd_b24 <- feols(death_rate ~ over21 + months,
                vcov = "HC1",
                data = subset(mlda_data, abs(months) <= 24))

rd_b12 <- feols(death_rate ~ over21 + months,
                vcov = "HC1",
                data = subset(mlda_data, abs(months) <= 12))

rd_b06 <- feols(death_rate ~ over21 + months,
                vcov = "HC1",
                data = subset(mlda_data, abs(months) <= 6))
```

. . .

```{r}
#| label: fig-bandwidths
#| fig-width: 9
#| fig-height: 3.4

bw_results <- tibble(
    Bandwidth = factor(c("24 months", "12 months", "6 months"),
                       levels = c("24 months", "12 months", "6 months")),
    Estimate  = c(coef(rd_b24)["over21"],
                  coef(rd_b12)["over21"],
                  coef(rd_b06)["over21"]),
    SE        = c(summary(rd_b24)$se["over21"],
                  summary(rd_b12)$se["over21"],
                  summary(rd_b06)$se["over21"])
) |>
    mutate(lo = Estimate - 1.96 * SE,
           hi = Estimate + 1.96 * SE)

ggplot(bw_results, aes(x = Bandwidth, y = Estimate)) +
    geom_hline(yintercept = 7.7, linetype = "dashed",
               color = "darkgreen", linewidth = 1) +
    geom_point(color = "#0072B2", size = 5) +
    geom_errorbar(aes(ymin = lo, ymax = hi),
                  width = 0.15, color = "#0072B2", linewidth = 1) +
    annotate("text", x = 3.4, y = 7.7, label = "Truth = 7.7",
             color = "darkgreen", hjust = 0, size = 4.5) +
    coord_flip(clip = "off") +
    labs(title = "Estimates stable; SEs grow as bandwidth shrinks",
         x = NULL, y = "MLDA effect on death rate") +
    theme(plot.margin = margin(10, 80, 10, 10))
```


# Validity Checks


## Falsification: Cause-Specific Effects {.smaller}

- If we're really picking up the effect of *legal drinking*, the jump should appear in alcohol-driven causes (motor vehicles, suicide) but **not** in unrelated causes (heart disease, cancer)

. . .

```{r}
#| label: fig-falsification
#| fig-width: 9
#| fig-height: 3.7

set.seed(99)
cause_data <- tibble(
    months = seq(-24, 23, by = 1),
    over21 = as.numeric(months >= 0),
    mva    = 25 + 0.05 * months + 4.5 * over21 +
             rnorm(length(months), 0, 1.2),
    intern = 15 + 0.10 * months + 0.4 * over21 +
             rnorm(length(months), 0, 1.0)
) |>
    pivot_longer(c(mva, intern), names_to = "cause", values_to = "rate") |>
    mutate(cause = recode(cause,
                          mva    = "Motor vehicle deaths",
                          intern = "Internal-cause deaths"))

ggplot(cause_data, aes(x = months, y = rate, color = cause)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(aes(group = interaction(cause, over21)),
                method = "lm", formula = y ~ poly(x, 2),
                se = FALSE, linewidth = 1.1) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_color_manual(values = c("Motor vehicle deaths" = "#0072B2",
                                  "Internal-cause deaths" = "grey50")) +
    labs(x = "Age in months", y = "Rate per 100,000",
         color = "Cause") +
    theme(legend.position = "bottom")
```

- Big jump where alcohol matters; nothing where it doesn't $\to$ design looks credible


## Manipulation: Density Tests {.smaller}

- If units can **engineer** their position relative to the cutoff (think: a manager rounding test scores up), the design breaks
- Diagnostic: plot the density of the running variable. A pile-up just on one side is a red flag.

. . .

```{r}
#| label: fig-mccrary
#| fig-width: 9
#| fig-height: 3.4

set.seed(3)
smooth_x <- tibble(x = rnorm(4000),
                   scenario = "Smooth density (no manipulation)")
manip_x <- tibble(x = c(rnorm(3000),
                        rnorm(1000, 0.6, 0.3)),
                  scenario = "Pile-up above cutoff (manipulation)")

bind_rows(smooth_x, manip_x) |>
    ggplot(aes(x = x)) +
    geom_histogram(bins = 40, fill = "#0072B2", color = "white",
                   alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    facet_wrap(~ scenario) +
    labs(x = "Running variable (centered)",
         y = "Number of observations") +
    theme(strip.text = element_text(face = "bold"))
```

- Formal test: McCrary (2008); modern version in the `rddensity` package


## Other Validity Checks {.smaller}

::: {.callout-important}
## A credible RD paper shows...
1. **Covariate balance**: pre-determined characteristics (age, gender, prior outcomes) don't jump at the cutoff
2. **Density smoothness**: no pile-up of the running variable on either side
3. **Robustness**: the jump survives different bandwidths and polynomial orders
4. **Placebo cutoffs**: estimating "effects" at fake cutoffs gives ~ 0
5. **Falsification outcomes**: outcomes treatment shouldn't affect... aren't affected
:::



# Wrapping Up


## Key Takeaways {.smaller}

- **RD turns rigid rules into natural experiments** by comparing units just above and just below a cutoff
- **Sharp RD**: treatment is a deterministic function of a running variable
- The estimate $\rho$ is the **vertical jump** in the outcome at the cutoff
- Two strategies for handling nonlinearity:
  - **Parametric** (polynomials & interactions; stop at quadratic)
  - **Nonparametric** (zoom in on a bandwidth; use `rdrobust` in practice)
- Validate with covariate balance, density smoothness, falsification outcomes, and placebo cutoffs

. . .

::: {.callout-tip}
## The RD pitch in one sentence
*Here is the rule, here is the cutoff, here is the jump --- and here is why nothing else can explain it.*
:::
