---
title: "Bivariate Regression and OLS"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 4 | Bivariate Regression and OLS"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


# From RCTs to Regression

## Where We've Been

- In Unit 3, we learned that **Randomized Control Trials** provide the gold standard for causal inference

- Random assignment eliminates selection bias: the simple difference in means equals the causal effect

. . .

- But most economic questions **cannot** be answered with experiments

- We typically work with **observational data** where treatment is not randomly assigned

## Where We're Going

- Today: **Ordinary Least Squares (OLS)** regression -- the workhorse tool of econometrics

. . .

- We focus on the **bivariate** case (one independent variable) to build intuition

. . .

::: {.callout-important}
## The Big Question
When can we interpret a regression coefficient as a **causal effect** vs. a mere **correlation**?
:::

# Modeling Relationships Between Variables

## What Do We Mean by "Relationship"?

- Suppose we have two variables, $y$ and $x$

- We want to study "how $x$ is related to $y$"

. . .

- We need a precise, mathematical way to do this

. . .

- The most natural approach: what is the **average** value of $y$ for each different value of $x$?

## Conditional Expectations

- This idea is called the **conditional expectation**

. . .

$$E[Y | X = x]$$

. . .

- "If we knew that $X$ took on some specific value $x$, what would we expect $Y$ to be, **on average**?"

. . .

- The conditional expectation is the central quantity we want to model and estimate

## Example: Education and Wages {.smaller}

- $X$ = education level (HS degree, College degree, Professional degree)
- $Y$ = hourly wage

. . .

- For each education level, compute the average wage among people in that group

. . .

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(42)
n <- 150
ed_levels <- c("HS Degree", "College Degree", "Prof. Degree")
education <- sample(ed_levels, size = n, replace = TRUE, prob = c(0.4, 0.4, 0.2))
wage <- case_when(
  education == "HS Degree" ~ rnorm(n, mean = 18, sd = 5),
  education == "College Degree" ~ rnorm(n, mean = 30, sd = 7),
  education == "Prof. Degree" ~ rnorm(n, mean = 50, sd = 10)
)
wage <- pmax(wage, 5)

ed_data <- tibble(education = factor(education, levels = ed_levels), wage = wage)

cond_means <- ed_data |>
  group_by(education) |>
  summarize(mean_wage = mean(wage))

ggplot(ed_data, aes(x = education, y = wage)) +
  geom_jitter(alpha = 0.4, width = 0.2, color = "steelblue") +
  geom_point(data = cond_means, aes(x = education, y = mean_wage),
             color = "red", size = 5, shape = 18) +
  labs(x = "Education Level", y = "Hourly Wage ($)",
       title = "Wages by Education Level",
       subtitle = "Red diamonds show conditional expectations (group averages)") +
  theme_minimal(base_size = 16)
```

## What Do We See?

- The **red diamonds** show the conditional expectations: $E[Y|X=\text{HS}]$, $E[Y|X=\text{College}]$, $E[Y|X=\text{Prof}]$

. . .

- These averages tell us how wages **change** as education increases

. . .

- This is precisely what we want to estimate!


# Visualizing Relationships: Scatter Plots

## Why Scatter Plots? {.smaller}

- When both $x$ and $y$ are **continuous**, a scatter plot is the best way to visualize their relationship

. . .

- Each point represents one observation: its $x$ value and its $y$ value

. . .

- Scatter plots help us:
  - See the **direction** of a relationship (positive or negative)
  - Identify **outliers** or unusual patterns
  - Judge whether a **linear** model is appropriate

## Introduction to `ggplot2` {.smaller}

- We'll use the `ggplot2` package (part of the `tidyverse`) for all our visualizations

. . .

- `ggplot2` builds plots **layer by layer** -- you start with data, then add visual elements on top:

. . .

1. **Start with data:** `ggplot(data = your_data)` -- creates an empty canvas
2. **Add a geometry:** `+ geom_point(...)` for scatter plots, `+ geom_histogram(...)` for histograms, etc.
3. **Map variables to aesthetics:** inside `aes()`, tell R which columns go on each axis (and optionally `color`, `size`, `shape`, etc.)
4. **Polish:** add labels with `labs()` and a clean theme with `theme_minimal()`

. . .

::: {.callout-tip}
## Want More?
The Unit 2 lecture notes cover `ggplot2` in much more detail, including histograms, bar charts, and more customization options. Refer back to those as needed!
:::

## Building a Scatter Plot {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- Let's build a scatter plot step by step. Start with data and add points using `geom_point()`:

. . .

```{r}
#| output-location: fragment
#| fig-height: 3.5

library(tidyverse)

ggplot(data = mtcars) +
  geom_point(aes(x = hp, y = mpg))
```

## Making Publication-Quality Plots {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

Always add proper labels and a clean theme:

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(data = mtcars) +
  geom_point(aes(x = hp, y = mpg)) +
  labs(
    title = "Fuel Efficiency vs. Engine Power",
    x = "Horsepower",
    y = "Miles per Gallon",
  ) +
  theme_minimal()
```

## Adding a Line of Best Fit {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- `geom_smooth(method = "lm")` adds a straight line that "best fits" the data

. . .

```{r}
#| output-location: fragment
#| fig-height: 3

ggplot(data = mtcars) +
  geom_point(aes(x = hp, y = mpg)) +
  geom_smooth(aes(x = hp, y = mpg), method = "lm", se = FALSE) +
  labs(x = "Horsepower", y = "Miles per Gallon",
       title = "MPG vs. Horsepower with Line of Best Fit") +
  theme_minimal()
```

## Try It: Scatter Plot Practice {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**Create a scatter plot of `wt` (weight) vs. `mpg` (fuel efficiency). Add proper labels.**

- *Hint:* Use `geom_point()` + `labs()`

. . .

```{r}
#| output-location: fragment
#| fig-height: 2.5

ggplot(mtcars) +
  geom_point(aes(x = wt, y = mpg)) +
  labs(x = "Weight (1000 lbs)", y = "Miles per Gallon",
       title = "Heavier Cars Get Worse Fuel Efficiency") +
  theme_minimal()
```


# The Population Regression Model

## Writing Down the Population Model {.smaller}

- We start by describing the **true relationship** in the population -- even though we can't observe it directly

. . .

- The simplest regression model relates $y$ to a single variable $x$:

. . .

$$y = \beta_0 + \beta_1 x + \mu$$

. . .

where:

- $y$ is the **dependent variable** (outcome)
- $x$ is the **independent variable** (explanatory)
- $\mu$ is the **error term** (everything else that affects $y$)
- $\beta_0$ and $\beta_1$ are **population parameters** (unknown)

## The Error Term

- The error term $\mu$ is crucial -- it captures **everything** that affects $y$ other than $x$

. . .

- In a wage equation: $wage = \beta_0 + \beta_1 \cdot educ + \mu$

. . .

- $\mu$ might include:
  - Ability and motivation
  - Family connections
  - Luck
  - Work experience

## Example: Education and Wages {.smaller}

- With some assumptions, the regression line traces out the **conditional expectation**: $E[y|x] = \beta_0 + \beta_1 x$

. . .

- We can interpret the coefficients like we would any linear equation:

- $\beta_0$ is the value of $y$ when $x=0$

- $\beta_1$ is the **marginal effect**: a one-unit increase in $x$ changes $y$ by $\beta_1$ units, on average

. . .

- Let's make this concrete:

$$wage = \beta_0 + \beta_1 \cdot educ + \mu$$

- $wage$ = dollars per hour
- $educ$ = years of education

. . .

- If $\beta_1 = 2.50$: each additional year of education is associated with **$2.50 higher hourly wages**, on average


# From Population to Sample

## We Can't See the Population

- The population model $y = \beta_0 + \beta_1 x + \mu$ describes the **true** relationship present in the **population**

. . .

- But $\beta_0$ and $\beta_1$ are **unknown** -- we never observe the whole population

. . .

- Instead, we use our **sample** to compute **estimates**, which we write with "hats":

. . .

- $\hat{\beta}_0$ = our estimate of $\beta_0$
- $\hat{\beta}_1$ = our estimate of $\beta_1$

## Predicted Values and the Fitted Line

- Using our estimates, we construct the **regression line**:

. . .

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$$

. . .

- For any value of $x$, this gives us $\hat{y}$, the **predicted value** of $y$

- This is exactly what `geom_smooth(method = "lm")` draws!

## Residuals: The Prediction Errors {.smaller}

- Our predictions won't be perfect. The **residual** for observation $i$ is:

. . .

$$\hat{\mu}_i = y_i - \hat{y}_i = y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i)$$

. . .

- Positive residual: we **underpredicted** (actual > predicted)
- Negative residual: we **overpredicted** (actual < predicted)

## Visualizing Residuals {.smaller}

```{r}
#| echo: false
#| fig-height: 4.5
#| fig-align: center

reg_fit <- lm(mpg ~ hp, data = mtcars)
mtcars_aug <- mtcars |>
  mutate(fitted = fitted(reg_fit), resid = residuals(reg_fit))

highlight_idx <- c(6, 15, 20)
highlight_data <- mtcars_aug[highlight_idx, ]

ggplot(mtcars_aug, aes(x = hp, y = mpg)) +
  geom_segment(data = highlight_data,
               aes(x = hp, xend = hp, y = fitted, yend = mpg),
               color = "salmon", linewidth = 1.2, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  geom_point(size = 2.5, alpha = 0.6) +
  geom_point(data = highlight_data, size = 3.5, color = "red") +
  annotate("text", x = 220, y = 28, label = "Residuals\n(prediction errors)",
           color = "salmon", fontface = "italic", size = 5) +
  labs(x = "Horsepower", y = "Miles per Gallon",
       title = "Regression Line with Residuals",
       subtitle = "Dashed lines show the gap between actual and predicted values") +
  theme_minimal(base_size = 16)
```

## The Goal: Minimize the Residuals

- We want to make residuals as small as possible

. . .

- Can't just minimize $\sum \hat{\mu}_i$ -- positive and negative residuals would cancel out!

. . .

- Instead, minimize the **Sum of Squared Residuals (SSR)**:

. . .

$$SSR = \sum_{i=1}^{n} \hat{\mu}_i^2 = \sum_{i=1}^{n} (y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i)^2$$

. . .

- This is where the name **"Ordinary Least Squares"** comes from!


# The OLS Estimator

## The Minimization Problem

- Given $n$ observations $\{(x_i, y_i): i = 1, ..., n\}$, find $\hat{\beta}_0$ and $\hat{\beta}_1$ that minimize:

. . .

$$\min_{\hat{\beta}_0, \hat{\beta}_1} \sum_{i=1}^{n} (y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i)^2$$

. . .

- This is a calculus optimization problem -- take partial derivatives, set equal to zero, solve

. . .

::: {.callout-tip}
## Full Derivation
The step-by-step derivation using first-order conditions is in Section 4.4 of the lecture notes. We'll skip the algebra here and focus on the **result** and its **intuition**.
:::

## The OLS Formulas {.smaller}

- Solving the minimization problem gives us:

. . .

::: {.callout-important}
## The OLS Slope
$$\hat{\beta}_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2} = \frac{Cov(x, y)}{Var(x)}$$
:::

. . .

**The OLS Intercept:**

$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$$

. . .

- The regression line always passes through the point $(\bar{x}, \bar{y})$ -- the "center" of the data!

## An Important Observation

- The slope can also be written as:

. . .

$$\hat{\beta}_1 = \rho_{xy} \frac{\sigma_y}{\sigma_x}$$

. . .

- Bivariate OLS is essentially a **scaled version of the correlation** between $x$ and $y$

. . .

- And as we discussed in Unit 1: **correlation does not imply causation!**

. . .

- We must be very careful about interpreting bivariate OLS results causally

## Visualizing OLS: Step 1 -- The Data {.smaller}

Let's build up the intuition for what OLS does, step by step.

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(15)
demo_x <- runif(20, 1, 10)
demo_y <- 2 + 1.5 * demo_x + rnorm(20, 0, 2)
demo_data <- tibble(x = demo_x, y = demo_y)
demo_fit <- lm(y ~ x, data = demo_data)
demo_data <- demo_data |> mutate(fitted = fitted(demo_fit))

ggplot(demo_data, aes(x = x, y = y)) +
  geom_point(size = 3) +
  labs(x = "x", y = "y",
       title = "Step 1: Start with the data") +
  theme_minimal(base_size = 16)
```

- We have 20 observations. How should we draw a line to summarize this relationship?

## Visualizing OLS: Step 2 -- Try a Line {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

ggplot(demo_data, aes(x = x, y = y)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  geom_point(size = 3) +
  labs(x = "x", y = "y",
       title = "Step 2: Fit a line through the data") +
  theme_minimal(base_size = 16)
```

- The blue line is our "best guess" -- but what makes it "best"?

## Visualizing OLS: Step 3 -- One Residual {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

# Pick one point to highlight
hi <- demo_data |> slice(5)

ggplot(demo_data, aes(x = x, y = y)) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  geom_segment(data = hi, aes(x = x, xend = x, y = fitted, yend = y),
               color = "salmon", linewidth = 1.5, linetype = "dashed") +
  geom_point(size = 3, alpha = 0.4) +
  geom_point(data = hi, size = 4, color = "red") +
  annotate("text", x = hi$x + 0.5, y = (hi$y + hi$fitted) / 2,
           label = paste0("Residual\nû = ", round(hi$y - hi$fitted, 1)),
           color = "salmon", fontface = "bold", size = 4.5) +
  labs(x = "x", y = "y",
       title = "Step 3: Measure the prediction error (residual) for one point") +
  theme_minimal(base_size = 16)
```

- The **residual** is the vertical distance between the point and the line

## Visualizing OLS: Step 4 -- Square the Residual {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

sq_size <- abs(hi$y - hi$fitted)

ggplot(demo_data, aes(x = x, y = y)) +
  geom_rect(data = hi,
            aes(xmin = x, xmax = x + sq_size,
                ymin = pmin(y, fitted), ymax = pmax(y, fitted)),
            fill = "salmon", alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  geom_point(size = 3, alpha = 0.4) +
  geom_point(data = hi, size = 4, color = "red") +
  annotate("text", x = hi$x + sq_size / 2, y = (hi$y + hi$fitted) / 2,
           label = paste0("û² = ", round((hi$y - hi$fitted)^2, 1)),
           color = "darkred", fontface = "bold", size = 4.5) +
  labs(x = "x", y = "y",
       title = "Step 4: Square it -- this is what OLS minimizes") +
  theme_minimal(base_size = 16)
```

- We **square** the residual so that positive and negative errors don't cancel

## Visualizing OLS: Step 5 -- All Squared Residuals {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

ggplot(demo_data, aes(x = x, y = y)) +
  geom_rect(aes(xmin = x, xmax = x + abs(y - fitted),
                ymin = pmin(y, fitted), ymax = pmax(y, fitted)),
            fill = "salmon", alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  geom_point(size = 3) +
  labs(x = "x", y = "y",
       title = "Step 5: OLS minimizes the TOTAL area of all squares",
       subtitle = "Each pink square = one squared residual. OLS makes the total area as small as possible.") +
  theme_minimal(base_size = 16)
```

. . .

- OLS chooses $\hat{\beta}_0$ and $\hat{\beta}_1$ to make the **total pink area** as small as possible!


# Computing OLS in R

## The `lm()` Function {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

- In R, the `lm()` function ("linear model") performs OLS regression:

. . .

```{r}
#| echo: true
#| eval: false

# General syntax
lm(y ~ x, data = your_data)
```

. . .

- The formula `y ~ x` says: "$y$ is the dependent variable, $x$ is the independent variable"
- An **intercept** is included by default

## Example: MPG and Horsepower {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

```{r}
#| output-location: fragment

# Estimate the regression
lm_mpg <- lm(mpg ~ hp, data = mtcars)

# View the results
summary(lm_mpg)
```

## Reading the Output {.smaller}

- $\hat{\beta}_0 = 30.10$: A car with **zero** horsepower is predicted to get 30.1 MPG (not meaningful here!)

. . .

- $\hat{\beta}_1 = -0.068$: Each additional unit of horsepower is associated with about **0.07 fewer miles per gallon**

. . .

- **R-squared = 0.60**: Horsepower explains about **60%** of the variation in MPG

. . .

- We'll discuss standard errors and p-values in the next unit

## Visualizing the Regression Output {.smaller}

```{r}
#| fig-height: 3.5
#| echo: false

ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(x = "Horsepower", y = "Miles per Gallon",
       title = "OLS Regression: MPG on Horsepower",
       subtitle = paste0("ŷ = ", round(coef(lm_mpg)[1], 1),
                         " + (", round(coef(lm_mpg)[2], 3), ") × hp")) +
  theme_minimal()
```

## Try It: Run Your Own Regression {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**Regress `mpg` on `wt` (weight) using the `mtcars` data. What does the slope tell you?**

- *Hint:* Use `lm(mpg ~ wt, data = mtcars)` and `summary()`

. . .

```{r}
#| output-location: fragment

lm_wt <- lm(mpg ~ wt, data = mtcars)
coef(lm_wt)
```

. . .

- Each additional 1,000 lbs of weight is associated with about **5.3 fewer** miles per gallon


# Goodness-of-Fit: R-Squared

## How Well Does Our Line Fit?

- **R-squared** ($R^2$) measures the fraction of variation in $y$ explained by the regression

. . .

$$R^2 = 1 - \frac{SSR}{SST}$$

. . .

where:

- $SST = \sum(y_i - \bar{y})^2$ = **total** variation in $y$
- $SSR = \sum\hat{\mu}_i^2$ = **unexplained** variation (residuals)

## R-Squared: Nearly Useless Regression {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(99)
n_demo <- 80
x_demo <- runif(n_demo, 0, 10)

y_noise <- 5 + 0.3 * x_demo + rnorm(n_demo, 0, 8)
r2_noise <- round(summary(lm(y_noise ~ x_demo))$r.squared, 2)

ggplot(tibble(x = x_demo, y = y_noise), aes(x, y)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  labs(title = paste0("R² = ", r2_noise, " — Almost no explanatory power"),
       subtitle = "x explains virtually none of the variation in y",
       x = "x", y = "y") +
  theme_minimal(base_size = 16)
```

- The data is a **cloud** -- knowing $x$ barely helps predict $y$

## R-Squared: Moderate Fit {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

y_mod <- 5 + 1.5 * x_demo + rnorm(n_demo, 0, 3)
r2_mod <- round(summary(lm(y_mod ~ x_demo))$r.squared, 2)

ggplot(tibble(x = x_demo, y = y_mod), aes(x, y)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  labs(title = paste0("R² = ", r2_mod, " — Clear trend, but lots of noise"),
       subtitle = "x explains some variation, but many points are far from the line",
       x = "x", y = "y") +
  theme_minimal(base_size = 16)
```

- A clear **upward trend**, but individual predictions would be imprecise

## R-Squared: Near-Perfect Fit {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

y_tight <- 5 + 1.5 * x_demo + rnorm(n_demo, 0, 0.3)
r2_tight <- round(summary(lm(y_tight ~ x_demo))$r.squared, 2)

ggplot(tibble(x = x_demo, y = y_tight), aes(x, y)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1.2) +
  labs(title = paste0("R² = ", r2_tight, " — Nearly perfect fit"),
       subtitle = "Almost all variation in y is explained by x",
       x = "x", y = "y") +
  theme_minimal(base_size = 16)
```

- Points hug the line tightly -- $x$ explains almost everything about $y$

## Caution About R-Squared

::: {.callout-warning}
## Don't Over-Interpret R-Squared

- A **high** $R^2$ does NOT mean we have a causal estimate -- it just means $x$ is correlated with $y$

- A **low** $R^2$ does NOT mean our estimate is wrong or unimportant

- In cross-sectional data, $R^2$ values of 0.1 to 0.3 are common and can still represent economically meaningful relationships
:::


# What Makes a Good Estimator?

## Can We Trust Our Estimates? {.smaller}

- We now know how to compute $\hat{\beta}_0$ and $\hat{\beta}_1$ -- but should we trust the numbers?

. . .

- Remember: population parameters ($\beta$) are fixed, but *different samples give different estimates* ($\hat{\beta}$)

. . .

- The property we want is **unbiasedness**: on average, our estimates hit the target

. . .

::: {.callout-important}
## Unbiasedness
$E[\hat{\beta}_1] = \beta_1$ — no systematic tendency to over- or under-estimate. Individual shots miss, but the **center** of all shots lands on the bullseye.
:::

. . .

- The key question: **under what conditions** is OLS unbiased? That depends on **assumptions**.


# The Gauss-Markov Assumptions

## Assumptions 1--3: The "Mechanical" Requirements {.smaller}

These first three assumptions are relatively easy to check. See the lecture notes for details and examples.

. . .

| Assumption | What it says | In plain English |
|------------|-------------|-----------------|
| 1. Linear in parameters | $y = \beta_0 + \beta_1 x + \mu$ | The model is linear in the $\beta$'s (not necessarily in $x$!) |
| 2. Random sampling | Obs. are i.i.d. draws | Our sample is representative of the population |
| 3. Variation in $x$ | $\sum(x_i - \bar{x})^2 > 0$ | $x$ isn't the same value for everyone |

. . .

- These are usually satisfied in practice -- the **hard** assumption is next

## Assumption 4: Zero Conditional Mean

$$E[\mu | x] = 0$$

. . .

::: {.callout-important}
## The Most Important Assumption
The average value of all unobserved factors affecting $y$ is the **same** regardless of the value of $x$.
:::

## What Does "Zero Conditional Mean" Mean? {.smaller}

- In words: $E[\mu | x] = 0$ means the **error term is unrelated to** $x$

. . .

- For *every* possible value of $x$, the average of all unobserved factors is zero

. . .

- Think of it this way: if you could gather all people with $x = 10$ and average their $\mu$ values, you'd get zero. Same for $x = 20$. Same for any $x$.

## Visualizing the Zero Conditional Mean {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(50)
n_zcm <- 300
x_zcm <- runif(n_zcm, 0, 10)
mu_good <- rnorm(n_zcm, 0, 2)
y_zcm <- 3 + 2 * x_zcm + mu_good

zcm_data <- tibble(x = x_zcm, y = y_zcm, mu = mu_good)

# Bin the data and compute mean residuals
zcm_bins <- zcm_data |>
  mutate(x_bin = cut(x, breaks = 5)) |>
  group_by(x_bin) |>
  summarize(mean_mu = mean(mu), mid_x = mean(x), .groups = "drop")

ggplot(zcm_data, aes(x = x, y = mu)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_point(data = zcm_bins, aes(x = mid_x, y = mean_mu),
             color = "red", size = 5, shape = 18) +
  labs(x = "x", y = "Error term (μ)",
       title = "When E[μ | x] = 0 holds",
       subtitle = "Red diamonds: average μ within each bin ≈ 0 regardless of x") +
  theme_minimal(base_size = 16)
```

- No matter where we look along the x-axis, errors average to **zero**

## When the Zero Conditional Mean Fails {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(50)
mu_bad <- rnorm(n_zcm, 0, 2) + 0.8 * x_zcm - 4  # mu is correlated with x
y_bad <- 3 + 2 * x_zcm + mu_bad

zcm_bad <- tibble(x = x_zcm, y = y_bad, mu = mu_bad)

zcm_bins_bad <- zcm_bad |>
  mutate(x_bin = cut(x, breaks = 5)) |>
  group_by(x_bin) |>
  summarize(mean_mu = mean(mu), mid_x = mean(x), .groups = "drop")

ggplot(zcm_bad, aes(x = x, y = mu)) +
  geom_point(alpha = 0.3, color = "coral") +
  geom_hline(yintercept = 0, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_point(data = zcm_bins_bad, aes(x = mid_x, y = mean_mu),
             color = "darkred", size = 5, shape = 18) +
  labs(x = "x", y = "Error term (μ)",
       title = "When E[μ | x] = 0 FAILS",
       subtitle = "Average μ is negative at low x and positive at high x — errors are correlated with x!") +
  theme_minimal(base_size = 16)
```

- The average error **rises** with $x$ -- the error is correlated with our independent variable

## Why Does This Matter for Wages? {.smaller}

- Consider the wage equation: $wage = \beta_0 + \beta_1 \cdot educ + \mu$

. . .

- For $E[\mu | educ] = 0$ we need:

. . .

- Among people with 8 years of education: average $\mu$ = 0
- Among people with 12 years of education: average $\mu$ = 0
- Among people with 16 years of education: average $\mu$ = 0

. . .

- Is this plausible?

- People with more education tend to have higher **ability** (which is in $\mu$)

. . .

- So $E[\mu | educ = 16] > E[\mu | educ = 8]$ -- the assumption **fails**

. . .

- This is the fundamental challenge of causal inference with observational data!

## Summary: The Four Assumptions for Unbiasedness

| Assumption | What it says |
|------------|-------------|
| 1. Linear in parameters | $y = \beta_0 + \beta_1 x + \mu$ |
| 2. Random sampling | Observations are independent draws |
| 3. Variation in $x$ | $\sum(x_i - \bar{x})^2 > 0$ |
| **4. Zero conditional mean** | $E[\mu \| x] = 0$ |

. . .

- Under these four assumptions, **OLS is unbiased**: $E[\hat{\beta}_1] = \beta_1$


# Seeing Unbiasedness in Action

## Simulating Unbiasedness {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

Let's prove unbiasedness to ourselves! First, create a "population" with **known** parameters:

. . .

```{r}
#| output-location: fragment

set.seed(1248)
n_pop <- 1000
pop_data <- tibble(
  x = sample(1:100, size = n_pop, replace = TRUE),
  y = 3.5 + 2.21 * x + rnorm(n_pop)
)

# The TRUE slope is 2.21
```

## Drawing Many Samples {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

Now draw 500 random samples and estimate $\hat{\beta}_1$ each time. We use a `for()` loop to repeat the same set of steps a fixed number of times. The loop runs the code inside `{}` once for each value of `i` (from 1 to 500), so we don't have to copy and paste the same code 500 times:

```{r}
#| output-location: fragment

set.seed(812476)
n_sims <- 500
sample_size <- 25

# Create a dataframe to store our results
sim_results <- tibble(iteration = integer(), b1_hat = numeric())

# Run the simulation
for (i in 1:n_sims) {
  samp <- slice_sample(pop_data, n = sample_size, replace = TRUE)
  b1_hat <- coef(lm(y ~ x, data = samp))[2]
  sim_results <- sim_results |> add_row(iteration = i, b1_hat = b1_hat)
}

mean(sim_results$b1_hat)
```

## Visualizing Unbiasedness {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

ggplot(sim_results, aes(x = b1_hat)) +
  geom_histogram(binwidth = 0.015, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 2.21, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(sim_results$b1_hat), color = "forestgreen",
             linewidth = 1.2) +
  annotate("text", x = 2.275, y = 40, label = paste0("True β = 2.21"),
           color = "red", size = 5, fontface = "bold") +
  annotate("text", x = 2.145, y = 35,
           label = paste0("Mean = ", round(mean(sim_results$b1_hat), 3)),
           color = "forestgreen", size = 5, fontface = "bold") +
  labs(x = expression(hat(beta)[1]), y = "Count",
       title = "Distribution of OLS Slope Estimates Across 500 Samples",
       subtitle = "Each bar represents estimates that fell in a narrow range") +
  theme_minimal(base_size = 16)
```

. . .

- Individual estimates vary, but the **average** equals the true value!


# So What? Bias, OVB, and Causality

## When Assumption 4 Fails {.smaller}

- When $E[\mu | x] \neq 0$, OLS is **biased** -- and the most common reason is **omitted variable bias**

. . .

- If a variable that belongs in the model is left out AND is correlated with $x$, our slope picks up the effect of both

. . .

- Example: estimating the return to education while omitting ability → $\hat{\beta}_1$ **overstates** the causal effect

. . .

- Remember from Unit 3: this is essentially **selection bias** -- people who choose more education also tend to have higher ability

## The Big Picture {.smaller}

- When $E[\mu | x] = 0$: $\hat{\beta}_1$ is an unbiased estimate of the **causal effect**

- When $E[\mu | x] \neq 0$: $\hat{\beta}_1$ measures **correlation**, not causation

. . .

::: {.callout-important}
## Correlation vs. Causation
OLS always estimates *something*. Whether that something is a **causal effect** or merely a **correlation** depends entirely on whether the assumptions hold. Most of econometrics is about finding clever ways to make the zero conditional mean assumption more plausible.
:::

. . .

::: {.callout-tip}
## Coming Up Next
**Multivariate regression** gives us the tools to **control for** confounding variables by adding more $x$'s to our model -- directly addressing omitted variable bias.
:::


# Variance, Standard Errors, and BLUE

## Beyond Unbiasedness

- Unbiasedness tells us OLS is "right on average" -- but individual estimates can vary a lot

. . .

- We also care about **precision**: how much do estimates bounce around from sample to sample?

. . .

- This is measured by the **variance** of the estimator

## Assumption 5: Homoskedasticity {.smaller}

- To derive the variance of OLS, we need one more assumption:

. . .

$$Var(\mu | x) = \sigma^2$$

. . .

- The "spread" of errors is **constant** across all values of $x$

. . .

```{r}
#| echo: false
#| fig-height: 3
#| fig-align: center
#| layout-ncol: 2

set.seed(77)
n_h <- 100
x_h <- runif(n_h, 1, 10)
y_homo <- 2 + 1.5 * x_h + rnorm(n_h, 0, 2)
y_hetero <- 2 + 1.5 * x_h + rnorm(n_h, 0, 1.2 * x_h)

ggplot(tibble(x = x_h, y = y_homo), aes(x, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(title = "Homoskedasticity", subtitle = "Constant spread",
       x = "x", y = "y") +
  theme_minimal(base_size = 14)

ggplot(tibble(x = x_h, y = y_hetero), aes(x, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(title = "Heteroskedasticity", subtitle = "Spread increases with x",
       x = "x", y = "y") +
  theme_minimal(base_size = 14)
```

## Deriving the Variance of $\hat{\beta}_1$: Sketch {.smaller}

- Start from the slope formula. Using the fact that $y_i = \beta_0 + \beta_1 x_i + \mu_i$:

. . .

$$\hat{\beta}_1 = \beta_1 + \frac{\sum(x_i - \bar{x})\mu_i}{\sum(x_i - \bar{x})^2}$$

. . .

- The only random piece is $\mu_i$, so the variance comes from the second term:

. . .

$$Var(\hat{\beta}_1) = Var\left(\frac{\sum(x_i - \bar{x})\mu_i}{\sum(x_i - \bar{x})^2}\right)$$

## Deriving the Variance: Result {.smaller}

- Since $\sum(x_i - \bar{x})^2$ is a constant (fixed in our sample) and the $\mu_i$ are independent with constant variance $\sigma^2$:

. . .

$$Var(\hat{\beta}_1) = \frac{\sum(x_i - \bar{x})^2 \cdot \sigma^2}{\left[\sum(x_i - \bar{x})^2\right]^2} = \frac{\sigma^2}{\sum(x_i - \bar{x})^2}$$

. . .

::: {.callout-important}
## Variance of the OLS Slope
$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$$

where $\sigma^2 = Var(\mu)$ is the error variance and $SST_x = \sum(x_i - \bar{x})^2$ is the total variation in $x$.
:::

## What Affects Precision? {.smaller}

$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$$

. . .

Two key insights:

. . .

1. **More noise** ($\sigma^2$ ↑) → **less** precise estimates (wider distribution)

. . .

2. **More variation in $x$** ($SST_x$ ↑) → **more** precise estimates (tighter distribution)

. . .

- Let's see this with simulations!

## Effect of Error Variance on Precision {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(444)
n_var <- 500

# Population with low noise
pop_low <- tibble(x = sample(1:100, 1000, replace = TRUE),
                  y = 5 + 2 * x + rnorm(1000, 0, 5))
# Population with high noise
pop_high <- tibble(x = sample(1:100, 1000, replace = TRUE),
                   y = 5 + 2 * x + rnorm(1000, 0, 50))

sim_var <- tibble(
  iteration = 1:n_var,
  low_noise = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_low, n = 30, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  }),
  high_noise = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_high, n = 30, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  })
)

sim_var_long <- sim_var |>
  pivot_longer(cols = c(low_noise, high_noise),
               names_to = "scenario", values_to = "estimate") |>
  mutate(scenario = ifelse(scenario == "low_noise",
                           paste0("Low noise (σ = 5), SD = ", round(sd(sim_var$low_noise), 2)),
                           paste0("High noise (σ = 50), SD = ", round(sd(sim_var$high_noise), 2))))

ggplot(sim_var_long, aes(x = estimate, fill = scenario)) +
  geom_histogram(binwidth = 0.06, alpha = 0.6, position = "identity", color = "white") +
  geom_vline(xintercept = 2, color = "red", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 2, y = 75, label = "True β = 2", color = "red",
           size = 5, fontface = "bold", hjust = -0.1) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(x = expression(hat(beta)[1]), y = "Count", fill = "",
       title = "More noise → less precise estimates",
       subtitle = "Both are unbiased (centered on 2), but high noise has much wider spread") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## Effect of Variation in $x$ on Precision {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(555)

# Population with low variation in x
pop_low_x <- tibble(x = sample(40:60, 1000, replace = TRUE),
                    y = 5 + 2 * x + rnorm(1000, 0, 15))
# Population with high variation in x
pop_high_x <- tibble(x = sample(1:100, 1000, replace = TRUE),
                     y = 5 + 2 * x + rnorm(1000, 0, 15))

sim_x <- tibble(
  iteration = 1:n_var,
  low_var_x = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_low_x, n = 30, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  }),
  high_var_x = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_high_x, n = 30, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  })
)

sim_x_long <- sim_x |>
  pivot_longer(cols = c(low_var_x, high_var_x),
               names_to = "scenario", values_to = "estimate") |>
  mutate(scenario = ifelse(scenario == "low_var_x",
                           paste0("x range: 40-60, SD = ", round(sd(sim_x$low_var_x), 2)),
                           paste0("x range: 1-100, SD = ", round(sd(sim_x$high_var_x), 2))))

ggplot(sim_x_long, aes(x = estimate, fill = scenario)) +
  geom_histogram(binwidth = 0.06, alpha = 0.6, position = "identity", color = "white") +
  geom_vline(xintercept = 2, color = "red", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 2, y = 70, label = "True β = 2", color = "red",
           size = 5, fontface = "bold", hjust = -0.1) +
  scale_fill_manual(values = c("coral", "forestgreen")) +
  labs(x = expression(hat(beta)[1]), y = "Count", fill = "",
       title = "More variation in x → more precise estimates",
       subtitle = "Both are unbiased, but wider x range gives tighter estimates") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## Effect of Sample Size on Precision {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(666)

pop_n <- tibble(x = sample(1:100, 2000, replace = TRUE),
                y = 5 + 2 * x + rnorm(2000, 0, 20))

sim_n <- tibble(
  iteration = 1:n_var,
  n_15 = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_n, n = 15, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  }),
  n_200 = map_dbl(1:n_var, function(i) {
    samp <- slice_sample(pop_n, n = 200, replace = TRUE)
    coef(lm(y ~ x, data = samp))[2]
  })
)

sim_n_long <- sim_n |>
  pivot_longer(cols = c(n_15, n_200),
               names_to = "scenario", values_to = "estimate") |>
  mutate(scenario = ifelse(scenario == "n_15",
                           paste0("n = 15, SD = ", round(sd(sim_n$n_15), 2)),
                           paste0("n = 200, SD = ", round(sd(sim_n$n_200), 2))))

ggplot(sim_n_long, aes(x = estimate, fill = scenario)) +
  geom_histogram(binwidth = 0.06, alpha = 0.6, position = "identity", color = "white") +
  geom_vline(xintercept = 2, color = "red", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 2, y = 80, label = "True β = 2", color = "red",
           size = 5, fontface = "bold", hjust = -0.1) +
  scale_fill_manual(values = c("coral", "forestgreen")) +
  labs(x = expression(hat(beta)[1]), y = "Count", fill = "",
       title = "Larger samples → more precise estimates",
       subtitle = "Larger n means more variation in x (higher SST_x) → tighter distribution") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## Standard Errors

- The **standard error** is the estimated standard deviation of $\hat{\beta}_1$:

. . .

$$se(\hat{\beta}_1) = \sqrt{Var(\hat{\beta}_1)} = \frac{\hat{\sigma}}{\sqrt{SST_x}}$$

. . .

- where $\hat{\sigma}$ estimates the error standard deviation from the residuals

. . .

- Standard errors quantify the **precision** of our estimates

- They are essential for hypothesis testing and confidence intervals (next unit!)

## OLS is BLUE

Under all five Gauss-Markov assumptions, OLS is **BLUE**:

. . .

- **B**est: Lowest variance among all linear unbiased estimators
- **L**inear: A linear function of the data
- **U**nbiased: $E[\hat{\beta}] = \beta$
- **E**stimator: A rule for using sample data to estimate parameters

. . .

- You cannot do better without accepting some bias or using non-linear methods!

## Which Assumptions Matter for What? {.smaller}

| Assumption | Needed for Unbiasedness? | Needed for BLUE? |
|------------|:------------------------:|:----------------:|
| 1. Linear in parameters | Yes | Yes |
| 2. Random sampling | Yes | Yes |
| 3. Variation in $x$ | Yes | Yes |
| **4. Zero conditional mean** | **Yes** | Yes |
| 5. Homoskedasticity | No | **Yes** |

. . .

- Assumptions 1-4 → OLS is **unbiased**
- All five → OLS is **BLUE** (Best Linear Unbiased Estimator)


# Practice Exercises

## Practice: Interpreting Regression Output {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**A researcher estimates the relationship between test scores and class size:**

```
             Estimate  Std. Error
(Intercept)   698.93      10.36
str            -2.28       0.52

n = 420, R-squared = 0.051
```

*(`str` = student-teacher ratio)*

. . .

**Questions:**

1. What happens to predicted scores with 5 more students per teacher?
2. What is the predicted score when `str = 20`?
3. What does the R-squared tell us?

## Practice: Answers {.smaller}

1. $5 \times (-2.28) = -11.4$ points lower

. . .

2. $698.93 + (-2.28)(20) = 698.93 - 45.6 = 653.33$

. . .

3. About 5.1% of the variation in test scores is explained by class size. This is low, but does NOT mean class size is unimportant -- it just means other factors also matter.

## Practice: Run a Full Analysis {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using `mtcars`, complete the following:

1. Create a scatter plot of `disp` (displacement) vs. `mpg` with a line of best fit
2. Run `lm(mpg ~ disp, data = mtcars)` and interpret the slope
3. What is the R-squared? What does it mean?

. . .

```{r}
#| output-location: fragment
#| fig-height: 2.5

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(x = "Displacement (cu.in.)", y = "MPG",
       title = "MPG vs. Engine Displacement") +
  theme_minimal()
```

## Practice: Results {.smaller}

```{r}
#| output-location: fragment

summary(lm(mpg ~ disp, data = mtcars))
```

## Practice: Identifying Assumption Violations {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

**For each scenario, which Gauss-Markov assumption is most likely violated? Discuss with your neighbor!**

. . .

1. You study the effect of education on wages, but can't observe ability

. . .

- *Assumption 4 (zero conditional mean): ability is in the error term and correlated with education*

. . .

2. You survey only college students about the effect of income on happiness

. . .

- *Assumption 2 (random sampling): your sample is not representative of the general population*


# Summary and Key Takeaways

## What We Learned {.smaller}

1. The **conditional expectation** $E[Y|X]$ is what we want to estimate

. . .

2. **Scatter plots** with `ggplot2` visualize relationships and regression lines

. . .

3. The **OLS slope** = $Cov(x,y)/Var(x)$ -- essentially a scaled correlation

. . .

4. The **Gauss-Markov assumptions** (especially zero conditional mean) determine when OLS is unbiased

. . .

5. **Unbiasedness** means OLS hits the target on average; **precision** depends on noise and variation in $x$

. . .

6. **Causality requires unbiasedness** -- only when $E[\mu|x] = 0$ can we interpret $\hat{\beta}_1$ causally

## The Key Equations

**The OLS Slope:**

$$\hat{\beta}_1 = \frac{Cov(x, y)}{Var(x)} = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sum(x_i - \bar{x})^2}$$

. . .

**The OLS Intercept:**

$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$$

. . .

**The Variance of the Slope:**

$$Var(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$$

## Looking Ahead

- Next up: **Multivariate Regression** -- adding more $x$ variables to address omitted variable bias

. . .

- Then: **Statistical Inference** -- hypothesis testing, confidence intervals, and when we can say a result is "statistically significant"
