---
title: "Statistical Inference"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 6 | Statistical Inference"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
```


# From Point Estimates to Uncertainty

## Where We've Been {.smaller}

- We know how to estimate regression parameters using OLS

. . .

- Under the Gauss-Markov assumptions, OLS is **unbiased:** $E[\hat{\beta}_j] = \beta_j$

. . .

- But a single estimate from a single sample is just one draw from a **sampling distribution**

. . .

- **The problem:** How do we know if our estimate reflects a *real* effect or is just due to random chance?

. . .

> Statistical inference gives us the tools to answer this question.


## The Sampling Distribution {.smaller}

- If we could draw many samples and estimate $\hat{\beta}_1$ each time, we'd get a **distribution** of estimates

. . .

- Some estimates would be larger, some smaller, but they'd center around $\beta_1$ (by unbiasedness)

. . .

- We already know two properties of this distribution:

. . .

**Mean:** $E[\hat{\beta}_j] = \beta_j$

. . .

**Variance:** $Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R^2_j)}$

. . .

- But we need to know the **shape** of the distribution to do inference

. . .

- Recall from last unit: we estimate $\sigma^2$ with $\hat{\sigma}^2 = SSR/(n-k-1)$, giving us $se(\hat{\beta}_j) = \hat{\sigma}/\sqrt{SST_j(1-R_j^2)}$


## Quick Recap: The Standard Error {.smaller}

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)} \xrightarrow{\text{estimate } \sigma^2 \text{ with residuals}} se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$$

. . .

- The **standard error** tells us how much $\hat{\beta}_j$ would typically vary across repeated samples

. . .

- Now: what **shape** does the sampling distribution have? And how do we use it for inference?


# The Normality Assumption and the Sampling Distribution

## The Normality Assumption {.smaller}

To fully characterize the **shape** of the sampling distribution, we add one more assumption:

. . .

::: {.callout-important}
## CLM Assumption 6: Normality

The population error $\mu$ is independent of the explanatory variables and normally distributed:

$$\mu \sim Normal(0, \sigma^2)$$
:::

. . .

- This is the final piece of the **Classical Linear Model (CLM)** assumptions

. . .

- CLM = Gauss-Markov assumptions + normality


## What Normality Gives Us {.smaller}

Under the CLM assumptions, the OLS estimator is **normally distributed:**

$$\hat{\beta}_j \sim Normal\left(\beta_j, \frac{\sigma^2}{SST_j(1 - R^2_j)}\right)$$

. . .

- Centered on the true value $\beta_j$ (unbiasedness)
- Spread determined by the variance formula
- **Shape** is the bell curve

. . .

When we replace $\sigma^2$ with $\hat{\sigma}^2$, we get a **t-distribution** instead of normal:

$$\frac{\hat{\beta}_j - \beta_j}{se(\hat{\beta}_j)} \sim t_{n-k-1}$$


# Hypothesis Testing

## The Core Question {.smaller}

- You estimate $\hat{\beta}_1 = 0.065$ for the return to education

. . .

- Is this reflecting a real effect? Or did you just get a noisy sample?

. . .

- **Hypothesis testing** provides a formal framework for answering this

. . .

**Null hypothesis:** $H_0: \beta_j = 0$ (no effect)

**Alternative hypothesis:** $H_1: \beta_j \neq 0$ (some effect)

. . .

> The null hypothesis is the "boring" explanation: any pattern in the data is just due to chance.


## The t-Statistic {.smaller}

To test $H_0: \beta_j = 0$, compute the **t-statistic:**

$$t_{\hat{\beta}_j} = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}$$

. . .

- This is the ratio of our estimate to its standard error

. . .

- **Interpretation:** How many standard errors away from zero is our estimate?

. . .

- Under $H_0$, the t-stat follows a $t_{n-k-1}$ distribution


## Properties of the t-Statistic {.smaller}

1. **Same sign as the estimate:** Since $se(\hat{\beta}_j) > 0$, the t-stat has the same sign as $\hat{\beta}_j$

. . .

2. **Magnitude matters:** Larger $|\hat{\beta}_j|$ relative to $se(\hat{\beta}_j)$ $\rightarrow$ larger $|t|$

. . .

3. **Signal-to-noise ratio:** How large is our estimate relative to the noise (uncertainty) in our data?

. . .

- **Large |t|:** Evidence against $H_0$ --- our estimate is unlikely if there were truly no effect

- **Small |t|:** Consistent with $H_0$ --- we can't rule out that the true effect is zero


## Example: Returns to Education {.smaller}

```{r}
#| echo: true
data(wage1, package = "wooldridge")
reg <- lm(lwage ~ educ + exper + tenure, data = wage1)
coef_table <- summary(reg)$coefficients
round(coef_table, 4)
```

. . .

- $\hat{\beta}_{educ} = `r round(coef(reg)["educ"], 4)`$ with $se = `r round(coef_table["educ", "Std. Error"], 4)`$

. . .

- $t = `r round(coef_table["educ", "t value"], 2)`$ --- the estimate is about `r round(abs(coef_table["educ", "t value"]), 0)` standard errors from zero

. . .

- This is strong evidence that education has a real effect on wages


## Critical Values and Rejection Regions {.smaller}

- We **reject $H_0$** if $|t_{\hat{\beta}_j}|$ exceeds a **critical value** $c_\alpha$

. . .

- The critical value depends on:
  - The **significance level** $\alpha$ (usually 5%)
  - The **degrees of freedom** $n - k - 1$

. . .

- At the 5% significance level with a two-sided test:
  - Reject $H_0$ if $|t_{\hat{\beta}_j}| > c_{0.05}$
  - Fail to reject $H_0$ if $|t_{\hat{\beta}_j}| \leq c_{0.05}$


## The "Rule of 2" {.smaller}

::: {.callout-tip}
## Quick Rule of Thumb

With reasonably large samples ($n > 30$ or so), the critical value at the 5% level is approximately **2**.

So: **reject $H_0$ at the 5% level if $|t| > 2$**
:::

. . .

- This is because the $t$-distribution approaches the standard normal as $n$ grows

. . .

- For the standard normal: $P(|Z| > 1.96) = 0.05$

. . .

- $1.96 \approx 2$ --- hence the rule of thumb


# P-Values

## What is a P-Value? {.smaller}

::: {.callout-note}
## Definition

The **p-value** is the probability of observing a t-statistic at least as extreme as the one we calculated, *assuming the null hypothesis is true.*
:::

. . .

- Small p-value $\rightarrow$ our result would be very unlikely under $H_0$ $\rightarrow$ evidence against $H_0$

. . .

- Large p-value $\rightarrow$ our result is plausible under $H_0$ $\rightarrow$ we can't reject $H_0$

. . .

- **Decision rule:** Reject $H_0$ if $p < \alpha$ (typically $\alpha = 0.05$)


## Interpreting P-Values {.smaller}

**Common (mis)interpretation:** "The p-value is the probability that $H_0$ is true."

. . .

**Correct interpretation:** "The p-value is the probability of getting data this extreme *if* $H_0$ were true."

. . .

Think of it like a legal trial:

- $H_0$: The defendant is innocent
- Small p-value: The evidence would be very surprising if the defendant were innocent
- We reject $H_0$ = "guilty beyond reasonable doubt"

. . .

- Failing to reject $H_0$ $\neq$ proving $H_0$ is true (just like "not guilty" $\neq$ "innocent")


## Statistical vs. Practical Significance {.smaller}

::: {.callout-warning}
## Important Distinction

**Statistical significance** ($p < 0.05$) tells you the estimate is unlikely due to chance.

**Practical significance** tells you the effect is large enough to matter.
:::

. . .

- With a huge sample, even a *tiny* effect can be statistically significant

. . .

- Example: An extra year of education raises wages by 0.1%
  - With $n = 1{,}000{,}000$, this might have $p < 0.001$
  - But is 0.1% economically meaningful?

. . .

- **Always report and discuss the magnitude of the coefficient, not just whether it's "significant"**


# Confidence Intervals

## From Point Estimates to Intervals {.smaller}

- A point estimate $\hat{\beta}_j$ gives our single best guess of $\beta_j$

. . .

- But we know there's uncertainty --- the standard error tells us how much

. . .

- A **confidence interval** provides a range of plausible values for $\beta_j$

. . .

$$\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)$$

. . .

- For a **95% confidence interval** (with large $n$): $c_\alpha \approx 1.96 \approx 2$


## Interpreting Confidence Intervals {.smaller}

::: {.callout-note}
## What a 95% CI Means

If we were to draw many samples and compute a 95% CI each time, about **95% of those intervals would contain the true parameter** $\beta_j$.
:::

. . .

- It is **not** "there is a 95% probability that $\beta_j$ is in this interval"
  - The true $\beta_j$ is fixed --- it's either in the interval or it's not

. . .

- Instead: our *procedure* captures the truth 95% of the time

. . .

- **Connection to hypothesis testing:** If zero is outside the 95% CI, we reject $H_0: \beta_j = 0$ at the 5% level


## Example: Confidence Intervals in R {.smaller}

```{r}
#| echo: true
data(wage1, package = "wooldridge")
reg <- lm(lwage ~ educ + exper + tenure, data = wage1)

# 95% confidence intervals
confint(reg)
```

. . .

- The CI for `educ` is entirely above zero --- we're confident education has a positive effect on wages

. . .

- The width of the interval reflects our uncertainty


## CIs, T-Tests, and P-Values {.smaller}

These three approaches to inference are **equivalent** --- they always give the same answer:

. . .

| Method | Reject $H_0: \beta_j = 0$ when... |
|:-------|:-----------------------------------|
| **t-test** | $|t_{\hat{\beta}_j}| > c_\alpha$ |
| **p-value** | $p < \alpha$ |
| **CI** | Zero is outside the $100(1-\alpha)\%$ CI |

. . .

> You'll often see all three reported in regression output and published research. They tell the same story in different ways.


# Summary

## Key Formulas {.smaller}

| Formula | What It Is |
|:--------|:-----------|
| $Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)}$ | Variance of the estimator |
| $\hat{\sigma}^2 = \frac{SSR}{n-k-1}$ | Estimated error variance (Unit 5) |
| $se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$ | Standard error (Unit 5) |
| $t_{\hat{\beta}_j} = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}$ | t-statistic |
| $\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)$ | Confidence interval |


## Key Concepts {.smaller}

1. The **normality assumption** + Gauss-Markov = the Classical Linear Model (CLM)

. . .

2. The **t-statistic** measures how many standard errors an estimate is from zero

. . .

3. The **p-value** is the probability of seeing data this extreme under $H_0$

. . .

4. A **confidence interval** gives a range of plausible values for the true parameter

. . .

5. **Statistical significance** ($p < 0.05$) is not the same as **practical significance**

. . .

6. t-tests, p-values, and CIs are **three equivalent ways** of doing inference
