---
title: "Statistical Inference"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Unit 6 | Statistical Inference"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
<<<<<<< HEAD
library(wooldridge)
library(modelsummary)
```


# From Point Estimates to Uncertainty

## Where We've Been {.smaller}

- We know how to estimate regression parameters using OLS

. . .

- Under the Gauss-Markov assumptions, OLS is **unbiased:** $E[\hat{\beta}_j] = \beta_j$

. . .

- But a single estimate from a single sample is just one draw from a **sampling distribution**

. . .

- **The problem:** How do we know if our estimate reflects a *real* effect or is just due to random chance?

. . .

> Statistical inference gives us the tools to answer this question.


## The Sampling Distribution {.smaller}

- If we could draw many samples and estimate $\hat{\beta}_1$ each time, we'd get a **distribution** of estimates

. . .

- Some estimates would be larger, some smaller, but they'd center around $\beta_1$ (by unbiasedness)

. . .

- We already know two properties of this distribution:

. . .

**Mean:** $E[\hat{\beta}_j] = \beta_j$

. . .

**Variance:** $Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R^2_j)}$

. . .

- But we need to know the **shape** of the distribution to do inference

. . .

- Recall from last unit: we estimate $\sigma^2$ with $\hat{\sigma}^2 = SSR/(n-k-1)$, giving us $se(\hat{\beta}_j) = \hat{\sigma}/\sqrt{SST_j(1-R_j^2)}$


## Quick Recap: The Standard Error {.smaller}

$$Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)} \xrightarrow{\text{estimate } \sigma^2 \text{ with residuals}} se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$$

. . .

- The **standard error** tells us how much $\hat{\beta}_j$ would typically vary across repeated samples

. . .

- Now: what **shape** does the sampling distribution have? And how do we use it for inference?


# The Normality Assumption and the Sampling Distribution

## The Normality Assumption {.smaller}

To fully characterize the **shape** of the sampling distribution, we add one more assumption:
=======
library(patchwork)
library(modelsummary)
options(scipen = 999)
```


# From Estimates to Evidence

## Where We've Been

- We know how to **estimate** $\hat{\beta}_j$ using OLS

. . .

- We know the Gauss-Markov assumptions guarantee OLS is **BLUE**

. . .

- But every estimate comes from *one sample* — a different sample gives a different $\hat{\beta}_j$
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb

. . .

::: {.callout-important}
<<<<<<< HEAD
## CLM Assumption 6: Normality

The population error $\mu$ is independent of the explanatory variables and normally distributed:

$$\mu \sim Normal(0, \sigma^2)$$
:::

. . .

- This is the final piece of the **Classical Linear Model (CLM)** assumptions

. . .

- CLM = Gauss-Markov assumptions + normality


## What Normality Gives Us {.smaller}

Under the CLM assumptions, the OLS estimator is **normally distributed:**

$$\hat{\beta}_j \sim Normal\left(\beta_j, \frac{\sigma^2}{SST_j(1 - R^2_j)}\right)$$

. . .

- Centered on the true value $\beta_j$ (unbiasedness)
- Spread determined by the variance formula
- **Shape** is the bell curve

. . .

When we replace $\sigma^2$ with $\hat{\sigma}^2$, we get a **t-distribution** instead of normal:

$$\frac{\hat{\beta}_j - \beta_j}{se(\hat{\beta}_j)} \sim t_{n-k-1}$$
=======
## The Big Question Today
How do we know if our estimate reflects a **real** relationship — or just **noise** from sampling?
:::


## The Roadmap

Today we build four tools for answering that question:

. . .

1. **The sampling distribution** — what would happen across many samples?

. . .

2. **Hypothesis testing** — is there evidence of a real effect?

. . .

3. **P-values** — how strong is that evidence?

. . .

4. **Confidence intervals** — what's the plausible range for $\beta_j$?


# The Sampling Distribution

## One Sample, One Estimate {.smaller}

Let's build intuition with a simulation. Suppose the **true** model is:

$$y = 2.45 + 0 \cdot x + \mu$$

. . .

- The true effect of $x$ on $y$ is **zero** ($\beta_1 = 0$)

. . .

- But if we draw a sample and run OLS, will we get exactly zero?

. . .

```{r}
#| echo: true
set.seed(42)
pop_data <- tibble(
  x = sample(1:100, size = 1000, replace = TRUE),
  y = 2.45 + 0 * x + rnorm(1000)
)

lm(y ~ x, data = pop_data) |> coef()
```

. . .

- Not exactly zero! We got `r round(coef(lm(y ~ x, data = pop_data))["x"], 4)` — close, but not zero


## What If We Drew 1,000 Samples? {.smaller}

```{r}
#| echo: true
#| code-fold: true

set.seed(42)
n_sims <- 1000

sim_results <- map_dfr(1:n_sims, function(i) {
  sample_data <- pop_data |> slice_sample(n = 50, replace = TRUE)
  reg <- lm(y ~ x, data = sample_data)
  tibble(iteration = i, b1_hat = coef(reg)["x"])
})
```

. . .

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

ggplot(sim_results, aes(x = b1_hat)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "black", bins = 40) +
  geom_vline(xintercept = 0, color = "#E63946", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 0.012, y = 40,
           label = "'True' ~ beta[1] == 0", parse = TRUE, color = "#E63946", size = 5) +
  labs(x = expression(hat(beta)[1]), y = "Density") +
  theme_minimal(base_size = 16)
```

This is the **sampling distribution** — it shows how $\hat{\beta}_1$ varies across repeated samples.


## Key Properties {.smaller}

```{r}
#| echo: false
#| fig-height: 3
#| fig-align: center

ggplot(sim_results, aes(x = b1_hat)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "black", bins = 40) +
  stat_function(fun = dnorm,
                args = list(mean = mean(sim_results$b1_hat),
                            sd = sd(sim_results$b1_hat)),
                color = "#E63946", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "#002169", linewidth = 1.2, linetype = "dashed") +
  labs(x = expression(hat(beta)[1]), y = "Density") +
  theme_minimal(base_size = 16)
```

. . .

1. **Centered on the truth** — the average estimate $\approx \beta_1 = 0$ (unbiasedness!)

. . .

2. **Bell-shaped** — follows a normal distribution

. . .

3. **The spread** is the **standard error** — how much estimates typically vary


## What Is the Standard Error? {.smaller}

- The **standard error** $se(\hat{\beta}_j)$ is the standard deviation of the sampling distribution

. . .

- It measures **how precise** our estimate is

. . .

| Scenario | Standard Error | Meaning |
|----------|----------------|---------|
| Small SE | 0.02 | Estimates barely vary — very precise |
| Large SE | 2.50 | Estimates vary a lot — imprecise |

. . .

- You see the SE in the `Std. Error` column of `summary()` output

. . .

- SE depends on: sample size ($n$), variation in $x$, and noise ($\sigma^2$)


## Asymptotic Normality: Why It Works {.smaller}

Under the CLM assumptions (GM + normality), $\hat{\beta}_j$ is exactly normal.

. . .

But what if errors **aren't** normal? (Wages are skewed, health spending has a long tail...)

. . .

**The Central Limit Theorem saves us:**

$$\frac{\hat{\beta}_j - \beta_j}{se(\hat{\beta}_j)} \xrightarrow{d} Normal(0, 1) \quad \text{as } n \to \infty$$

. . .

- $\hat{\beta}_j$ is a weighted average of the $y_i$'s — and averages become normal

- Works even with non-normal errors, as long as $n$ is "large enough"


## Simulation: Normal vs. Skewed Errors {.smaller}

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-align: center

set.seed(42)
n_obs <- 50
n_sims_asym <- 2000

errors_df <- tibble(
  Normal = rnorm(500, 0, 2),
  `Shifted Exponential` = rexp(500, rate = 0.5) - 2
) |>
  pivot_longer(everything(), names_to = "Distribution", values_to = "u")

ggplot(errors_df, aes(x = u)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightgrey", color = "black", bins = 40) +
  facet_wrap(~Distribution, scales = "free") +
  labs(x = "Error (u)", y = "Density",
       title = "The error distributions look VERY different") +
  theme_minimal(base_size = 14)
```

. . .

Left: symmetric bell curve. Right: heavily right-skewed (like wages).

Will $\hat{\beta}_1$ still be normally distributed with skewed errors?


## Both Sampling Distributions Are Normal! {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

sim_normal <- tibble(b1_hat = numeric(n_sims_asym))
sim_skewed <- tibble(b1_hat = numeric(n_sims_asym))
beta1_true <- 3

for (s in 1:n_sims_asym) {
  x <- runif(n_obs, 1, 10)
  u_norm <- rnorm(n_obs, 0, 2)
  u_skew <- rexp(n_obs, rate = 0.5) - 2
  y_norm <- 2 + beta1_true * x + u_norm
  y_skew <- 2 + beta1_true * x + u_skew
  sim_normal$b1_hat[s] <- coef(lm(y_norm ~ x))["x"]
  sim_skewed$b1_hat[s] <- coef(lm(y_skew ~ x))["x"]
}

combined <- bind_rows(
  sim_normal |> mutate(DGP = "Normal Errors"),
  sim_skewed |> mutate(DGP = "Skewed Errors")
)

ggplot(combined, aes(x = b1_hat)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "lightblue", color = "black", bins = 40) +
  stat_function(
    data = combined |> filter(DGP == "Normal Errors"),
    fun = dnorm,
    args = list(mean = mean(sim_normal$b1_hat), sd = sd(sim_normal$b1_hat)),
    color = "#E63946", linewidth = 1.2) +
  stat_function(
    data = combined |> filter(DGP == "Skewed Errors"),
    fun = dnorm,
    args = list(mean = mean(sim_skewed$b1_hat), sd = sd(sim_skewed$b1_hat)),
    color = "#E63946", linewidth = 1.2) +
  geom_vline(xintercept = beta1_true, color = "#002169",
             linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~DGP) +
  labs(x = expression(hat(beta)[1]), y = "Density") +
  theme_minimal(base_size = 14)
```

. . .

The normal curve fits **both** — even with $n = 50$ and skewed errors!

This is **asymptotic normality** at work.


## Review: Sampling Distribution {.smaller}

::: {.callout-note icon=false}
## ✅ Check your understanding
:::

**True or False:**

1. If we draw many samples and compute $\hat{\beta}_1$ from each, the average will equal $\beta_1$.

. . .

> **True** — this is unbiasedness.

. . .

2. The standard error tells us how far our estimate is from the true value.

. . .

> **False** — it tells us how much estimates *typically vary* across samples. Any single estimate could be close or far.

. . .

3. Asymptotic normality requires that the error term is normally distributed.

. . .

> **False** — that's the whole point! The CLT makes $\hat{\beta}_j$ approximately normal even with non-normal errors.
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb


# Hypothesis Testing

<<<<<<< HEAD
## The Core Question {.smaller}

- You estimate $\hat{\beta}_1 = 0.065$ for the return to education

. . .

- Is this reflecting a real effect? Or did you just get a noisy sample?

. . .

- **Hypothesis testing** provides a formal framework for answering this

. . .

**Null hypothesis:** $H_0: \beta_j = 0$ (no effect)

**Alternative hypothesis:** $H_1: \beta_j \neq 0$ (some effect)

. . .

> The null hypothesis is the "boring" explanation: any pattern in the data is just due to chance.


## The t-Statistic {.smaller}

To test $H_0: \beta_j = 0$, compute the **t-statistic:**

=======
## The Logic of the Test

We want to know: **does $x_j$ actually affect $y$?**

. . .

We formalize this as:

- **Null hypothesis:** $H_0: \beta_j = 0$ — no effect

- **Alternative:** $H_1: \beta_j \neq 0$ — there is an effect

. . .

::: {.callout-tip}
## The Strategy
1. **Assume** $H_0$ is true
2. Ask: "Would my estimate be surprising under this assumption?"
3. If yes → reject $H_0$ (evidence of a real effect)
:::


## Step 1: What Does the World Look Like Under $H_0$? {.smaller}

If $\beta_j = 0$, our estimate is just **noise** — it should hover around zero:

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

set.seed(42)
df_value <- 100
t_sample <- tibble(t_value = rt(10000, df = df_value))

ggplot(t_sample, aes(x = t_value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "lightblue", color = "black") +
  stat_function(fun = dt, args = list(df = df_value),
                color = "#E63946", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "#002169", linewidth = 1.2) +
  annotate("text", x = 1.8, y = 0.35,
           label = "'Distribution under' ~ H[0]", parse = TRUE, color = "#002169", size = 5) +
  labs(x = "Standardized values", y = "Density") +
  coord_cartesian(xlim = c(-4, 4)) +
  theme_minimal(base_size = 16)
```

. . .

Most values cluster near zero. Values far from zero are **rare**.

**Question:** Does our estimate look like it belongs here, or is it an outlier?


## Step 2: The t-Statistic {.smaller}

We need to **standardize** our estimate so we can place it on that distribution:

. . .

>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb
$$t_{\hat{\beta}_j} = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}$$

. . .

<<<<<<< HEAD
- This is the ratio of our estimate to its standard error

. . .

- **Interpretation:** How many standard errors away from zero is our estimate?

. . .

- Under $H_0$, the t-stat follows a $t_{n-k-1}$ distribution


## Properties of the t-Statistic {.smaller}

1. **Same sign as the estimate:** Since $se(\hat{\beta}_j) > 0$, the t-stat has the same sign as $\hat{\beta}_j$

. . .

2. **Magnitude matters:** Larger $|\hat{\beta}_j|$ relative to $se(\hat{\beta}_j)$ $\rightarrow$ larger $|t|$

. . .

3. **Signal-to-noise ratio:** How large is our estimate relative to the noise (uncertainty) in our data?

. . .

- **Large |t|:** Evidence against $H_0$ --- our estimate is unlikely if there were truly no effect

- **Small |t|:** Consistent with $H_0$ --- we can't rule out that the true effect is zero


## Example: Returns to Education {.smaller}

```{r}
#| echo: true
data(wage1, package = "wooldridge")
reg <- lm(lwage ~ educ + exper + tenure, data = wage1)
coef_table <- summary(reg)$coefficients
round(coef_table, 4)
=======
- This tells us: **how many standard errors away from zero** is our estimate?

. . .

| $t$-stat | Interpretation |
|----------|---------------|
| $t = 0.5$ | Half a standard error from zero — could easily be noise |
| $t = 2.0$ | Two standard errors away — unusual under $H_0$ |
| $t = 4.0$ | Four standard errors away — very unlikely to be noise |

. . .

- Same sign as $\hat{\beta}_j$ (since $se > 0$)
- Bigger $|t|$ = stronger evidence against $H_0$


## Step 3: How Far Is "Too Far"? {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

p_close <- ggplot(t_sample, aes(x = t_value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "lightblue", color = "black") +
  stat_function(fun = dt, args = list(df = df_value),
                color = "#E63946", linewidth = 1.2) +
  geom_vline(xintercept = 0.85, color = "#002169", linewidth = 1.5) +
  annotate("text", x = 0.85, y = 0.42, label = "t = 0.85",
           color = "#002169", size = 4, hjust = -0.1) +
  labs(x = "t-values", y = "Density", title = expression("Consistent with " * H[0])) +
  coord_cartesian(xlim = c(-4, 4)) +
  theme_minimal(base_size = 14)

p_far <- ggplot(t_sample, aes(x = t_value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "lightblue", color = "black") +
  stat_function(fun = dt, args = list(df = df_value),
                color = "#E63946", linewidth = 1.2) +
  geom_vline(xintercept = 2.8, color = "#E63946", linewidth = 1.5) +
  annotate("text", x = 2.8, y = 0.15, label = "t = 2.8",
           color = "#E63946", size = 4, hjust = -0.1) +
  labs(x = "t-values", y = "Density", title = expression("Evidence against " * H[0])) +
  coord_cartesian(xlim = c(-4, 4)) +
  theme_minimal(base_size = 14)

p_close + p_far
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb
```

. . .

<<<<<<< HEAD
- $\hat{\beta}_{educ} = `r round(coef(reg)["educ"], 4)`$ with $se = `r round(coef_table["educ", "Std. Error"], 4)`$

. . .

- $t = `r round(coef_table["educ", "t value"], 2)`$ --- the estimate is about `r round(abs(coef_table["educ", "t value"]), 0)` standard errors from zero

. . .

- This is strong evidence that education has a real effect on wages
=======
- $t = 0.85$: well within the body — no reason to doubt $H_0$
- $t = 2.8$: out in the tail — this would be **rare** if $H_0$ were true


## Type I Error and the Significance Level {.smaller}

Even when $H_0$ is true, we might get an unlucky sample with a large $t$-stat.

. . .

- **Type I error** ("false positive"): rejecting $H_0$ when it's actually true

. . .

- The **significance level** $\alpha$ is our tolerance for this mistake

  - $\alpha = 0.05$: we accept a 5% chance of false rejection
  - $\alpha = 0.01$: we accept only a 1% chance

. . .

- The **critical value** $c$ is the threshold: reject if $|t| > c$
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb


## Critical Values and Rejection Regions {.smaller}

<<<<<<< HEAD
- We **reject $H_0$** if $|t_{\hat{\beta}_j}|$ exceeds a **critical value** $c_\alpha$

. . .

- The critical value depends on:
  - The **significance level** $\alpha$ (usually 5%)
  - The **degrees of freedom** $n - k - 1$

. . .

- At the 5% significance level with a two-sided test:
  - Reject $H_0$ if $|t_{\hat{\beta}_j}| > c_{0.05}$
  - Fail to reject $H_0$ if $|t_{\hat{\beta}_j}| \leq c_{0.05}$


## The "Rule of 2" {.smaller}

::: {.callout-tip}
## Quick Rule of Thumb

With reasonably large samples ($n > 30$ or so), the critical value at the 5% level is approximately **2**.

So: **reject $H_0$ at the 5% level if $|t| > 2$**
:::

. . .

- This is because the $t$-distribution approaches the standard normal as $n$ grows

. . .

- For the standard normal: $P(|Z| > 1.96) = 0.05$

. . .

- $1.96 \approx 2$ --- hence the rule of thumb


# P-Values

## What is a P-Value? {.smaller}

::: {.callout-note}
## Definition

The **p-value** is the probability of observing a t-statistic at least as extreme as the one we calculated, *assuming the null hypothesis is true.*
:::

. . .

- Small p-value $\rightarrow$ our result would be very unlikely under $H_0$ $\rightarrow$ evidence against $H_0$

. . .

- Large p-value $\rightarrow$ our result is plausible under $H_0$ $\rightarrow$ we can't reject $H_0$

. . .

- **Decision rule:** Reject $H_0$ if $p < \alpha$ (typically $\alpha = 0.05$)


## Interpreting P-Values {.smaller}

**Common (mis)interpretation:** "The p-value is the probability that $H_0$ is true."

. . .

**Correct interpretation:** "The p-value is the probability of getting data this extreme *if* $H_0$ were true."

. . .

Think of it like a legal trial:

- $H_0$: The defendant is innocent
- Small p-value: The evidence would be very surprising if the defendant were innocent
- We reject $H_0$ = "guilty beyond reasonable doubt"

. . .

- Failing to reject $H_0$ $\neq$ proving $H_0$ is true (just like "not guilty" $\neq$ "innocent")


## Statistical vs. Practical Significance {.smaller}

::: {.callout-warning}
## Important Distinction

**Statistical significance** ($p < 0.05$) tells you the estimate is unlikely due to chance.

**Practical significance** tells you the effect is large enough to matter.
:::

. . .

- With a huge sample, even a *tiny* effect can be statistically significant

. . .

- Example: An extra year of education raises wages by 0.1%
  - With $n = 1{,}000{,}000$, this might have $p < 0.001$
  - But is 0.1% economically meaningful?

. . .

- **Always report and discuss the magnitude of the coefficient, not just whether it's "significant"**


# Confidence Intervals

## From Point Estimates to Intervals {.smaller}

- A point estimate $\hat{\beta}_j$ gives our single best guess of $\beta_j$

. . .

- But we know there's uncertainty --- the standard error tells us how much

. . .

- A **confidence interval** provides a range of plausible values for $\beta_j$

. . .

$$\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)$$

. . .

- For a **95% confidence interval** (with large $n$): $c_\alpha \approx 1.96 \approx 2$


## Interpreting Confidence Intervals {.smaller}

::: {.callout-note}
## What a 95% CI Means

If we were to draw many samples and compute a 95% CI each time, about **95% of those intervals would contain the true parameter** $\beta_j$.
:::

. . .

- It is **not** "there is a 95% probability that $\beta_j$ is in this interval"
  - The true $\beta_j$ is fixed --- it's either in the interval or it's not

. . .

- Instead: our *procedure* captures the truth 95% of the time

. . .

- **Connection to hypothesis testing:** If zero is outside the 95% CI, we reject $H_0: \beta_j = 0$ at the 5% level


## Example: Confidence Intervals in R {.smaller}

```{r}
#| echo: true
data(wage1, package = "wooldridge")
reg <- lm(lwage ~ educ + exper + tenure, data = wage1)

# 95% confidence intervals
confint(reg)
=======
```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

c_val <- 1.96

t_grid <- seq(-4, 4, length.out = 500)
t_dens <- dt(t_grid, df = df_value)

shade_data <- tibble(t_value = t_grid, density = t_dens) |>
  mutate(region = case_when(
    t_value < -c_val ~ "reject",
    t_value > c_val ~ "reject",
    TRUE ~ "fail_to_reject"
  ))

ggplot() +
  geom_ribbon(data = shade_data |> filter(region == "fail_to_reject"),
              aes(x = t_value, ymin = 0, ymax = density),
              fill = "lightblue", alpha = 0.7) +
  geom_ribbon(data = shade_data |> filter(t_value <= -c_val),
              aes(x = t_value, ymin = 0, ymax = density),
              fill = "#E63946", alpha = 0.5) +
  geom_ribbon(data = shade_data |> filter(t_value >= c_val),
              aes(x = t_value, ymin = 0, ymax = density),
              fill = "#E63946", alpha = 0.5) +
  stat_function(fun = dt, args = list(df = df_value),
                color = "#002169", linewidth = 1.2) +
  geom_vline(xintercept = c(-c_val, c_val),
             color = "#EEB808", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 0, y = 0.22,
           label = "'Fail to reject' ~ H[0]", parse = TRUE, color = "#002169", size = 4) +
  annotate("text", x = 0, y = 0.17,
           label = "(95%)", color = "#002169", size = 3.5) +
  annotate("text", x = 3.2, y = 0.05,
           label = "Reject\n(2.5%)", color = "#E63946", size = 3) +
  annotate("text", x = -3.2, y = 0.05,
           label = "Reject\n(2.5%)", color = "#E63946", size = 3) +
  annotate("text", x = -c_val, y = 0.42, label = "-1.96",
           color = "#EEB808", size = 4, hjust = 1.1) +
  annotate("text", x = c_val, y = 0.42, label = "+1.96",
           color = "#EEB808", size = 4, hjust = -0.1) +
  labs(x = "t-values", y = "Density") +
  coord_cartesian(xlim = c(-4, 4)) +
  theme_minimal(base_size = 16)
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb
```

. . .

<<<<<<< HEAD
- The CI for `educ` is entirely above zero --- we're confident education has a positive effect on wages

. . .

- The width of the interval reflects our uncertainty


## CIs, T-Tests, and P-Values {.smaller}

These three approaches to inference are **equivalent** --- they always give the same answer:

. . .

| Method | Reject $H_0: \beta_j = 0$ when... |
|:-------|:-----------------------------------|
| **t-test** | $|t_{\hat{\beta}_j}| > c_\alpha$ |
| **p-value** | $p < \alpha$ |
| **CI** | Zero is outside the $100(1-\alpha)\%$ CI |

. . .

> You'll often see all three reported in regression output and published research. They tell the same story in different ways.


# Summary

## Key Formulas {.smaller}

| Formula | What It Is |
|:--------|:-----------|
| $Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)}$ | Variance of the estimator |
| $\hat{\sigma}^2 = \frac{SSR}{n-k-1}$ | Estimated error variance (Unit 5) |
| $se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$ | Standard error (Unit 5) |
| $t_{\hat{\beta}_j} = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}$ | t-statistic |
| $\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)$ | Confidence interval |


## Key Concepts {.smaller}

1. The **normality assumption** + Gauss-Markov = the Classical Linear Model (CLM)

. . .

2. The **t-statistic** measures how many standard errors an estimate is from zero

. . .

3. The **p-value** is the probability of seeing data this extreme under $H_0$

. . .

4. A **confidence interval** gives a range of plausible values for the true parameter

. . .

5. **Statistical significance** ($p < 0.05$) is not the same as **practical significance**

. . .

6. t-tests, p-values, and CIs are **three equivalent ways** of doing inference
=======
For a two-sided 5% test: reject $H_0$ if $|t| > 1.96$


## The Rule of 2 {.smaller}

With large samples, the critical value for a 5% test is always $\approx 1.96 \approx 2$

. . .

::: {.callout-tip}
## Quick Rule
An estimate is **statistically significant at 5%** if:

$$|\hat{\beta}_j| > 2 \times se(\hat{\beta}_j)$$

In other words: if the estimate is more than **twice** its standard error.
:::

. . .

- This is a handy shortcut when scanning regression output


## Example: Housing Prices {.smaller}

```{r}
#| echo: true

set.seed(2025)
n <- 500
housing_data <- tibble(
  lot_area = runif(n, 5000, 20000),
  pool_area = rbinom(n, 1, 0.15) * runif(n, 200, 600),
  garage_area = runif(n, 200, 800),
  year_built = sample(1960:2020, n, replace = TRUE),
  year_remod = pmax(year_built, sample(1980:2023, n, replace = TRUE)),
  sale_price = -2500000 + 1.50 * lot_area + 80 * pool_area +
               150 * garage_area + 500 * year_built +
               900 * year_remod + rnorm(n, 0, 50000)
)

reg_housing <- lm(sale_price ~ lot_area + pool_area + garage_area +
                    year_built + year_remod, data = housing_data)
```


## Reading the Output {.smaller}

```{r}
#| echo: true
summary(reg_housing)
```


## Decoding `summary()` {.smaller}

The **Coefficients table** has four columns:

| Column | What it shows |
|--------|--------------|
| `Estimate` | $\hat{\beta}_j$ — the point estimate |
| `Std. Error` | $se(\hat{\beta}_j)$ — how precise the estimate is |
| `t value` | $\hat{\beta}_j / se(\hat{\beta}_j)$ — the t-statistic |
| `Pr(>|t|)` | The p-value (more on this next!) |

. . .

**Significance stars:**

| Stars | Meaning |
|-------|---------|
| (none) | $p \geq 0.10$ |
| `.` | $p < 0.10$ |
| `*` | $p < 0.05$ |
| `**` | $p < 0.01$ |
| `***` | $p < 0.001$ |


## Try It: Manual t-Test {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the housing regression, compute the t-statistic for `lot_area` **by hand** (using `coef()` and the standard error from `summary()`). Does it match the `t value` in the output?

. . .

```{r}
#| echo: true
b1 <- coef(reg_housing)["lot_area"]
se1 <- summary(reg_housing)$coefficients["lot_area", "Std. Error"]
t_manual <- b1 / se1
cat("Manual t-stat:", round(t_manual, 2), "\n")
cat("From summary: ", round(summary(reg_housing)$coefficients["lot_area", "t value"], 2))
```


# P-Values

## Beyond "Reject" or "Fail to Reject" {.smaller}

The t-test with a fixed critical value gives a **binary** answer: significant or not.

. . .

But it doesn't tell us **how strongly** we reject — was it a narrow rejection or a knockout?

. . .

The **p-value** answers this:

::: {.callout-important}
## Definition
The p-value is the **probability of seeing a t-statistic as extreme as ours** (or more extreme), *if the null hypothesis were true*.

Equivalently: the **smallest significance level** at which we would reject $H_0$.
:::


## Visualizing the P-Value {.smaller}

For $t = 2.3$: what fraction of the t-distribution lies beyond $\pm 2.3$?

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center

df_viz <- 250
observed_t <- 2.3
p_val <- 2 * pt(abs(observed_t), df = df_viz, lower.tail = FALSE)

t_grid <- seq(-4, 4, length.out = 500)
t_dens <- dt(t_grid, df = df_viz)
shade_data <- tibble(t_value = t_grid, density = t_dens) |>
  mutate(in_tail = abs(t_value) >= abs(observed_t))

ggplot() +
  geom_line(data = shade_data, aes(x = t_value, y = density),
            color = "#002169", linewidth = 1.2) +
  geom_area(data = shade_data |> filter(in_tail & t_value > 0),
            aes(x = t_value, y = density), fill = "#E63946", alpha = 0.6) +
  geom_area(data = shade_data |> filter(in_tail & t_value < 0),
            aes(x = t_value, y = density), fill = "#E63946", alpha = 0.6) +
  geom_vline(xintercept = c(-observed_t, observed_t),
             color = "#002169", linewidth = 1, linetype = "dashed") +
  annotate("text", x = observed_t + 0.3, y = 0.05,
           label = paste0("t = ", observed_t), color = "#002169", size = 5, hjust = 0) +
  annotate("text", x = 0, y = 0.2,
           label = paste0("p = ", round(p_val, 4)),
           color = "#E63946", size = 6) +
  labs(x = "t-statistic", y = "Density") +
  coord_cartesian(xlim = c(-4, 4)) +
  theme_minimal(base_size = 16)
```

. . .

The red area = p-value = `r round(p_val, 4)`. Only a `r round(p_val * 100, 1)`% chance of seeing this if $H_0$ were true.


## Interpreting P-Values {.smaller}

- **p = 0.03:** Only a 3% chance of our estimate if $\beta_j = 0$. That's unlikely — reject $H_0$.

. . .

- **p = 0.35:** A 35% chance of our estimate if $\beta_j = 0$. Not unusual at all — fail to reject.

. . .

**Decision rule:** Reject $H_0$ if $p < \alpha$ (your chosen significance level)

. . .

::: {.callout-warning}
## Common Misconception
The p-value is **NOT** the probability that $H_0$ is true!

It's the probability of **seeing data this extreme** if $H_0$ *were* true. Big difference.
:::


## Statistical vs. Practical Significance {.smaller}

::: {.callout-important}
## Key Distinction
A **statistically significant** result ($p < 0.05$) is not necessarily **practically important**.
:::

. . .

- With a large enough sample, even **tiny** effects become significant

. . .

- Example: A drug lowers blood pressure by 0.1 mmHg ($p = 0.001$)
  - Statistically significant? ✅
  - Clinically meaningful? ❌

. . .

- **Always** look at the **magnitude** of $\hat{\beta}_j$, not just the p-value!


## Review: Hypothesis Testing & P-Values {.smaller}

::: {.callout-note icon=false}
## ✅ Check your understanding
:::

You estimate $\hat{\beta}_1 = 4.5$ with $se(\hat{\beta}_1) = 2.0$.

1. What is the t-statistic?

. . .

> $t = 4.5 / 2.0 = 2.25$

. . .

2. Is this significant at the 5% level?

. . .

> Yes — $|2.25| > 1.96$

. . .

3. A colleague's estimate has $p = 0.85$. What does that mean?

. . .

> Under $H_0$, there's an 85% chance of seeing an estimate this extreme. Very consistent with $H_0$ — no evidence of an effect.


## Try It: P-Values in R {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using our housing regression, compute the p-value for `lot_area` by hand:

. . .

```{r}
#| echo: true

# The t-stat
t_stat <- b1 / se1

# Degrees of freedom
df <- nobs(reg_housing) - length(coef(reg_housing))

# Two-sided p-value using pt()
# pt() gives P(T ≤ t). We want the probability in BOTH tails.
p_manual <- 2 * pt(abs(t_stat), df = df, lower.tail = FALSE)

cat("t-statistic:", round(t_stat, 4), "\n")
cat("p-value:    ", format(p_manual, scientific = TRUE), "\n")
```


# Confidence Intervals

## From Point Estimates to Ranges {.smaller}

A point estimate ($\hat{\beta}_j = 1.50$) gives our **best guess** — but how uncertain are we?

. . .

A **confidence interval** gives a **range of plausible values** for $\beta_j$:

. . .

$$CI = \hat{\beta}_j \pm c \cdot se(\hat{\beta}_j)$$

. . .

For a 95% CI with large $n$: $c \approx 1.96$

$$CI_{95\%} = \hat{\beta}_j \pm 1.96 \cdot se(\hat{\beta}_j)$$


## Computing CIs in R {.smaller}

::: {.callout-note icon=false}
## 🖥️ Try it yourself!
:::

```{r}
#| echo: true

# By hand
lower <- b1 - 1.96 * se1
upper <- b1 + 1.96 * se1
cat("Manual 95% CI: [", round(lower, 4), ",", round(upper, 4), "]\n")
```

. . .

Or use R's `confint()` function:

```{r}
#| echo: true

# confint() computes CIs directly
# First arg = model, second = which coefficient, level = confidence level
confint(reg_housing, "lot_area", level = 0.95)
```


## What Does "95% Confident" Mean? {.smaller}

It does **not** mean "95% probability that $\beta_j$ is in this interval."

. . .

The true $\beta_j$ is a fixed number — it's either in the interval or it isn't.

. . .

::: {.callout-important}
## The Correct Interpretation
If we drew 100 different samples and computed a 95% CI from each, about **95 of them** would contain the true $\beta_j$.
:::

. . .

Any single interval is one draw from this process — we don't know if ours is one of the lucky 95 or the unlucky 5.


## Simulation: 50 Confidence Intervals {.smaller}

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center

set.seed(321)
true_b1_ci <- 3
n_per_sample <- 30
n_ci_sims <- 50

ci_data <- map_dfr(1:n_ci_sims, function(i) {
  x <- runif(n_per_sample, 0, 10)
  u <- rnorm(n_per_sample, 0, 8)
  y <- 5 + true_b1_ci * x + u
  reg <- lm(y ~ x)
  ci <- confint(reg, "x", level = 0.95)
  tibble(sample = i, estimate = coef(reg)["x"],
         lower = ci[1], upper = ci[2],
         covers_true = lower <= true_b1_ci & upper >= true_b1_ci)
})

n_covered <- sum(ci_data$covers_true)

ggplot(ci_data, aes(y = reorder(factor(sample), sample))) +
  geom_vline(xintercept = true_b1_ci, linetype = "dashed",
             color = "#002169", linewidth = 1) +
  geom_errorbar(aes(xmin = lower, xmax = upper, color = covers_true),
                width = 0.3, linewidth = 0.6, orientation = "y") +
  geom_point(aes(x = estimate, color = covers_true), size = 1.5) +
  scale_color_manual(values = c("TRUE" = "#5D8EE5", "FALSE" = "#E63946"),
                     labels = c("TRUE" = "Contains true value",
                                "FALSE" = "Misses true value")) +
  annotate("text", x = true_b1_ci, y = 53,
           label = paste0("True value = ", true_b1_ci),
           color = "#002169", size = 4) +
  annotate("text", x = max(ci_data$upper) - 0.5, y = 3,
           label = paste0(n_covered, " of ", n_ci_sims, " intervals\ncontain the true value"),
           color = "grey30", size = 3.5, hjust = 1) +
  labs(x = expression(hat(beta)[1]), y = "Sample", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", axis.text.y = element_text(size = 6))
```


## The CI ↔ Hypothesis Test Connection {.smaller}

There's a direct link between CIs and hypothesis testing:

. . .

| If the 95% CI... | Then at the 5% level... |
|-------------------|------------------------|
| **Excludes** zero | We **reject** $H_0: \beta_j = 0$ |
| **Includes** zero | We **fail to reject** $H_0: \beta_j = 0$ |

. . .

This makes CIs a handy visual shortcut — just check if zero is inside the interval!

. . .

::: {.callout-tip}
## Three Tools, One Answer
$|t| > 1.96$ ⟺ $p < 0.05$ ⟺ 95% CI excludes zero

They always agree.
:::


## Review: Confidence Intervals {.smaller}

::: {.callout-note icon=false}
## ✅ Check your understanding
:::

1. You estimate $\hat{\beta}_1 = 3.0$ with $se = 1.5$. What's the 95% CI?

. . .

> $3.0 \pm 1.96 \times 1.5 = [0.06, 5.94]$

. . .

2. Does this CI suggest the effect is statistically significant at 5%?

. . .

> Yes — zero is (barely) excluded from the interval.

. . .

3. Your friend says "there's a 95% probability that $\beta_1$ is between 0.06 and 5.94." Is this correct?

. . .

> No! The correct interpretation is about the *procedure*, not this particular interval.


# Putting It All Together

## The Three Tools: A Comparison {.smaller}

| Method | Question |
|--------|----------|
| **t-statistic** | Is my estimate large relative to the noise? |
| **P-value** | How surprising would my data be if $H_0$ were true? |
| **Confidence interval** | What range of values is plausible for $\beta_j$? |

. . .

All three are mathematically connected:

- If $|t| > 1.96$, then $p < 0.05$, and the 95% CI excludes zero
- If $p < 0.05$, then $|t| > 1.96$, and the 95% CI excludes zero
- If the 95% CI excludes zero, then $p < 0.05$ and $|t| > 1.96$


## Reading a Regression Table {.smaller}

```{r}
#| echo: false
#| label: tbl-housing-final

modelsummary(
  list("Sale Price" = reg_housing),
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  coef_rename = c(
    "lot_area" = "Lot Area (sq ft)",
    "pool_area" = "Pool Area",
    "garage_area" = "Garage Area",
    "year_built" = "Year Built",
    "year_remod" = "Year Remodeled"
  ),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  notes = "Standard errors in parentheses. * p < 0.1, ** p < 0.05, *** p < 0.01"
)
```

. . .

- Numbers in parentheses are **standard errors** (not t-stats, not CIs!)
- Stars tell you significance at a glance
- Always check the **notes** at the bottom for the star convention


## Try It: Full Inference Workflow {.smaller}

::: {.callout-note icon=false}
## 🖥️ Your turn!
:::

Using the `mtcars` dataset, estimate `mpg ~ hp + wt`. For the coefficient on `wt`:

1. What is $\hat{\beta}_{wt}$ and its standard error?
2. Compute the t-statistic. Is it significant at 5%?
3. Compute the 95% confidence interval (by hand and using `confint()`).
4. Interpret the CI in a sentence.

. . .

```{r}
#| echo: true
#| output-location: fragment

reg_mtcars <- lm(mpg ~ hp + wt, data = mtcars)
summary(reg_mtcars)$coefficients["wt", ]
confint(reg_mtcars, "wt", level = 0.95)
```


## Key Takeaways {.smaller}

::: {.callout-tip}
## What You Should Remember

1. **Sampling distribution:** $\hat{\beta}_j$ varies across samples; the SE measures how much

2. **Hypothesis test:** Compare your t-stat to 1.96 (or use the Rule of 2)

3. **P-value:** The probability of your data under $H_0$ — small $p$ = strong evidence against $H_0$

4. **Confidence interval:** A range of plausible values for $\beta_j$

5. **Statistical ≠ practical significance** — always look at the *magnitude*, not just the stars!
:::


## Next Time

- **F-tests:** Testing multiple hypotheses simultaneously

. . .

- **Heteroskedasticity:** What happens when the variance of $\mu$ isn't constant?

. . .

- **Robust standard errors:** Inference that works even when assumptions fail
>>>>>>> 19de0831900ec22b7b1fa5ab3a06accbcb11dddb
