---
title: "Midterm Exam Review"
subtitle: "EC 325: Econometrics"
author: "Prof. Caraher"
institute: "Colby College"
date: "today"
date-format: "medium"
embed-resources: true
format:
  revealjs:
    theme: colby-theme.scss
    slide-number: true
    incremental: true
    footer: "EC325 | Midterm Review"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(wooldridge)
library(modelsummary)
```

## Roadmap for Today

- **Part I:** Causality, RCTs, and the Potential Outcomes Framework (~15 min)
- **Part II:** Good Data Practices and R/Tidyverse (~20 min)
- **Part III:** Bivariate OLS — Model, Estimation, Interpretation (~20 min)
- **Part IV:** Multivariate OLS and Omitted Variable Bias (~20 min)
- **Part V:** Properties of the OLS Estimator — Assumptions and Variance (~20 min)
- **Part VI:** Statistical Inference — Hypothesis Testing through CIs (~20 min)
- **Q&A / Practice** (~5 min)

# Part I: Causality and Experimental Design

## What is Econometrics?

**Econometrics:** The practice of using statistical, quantitative methods to study social phenomena and provide insights into public policy.

- Modern econometrics focuses on identifying **causal effects** — not mere correlation
- This focus on causality is called **causal inference**

. . .

**Examples of Causal Inference Questions:**

- Does having health insurance cause you to be healthier?
- Does a college degree increase your earnings?
- Do tariffs increase domestic production?

## Steps of an Econometric Project

1. Write an econometric model
2. Get the data
3. Estimate the relationship
4. Interpret the estimates

. . .

**Key takeaway:** Econometrics is not just "running a regression" — it requires careful thinking about *what* we are estimating and *why* at every stage.

## The Core Problem: Correlation $\neq$ Causation {.smaller}

- **Correlation vs. Causation:** We cannot infer causality from correlation alone
  - *Example:* Health insurance coverage is correlated with better health, but this does not mean insurance *causes* better health

- **Selection Bias (Apples-to-Oranges):** Simple comparisons fail because the groups being compared differ in underlying characteristics
  - Those with insurance may also be wealthier, which independently improves health

- **Ceteris Paribus:** If we compare "apples-to-apples" (all else equal), then the difference in outcomes *can* be interpreted as a causal effect

## Concept Check: Selection Bias

> Suppose we compare average test scores of students who attend private tutoring vs. those who do not. We find tutored students score 15 points higher.

. . .

**Can we conclude tutoring causes a 15-point increase?**

. . .

**No.** Students who seek tutoring may differ in motivation, parental involvement, prior ability, etc. These are sources of **selection bias**.

## Formalizing Causality: Potential Outcomes {.smaller}

- **The Framework:** We denote:
  - $Y_{1i}$: treated outcome if unit $i$
  - $Y_{0i}$: untreated outcome if unit $i$
  - Causal effect: $\kappa_i = Y_{1i} - Y_{0i}$

- **Fundamental Problem of Causal Inference:** We can never observe *both* $Y_{1i}$ and $Y_{0i}$ for the same unit — we only see one outcome

## Mean Comparisons and Selection Bias {.smaller}

- A simple difference in group means between units $i$ in the treatment group ($D_i=1$) and control group ($D_i=0$):

$$
E[Y_i | D_i = 1] - E[Y_i | D_i = 0]
$$

- This decomposes into:

$$
\underbrace{E[Y_{1i} | D_i = 1] - E[Y_{0i} | D_i = 1]}_{\text{Causal Effect } (\kappa)} + \underbrace{E[Y_{0i} | D_i = 1] - E[Y_{0i} | D_i = 0]}_{\text{Selection Bias}}
$$

. . .

- **Selection Bias Term:** Captures how the treated group *would have fared without treatment* ($E[Y_{0i} | D_i = 1]$) vs. the control group ($E[Y_{0i} | D_i = 0]$) — driven by pre-existing differences (e.g., wealth, health preferences)

## The Gold Standard: RCTs {.smaller}

- **Randomized Control Trials (RCTs):** Randomly assign units to treatment or control
- **Why Randomization Works:**
  - The **Law of Large Numbers (LLN)** ensures that, with a large enough sample, background characteristics are balanced across groups
  - Formally: $E[Y_{0i} | D_i = 1] = E[Y_{0i} | D_i = 0]$, so selection bias = 0

. . .

- **Example:** The **Oregon Health Insurance Experiment** used a lottery to randomly assign Medicaid eligibility, allowing researchers to estimate the causal effect of health insurance

## Quick Recap: Part I {.smaller}

| Concept | Key Idea |
|:--------|:---------|
| Econometrics | Causal inference using statistical methods |
| Selection Bias | Groups differ in unobservable ways |
| Potential Outcomes | $\kappa = Y_{1i} - Y_{0i}$, but we only see one |
| RCTs | Randomization eliminates selection bias via LLN |

# Part II: Good Data Practices and R

## Why Good Data Practices Matter

- Before running any regression, you must **get to know your data**
- Reinhart-Rogoff (2010): influential research on debt and economic growth contained spreadsheet errors
  - Errors discovered by a grad student attempting to replicate the results
  - Corrected results told a **very different** policy story

. . .

::: {.callout-important}
## The Lesson
No sophisticated econometric technique can compensate for errors in the underlying data.
:::

## Key Questions to Ask About Your Data {.smaller}

Before any analysis:

1. **How many observations?** Is anything missing?
2. **Is the distribution reasonable?** Do the mean, min, and max make sense?
3. **Are there outliers or impossible values?**

. . .

Tools:

- **Descriptive statistics:** `mean()`, `sd()`, `min()`, `max()`, `summary()`
- **Visualizations:** histograms, scatterplots via `ggplot2`

## Missing Data {.smaller}

- Many R functions return `NA` if any value is missing

. . .

- Use `is.na()` and `sum(is.na())` to detect missing values
- Use `na.rm = TRUE` to ignore them in calculations:

. . .

```{r}
#| eval: false

mean(airquality$Ozone)             # Returns NA
mean(airquality$Ozone, na.rm = TRUE)  # Ignores NAs
```

. . .

- **Caution:** Data can be **missing systematically** (not at random)
  - High earners refusing to report income $\rightarrow$ underestimates average income
  - Sicker patients dropping out of a study $\rightarrow$ remaining sample looks healthier

## R Basics You Should Know {.smaller}

| Concept | Syntax | What It Does |
|---------|--------|-------------|
| Assignment | `x <- 3` | Stores value as an object |
| Vectors | `c(1, 2, 3)` | Combines values into a sequence |
| Functions | `mean(x)` | Applies a pre-built operation |
| Data frames | `mtcars` | Rows = observations, columns = variables |
| Column access | `mtcars$mpg` | Extracts a single column |
| Data types | numeric, character, logical, factor | Determines what operations are valid |
| Packages | `library(tidyverse)` | Loads a package of functions |

## The Pipe Operator: `|>` {.smaller}

- Takes whatever is on the **left** and passes it as the **first argument** to the function on the **right**
- Read it as **"and then"**

. . .

```{r}
#| output-location: fragment

# Take mtcars AND THEN filter to high-mpg cars
mtcars |>
  filter(mpg > 25)
```

## Key `dplyr` Functions {.smaller}

| Function | Purpose | Example |
|----------|---------|---------|
| `filter()` | Subset **rows** by condition | `filter(mpg > 25)` |
| `select()` | Choose **columns** | `select(mpg, cyl, wt)` |
| `mutate()` | Create/modify **variables** | `mutate(kmpl = mpg * 0.425)` |
| `summarize()` | Collapse to **summary stats** | `summarize(avg = mean(mpg))` |
| `group_by()` | Group data before summarizing | `group_by(cyl)` |
| `arrange()` | Sort rows | `arrange(desc(mpg))` |
| `count()` | Count rows per group | `count(cyl)` |

. . .

Each verb does **one thing** well. The pipe `|>` chains them together.

## `filter()` and `select()` {.smaller}

**`filter()`:** Keep rows meeting a condition

```{r}
#| output-location: fragment

mtcars |> filter(cyl == 4, mpg > 25)
```

. . .

**`select()`:** Choose (or drop) specific columns

```{r}
#| output-location: fragment

mtcars |> select(mpg, cyl, wt) |> head(4)
```

## `mutate()` {.smaller}

**Purpose:** Create new variables or transform existing ones

```{r}
#| output-location: fragment

mtcars |>
  mutate(kmpl = mpg * 0.425) |>
  select(mpg, kmpl) |>
  head(5)
```

## Chaining Verbs Together {.smaller}

- The real power: **chain multiple operations** in sequence
- Read like a sentence: "Take the data, *then* filter, *then* mutate, *then* select..."

. . .

```{r}
#| output-location: fragment

mtcars |>
  filter(cyl == 4) |>
  mutate(kmpl = mpg * 0.425) |>
  select(mpg, kmpl, wt)
```

## `summarize()` and `group_by()` {.smaller}

**`summarize()`:** Collapse a data frame into summary statistics

```{r}
#| output-location: fragment

mtcars |>
  summarize(mean_mpg = mean(mpg), sd_mpg = sd(mpg), n = n())
```

. . .

**`group_by()` + `summarize()`:** Compute stats **by group**

```{r}
#| output-location: fragment

mtcars |>
  group_by(cyl) |>
  summarize(mean_mpg = mean(mpg), mean_hp = mean(hp), n = n())
```

## `count()` and `arrange()` {.smaller}

**`count()`:** Quickly tally rows per group

```{r}
#| output-location: fragment

mtcars |> count(cyl)
```

. . .

**`arrange()`:** Sort rows (use `desc()` for descending)

```{r}
#| output-location: fragment

mtcars |> arrange(desc(mpg)) |> select(mpg, cyl, hp) |> head(4)
```

## Quick Overview Tools {.smaller}

**`summary()`:** Min, quartiles, mean, max for every column

```{r}
#| output-location: fragment

summary(mtcars[, 1:4])
```

. . .

**`glimpse()`:** Variable names, types, and first few values

```{r}
#| output-location: fragment

glimpse(mtcars)
```

## Visualization with `ggplot2` {.smaller}

- Every ggplot is built **layer by layer:**

1. Start with data: `ggplot(data)`
2. Add geometry: `geom_point()`, `geom_histogram()`, `geom_smooth()`, `geom_col()`
3. Map variables to aesthetics: `aes(x = ..., y = ..., color = ...)`
4. Add labels: `labs(title = ..., x = ..., y = ...)`

. . .

| Function | What It Does |
|----------|--------------|
| `ggplot(data)` | Initialize plot |
| `aes(x, y, color)` | Map variables to aesthetics |
| `geom_point()` | Scatterplot |
| `geom_smooth(method = "lm")` | Line of best fit |
| `geom_histogram()` | Histogram |
| `geom_col()` | Bar chart |
| `labs()` | Labels and titles |

## `ggplot2` Example {.smaller}

```{r}
#| echo: true
#| fig-height: 3.5

ggplot(data = mtcars) +
  geom_point(aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_smooth(aes(x = wt, y = mpg), method = "lm") +
  labs(title = "MPG vs. Weight", x = "Weight (1000 lbs)",
       y = "Miles per Gallon", color = "Cylinders") +
  theme_minimal()
```

## Saving Your Work {.smaller}

- Always work in **scripts** (`.R` files), not the console — scripts are saveable and replicable
- Use **R Projects** (`.Rproj`) so your working directory is set automatically
- **Comment your code** with `#` to explain what it does and *why*

. . .

::: {.callout-tip}
## The Replication Standard
Research meets the replication standard when an independent researcher can exactly duplicate the results using only the information provided.
:::

# Part III: Bivariate OLS

## Conditional Expectations

- **Conditional expectation:** Our "best guess" of $Y$ given that $X$ takes some value $x$:

$$
E[Y | X = x]
$$

- *Example:* What is the expected wage for those with a college degree ($X = BA$)?

- This is the quantity that OLS attempts to estimate

## The Population Regression Model (Bivariate)

- When RCTs are not feasible, we use **Ordinary Least Squares (OLS)** regression

$$
y = \beta_0 + \beta_1 x + \mu
$$

- $y$: Dependent variable (outcome/response)
- $x$: Independent variable (explanatory/predictor)
- $\mu$: Error term (all unobserved factors affecting $y$)
- $\beta_0, \beta_1$: Population parameters (true, unknown values)

## The OLS Regression Line {.smaller}

- We estimate the unknown population parameters using sample data, yielding $\hat{\beta}_0$ and $\hat{\beta}_1$:

$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x
$$

- $\hat{\beta}_0$: Predicted value of $y$ when $x = 0$ (the intercept)
- $\hat{\beta}_1$: **Marginal effect** — the predicted change in $y$ when $x$ increases by 1 unit

. . .

$$
\Delta \hat{y} = \hat{\beta}_1 \cdot \Delta x
$$

## Deriving the OLS Estimates {.smaller}

- OLS chooses $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimize the **Sum of Squared Residuals (SSR)**:

$$
SSR = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2 = \sum_{i=1}^{n}\hat{\mu}_i^2
$$

- The slope estimate:

$$
\hat{\beta}_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2} = \frac{Cov(x,y)}{Var(x)}
$$

- The intercept: $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$

## Correlation and the Slope Coefficient

- The bivariate OLS slope is the **scaled correlation** between $x$ and $y$:

$$
\hat{\beta}_1 = \rho_{xy} \frac{\sigma_y}{\sigma_x}
$$

- Because $\hat{\beta}_1$ relies heavily on correlation, **caution is needed** when inferring causality from bivariate OLS

## Log-Linear Models {.smaller}

- When the dependent variable is logged:

$$
\log(y) = \beta_0 + \beta_1 x + \mu
$$

- Interpretation changes: a 1-unit increase in $x$ is associated with a $100 \times \hat{\beta}_1$ **percent** change in $y$

$$
\%\Delta y \approx 100 \times \hat{\beta}_1 \cdot \Delta x
$$

## Goodness-of-Fit: $R^2$ {.smaller}

- **$R^2$** measures the fraction of total variation in $y$ explained by the model:

$$
R^2 = \frac{SSE}{SST} = 1 - \frac{SSR}{SST}
$$

- $SST = \sum(y_i - \bar{y})^2$: Total Sum of Squares (total variation in $y$)
- $SSE = \sum(\hat{y}_i - \bar{y})^2$: Explained Sum of Squares
- $SSR = \sum(y_i - \hat{y}_i)^2$: Residual (unexplained) Sum of Squares
- $R^2$ ranges from 0 to 1; higher values indicate a better fit

## Concept Check: Bivariate OLS {.smaller}

> You estimate: $\widehat{wage} = 5.2 + 1.8 \cdot educ$

. . .

- What is the predicted wage for someone with 16 years of education?
  - $5.2 + 1.8(16) = 34.0$

. . .

- What is the marginal effect of one additional year of education?
  - $\$1.80$ per hour

. . .

- Can we interpret 1.8 as the *causal effect* of education on wages?
  - **Not necessarily** — bivariate OLS is susceptible to **omitted variable bias**

# Part IV: Multivariate OLS and OVB

## Why Multivariate OLS? {.smaller}

- Bivariate OLS is susceptible to **Omitted Variable Bias (OVB)**
- OVB occurs when:
  1. A variable affecting $y$ is excluded from the model, **AND**
  2. That excluded variable is correlated with an included variable ($x$)
- This violates the **Zero Conditional Mean** assumption

. . .

- **Solution:** Include additional control variables using multivariate OLS

## The Multivariate OLS Model

$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_k x_k + \mu
$$

- **Interpretation:** Each $\hat{\beta}_j$ gives the **ceteris paribus** effect of $x_j$ on $y$, holding all other included variables constant

- By controlling for confounders, multivariate OLS moves us closer to plausibly causal estimates

## Multivariate OLS as Matching

- Multivariate OLS can be thought of as a **matching estimator**:
  - It matches observations on control variables and then computes a weighted average of the effect across groups

. . .

- *Example (Gender Wage Gap):* Controlling for industry, the OLS coefficient on "female" approximates the weighted average of the within-industry gender wage gap

## The OVB Formula {.smaller}

- If we omit $x_2$ from a model that should include it, the biased estimate $\tilde{\beta}_1$ from the short regression relates to the true $\beta_1$:

$$
E[\tilde{\beta}_1] = \beta_1 + \beta_2 \cdot \delta_1
$$

where $Bias(\tilde{\beta}_1) = \beta_2 \cdot \delta_1$

- $\beta_2$: Effect of the omitted variable on $y$
- $\delta_1$: Correlation between the omitted variable and the included variable

## Direction of OVB {.smaller}

| $Corr(x_1, x_2)$ | $\beta_2 > 0$ | $\beta_2 < 0$ |
|:-----------------:|:-------------:|:-------------:|
| $\delta_1 > 0$ (Positive) | **Positive Bias** ($\tilde{\beta}_1$ too large) | **Negative Bias** ($\tilde{\beta}_1$ too small) |
| $\delta_1 < 0$ (Negative) | **Negative Bias** | **Positive Bias** |

. . .

**Example (Wages & Education):**

- Omitted: Ability. Included: Education. Outcome: Wage
- $\beta_2 > 0$ (ability $\uparrow$ wages) and $\delta_1 > 0$ (ability $\uparrow$ education)
- Result: **Positive bias** — bivariate OLS *overstates* the return to education

## Concept Check: OVB {.smaller}

> You estimate a bivariate regression of health outcomes on exercise frequency. You suspect that income is omitted.

. . .

**What is the likely direction of the bias?**

- $\beta_2 > 0$: Higher income $\rightarrow$ better health (through nutrition, healthcare access, etc.)
- $\delta_1 > 0$: Higher income $\rightarrow$ more exercise (gym memberships, leisure time)
- Bias $= \beta_2 \cdot \delta_1 > 0$: **Positive bias** — the bivariate estimate overstates the effect of exercise

## Adjusted $R^2$ {.smaller}

- In multivariate OLS, $R^2$ mechanically increases when adding variables — even irrelevant ones

- **Adjusted $R^2$** penalizes for additional parameters:

$$
\bar{R}^2 = 1 - \frac{SSR/(n-k-1)}{SST/(n-1)}
$$

- $\bar{R}^2 \leq R^2$ always, and can decrease if an added variable is not useful
- Preferred measure of fit in multivariate models

# Part V: Properties of the OLS Estimator

## The Gauss-Markov Assumptions

The first four assumptions ensure OLS is **unbiased**: $E[\hat{\beta}_j] = \beta_j$

## GM Assumption 1: Linearity in Parameters

::: {.callout-note}
## GM 1: Linearity in Parameters
The model is linear in the $\beta$ parameters:
$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_k x_k + \mu
$$
:::

- Non-linear transformations of the *data* are fine (e.g., $\log(y)$, $x^2$)
- Only need linearity in the $\beta$'s

## GM Assumption 2: Random Sampling

::: {.callout-note}
## GM 2: Random Sampling
We have a random sample of size $n$: $\{(y_i, x_{1i}, x_{2i}, \dots, x_{ki}) : i = 1, 2, \dots, n\}$ from the population.
:::

## GM Assumption 3: No Perfect Multicollinearity {.smaller}

::: {.callout-note}
## GM 3: No Perfect Multicollinearity
None of the independent variables are constant, and there is no exact linear relationship among the independent variables.
:::

- **Violation** occurs if one variable is an exact linear function of another
- *Example:* Including both "age" and "birth year" in a cross-section
- Note: High (but imperfect) correlation is allowed but inflates variance

## GM Assumption 4: Zero Conditional Mean

::: {.callout-note}
## GM 4: Zero Conditional Mean (ZCM)
$$
E[\mu | x_1, x_2, \dots, x_k] = 0
$$
:::

- **The most important assumption for causal inference**
- All factors in $\mu$ must be uncorrelated with the $x$ variables
- Violated most commonly by **omitted variable bias**

## GM 1–4: Unbiasedness

- If GM 1–4 hold, the OLS estimator is **unbiased:**

$$
E[\hat{\beta}_j] = \beta_j
$$

- On average (across many hypothetical samples), the OLS estimate equals the true population parameter

## GM Assumption 5: Homoskedasticity

::: {.callout-note}
## GM 5: Homoskedasticity
$$
Var(\mu | x_1, \dots, x_k) = \sigma^2
$$
:::

- The error variance is constant across all values of $x$
- **Heteroskedasticity:** When this assumption fails (variance of $\mu$ depends on $x$)

. . .

- If **GM 1–5** all hold, OLS is **BLUE** (Best Linear Unbiased Estimator) — it has the lowest variance among all linear unbiased estimators

## Variance of $\hat{\beta}_j$ {.smaller}

- Under GM 1–5, the variance of the multivariate OLS estimator is:

$$
Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1 - R_j^2)} \quad \text{for } j = 1, 2, \dots, k
$$

where:

- $\sigma^2$: Population error variance
- $SST_j = \sum(x_{ji} - \bar{x}_j)^2$: Sample variation in $x_j$
- $R_j^2$: $R^2$ from regressing $x_j$ on all other independent variables

## What Affects Estimator Variance? {.smaller}

Three factors determine the precision of $\hat{\beta}_j$:

1. **Error variance ($\sigma^2$):** More noise in $y$ $\rightarrow$ higher variance
2. **Sample variation in $x_j$ ($SST_j$):** More variation in $x_j$ $\rightarrow$ lower variance (more information)
3. **Multicollinearity ($R_j^2$):** Higher correlation among $x$'s $\rightarrow$ higher variance (less independent info in $x_j$)

. . .

**Also:** Larger sample size ($n$) increases $SST_j$, reducing variance

## The Problem: $\sigma^2$ is Unknown {.smaller}

- We want to compute the standard error, but the variance formula contains $\sigma^2$ — a **population parameter** we cannot observe
- We need to estimate $\sigma^2$ from our sample data

. . .

- **Starting point:** From homoskedasticity (GM 5) and zero conditional mean (GM 4):

$$
Var(\mu_i | x_i) = \sigma^2 \quad \text{and} \quad E[\mu_i] = 0
$$

- Combining these:

$$
Var(\mu_i) = E[(\mu_i - E[\mu_i])^2] = E[(\mu_i - 0)^2] = E[\mu_i^2]
$$

- So to estimate $\sigma^2$, we need a good estimate of the **error terms** $\mu_i$

## Estimating $\sigma^2$ from the Residuals {.smaller}

- We can't observe the true errors $\mu_i$, but we *can* compute the **residuals:**

$$
\hat{\mu}_i = y_i - \hat{y}_i
$$

- The residuals are our sample analog of the population errors

. . .

- It turns out that the average squared residual (adjusted for degrees of freedom) is an **unbiased estimator** of $\sigma^2$:

$$
\hat{\sigma}^2 = \frac{\sum_{i=1}^{n}\hat{\mu}_i^2}{n - k - 1} = \frac{SSR}{n - k - 1}
$$

- We divide by $n - k - 1$ (not $n$) because estimating $k + 1$ parameters ($\beta_0, \beta_1, \dots, \beta_k$) "uses up" $k + 1$ degrees of freedom

## From $\hat{\sigma}^2$ to the Standard Error {.smaller}

- Now we can plug $\hat{\sigma}^2$ into our variance formula and take the square root to get the **standard error:**

$$
se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1 - R_j^2)}}
$$

- The standard error is the estimated standard deviation of the sampling distribution of $\hat{\beta}_j$
- It measures the **precision** of our estimate

. . .

::: {.callout-tip}
## The Key Chain of Logic
$$
Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)} \xrightarrow{\text{estimate } \sigma^2 \text{ with residuals}} se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}
$$
We go from a formula involving an unknown population parameter ($\sigma^2$) to a computable standard error by replacing $\sigma^2$ with its sample estimate $\hat{\sigma}^2 = SSR/(n-k-1)$.
:::

## The Bias/Variance Tradeoff

- Including more controls: reduces bias (satisfies ZCM) but can increase variance if controls are correlated with $x_1$
- Omitting relevant controls: reduces variance but introduces bias
- **Larger samples** alleviate the tradeoff: variance shrinks with $n$, making it safer to include more controls

# Part VI: Statistical Inference

## The Big Picture

- We have one estimate $\hat{\beta}_j$ from one sample
- **Question:** Does this estimate reflect something "real" about the population, or is it just noise from random sampling?
- **Statistical inference** provides the tools to answer this question

## The Normality Assumption (GM 6)

::: {.callout-note}
## Normality
The population error $\mu$ is independent of $x_1, \dots, x_k$ and is normally distributed:
$$
\mu \sim Normal(0, \sigma^2)
$$
:::

. . .

- This implies a **normal sampling distribution** for the OLS estimator:

$$
\hat{\beta}_j \sim Normal\left[\beta_j, \; Var(\hat{\beta}_j)\right]
$$

## Hypothesis Testing: The Logic {.smaller}

- **Null Hypothesis** ($H_0$): Typically $\beta_j = 0$ (the variable has no effect)
- **Alternative Hypothesis** ($H_A$): Typically $\beta_j \neq 0$

. . .

**The procedure:**

1. **Assume** the null is true ($\beta_j = 0$)
2. **Construct** the distribution of estimates we would expect *if the null were true*
3. **Check** whether our actual estimate is unlikely under the null
4. If very unlikely $\rightarrow$ **reject $H_0$** in favor of $H_A$

## The T-Statistic

$$
t_{\hat{\beta}_j} = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}
$$

- Measures how many standard errors the estimate is from zero
- If $|t|$ is "large," our estimate is unlikely under $H_0$

. . .

**Rule-of-Thumb (Large Samples):** If $|t| > 2$, the estimate is statistically significant at the 5% level

## P-Values {.smaller}

- **Significance Level ($\alpha$):** Pre-chosen threshold (usually 10%, 5%, or 1%) — the probability of rejecting $H_0$ when it is actually true (Type I error)

- **P-Value:** The smallest $\alpha$ at which we would reject $H_0$
  - *Interpretation:* The probability of observing a $t$-statistic as extreme as ours, *if the null were true*

. . .

- **Rejection Rule:** If p-value $< \alpha$, reject $H_0$

## Statistical Significance: Key Points

- **Language matters:** When we cannot reject $H_0$, we say we **"fail to reject the null"** — *not* that we "accept the null"

- **Statistical vs. Practical Significance:** A small p-value tells us the effect is statistically distinguishable from zero, but the *magnitude* of the coefficient determines whether the effect is economically meaningful

## Concept Check: Hypothesis Testing {.smaller}

> You estimate: $\hat{\beta}_1 = 0.45$ with $se(\hat{\beta}_1) = 0.20$

. . .

- $t = 0.45 / 0.20 = 2.25$
- Since $|t| > 2$: statistically significant at the 5% level
- We **reject** $H_0: \beta_1 = 0$

. . .

> Now suppose $\hat{\beta}_1 = 0.003$ with $se(\hat{\beta}_1) = 0.001$, so $t = 3.0$

- Statistically significant, but is $0.003$ economically meaningful? **Context matters.**

## Confidence Intervals {.smaller}

- CIs provide a **range of plausible values** for the true population parameter

$$
\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)
$$

where $c_\alpha$ is the critical value (e.g., $c_{0.05} \approx 1.96$ for large samples)

. . .

- **Interpretation (95% CI):** "We are 95% confident that the true parameter lies between [Lower Bound] and [Upper Bound]"
- If the CI contains zero $\rightarrow$ the estimate is **not** statistically significant at that level

## CIs, P-Values, and T-Tests: The Connection {.smaller}

- All three are **equivalent** ways to assess statistical significance:

  - **T-test:** $|t| > c_\alpha$ $\Rightarrow$ reject $H_0$
  - **P-value:** $p < \alpha$ $\Rightarrow$ reject $H_0$
  - **CI:** Zero not in the interval $\Rightarrow$ reject $H_0$

- They will always give the same conclusion for the same $\alpha$ level

## Concept Check: Confidence Intervals {.smaller}

> $\hat{\beta}_1 = 3.2$ with $se(\hat{\beta}_1) = 1.5$

. . .

**95% CI:** $3.2 \pm 1.96 \times 1.5 = 3.2 \pm 2.94 = [0.26, \; 6.14]$

- Does the CI contain zero? **No** $\rightarrow$ statistically significant at 5%

. . .

**90% CI:** $3.2 \pm 1.645 \times 1.5 = 3.2 \pm 2.47 = [0.73, \; 5.67]$

- Still does not contain zero $\rightarrow$ significant at 10% as well

# Putting It All Together

## The Big Picture: From Causality to Inference {.smaller}

1. **Start** with a causal question
2. **Check your data** — descriptive stats, visualizations, missing values
3. **Ideally** use an RCT; if not, use multivariate OLS with controls
4. **Verify** assumptions (especially ZCM — think about OVB)
5. **Estimate** and conduct inference (t-tests, p-values, CIs)
6. **Interpret** carefully: statistical significance $\neq$ practical significance

## Key Formulas to Know {.smaller}

| Formula | What It Measures |
|:--------|:-----------------|
| $\hat{\beta}_1 = \frac{Cov(x,y)}{Var(x)}$ | OLS slope estimate |
| $Bias = \beta_2 \cdot \delta_1$ | Omitted variable bias |
| $Var(\hat{\beta}_j) = \frac{\sigma^2}{SST_j(1-R_j^2)}$ | Estimator variance |
| $se(\hat{\beta}_j) = \frac{\hat{\sigma}}{\sqrt{SST_j(1-R_j^2)}}$ | Standard error |
| $t = \frac{\hat{\beta}_j}{se(\hat{\beta}_j)}$ | T-statistic |
| $\hat{\beta}_j \pm c_\alpha \cdot se(\hat{\beta}_j)$ | Confidence interval |

## Key R Functions to Know {.smaller}

| Task | Functions |
|:-----|:---------|
| Subset rows | `filter()` |
| Choose columns | `select()` |
| Create variables | `mutate()` |
| Summarize data | `summarize()`, `group_by()` |
| Count observations | `count()`, `n()` |
| Sort data | `arrange()`, `desc()` |
| Descriptive stats | `mean()`, `sd()`, `min()`, `max()`, `summary()` |
| Handle missing data | `is.na()`, `sum(is.na())`, `na.rm = TRUE` |
| Visualize | `ggplot()`, `geom_point()`, `geom_histogram()`, `geom_smooth()` |
| Chain operations | `|>` (pipe) |


## Good Luck!

- Review the study guide for a complete topic list
- Practice interpreting regression output
- Think through OVB examples — they are a favorite exam topic
- Practice writing and reading `tidyverse` code
- Remember: the goal is to understand the *logic* behind each method, not just memorize formulas
