# Statistics and Probability Review {#sec-stats-review}

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)

```

```{=html}

<style>
.problem {
    background-color: #f8f9fa;
    border: 1px solid #dee2e6;
    border-radius: 8px;
    padding: 20px;
    margin: 20px 0;
}
.context {
    background-color: #e8f4f8;
    padding: 12px;
    border-radius: 5px;
    margin-bottom: 15px;
    border-left: 4px solid #2874a6;
}
.question {
    margin: 15px 0;
    padding: 10px;
    background-color: #fff3cd;
    border-radius: 5px;
}
.answer-row {
    display: flex;
    align-items: center;
    margin: 10px 0;
    flex-wrap: wrap;
    gap: 10px;
}
.answer-row label {
    min-width: 320px;
    font-weight: 500;
}
select {
    padding: 8px 12px;
    font-size: 14px;
    border: 2px solid #ccc;
    border-radius: 5px;
    min-width: 300px;
    cursor: pointer;
}
select:focus {
    border-color: #2874a6;
    outline: none;
}
.correct {
    border-color: #28a745 !important;
    background-color: #d4edda !important;
}
.incorrect {
    border-color: #dc3545 !important;
    background-color: #f8d7da !important;
}
.explanation {
    margin-top: 10px;
    padding: 15px;
    background-color: #e7f3ff;
    border-radius: 5px;
    border-left: 4px solid #2874a6;
}
.note {
    font-size: 0.85em;
    color: #666;
    font-style: italic;
}
</style>

<script>
function checkAnswer(selectElement, correctValue) {
    if (selectElement.value === '') {
        selectElement.classList.remove('correct', 'incorrect');
    } else if (selectElement.value === correctValue) {
        selectElement.classList.remove('incorrect');
        selectElement.classList.add('correct');
    } else {
        selectElement.classList.remove('correct');
        selectElement.classList.add('incorrect');
    }
}
</script>

```


::: {.callout-tip collapse="true"}
## **Key Questions**
- What are the properties of the summation operator?
- How do we interpret derivatives and partial derivatives?
- How do we find the maximum or minimum of a function?
- What is the difference between a population and a sample?
- What is a random variable and how do we describe its distribution?
- How do we measure central tendency, variability, and relationships between variables?
- What is the normal distribution and why is it important?
:::

::: {.callout-note collapse="true"}
## **Suggested Readings**
- @wooldridge_introductory_2019, Appendix A, B, and C
:::


This appendix reviews the essential mathematical and statistical concepts that form the foundation for econometrics.
If you've taken introductory statistics or calculus,
much of this material will be familiar.
However,
econometrics uses these tools in specific ways,
so it's worth reviewing them with an eye toward how they'll appear throughout the course.


## Basic Mathematical Tools

### The Summation Operator

The summation operator provides a compact way to write the sum of a sequence of numbers.
If $\{x_i: i = 1, 2, ..., n\}$ is a sequence of $n$ numbers,
then the sum of these numbers is written as:

$$
x_1 + x_2 + ... + x_n = \sum_{i=1}^{n} x_i
$$

The symbol $\Sigma$ (capital Greek letter sigma) tells us to add up the terms.
The expression $i=1$ below the sigma indicates that we start counting at $i=1$,
and the $n$ above tells us to stop at $i=n$.

#### Properties of Summation

The summation operator has several useful properties that we will use frequently.

**Property 1: Summing a constant.**
If we add a constant $c$ to itself $n$ times,
we get $n$ times the constant:

$$
\sum_{i=1}^{n} c = nc
$$

**Property 2: Factoring out constants.**
We can pull constants outside the summation:

$$
\sum_{i=1}^{n} cx_i = c \sum_{i=1}^{n} x_i
$$

**Property 3: Separating additive terms.**
Sums can be split across addition:

$$
\sum_{i=1}^{n} (ax_i + by_i) = a\sum_{i=1}^{n} x_i + b\sum_{i=1}^{n} y_i
$$

::: {.callout-warning}
## Common Mistakes with Summation
These properties do **not** extend to division or multiplication in the ways you might expect.

We **cannot** break up divided expressions:
$$
\sum_{i=1}^{n} \frac{x_i}{y_i} \neq \frac{\sum_{i=1}^{n} x_i}{\sum_{i=1}^{n} y_i}
$$

We **cannot** break up multiplicative expressions:
$$
\sum_{i=1}^{n} x_i^2 \neq \left(\sum_{i=1}^{n} x_i\right)^2
$$
:::

#### The Sample Mean

Given $n$ numbers $\{x_i: i = 1, 2, ..., n\}$,
we can compute their **average** or **mean** value by adding them up and dividing by $n$.
The average value,
denoted $\bar{x}$ (read "x-bar"),
is written as:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

The average is an example of a **descriptive statistic**:
a value that describes or summarizes a characteristic of a set of data points.
In this case,
the mean describes the **central tendency** of $x_i$---where the "middle" of the data lies.

#### Useful Summation Results

Two results involving summations will appear repeatedly in this course.

**Result 1: Deviations from the mean sum to zero.**
The sum of the deviations $(x_i - \bar{x})$ from the mean always equals zero:

$$
\sum_{i=1}^{n} (x_i - \bar{x}) = 0
$$

This can be proven using the properties above:

$$
\sum_{i=1}^{n} (x_i - \bar{x}) = \sum_{i=1}^{n} x_i - \sum_{i=1}^{n} \bar{x} = \sum_{i=1}^{n} x_i - n\bar{x} = n\bar{x} - n\bar{x} = 0
$$

The last step follows because $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ implies $n\bar{x} = \sum_{i=1}^{n} x_i$.

**Result 2: Sum of squared deviations.**
The sum of squared deviations can be rewritten as:

$$
\sum_{i=1}^{n} (x_i - \bar{x})^2 = \sum_{i=1}^{n} x_i^2 - n\bar{x}^2
$$

More generally,
for two variables $x$ and $y$:

$$
\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y}) = \sum_{i=1}^{n} x_i y_i - n\bar{x}\bar{y}
$$

These identities are useful for computing variances and covariances,
which we will discuss shortly.


### Derivatives

The derivative $\frac{dy}{dx}$ of a function $y = f(x)$ measures its **instantaneous rate of change**.
It tells you how sensitive the function's output is to a small change in its input.
In econometrics,
derivatives help us interpret how changes in independent variables affect dependent variables.

Here are derivatives of some common functional forms in econometrics:

| Function | Derivative |
|----------|------------|
| $y = \beta_0 + \beta_1 x$ | $\frac{dy}{dx} = \beta_1$ |
| $y = \beta_0 + \beta_1 x + \beta_2 x^2$ | $\frac{dy}{dx} = \beta_1 + 2\beta_2 x$ |
| $y = \beta_0 + \frac{\beta_1}{x}$ | $\frac{dy}{dx} = -\frac{\beta_1}{x^2}$ |
| $y = \beta_0 + \beta_1 \ln(x)$ | $\frac{dy}{dx} = \frac{\beta_1}{x}$ |

Notice that in the linear case,
the derivative is simply the coefficient $\beta_1$---the effect of $x$ on $y$ is constant.
In the other cases,
the derivative depends on the value of $x$,
meaning the effect of $x$ on $y$ varies depending on where you start.


### Partial Derivatives

When $y$ is a function of multiple variables,
say $y = f(x_1, x_2)$,
we use **partial derivatives** to measure how $y$ changes with respect to one variable while holding the others constant.

The partial derivative of $y$ with respect to $x_1$,
holding $x_2$ constant,
is denoted $\frac{\partial y}{\partial x_1}$.
Similarly,
the partial derivative with respect to $x_2$,
holding $x_1$ constant,
is $\frac{\partial y}{\partial x_2}$.

For example,
if $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2$,
then:

$$
\frac{\partial y}{\partial x_1} = \beta_1, \qquad \frac{\partial y}{\partial x_2} = \beta_2
$$

The concept of "holding other variables constant" is central to econometrics.
When we estimate a regression coefficient,
we are trying to measure the partial effect of one variable on the outcome,
controlling for other factors.


### Optimization: Maxima and Minima

In econometrics,
we often need to find the value of a variable that maximizes or minimizes a function.
For example,
Ordinary Least Squares (OLS) regression finds coefficient estimates by minimizing the sum of squared residuals.
Understanding optimization is essential for understanding how these estimates are derived.

#### Finding Critical Points

To find the maximum or minimum of a function $f(x)$,
we take the derivative,
set it equal to zero,
and solve for $x$.
The solutions are called **critical points**.

For example,
consider the quadratic function:

$$
f(x) = ax^2 + bx + c
$$

Taking the derivative and setting it equal to zero:

$$
\frac{df}{dx} = 2ax + b = 0
$$

Solving for $x$:

$$
x^* = -\frac{b}{2a}
$$

This critical point $x^*$ is where the function reaches its maximum or minimum.

#### Determining Maximum vs. Minimum

How do we know if a critical point is a maximum or a minimum?
We use the **second derivative test**.
The second derivative tells us about the curvature of the function:

$$
\frac{d^2f}{dx^2} = 2a
$$

The rule is:

- If $\frac{d^2f}{dx^2} > 0$ at the critical point, the function is **concave up** (bowl-shaped), and we have a **minimum**
- If $\frac{d^2f}{dx^2} < 0$ at the critical point, the function is **concave down** (hill-shaped), and we have a **maximum**

For our quadratic function $f(x) = ax^2 + bx + c$:

- If $a > 0$: The parabola opens upward, and $x^* = -\frac{b}{2a}$ is a **minimum**
- If $a < 0$: The parabola opens downward, and $x^* = -\frac{b}{2a}$ is a **maximum**

```{r}
#| echo: false
#| label: fig-quadratic-optimization
#| fig-cap: "Left: When a > 0, the parabola opens upward and the critical point is a minimum. Right: When a < 0, the parabola opens downward and the critical point is a maximum."
#| fig-width: 8
#| fig-height: 3.5
#| out-width: "85%"

# Minimum case (a > 0)
x <- seq(-3, 3, length.out = 100)
y_min <- x^2 - 2*x + 1  # a=1, b=-2, c=1, min at x=1

p1 <- ggplot(data.frame(x = x, y = y_min), aes(x = x, y = y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(aes(x = 1, y = 0), color = "red", size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 1.5, y = 0.5, label = "Minimum", color = "red", size = 4) +
  labs(
    title = "a > 0: Minimum",
    x = "x",
    y = "f(x)"
  ) +
  theme_minimal()

# Maximum case (a < 0)
y_max <- -x^2 + 2*x + 1  # a=-1, b=2, c=1, max at x=1

p2 <- ggplot(data.frame(x = x, y = y_max), aes(x = x, y = y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(aes(x = 1, y = 2), color = "red", size = 3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 1.5, y = 1.5, label = "Maximum", color = "red", size = 4) +
  labs(
    title = "a < 0: Maximum",
    x = "x",
    y = "f(x)"
  ) +
  theme_minimal()

p1 + p2
```

::: {.callout-tip}
## Optimization in OLS
In OLS regression,
we minimize the sum of squared residuals,
which is a quadratic function of the coefficients.
The second derivative is positive (the function is convex),
confirming that the solution is indeed a minimum.
This is why setting the first derivative equal to zero gives us the best-fitting line.
:::

#### Optimization with Multiple Variables

When a function depends on multiple variables,
we find critical points by setting all partial derivatives equal to zero simultaneously.
For a function $f(x_1, x_2)$:

$$
\frac{\partial f}{\partial x_1} = 0 \quad \text{and} \quad \frac{\partial f}{\partial x_2} = 0
$$

Determining whether a critical point is a maximum,
minimum,
or neither (a saddle point) requires examining the second partial derivatives,
which we will encounter when deriving the OLS estimator.


## Population vs. Sample

Before diving into probability,
it's important to understand the distinction between a **population** and a **sample**,
as this distinction underlies all of statistical inference.

The **population** is the entire group of individuals, objects, or observations that we are interested in studying.
For example,
if we want to understand the wages of all U.S. workers,
the population is every person employed in the United States.
If we want to know the effect of a drug on blood pressure,
the population might be all adults with hypertension.

In practice,
we almost never observe the entire population.
Instead,
we collect data on a **sample**---a subset of the population.
We might survey 5,000 workers rather than all 160 million,
or run a clinical trial with 500 patients rather than everyone with hypertension.

::: {.callout-important}
## The Goal of Statistical Inference
The fundamental goal of statistics is to use information from a **sample** to draw conclusions about the **population**.
We observe sample statistics (like the sample mean $\bar{x}$) and use them to make inferences about population parameters (like the population mean $\mu$).
:::

This distinction affects our notation:

| Concept | Population Parameter | Sample Statistic |
|---------|---------------------|------------------|
| Mean | $\mu$ | $\bar{x}$ |
| Variance | $\sigma^2$ | $s^2$ |
| Standard deviation | $\sigma$ | $s$ |
| Correlation | $\rho$ | $r$ |

Population parameters are typically denoted with Greek letters and are usually unknown.
Sample statistics are computed from our data and are used to estimate the population parameters.

For example,
if we want to know the average wage of all U.S. workers ($\mu$),
we collect a sample and compute the sample mean ($\bar{x}$).
Under certain conditions,
$\bar{x}$ is a good estimate of $\mu$.
Much of econometrics is about understanding when and how well sample statistics estimate population parameters,
and quantifying the uncertainty in those estimates.


## Probability Basics

### Random Variables

A **random variable** is a variable whose value is determined by a random phenomenon.
It represents a number,
but the exact value is determined by chance.
Examples include the outcome of a dice roll,
whether a coin lands heads or tails,
or the percentage of flights that arrive on time.

Most variables we work with in econometrics can be thought of as random variables.
A person's wage,
years of education,
or health status all have elements of randomness---we cannot perfectly predict them in advance.

### Probability

**Probability** quantifies uncertainty.
If $X$ is a random variable that takes on $k$ possible values $\{x_1, x_2, ..., x_k\}$,
then the probability that $X$ takes a particular value $x_j$ is denoted:

$$
p_j = P(X = x_j)
$$

Each probability $p_j$ must be between 0 and 1,
and the probabilities must sum to 1:
$p_1 + p_2 + ... + p_k = 1$.

### Probability Distributions

The **probability density function (pdf)**,
denoted $f(x)$,
gives the probability that a random variable takes on a particular value $x$.
For discrete random variables,
$f(x) = P(X = x)$.

The **cumulative distribution function (cdf)**,
denoted $F(x)$,
gives the probability that the random variable is less than or equal to $x$:

$$
F(x) = P(X \leq x)
$$

The cdf is useful for answering questions like "what is the probability that $X$ is below some threshold?"


### Expected Values

The **expected value** (or **mathematical expectation**) of a random variable $X$,
denoted $E[X]$,
represents the "average" outcome we would expect if we could repeat the random process many times.
It is our population "best guess" for the value of $X$.
We often denote it with $\mu$ or $\mu_X$.

For a discrete random variable with pdf $f(x)$,
the expected value is the weighted average of all possible values,
where the weights are the probabilities:

$$
E[X] = \sum_{j=1}^{k} x_j f(x_j) = \mu
$$

**Example:** Let $X$ be the outcome of rolling a fair six-sided die.
Each value (1 through 6) has probability $1/6$.
The expected value is:

$$
E[X] = \frac{1}{6}(1) + \frac{1}{6}(2) + \frac{1}{6}(3) + \frac{1}{6}(4) + \frac{1}{6}(5) + \frac{1}{6}(6) = 3.5
$$

Notice that 3.5 is not a possible outcome of any single roll,
but it is the average we would expect over many rolls.

#### Properties of Expected Values

The properties of expected values mirror those of the summation operator:

**Property 1:** For any constant $c$, $E[c] = c$.

**Property 2:** For any constants $a$ and $b$, $E[aX + b] = aE[X] + b$.

**Property 3:** If $\{a_1, a_2, ..., a_n\}$ are constants and $\{X_1, X_2, ..., X_n\}$ are random variables, then:

$$
E\left[\sum_{i=1}^{n} a_i X_i\right] = \sum_{i=1}^{n} a_i E[X_i]
$$


## Measures of Variability

Knowing the central tendency of a variable is useful,
but it doesn't tell us everything.
Two distributions can have the same mean but look very different if one is tightly clustered around the mean while the other is spread out.
We need measures of **variability** or **dispersion**.

### Variance

Let $X$ be a random variable with mean $\mu = E[X]$.
The **variance** of $X$,
denoted $\sigma^2$ or $\text{Var}(X)$,
measures how spread out the distribution is:

$$
\text{Var}(X) = E[(X - \mu)^2] = \sigma^2
$$

The variance is the expected squared deviation from the mean.
Squaring serves two purposes:
it ensures all deviations are positive (so positive and negative deviations don't cancel out),
and it gives more weight to observations far from the mean.

### Standard Deviation

The **standard deviation** of $X$,
denoted $\text{sd}(X)$ or $\sigma$,
is the positive square root of the variance:

$$
\text{sd}(X) = \sqrt{\text{Var}(X)} = \sigma
$$

The standard deviation is often more interpretable than the variance because it is measured in the same units as $X$ itself.
If $X$ is measured in dollars,
the standard deviation is also in dollars,
whereas the variance would be in "dollars squared."

### Covariance

A central concern in econometrics is understanding how variables move together.
When $X$ is large,
is $Y$ also large?
Or is it small?
Or is there no relationship at all?

The **covariance** quantifies the linear relationship between two random variables.
Let $X$ and $Y$ be random variables with means $\mu_X = E[X]$ and $\mu_Y = E[Y]$.
The covariance is defined as:

$$
\text{Cov}(X, Y) = E[(X - \mu_X)(Y - \mu_Y)]
$$

The covariance measures how $X$ and $Y$ move together:

- **Positive covariance:** When $X$ is above its mean, $Y$ tends to be above its mean (and vice versa). The variables move in the same direction.
- **Negative covariance:** When $X$ is above its mean, $Y$ tends to be below its mean. The variables move in opposite directions.
- **Zero covariance:** There is no linear relationship between $X$ and $Y$.

### Correlation Coefficient

The covariance tells us the direction of the relationship,
but its magnitude depends on the scales of $X$ and $Y$,
making it hard to interpret.
The **correlation coefficient** standardizes the covariance to produce a unitless measure:

$$
\text{Corr}(X, Y) = \frac{\text{Cov}(X, Y)}{\text{sd}(X) \cdot \text{sd}(Y)} = \rho_{XY}
$$

The correlation coefficient is always between $-1$ and $1$:

- $\rho_{XY} = 1$: Perfect positive linear relationship
- $\rho_{XY} = -1$: Perfect negative linear relationship
- $\rho_{XY} = 0$: No linear relationship

Perfect correlations rarely occur in practice.
Values of $\rho_{XY}$ closer to $1$ or $-1$ indicate stronger linear relationships.

::: {.callout-note}
## Correlation vs. Causation
Remember from @sec-introduction: correlation does not imply causation!
A strong correlation between $X$ and $Y$ tells us the variables move together,
but it does not tell us whether $X$ causes $Y$,
$Y$ causes $X$,
or some third factor causes both.
:::


### Conditional Expectations

While correlation measures the overall linear relationship between two variables,
we often want to describe how the expected value of one variable changes with another.
This is where **conditional expectations** come in.

Let $X$ and $Y$ be random variables.
The conditional expectation of $Y$ given that $X$ takes a specific value $x$ is denoted:

$$
E[Y | X = x]
$$

This asks:
what is our "best guess" for $Y$ when we know that $X$ equals $x$?

**Example:** What is the expected wage ($Y$) for people with a college degree ($X = 1$) versus those without ($X = 0$)?
$E[Y | X = 1]$ and $E[Y | X = 0]$ answer these questions.

Conditional expectations are fundamental to regression analysis.
When we estimate a regression of $Y$ on $X$,
we are essentially modeling how $E[Y | X]$ changes with $X$.


## Distributions

### The Normal Distribution

The **normal distribution** (also called the Gaussian distribution) is the most important distribution in statistics and econometrics.
Many naturally occurring phenomena are approximately normally distributed,
and many statistical procedures assume normality.

A random variable $X$ with mean $\mu$ and variance $\sigma^2$ that follows a normal distribution is denoted:

$$
X \sim N(\mu, \sigma^2)
$$

The normal distribution has a characteristic "bell curve" shape:
symmetric around the mean,
with most observations clustered near the center and fewer observations in the tails.

```{r}
#| echo: false
#| label: fig-normal-dist
#| fig-cap: "A histogram of 10,000 draws from a normal distribution with mean 20 and standard deviation 10."
#| fig-width: 6
#| fig-height: 4
#| out-width: "70%"

set.seed(1247)
Y <- rnorm(10000, mean = 20, sd = 10)

ggplot(data.frame(Y = Y), aes(x = Y)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  labs(
    x = "Y",
    y = "Frequency",
    title = expression(paste("Normal Distribution: ", mu, " = 20, ", sigma, " = 10"))
  ) +
  theme_minimal()
```

Two key facts about the normal distribution:

- Approximately **68%** of observations lie within one standard deviation of the mean (between $\mu - \sigma$ and $\mu + \sigma$)
- Approximately **95%** of observations lie within two standard deviations of the mean (between $\mu - 2\sigma$ and $\mu + 2\sigma$)

### The Standard Normal Distribution

The **standard normal distribution** is a special case of the normal distribution with mean 0 and variance 1:

$$
Z \sim N(0, 1)
$$

The standard normal is important because any normal random variable can be converted to a standard normal through **standardization**.
If $X \sim N(\mu, \sigma^2)$,
then:

$$
Z = \frac{X - \mu}{\sigma} \sim N(0, 1)
$$

Standardization subtracts the mean (centering the distribution at zero) and divides by the standard deviation (scaling it to have unit variance).

```{r}
#| echo: false
#| label: fig-standard-normal
#| fig-cap: "A histogram of 10,000 draws from a standard normal distribution (mean 0, standard deviation 1)."
#| fig-width: 6
#| fig-height: 4
#| out-width: "70%"

set.seed(1247)
Z <- rnorm(10000, mean = 0, sd = 1)

ggplot(data.frame(Z = Z), aes(x = Z)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  labs(
    x = "Z",
    y = "Frequency",
    title = expression(paste("Standard Normal Distribution: ", mu, " = 0, ", sigma, " = 1"))
  ) +
  theme_minimal()
```

The standard normal is what z-tables (or statistical software) use to compute probabilities.
Once you standardize a variable,
you can look up the probability of observing values in any range.


## Summary

This appendix reviewed the mathematical and statistical foundations needed for econometrics.
We covered the summation operator and its properties,
which are essential for understanding formulas throughout the course.
We introduced derivatives and partial derivatives,
which help us interpret how changes in one variable affect another,
and learned how to find and classify maxima and minima of functions.
We distinguished between populations and samples---a fundamental concept for statistical inference.
We then turned to probability,
defining random variables,
expected values,
variance,
covariance,
and correlation.
Finally,
we introduced the normal distribution,
which plays a central role in statistical inference.

### Key Formulas

| Concept | Formula |
|---------|---------|
| Sample mean | $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ |
| Critical point (quadratic) | $x^* = -\frac{b}{2a}$ for $f(x) = ax^2 + bx + c$ |
| Second derivative test | Max if $f''(x^*) < 0$; Min if $f''(x^*) > 0$ |
| Expected value | $E[X] = \sum_{j=1}^{k} x_j f(x_j) = \mu$ |
| Variance | $\text{Var}(X) = E[(X-\mu)^2] = \sigma^2$ |
| Standard deviation | $\text{sd}(X) = \sqrt{\text{Var}(X)} = \sigma$ |
| Covariance | $\text{Cov}(X,Y) = E[(X-\mu_X)(Y-\mu_Y)]$ |
| Correlation | $\rho_{XY} = \frac{\text{Cov}(X,Y)}{\text{sd}(X) \cdot \text{sd}(Y)}$ |
| Standardization | $Z = \frac{X - \mu}{\sigma}$ |


## Check Your Understanding

:::: {.content-visible when-format="html"}

For each question below, 
select the best answer from the dropdown menu. 
The dropdown will turn green if correct and red if incorrect. 
Click the "Show Explanation" toggle to see a full explanation of the answer after attempting each question.

```{=html}
<div class="question">
<div class="answer-row">
    <label>i. What is <span style="font-size:1.2em;">∑</span><sub>i=1</sub><sup>5</sup> 3? </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">3</option>
        <option value="b">8</option>
        <option value="c">15</option>
        <option value="d">243</option>
    </select>
</div>

<div class="answer-row">
    <label>ii. If the deviations from the mean always sum to zero, what is <span style="font-size:1.2em;">∑</span><sub>i=1</sub><sup>n</sup>(x<sub>i</sub> − x̄)? </label>
    <select onchange="checkAnswer(this, 'a')">
        <option value="">-- Select Answer --</option>
        <option value="a">0</option>
        <option value="b">n</option>
        <option value="c">nx̄</option>
        <option value="d">Cannot be determined</option>
    </select>
</div>

<div class="answer-row">
    <label>iii. If y = β₀ + β₁x₁ + β₂x₂, what is ∂y/∂x₁? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">β₀</option>
        <option value="b">β₁</option>
        <option value="c">β₂</option>
        <option value="d">β₁ + β₂</option>
    </select>
</div>

<div class="answer-row">
    <label>iv. For the function f(x) = −3x² + 12x + 5, the critical point x* = 2 is a: </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">Minimum, because the coefficient on x² is negative</option>
        <option value="b">Maximum, because the coefficient on x² is negative</option>
        <option value="c">Minimum, because the coefficient on x² is positive</option>
        <option value="d">Maximum, because the coefficient on x² is positive</option>
    </select>
</div>

<div class="answer-row">
    <label>v. The sample mean x̄ is a _____, while the population mean μ is a _____: </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">Parameter; statistic</option>
        <option value="b">Population; sample</option>
        <option value="c">Statistic; parameter</option>
        <option value="d">Variable; constant</option>
    </select>
</div>

<div class="answer-row">
    <label>vi. The variance measures: </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">The central tendency of a distribution</option>
        <option value="b">The linear relationship between two variables</option>
        <option value="c">The spread or dispersion of a distribution</option>
        <option value="d">The probability of a specific outcome</option>
    </select>
</div>

<div class="answer-row">
    <label>vii. If Cov(X, Y) > 0, this means: </label>
    <select onchange="checkAnswer(this, 'a')">
        <option value="">-- Select Answer --</option>
        <option value="a">X and Y tend to move in the same direction</option>
        <option value="b">X and Y tend to move in opposite directions</option>
        <option value="c">X causes Y</option>
        <option value="d">There is no relationship between X and Y</option>
    </select>
</div>

<div class="answer-row">
    <label>viii. If X ~ N(100, 25), what are the mean and standard deviation of X? </label>
    <select onchange="checkAnswer(this, 'b')">
        <option value="">-- Select Answer --</option>
        <option value="a">Mean = 100, SD = 25</option>
        <option value="b">Mean = 100, SD = 5</option>
        <option value="c">Mean = 25, SD = 100</option>
        <option value="d">Mean = 100, SD = 625</option>
    </select>
</div>

<div class="answer-row">
    <label>ix. <strong>(Challenge)</strong> To standardize a variable X with mean μ and standard deviation σ, you compute: </label>
    <select onchange="checkAnswer(this, 'c')">
        <option value="">-- Select Answer --</option>
        <option value="a">(X + μ) / σ</option>
        <option value="b">(X − μ) × σ</option>
        <option value="c">(X − μ) / σ</option>
        <option value="d">X / (μ + σ)</option>
    </select>
</div>
</div>

```

::: {.callout-tip collapse="true"}
## Show Explanation

i. By Property 1 of summation, $\sum_{i=1}^{n} c = nc$. Here, $c = 3$ and $n = 5$, so the sum is $5 \times 3 = 15$.

ii. This is a key result: deviations from the mean always sum to zero. This follows from the definition of the mean and the properties of summation.

iii. The partial derivative $\frac{\partial y}{\partial x_1}$ measures how $y$ changes with $x_1$ holding $x_2$ constant. In a linear function, this is simply the coefficient on $x_1$, which is $\beta_1$.

iv. For a quadratic function $f(x) = ax^2 + bx + c$, the sign of $a$ determines whether the critical point is a maximum or minimum. Here $a = -3 < 0$, so the parabola opens downward and the critical point is a maximum. The second derivative is $f''(x) = 2a = -6 < 0$, confirming concavity downward.

v. Sample statistics (like $\bar{x}$) are computed from data and used to estimate population parameters (like $\mu$). Parameters describe the population and are typically unknown; statistics describe the sample and are observable.

vi. Variance measures how spread out a distribution is around its mean. The mean measures central tendency, covariance measures the relationship between two variables, and the pdf gives probabilities.

vii. Positive covariance means that when $X$ is above its mean, $Y$ tends to be above its mean as well. Remember: covariance tells us about association, not causation!

viii. The notation $N(\mu, \sigma^2)$ specifies the mean and **variance**. Here, $\mu = 100$ and $\sigma^2 = 25$, so $\sigma = \sqrt{25} = 5$.

ix. Standardization centers the variable at zero (by subtracting $\mu$) and scales it to have unit variance (by dividing by $\sigma$). The formula is $Z = \frac{X - \mu}{\sigma}$.

:::

::::

:::: {.content-hidden when-format="html"}

*Note: This section contains interactive content only available in the HTML version.*

::::
